### 2017-05-22 tex/MercuryNumerical.tex
Initial commit

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%    INSTITUTE OF PHYSICS PUBLISHING                                   %
%                                                                      %
%   `Preparing an article for publication in an Institute of Physics   %
%    Publishing journal using LaTeX'                                   %
%                                                                      %
%    LaTeX source code `ioplau2e.tex' used to generate `author         %
%    guidelines', the documentation explaining and demonstrating use   %
%    of the Institute of Physics Publishing LaTeX preprint files       %
%    `iopart.cls, iopart12.clo and iopart10.clo'.                      %
%                                                                      %
%    `ioplau2e.tex' itself uses LaTeX with `iopart.cls'                %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
% First we have a character check
%
% ! exclamation mark    " double quote  
% # hash                ` opening quote (grave)
% & ampersand           ' closing quote (acute)
% $ dollar              % percent       
% ( open parenthesis    ) close paren.  
% - hyphen              = equals sign
% | vertical bar        ~ tilde         
% @ at sign             _ underscore
% { open curly brace    } close curly   
% [ open square         ] close square bracket
% + plus sign           ; semi-colon    
% * asterisk            : colon
% < open angle bracket  > close angle   
% , comma               . full stop
% ? question mark       / forward slash 
% \ backslash           ^ circumflex
%
% ABCDEFGHIJKLMNOPQRSTUVWXYZ 
% abcdefghijklmnopqrstuvwxyz 
% 1234567890
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\documentclass[12pt]{iopart}
\newcommand{\gguide}{{\it Preparing graphics for IOP Publishing journals}}
%Uncomment next line if AMS fonts required
%\usepackage{iopams}  
\begin{document}

\title[A primer to numerical simulations]{A primer to numerical simulations: The perihelion
motion of Mercury}

\author{I. Hammer, C. Hanhart, C. K\"orber and C. M\"uller}

\address{Forschungszentrum J\"ulich}
\ead{c.hanhart@fz-juelich.de}
\vspace{10pt}
%\begin{indented}
%\item[]February 2014
%\end{indented}

\begin{abstract}
Numerical simulations are playing an increasingly important role in modern science. In this work it is
suggested to use a numerical study of the famous perihelion motion of the planet Mercury (one of the prime observables
supporting Einsteins General Relativity) as a test case to teach numerical simulations to high school students.
The paper includes details about the development of the code as well as a discussion of the visualization 
of the results. In addition a method is discussed how to estimate the size of the effect a priori. 
\end{abstract}

% Uncomment for PACS numbers
%\pacs{00.00, 20.00, 42.10}
%
% Uncomment for keywords
%\vspace{2pc}
%\noindent{\it Keywords}: XXXXXX, YYYYYYYY, ZZZZZZZZZ
%
% Uncomment for Submitted to journal title message
%\submitto{\JPA}
%
% Uncomment if a separate title page is required
%\maketitle
% 
% For two-column output uncomment the next line and choose [10pt] rather than [12pt] in the \documentclass declaration
%\ioptwocol
%

\section{Introduction}

Numerical simulations play a key role in modern physics for they allow one to tackle theoretical problems
not accessible otherwise, e.g., since there are too many particle participating in the system (as in
simulations for weather predictions) or the interactions are too complicated to allow for a systematic,
perturbative approach (as in theoretical descriptions of nuclear particles at the fundamental level). 
 The goal of this paper is it to
introduce a project that could be used to make numerical simulations themselves the topic of the class.
On the example of the perihelion motion of the planet Mercury the students are supposed to get in touch with
\begin{itemize}
\item the importance of differential equations in theoretical physics;
\item the numerical implementation of Newtonian dynamics;
\item systematic test and optimization of computer codes;
\item effective tools to estimate the result a priori as an important cross check;
\item the visualization of numerical results using V-Phython.
\end{itemize}
The course as well as this paper is structured as follows: after an introduction to Newtonian dynamics and the 
concepts of differential equations their discretization is discussed based on Newtons law of gravitation and possible
extensions thereof. Afterwards the visualization of the resulting trajectories using V-Phython is introduced and
applied to the problem at hand. In particular tools are developed to extract the relevant quantity from the result
of the simulation.
Finally the principle of dimensional analysis is presented as a tool to cross check, if the results of the simulation
are sensible.

\section{Trajectories, velocities, accelerations and Newtons second law}
\label{sec:tva}

We can say that we understand a physical system if we can demonstrate that an assumed force leads to the trajectories
observed, where here trajectory means that we can calculate the location in space of the object of interest at any point
in time, once the initial conditions are fixed properly. If we can neglect the finite size of this object the location is parametrized
by a single three vector $\vec x(t)$. As will become clear below in order to describe and control the dynamics of a physical
object in addition we need to be able to calculate its velocity, $\vec v(t)$, related to the location via
\begin{equation}
\vec x(t+\Delta t) = \vec x(t) + \vec v(t) \Delta t + ... \ , \label{eq:vdef}
\end{equation}
for some infinitesimally small $\Delta t$ and the acceleration, $\vec a(t)$ that describes the change of the velocity
\begin{eqnarray}
\vec v(t+\Delta t) = \vec v(t) + \vec a(t) \Delta t + ...\ . \label{eq:adef}
\end{eqnarray}
The dots in the above expressions indicate that there are in general additional terms that may be expressed with
higher powers in $\Delta t$, however, for sufficiently small $\Delta t$ those can be safely neglected.
Thus we may define the time derivative via
\begin{equation}
\vec v(t) = \lim_{\Delta t\to 0} \frac{\Delta \vec x(t)}{\Delta t} =: \frac{d\vec x(t)}{dt}  = \dot{\vec  x}(t) \ ,
\end{equation}
where $\Delta \vec x(t)=\vec x(t+\Delta t)-\vec x(t)$ and we introduced with the last expression a common short hand notation
for time derivatives. Analogously we get
\begin{eqnarray}
\vec a(t) &=& \lim_{\Delta t\to 0} \frac{\Delta \vec v(t)}{\Delta t} =: \frac{d\vec v(t)}{dt}  = \dot{\vec v}(t) \ , \\
 &=& \frac{d^2\vec x(t)}{dt^2}  = \ddot{\vec x}(t) \ ,
\end{eqnarray}
where we introduced in the second line the second derivative.

It was Newton who observed that if a body is at rest it will remain at rest, and if it is in motion it will remain in motion in a 
constant velocity in a straight line, unless it is acted upon by some force --- this is known as Newtons first law. 
Said differently: a force $\vec F$shows up by changing the motion of some object. This is quantified in Newton's second law
\begin{equation}
\vec F(\vec x, t) = \frac{d}{dt}(m \vec v) \ .
\end{equation} 
If the mass does not change~\footnote{A well known example where $m$ does change with time is a 
rocket, whose mass decreases as the rocket rises.} with time this reduces to the well known
\begin{equation}
\vec F(\vec x) = m \vec a(t) = m\dot{\vec v}(t) = m\ddot{\vec{x}}(t) \ . \label{eq:newton2}
\end{equation} 
Note that in general the force could depend also on the time or the velocity, we here restrict ourselves to the
case relevant for our example where the force depends on the location only. 
Therefore, 
as soon as the force $\vec F(\vec x)$ is known for all $\vec x$ one can in principle can, e.g., Eq.~(\ref{eq:newton2})
for $\vec x(t)$.
Sometimes requires some advanced knowledge in math, sometimes no closed form solution exists. However, alternatively 
one can calculate the whole trajectory
of some test body that experiences this force by a successive application of the rules 
given in Eqs.~(\ref{eq:vdef}) and (\ref{eq:adef}):
\begin{enumerate}
\item For a given time $t$, where $\vec x(t)$ and $\vec v(t)$ are known, use Eq.~(\ref{eq:newton2}) to calculate $\vec a(t)$.
\item Use Eq.~(\ref{eq:vdef}) to calculate $\vec x(t+\Delta t)$ and
\item then use Eq.~(\ref{eq:adef}) to calculate $\vec v(t+\Delta t)$.
\item Go back to (i).
\end{enumerate}
Clearly to initiate the procedure at some time $t_0$ both $\vec x(t_0)$ as well as $\vec v(t_0)$ must be known --- the 
trajectories depend on these initial conditions.~\footnote{In general a differential equation of $n^{\rm th}$ degree (where
the highest derivative is $n$) needs $n$ initial conditions specified. For $n=2$ those are often chosen as location and velocity
at some defined time, but also to pick two location at different times is possible. }

Clearly for this procedure to work $\Delta t$ must be sufficiently small. What this means depends on the process
studied. One way to check, if $\Delta t$ is small enough is to verify, if the realation
\begin{equation}
\vec v(t) \gg \frac12\vec a(t)\Delta t
\label{eq:check}
\end{equation}
holds, for the next term neglected in Eq.~(\ref{eq:vdef}) reads $(1/2)a(t)(\Delta t)^2$.  Eq.~({\ref{eq:check}})
also shows that small (large) time steps are necessary (sufficient), if the force is strong (weak), since
the time steps need to be small enough that all changes induced by the force get resolved.

\section{How to treat differential equations on a computer}

\section{Visualization using V-Phython}

\section{Example: The perihelion motion of Mercury}

For this concrete example the starting point for the force is Newtons law
of gravitation
\begin{equation}
F_N(\vec x) = \frac{G_N m M_\odot}{r^2} \ ,
\end{equation}
where $G_N=6.67\times 10^{-11}$ m$^3$kg$^{-1}$s${-2}$ is the Newtonian constant of gravitation,
$m$ is the mass of Mercury and $M_\odot=2\times 10^{30}$ kg is the mass of the sun.
In addition $r=|\vec x(t)|$ denotes the distance between sun and Mercury, when we assume that the sun
is infinitely more heavy than the planet and located at the center of the coordinate system. Although this is not exact, since $m/M_\odot\sim 10^{-8}$ this is
a pretty good approximation. For later convenience we introduce the Schwarzschildradius of the sun
\begin{equation}
r_S=\frac{2G_N  M_\odot}{c^2} = 3 \ \mbox{km} \ ,
\end{equation}
where $c=3\times 10^8$ m/s denotes the speed of light.
With this Newtons second law reads 
\begin{equation}
\ddot{\vec x} = \frac{c^2}{2}\left(\frac{r_S}{r^2}\right) \ .
\label{eq:newton}
\end{equation}
\begin{itemize}
\item{\bf First task:}
Now the students should implement the program outlined in Sec.~\ref{sec:tva} and visualize the resulting
trajectories of Mercury for different start parameters.
Why is it sufficient to work in two dimensions using simply 
$\vec x(t)=(x(t),y(t))^T$?
 When the parameters are chosen appropriately close
orbits should emerge that are fixed in time --- in particular the perihelion (the point of closest approach
of planet and sun) stays fixed in space.
\end{itemize}
It is a well known feature of a $1/r^2$ force that elliptic orbits with fixed perihelion emerge. However, as soon
as the potential is different, the perihelion moves. This will be studied next:
\begin{itemize}
\item{\bf Second task:}
Now the students should add a new term to the right hand side of Eq.~(\ref{eq:newton}). In particular the equation should now
read 
\begin{equation}
\ddot{\vec x} = \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\left(1+\alpha\left(\frac{r_S}{r}\right)\right) \ ,
\label{eq:newton}
\end{equation}
where $\alpha$ is some parameter. What happens to the trajectories as $\alpha$ is varied?
\end{itemize}
Finally the start values need to be chosen appropriately such that the simulation indeed mimicks the trajectories
of Mercury ...

\section{Dimensional analysis}

Dimensional analysis is not only a tool that allows one to cross check if the results of some simulation is
of the right order of magnitude, it is also very helpful to identify unusual dynamics in some system.
Especially the latter aspect should become clear from the discussion in this section.

The idea of dimensional analysis is that in a system that can be controlled by expanding the relevant quantities
(like the force) in some small parameter(s), the coefficients in the expansion should turn out to be of order unity (that
means anything between about 0.1 and 10 is fine - but 0.01 or 100 is irritating) --- parameters in line with this
are called 'natural'. Applied to the problem at hand
given by Eq.~(\ref{eq:newton}) this statement implies that from naturalness one would expect that 
the parameter $\alpha$ is of order 1 when one uses
for $r$ the average distance Mercury-sun, namely $\bar r=6\times 10^7$ km.
Form this one estimates for the expected angular shift per orbit
\begin{equation}
\delta \phi \simeq 2\pi\left(\frac{r_S}{\bar r}\right) = (\pi \times 10^{-7}) \ \mbox{rad} = (2\times 10^{-5})^o = (7\times 10^{-2}) ^{''} \ ,
\end{equation}
which leads to a shift of  about $30^{''}$ in 100 earth years to be compared to the empirical value of $43^{''}$.
 Thus indeed the
amount of perihelion motion of Mercury is in line with expectations, $if$ --- and this is an important 'if' ---
the Newtonian dynamics is simply the leading term of some more general underlying theory. In particular,
no new scales enter in the correction terms.

Thanks to Einstein we indeed know that the conclusion formulated above is correct ---- the more general underlying
theory is General Relativity and indeed Einstein was able to quantitatively explain the perihelion motion of Mercury
from his equations. 
On the other hand had we found a dramatic deviation of $\alpha$ from unity one would have concluded that 
there is probably some other dynamics going on that drives this difference.

Indeed, we now several of those hierarchy problems in modern physics: e.g. the so called QCD $\theta$ term,
expected to be of natural size, is at present known to be at most $10^{-10}$. This smallness, called the
strong CP problem, is so irritating that physicists like S. Weinberg even proposed that there must exist an
additional particle, the axion, whose interactions are in charge of pushing $\theta$ even to zero and there
are now intense searches for this axion going on various labs.

\section{Summary and possible extensions}

%\begin{table}
%\caption{\label{jlab1}Journals to which this document applies, and macros for the abbreviated journal names in {\tt iopart.cls}. Macros for other journal titles are listed in appendix\,A.}
%\footnotesize
%\begin{tabular}{@{}llll}
%\br
%Short form of journal title&Macro name&Short form of journal title&Macro name\\
%\mr
%2D Mater.&\verb"\TDM"&Mater. Res. Express&\verb"\MRE"\\
%\br
%\end{tabular}\\
%$^{a}$UK spelling is required; $^{b}$MSC classification may be used as well as PACS; $^{c}$titles of articles are required in journal references; $^{d}$Harvard-style references must be used (see section \ref{except}); $^{e}$final page numbers of articles are required in journal references.
%
%\end{table}
%\normalsize

\subsection{Acknowledgments}
Here they come

\appendix
\section{Instructions to install V-phython}

\section{The code}
Here we should put the code

\bibliographystyle{plain}
\bibliography{MercuryNumerical}

\end{document}

### 2017-05-22 tex/MercuryNumerical.tex
Updated the Numerical implementation section and added an example script (which still needs to be adjusted to the text).

@@
 %\usepackage{iopams}
+
+\usepackage{graphicx}           % Import Pictures + PDFs
+\usepackage[svgnames]{xcolor}   % Enabling colors by their 'svgnames'
+\definecolor{fzjblue}{HTML}{005B81}
+\definecolor{light-gray}{gray}{0.97}
+\definecolor{mygreen}{rgb}{0,0.6,0}
+\usepackage{listings} %  See: https://en.wikibooks.org/wiki/LaTeX/Source_Code_Listings
+\lstset{
+  language=Python,
+  backgroundcolor=\color{light-gray},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}
+  basicstyle=\linespread{1}\footnotesize,        % the size of the fonts that are used for the code
+  commentstyle=\color{mygreen},
+  frame=single,
+  keywordstyle=\color{blue},       % keyword style
+  tabsize=4,	                   % sets default tabsize to 2 spaces
+  rulecolor=\color{light-gray},
+}
+
+\newcommand{\python}[0]{\texttt{Python} }
+\newcommand{\abs}[1]{\left\vert #1 \right\vert}
+
 \begin{document}
@@
 object in addition we need to be able to calculate its velocity, $\vec v(t)$, related to the location via
-\begin{equation}
+\begin{equation}\label{eq:x_update}
 \vec x(t+\Delta t) = \vec x(t) + \vec v(t) \Delta t + ... \ , \label{eq:vdef}
@@
 for some infinitesimally small $\Delta t$ and the acceleration, $\vec a(t)$ that describes the change of the velocity
-\begin{eqnarray}
+\begin{eqnarray}\label{eq:v_update}
 \vec v(t+\Delta t) = \vec v(t) + \vec a(t) \Delta t + ...\ . \label{eq:adef}
@@
 
-\section{How to treat differential equations on a computer}
-
-\section{Visualization using V-Phython}
-
 \section{Example: The perihelion motion of Mercury}
@@
 \ddot{\vec x} = \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\left(1+\alpha\left(\frac{r_S}{r}\right)\right) \ ,
-\label{eq:newton}
+\label{eq:newton_art}
 \end{equation}
@@
 of Mercury ...
+
+\section{Numerical Implementation}
+In this section we present the numerical implementation of the perihelion motion of Mercury.
+We select \python as the programming language of choice \cite{}, because \python is easy to learn, intuitive to understand and an open source language\footnote{See also appendix \ref{appendix:python} for an instruction how to set up python on different operating systems.}.
+No prior knowledge of \python or any other programming language is required because we explain all the essential steps in order to create the simulation.
+We also provide a working example \cite{} which can be used as template.
+
+The first step of a numerical simulation is the extraction of the relevant physical parameters and quantities in order to translate the problem to mathematical equations.
+The relevant objects for this example are the sun and mercury.
+Thus, to describe their motion, one needs their "initial" distances, velocities and accelerations which are the input to their equation of motions (Eq.~(\ref{eq:x_update}) and Eq.~(\ref{eq:v_update})).
+In the case of this problem, one can simplify the implementation by placing one planet (the "infinitely" heavy sun) in the center of the coordinate system and describe the motion of the other planet (mercury) in a two-dimensional plane.
+\textit{Why is this simplified picture sufficient to describe the motion}?
+
+Because the problem does not depend on the initial position of mercury on its orbit around the sun, we extract the parameters at the perihelion with $ \abs{\vec r_{MS}(0)} = 46 \cdot 10^6 km$ and $ \abs{\vec v_M(0)} = 39 km/s $\footnote{https://nssdc.gsfc.nasa.gov/planetary/factsheet/mercuryfact.html}.
+Since the computer does not understand physical units, one has to express each variable in their respective unit.
+Both, for the numerical treatment and the intuitive understanding, it is useful to select parameters in a "natural range".
+A particular useful choice for this problem is given by expressing time intervals in days $T_0 = 1d$ and distances in $R_0 = 10^{10}m$.
+With this choice, the initial distance of mercury to the sun, the initial velocity of mercury and the acceleration prefactor become
+\begin{equation}
+	r_{MS}(0) = 4.6 R_0 \, , \quad
+	v_{M}(0)  = 0.34 \frac{R_0}{T_0} \, ,  \quad
+	a_M       = 0.99 \frac{R_0}{T_0^2} \frac{1}{\left(r_{MS}/R_0\right)^2}
+	\, .
+\end{equation}
+
+\begin{figure}[htb]
+	\centering
+	\includegraphics[width=.4\textwidth]{figs/sun_merc.pdf}
+	\caption{\label{fig:sun_merc}Sun mercury system with relevant vectors.  Because the mercury is at its perihelion, its velocity is perpendicular to its direct connection vector with the sun.}
+\end{figure}
+
+In python, this reads
+\begin{lstlisting}
+	# Definition of parameters
+	rM0 = 4.6    # in units of R0
+	vM0 = 0.34   # in units of R0/T0
+	aM  = 0.99   # in units of R0/T0**2
+\end{lstlisting}
+Next, we set up the initial vectors which will describe the motion in the two-dimensional plane.
+We will build on the existing \python module: \texttt{Vpython}, which provides implementation for the vectorial implementation and animation.
+The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.
+Because of our choice of initial conditions, the velocity of mercury is perpendicular to the vector which connects mercury and the sun (see figure \ref{fig:sun_merc}):
+\begin{lstlisting}
+	# Import the class vector from python
+	from vpython import vector
+	# Initialize distance and velocity vectors
+	vec_rS0 = vector(0, 0, 0)
+	vec_rM0 = vector(0, rM0, 0)
+	vec_vM0 = vector(vm0, 0, 0)
+\end{lstlisting}
+According to Eq.~(\ref{eq:newton}), the force which acts on mercury changes the velocity of mercury which eventually changes the position.
+In first approximation we can numerically express the updates of the position and the velocity by
+\begin{lstlisting}
+	# Compute the strength of the acceleration
+	aMS = aM * ( 1 + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
+	# Multiply by the direction
+	vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
+	# Update velocity vector
+	vec_vM_new = vec_vM_old + vec_aMS * dt
+	# Update position vector
+	vec_rM_new = vec_rM_old + vec_vM_new * dt
+\end{lstlisting}
+Note the beauty of working with the predefined \texttt{vector} class: the basic vector operations are already implemented.
+The difference and sum of two vectors, or the scalar vector multiplication return vectors themselves.
+Also the magnitude of a vector -- \texttt{vector.mag} -- is a property of the vector and can be easily extracted.
+
+It is handy to use \texttt{Python}s \texttt{function}s to embed repeating structures ("DRY" -- Don't Repeat Yourself)
+\begin{lstlisting}
+	# Define the function
+	def evolve_mercury(vec_rM_old, vec_vM_old):
+		<...Code...>
+		return vec_rM_new, vec_vM_new
+
+	# Call the function
+	vec_rM_new, vec_vM_new = do_time_step(vec_rM_old, vec_vM_old)
+\end{lstlisting}
+Because of \python{}s syntax, it is necessary that the body of the function is indented relative to the definition statement.
+
+Finally, we can express the evolution by a \texttt{while}-loop
+\begin{lstlisting}
+	t = 0
+	# Execute the loop as long as t < T
+	while t < T:
+		vec_rM, vec_vM = evolve_mercury(vec_rM, vec_vM)
+		t = t + dt
+\end{lstlisting}
+Note the required indent of the loop structure similar to the indent of a function.
+In each iteration of the \texttt{while}-loop, the previous distance and velocity are used to compute the new values -- which directly overwrite the previous values and thus enter the next iteration.
+The time step \texttt{dt} needs to be sufficiently small in order to avoid numerical errors.
+We recommend $dt = ...$.
+The total runtime \texttt{T} is the amount of "virtual" days the simulations should run.
+To describe at least one full orbit, $T$ needs to be larger than $T>..$.
+
+Ideas for questions:
+\begin{itemize}
+	\item Mass and motion of sun
+	\item Estimate \texttt{dt} and also \texttt{T} in order to cover roughly two "mercury years".
+	\item What does happen when you change those units?
+	\item Compare values to the values in \cite{}.  Where does the difference come from?
+	\item Start with $\alpha=0$. What do you observe?
+\end{itemize}
+
+\section{Visualization using V-Phython}
 
@@
 \section{Instructions to install V-phython}
+\label{appendix:python}
 

### 2017-05-22 scripts/merkur_orbit.py

@@ -1,0 +1,154 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+### Darstellung des Merkurorbits mit VPython
+
+
+# ----------------------------------
+## Laden der Hilfsmodule
+# ----------------------------------
+
+# Da es unglaublich viel Zeit kostet ein Programm für 3-dimensionale graphische Darstellung selber 
+# zu schreiben, lassen wir uns die Arbeit von "Vpython" abnehmen. Doch dazu müssen wir zunächst 
+# dieses Modul bereitstellen. Das erreicht man mit Hilfe des Befehls "import" oder "from .. import".
+
+from visual import vector, sphere, color, curve, rate, display
+
+# Schreibt man 
+#   "from visual import *" 
+# bedeutet dies: "Lade alle Funktionen von Vpython so, dass man sie direkt verwenden kann: 
+#   "func = function(..)". 
+# Würde man 
+#   "import visual" 
+# schreiben, so müsste man die Funktion wie folgt aufrufen: 
+#   "func = visual.function(..)".
+
+# Eine Dokumentation von Modulen lässt sich durch den Befehl "help( modul_name )" aufrufen.
+#
+# Tipp: Will man eine Eingabe unterdrücken, so lässt sich dies auch mit "#" tun. Der darauf 
+# folgende Text wird dabei von Python überlesen.
+
+# help(visual)
+
+# ----------------------------------
+## Aufgabe 1: Merkurorbit)
+# ----------------------------------
+
+# Funktionen werden in Python wie folgt deklariert:
+# def func(x):
+#     .... Rechnung und Anderes ....
+# 
+# Diese wollen wir nun nutzen. Um die Umlaufbahn des Merkurs darzustellen, ist es sinvoll die 
+# Aufgabe in zwei Abschnitte zu Teilen:
+# (a) Berchnung des Orbits
+# (b) Zeichnung des Orbits
+# Für beide Abschnitte werden wir Funktionen nutzen.
+
+
+# ----------------------------------
+## Parameter
+# ----------------------------------
+
+# Physikalische Größen/Einheiten
+ms = 1.98892 * 1.e30 # Sonnenmassen
+AE = 1.49598 * 1.e11 # Astronomische Einheit
+T  = 60 * 60 * 24    # 1 Tag
[…]

### 2017-05-23 tex/MercuryNumerical.tex
Added some links for ball in the box (Christian)

@@
 	# Initialize distance and velocity vectors
-	vec_rS0 = vector(0, 0, 0)
 	vec_rM0 = vector(0, rM0, 0)
@@
 
-\bibliographystyle{plain}
+\bibliographystyle{alpha din}
 \bibliography{MercuryNumerical}

### 2017-06-30 tex/MercuryNumerical.tex
Included Christoph's suggestions

@@
 \item the numerical implementation of Newtonian dynamics;
-\item systematic test and optimization of computer codes;
+\item systematic tests and optimization of computer codes;
 \item effective tools to estimate the result a priori as an important cross check;
@@
 
+We are convinced that in order to excite the students for numerical simulations it is compulsory to
+demonstrate their power on an example that they catches their interest. This purpose is served 
+perfectly by the case chosen here, since Einsteins equations of General Relativity are fascinating to
+a very broad public. While their detailed study needs deeper knowledge in Differential Geometry, for
+the study outlined here very little math is necessary, such that high school students from about 
+10$^{\rm th}$ grade up should benefit from it. This was already demonstrated in the
+'Sch\"ulderakademie Teilchenphysik', where an early version of this course was tested successfully 
+on a goup of 10 German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade.
+
 \section{Trajectories, velocities, accelerations and Newtons second law}
@@
 
-We can say that we understand a physical system if we can demonstrate that an assumed force leads to the trajectories
-observed, where here trajectory means that we can calculate the location in space of the object of interest at any point
-in time, once the initial conditions are fixed properly. If we can neglect the finite size of this object the location is parametrized
+We can say that we understand a physical system, if we can demonstrate that the assumed force leads to the trajectories
+observed. In other words, we need to show that we can calculate the location in space of the object of interest at any point
+in time, once the initial conditions are fixed properly. If we can neglect the finite size of this object (and
+in particular its orientation in space) the location is parametrized
 by a single three vector $\vec x(t)$. As will become clear below in order to describe and control the dynamics of a physical
@@
 constant velocity in a straight line, unless it is acted upon by some force --- this is known as Newtons first law. 
-Said differently: a force $\vec F$shows up by changing the motion of some object. This is quantified in Newton's second law
+Said differently: a force $\vec F$ shows up by changing the motion of some object. This is quantified in Newton's second law
 \begin{equation}
@@
 \end{equation} 
-Note that in general the force could depend also on the time or the velocity, we here restrict ourselves to the
+Note that in general the force could depend also on the time or the velocity. We here restrict ourselves to the
 case relevant for our example where the force depends on the location only. 
 Therefore, 
-as soon as the force $\vec F(\vec x)$ is known for all $\vec x$ one can in principle can, e.g., Eq.~(\ref{eq:newton2})
+as soon as the force $\vec F(\vec x)$ is known for all $\vec x$ one can in principle solve, e.g., Eq.~(\ref{eq:newton2})
 for $\vec x(t)$.
-Sometimes requires some advanced knowledge in math, sometimes no closed form solution exists. However, alternatively 
+Sometimes this requires some advanced knowledge in math, sometimes no closed form solution exists. However, alternatively 
 one can calculate the whole trajectory
@@
 \item then use Eq.~(\ref{eq:adef}) to calculate $\vec v(t+\Delta t)$.
-\item Go back to (i).
+\item Go back to (i) with $t\to t+\Delta t$.
 \end{enumerate}
@@
 Clearly for this procedure to work $\Delta t$ must be sufficiently small. What this means depends on the process
-studied. One way to check, if $\Delta t$ is small enough is to verify, if the realation
+studied. One way to estimate, if $\Delta t$ is small enough, is to verify, if the relation
 \begin{equation}
-\vec v(t) \gg \frac12\vec a(t)\Delta t
+|\vec v(t)| \gg \frac12|\vec a(t)|\Delta t = \frac{1}{2m}\vec F(\vec x)\Delta t\ . 
 \label{eq:check}
@@
 also shows that small (large) time steps are necessary (sufficient), if the force is strong (weak), since
-the time steps need to be small enough that all changes induced by the force get resolved.
+the time steps need to be small enough that all changes induced by the force get resolved. Clearly,
+a relation as Eq.~({\ref{eq:check}}) can only provide guidance and can not replace a careful numerical check of
+the solutions: A valid result has to be insensitive the concrete value $\Delta t$ chosen --- in particular replacing
+$\Delta t$ by $\Delta t/2$ should not change the result significantly.
 
-\section{Example: The perihelion motion of Mercury}
+\section{Example: General Relativity and the perihelion motion of Mercury}
 
@@
 In addition $r=|\vec x(t)|$ denotes the distance between sun and Mercury, when we assume that the sun
-is infinitely more heavy than the planet and located at the center of the coordinate system. Although this is not exact, since $m/M_\odot\sim 10^{-8}$ this is
+is infinitely more heavy than the planet and located at the center of the coordinate system. 
+Although this is not exact, since $m/M_\odot\sim 10^{-8}$ this is
 a pretty good approximation. For later convenience we introduce the Schwarzschildradius of the sun
@@
 where $c=3\times 10^8$ m/s denotes the speed of light.
+Note the $r_S$ is the characteristic length scale of the gravitational field of the sun
+for --- up to the prefactor --- one can not form another quantity with dimensions of a length from $G_N$, $M_\odot$ and
+$c$. This is what we need $r_S$ for later in the discussion. That the Schwarzschild radius is
+also an important quantity to characterize black holes is not relevant for the discussion
+at hand.
 With this Newtons second law reads 
@@
 \end{equation}
-\begin{itemize}
-\item{\bf First task:}
-Now the students should implement the program outlined in Sec.~\ref{sec:tva} and visualize the resulting
-trajectories of Mercury for different start parameters.
-Why is it sufficient to work in two dimensions using simply 
-$\vec x(t)=(x(t),y(t))^T$?
- When the parameters are chosen appropriately close
-orbits should emerge that are fixed in time --- in particular the perihelion (the point of closest approach
-of planet and sun) stays fixed in space.
-\end{itemize}
-It is a well known feature of a $1/r^2$ force that elliptic orbits with fixed perihelion emerge. However, as soon
-as the potential is different, the perihelion moves. This will be studied next:
-\begin{itemize}
-\item{\bf Second task:}
-Now the students should add a new term to the right hand side of Eq.~(\ref{eq:newton}). In particular the equation should now
-read 
+%\begin{itemize}
+%\item{\bf First task:}
+%Now the students should implement the program outlined in Sec.~\ref{sec:tva} and visualize the resulting
+%trajectories of Mercury for different start parameters.
+%Why is it sufficient to work in two dimensions using simply 
+%$\vec x(t)=(x(t),y(t))^T$?
+% When the parameters are chosen appropriately close
+%orbits should emerge that are fixed in time --- in particular the perihelion (the point of closest approach
+%of planet and sun) stays fixed in space.
+%\end{itemize}
+In general an attractive force that vanishes at large distances leads,
+ depending on the initial conditions,
+  either to closed orbits or open orbits --- 
+  in the latter case the planet simply disappears from the sun. Depending on time available
+for the course  it might be helpful to explain to the students the origin of this, however, since the most
+transparent reasoning needs an integration as well as the concept of angular momentum
+not necessarily familiar to the whole class (details are given in the appendix {\bf to be done}), 
+it appears in general more appropriate to work out together with the students 
+that a moving body can not be kept by a force if its momentum is too large.
+ Then they may try to find closed orbits in the simulation, e.g., by varying
+the start velocity while keeping the start location fixed.
+
+The closed orbits that emerge from a force that scales as $1/r^2$
+are elliptic and in particular fixed in space. In particular the point of closest approach of the planet
+to the sun, the perihelion, does not move.
+  However, as soon
+as the potential is different, the perihelion moves. Because of this the behavior of the perihelion
+is a very sensitive probe of the gravitational potential.
+
+The observed perihelion motion of Mercury is nowadays determined as $(574.10\pm 0.65)''$ per
+100 earth years~\footnote{The symbol $''$ denotes 'arc seconds': 1$''=(1/3600)^o$. }, but it should be noted that the bulk of this number
+can be understood by the presence of other planets within the Newtonian theory, since
+their gravitational force also acts on Mercury. However, a
+residual motion of $(43)''$ in 100 earth years remained unexplained, until Einstein quantified the
+predictions of General Relativity to this particular observable.
+
+{\bf Here should be a brief description of the relation between Newtons dynamics and General Relativity}
+
+To allow for a movement of the perihelion also in the study presented here the
+potential of Eq.~(\ref{eq:newton}). We do this using the following ansatz
 \begin{equation}
@@
 \end{equation}
-where $\alpha$ is some parameter. What happens to the trajectories as $\alpha$ is varied?
-\end{itemize}
-Finally the start values need to be chosen appropriately such that the simulation indeed mimicks the trajectories
-of Mercury ...
+where $\alpha$ is some parameter. The idea is that the correction must be dimension less and
+should scale with some inverse power of $r$, for we still need to demand that the potential 
+vanishes at large distances. Since the only parameter of the system with dimension
+of a length is $r_S$, the correction should scale as $(r_S/r)$. When replacing $r$ by a typical
+distance of Mercury to the sun, $(r_S/r)\sim 10^{-7}$, it should be sufficient to include this kind
+of term to first order only. For comparison: the corresponding number for the venus, the neighbor
+of  Mercury, reads ...., which naturally explains why the effect is observable for Mercury only (IS IT - CHECK!)
+%\end{itemize}
 
 \section{Numerical Implementation}
-In this section we present the numerical implementation of the perihelion motion of Mercury.
-We select \python as the programming language of choice \cite{}, because \python is easy to learn, intuitive to understand and an open source language\footnote{See also appendix \ref{appendix:python} for an instruction how to set up python on different operating systems.}.
-No prior knowledge of \python or any other programming language is required because we explain all the essential steps in order to create the simulation.
-We also provide a working example \cite{} which can be used as template.
+In this section we present the numerical implementation as well as the visualization of planetary trajectories and in
+particular
+  the perihelion motion of Mercury.  We select \python~\cite{} as  programming language, because \python is easy to learn, 
+  intuitive to understand and an open source language\footnote{See also appendix \ref{appendix:python} for an instruction 
+  how to set up python on different operating systems.}.  No prior knowledge of \python or any other programming language 
+  is required because we explain all the steps necessary to create the simulation.  We also provide a working 
+  example \cite{} that can be used as template.
 
-The first step of a numerical simulation is the extraction of the relevant physical parameters and quantities in order to translate the problem to mathematical equations.
-The relevant objects for this example are the sun and mercury.
-Thus, to describe their motion, one needs their "initial" distances, velocities and accelerations which are the input to their equation of motions (Eq.~(\ref{eq:x_update}) and Eq.~(\ref{eq:v_update})).
-In the case of this problem, one can simplify the implementation by placing one planet (the "infinitely" heavy sun) in the center of the coordinate system and describe the motion of the other planet (mercury) in a two-dimensional plane.
+To start the simulation one needs the "initial" distances and velocities, which are needed as input to their equation of motions (Eq.~(\ref{eq:x_update}) and Eq.~(\ref{eq:v_update})).
+As described above, here  one can simplify the problem by placing one planet (the "infinitely" heavy sun) in the center of the coordinate system and describe the motion of the other planet (mercury) in a two-dimensional plane.
 \textit{Why is this simplified picture sufficient to describe the motion}?
 
-Because the problem does not depend on the initial position of mercury on its orbit around the sun, we extract the parameters at the perihelion with $ \abs{\vec r_{MS}(0)} = 46 \cdot 10^6 km$ and $ \abs{\vec v_M(0)} = 39 km/s $\footnote{https://nssdc.gsfc.nasa.gov/planetary/factsheet/mercuryfact.html}.
-Since the computer does not understand physical units, one has to express each variable in their respective unit.
+Because the problem does not depend on the initial position of mercury on its orbit around the sun, we extract the parameters at the perihelion with $ \abs{\vec r_{MS}(0)} = 46 \cdot 10^6 km$ and $ \abs{\vec v_M(0)} = 39$ km/s \footnote{https://nssdc.gsfc.nasa.gov/planetary/factsheet/mercuryfact.html}.
+Since the computer does not understand physical units, one has to express each variable in an appropriate unit.
 Both, for the numerical treatment and the intuitive understanding, it is useful to select parameters in a "natural range".
-A particular useful choice for this problem is given by expressing time intervals in days $T_0 = 1d$ and distances in $R_0 = 10^{10}m$.
-With this choice, the initial distance of mercury to the sun, the initial velocity of mercury and the acceleration prefactor become
+A particular useful choice for this problem is given by expressing time intervals in days $T_0 = 1d$ {\bf These are earth days, I assume, right?} and distances in $R_0 = 10^{10}$ m.
+With this choice, the initial distance of mercury to the sun, the size of the initial velocity of mercury and the acceleration prefactor become
 \begin{equation}
@@
 \end{lstlisting}
-Next, we set up the initial vectors which will describe the motion in the two-dimensional plane.
-We will build on the existing \python module: \texttt{Vpython}, which provides implementation for the vectorial implementation and animation.
+So far we only fixed the length of the vectors.
+Next, we set up the initial directions, which will describe the motion in the two-dimensional plane.
+We will build on the existing \python module \texttt{Vpython}, which provides  an implementation for treating vectors as well as their visualization.
 The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.
-Because of our choice of initial conditions, the velocity of mercury is perpendicular to the vector which connects mercury and the sun (see figure \ref{fig:sun_merc}):
+Because of our choice of initial conditions (we picked the initial vectors in the perihelion), the velocity of mercury is perpendicular to the vector which connects mercury and the sun (see figure \ref{fig:sun_merc}):
 \begin{lstlisting}
@@
 According to Eq.~(\ref{eq:newton}), the force which acts on mercury changes the velocity of mercury which eventually changes the position.
-In first approximation we can numerically express the updates of the position and the velocity by
+However, we first need to fix the time step $\Delta t$ and use the estimate
+of Eq.~(\ref{eq:check}) as a guidance {\bf Careful: Check units of dt}
+\begin{lstlisting}
+	# Definition of the time step
+        dt = 2 * vm0 / aM / 100
+\end{lstlisting}
+Here the factor $1/100$ makes sure that $\Delta t$ is indeed
+consistent with Eq.~(\ref{eq:check}).  {\bf We need to make sure that 'dt' is known
+to the subroutines}
+Now we are in the position to calculate
+location and velocity of the planet at $t_0+\Delta t$ using the following comands
 \begin{lstlisting}
@@
 	# Define the function
-	def evolve_mercury(vec_rM_old, vec_vM_old):
+	def evolve_mercury(vec_rM_old, vec_vM_old,alpha):
 		<...Code...>
@@
 \end{lstlisting}
-Because of \python{}s syntax, it is necessary that the body of the function is indented relative to the definition statement.
+Because of \python{$\!'$}s syntax, it is necessary that the body of the function is indented relative to the definition statement {\bf I do not understand this sentence}.
 
@@
 	t = 0
+	alpha=0
 	# Execute the loop as long as t < T
 	while t < T:
-		vec_rM, vec_vM = evolve_mercury(vec_rM, vec_vM)
+		vec_rM, vec_vM = evolve_mercury(vec_rM, vec_vM,alpha)
 		t = t + dt
 \end{lstlisting}
+where for the start we set the parameter $\alpha$ to zero in order first study the properties of the pure 
+$1/r^2$ force.
 Note the required indent of the loop structure similar to the indent of a function.
 In each iteration of the \texttt{while}-loop, the previous distance and velocity are used to compute the new values -- which directly overwrite the previous values and thus enter the next iteration.
-The time step \texttt{dt} needs to be sufficiently small in order to avoid numerical errors.
-We recommend $dt = ...$.
 The total runtime \texttt{T} is the amount of "virtual" days the simulations should run.
 To describe at least one full orbit, $T$ needs to be larger than $T>..$.
+
+{\bf Here we need to write something about visualizing the trajectories and how to run the code }
+
+\section{Tests of stability}
+
+\section{Tasks}
 
@@
 \end{itemize}
-
-\section{Visualization using V-Phython}
 

### 2017-06-30 tex/MercuryNumerical.tex
Included Inka's modifications

@@
 
+\usepackage{ amssymb }
 \usepackage{graphicx}           % Import Pictures + PDFs
 \usepackage[svgnames]{xcolor}   % Enabling colors by their 'svgnames'
+\usepackage{ulem}
 \definecolor{fzjblue}{HTML}{005B81}
@@
 Therefore, 
-as soon as the force $\vec F(\vec x)$ is known for all $\vec x$ one can in principle solve, e.g., Eq.~(\ref{eq:newton2})
+as soon as the force $\vec F(\vec x)$ is known for all $\vec x$ one can in principle calculate the trajectory, e.g. solving eq.~(\ref{eq:newton2})
 for $\vec x(t)$.
@@
 \end{equation}
-holds, for the next term neglected in Eq.~(\ref{eq:vdef}) reads $(1/2)a(t)(\Delta t)^2$.  Eq.~({\ref{eq:check}})
+holds, since the next term neglected in Eq.~(\ref{eq:vdef}) reads $(1/2)a(t)(\Delta t)^2$.  Eq.~({\ref{eq:check}})
 also shows that small (large) time steps are necessary (sufficient), if the force is strong (weak), since
@@
 \begin{equation}
-F_N(\vec x) = \frac{G_N m M_\odot}{r^2} \ ,
+\vec F_N(\vec x) = \frac{G_N m M_\odot}{r^2} \frac{\vec r}{r}\ ,
 \end{equation}
@@
 \section{Numerical Implementation}
+\label{sec:Numerical Implementation}
 In this section we present the numerical implementation as well as the visualization of planetary trajectories and in
@@
 	r_{MS}(0) = 4.6 R_0 \, , \quad
-	v_{M}(0)  = 0.34 \frac{R_0}{T_0} \, ,  \quad
+	v_{M}(0)  = 0.51 \frac{R_0}{T_0} \, ,  \quad
 	a_M       = 0.99 \frac{R_0}{T_0^2} \frac{1}{\left(r_{MS}/R_0\right)^2}
@@
 	# Define the function
-	def evolve_mercury(vec_rM_old, vec_vM_old,alpha):
+	def evolve_mercury(vec_rM_old, vec_vM_old, alpha):
 		<...Code...>
@@
 	while t < T:
-		vec_rM, vec_vM = evolve_mercury(vec_rM, vec_vM,alpha)
+		vec_rM, vec_vM = evolve_mercury(vec_rM, vec_vM, alpha)
 		t = t + dt
@@
 
-\section{Summary and possible extensions}
+\section{Discussion of Starting Values}
+
+\textit{I am not quite sure what to add here to Christopher's discussion.}\\
+
+If the starting values are chosen as advised in the previous section, then the students should end up with a trajectory as depicted in figure \ref{fig:MercuryOrbit}.
+To get the perihelion motion, the term from equation \ref{eq:newton_art} has to be be added.
+This is a good opportunity to let the students play with the size of $\alpha$ and get a feeling for the impact on the trajectories.
+However it might be advisable to not start out with the correct mercury values but e.g.
+a slightly higher initial distance between mercury and sun, since the perihelion motion is better visible then.
+This is illustrated in figure \ref{fig:MercuryOrbit}, where we used a value of $\alpha=10^5$ and $dt=0.1T_0$.
+
+\begin{figure}[htb]
+	\centering
+	\includegraphics[width=.25\textwidth]{figs/MercuryOrbit.png} \quad
+	\includegraphics[width=.273\textwidth]{figs/PerihelionMotionLargeV.png} \quad
+	\includegraphics[width=.26\textwidth]{figs/PerihelionMotion.png}
+	\caption{\label{fig:MercuryOrbit} Output for $\alpha=0$ (left), $\alpha=10^5$ and bigger starting distance $r_{MS}(0)=6R_0$ (middle) and $\alpha=10^5$ with original starting values.}
+\end{figure}
+
+\section{Extracting the Perihelion Motion}
+
+\textit{What exactly do we want the students / advise the teachers to do?
+I think a nice approach would be to let the students implement the code to extract the perihelion motion and make a list of some values of alpha and the corresponding angles.
+Then discuss, that $\alpha=1$ is the natural size for $\alpha$ and that to extract the value of angle they need to exploit the linear dependence.
+Finally they can think on whether the value they get is reasonable and the teacher can explain the rest of section?
+What do you think?}\\
+
+To calculate the expected size of the perihelion motion for a modified gravitational force, the students need to be abel to extract the angle.
+There are several ways to do this, but the easiest is to extract the point where the distance to the sun gets maximal, the aphelion, after $n$ turns and compare it to the original one.
+This can be done by using an $if$-statement in the $while$-loop.
+
+Suppose we want to save the number of turns in the variable $counter$, the maximal number of turns in $maxcounter$ (here e.g.
+10), the initial aphelion in the vector $a0$ and the final aphelion in the vector $a$.
+In addition to find out when the aphelion is reached, the last two positions $rlast$ and $rbeforelast$ have to be known.
+To set this up, before the while loop the following code needs to be included
+\begin{lstlisting}
+	r=r0
+	rlast=r0
+	counter=-1
+	maxcounter=10
+\end{lstlisting}
+Then the following code needs to be included in the $while$-loop
+\begin{lstlisting}
+    rbeforelast=rlast
+    rlast=r
+    r=r+v*dt
+    if (counter==-1): a0=rlast
+    if (rbeforelast.mag<rlast.mag and rlast.mag>r.mag): counter=counter+1
+    if counter==maxcounter:
+    	a=rlast
+    	break
+\end{lstlisting}
+and the condition for the $while$-loop should be set to $True$.
+The angle between two vectors is given by the formula
+ \begin{equation}
+ 	\sphericalangle(\vec{a}_0,\vec{a}) = \cos^{-1} \left( \frac{\vec{a}_0 \cdot \vec{a}}{|\vec{a}_0|\:|\vec{a}|} \right)
+ \end{equation}
+This is readily implemented in Vpython via
+\begin{lstlisting}
+	angle=acos(dot(a0,a)/(a0.mag*a.mag))/maxcounter*180/pi
+	print("angle: ", (angle))
+\end{lstlisting}
+
+\begin{figure}[htb]
+	\centering
+	\includegraphics[width=.5\textwidth]{figs/AlphaAngle.png}
+	\caption{\label{fig:AlphaAngle} Linear relation between $\alpha$ and the perihelion motion.}
+\end{figure}
+
+As explained above the natural value for $\alpha$ would be $1$.
+However if the student try out this value, they will find, that there is no visual change in the trajectories and the distance between the first an the last aphelion is of the order of magnitude of one step in the calculation $v_{aphelion}\cdot dt$.
+Therefore to estimate the size of the perihelion motion it is more advisable to use the fact, that there is an approximately linear dependence between $\alpha$ and the perihelion motion, if the angle is not to large.
+The students can convince themselves of this fact, by plotting some angles over $\alpha$.
+This can be done either by hand or using another program like e.g.
+Excel.
+One such plot can be found in figure \ref{fig:AlphaAngle}.
+Then to get to the angle at $\alpha=1$ one would ideally employ a line of best fit, but to get the order of magnitude right it is enough to consider only two points.
+Choosing $\alpha_1=1000$ and $\alpha_2=2000$ here, this would yield
+\begin{equation}
+	\Theta(\alpha=1) \approx \frac{0.201-0.100}{2000-1000} = 0.0001 = 0.36''
+\end{equation}
+so $3.63''\cdot 415\approx 150''$ per 100 earth years.
+To find out, whether this is a reasonable result, the students could first follow the dimensional analysis and then compare to the actual value for the mercury  as outlined above.
+
+\section{Possible extensions}
+
+\begin{itemize}
+
+\item \textbf{Explore Problem autonomously}
+
+In chapter \ref{sec:Numerical Implementation} we suggested a certain way to present the material.
+By using a template as well as a step by step instruction, the students are guided rather strictly through the problem solution.
+We presented this approach with young students and limited time in mind.
+However if the students are more experienced and enough time is available, it is certainly advisable to give the students space for exploring the problem independently.
+This could be achieved by the following changes or additions to the concept presented in chapter \ref{sec:Numerical Implementation}
+
+\begin{itemize}
+\item \textbf{Build code from scratch}
+
+Instead of providing the template to the students, they could build the program from scratch.
+Of course, this requires some basic knowledge in VPython, as they could acquire for example by working through a ball in the box example or the like.
+They could even look up the needed parameters on their own.
+\item \textbf{Why can we work in one plane?}
+
+In the code the third coordinate is never used. On the first glance, this could seem like a simplification. Let the students work out on their own, why this does not imply any loss of generality.
+\item \textbf{What is the impact of the different parameters?}
+
+Especially if the needed parameters are not specified beforehand, the students might have to experiment a bit, before getting the correct trajectories.
+But even if they are given, it might be beneficial to encourage the students to play with a few parameters, like mass or starting velocity, and track their impact on the trajectories.
+This way the students get a much better feeling for the physics involved.
+\end{itemize}
+
+\item \textbf{Optimizing performance}
+
+Simulations always involve a balance between the time needed for the calculations and the accuracy of the results achieved.
+Even though this is a rather simple example, it contains some opportunities to make this concept accessible to the students.
+
+\begin{itemize}
+\item \textbf{Consider error due to finite time steps}
+
+To make the students realise, that the finite time steps lead to inaccuracies in the trajectories encourage them to increase the time steps.
+This is best done in a program with unmodified gravitational force, as in this case the trajectories are closed ellipses and a deviation is most prominent.
+If the students choose the time steps big enough, they should witness big deviations.
+This exercise stresses the point, that by choosing increasingly smaller time steps, deviations from the physical trajectories can be reduced.
+\item \textbf{Measure calculation time}
+
+However in practice there is a limit to decreasing the time steps, because the time needed for the calculation grows simultaneously. By including
+import time
+\begin{lstlisting}
+	start_time = time.time()
+	main()
+	print("--- %s seconds ---" % (time.time() - start_time))
+\end{lstlisting}
+the students can measure the time needed by their program.
+By varying dt they can validate, that there is indeed approximately an anti-proportional dependence.
+(Note: This only works if the time in the loop is increased by dt, so t=t+dt.)
+\item \textbf{Verlet integration}
+
+Of course by optimising the code, an improvement in accuracy can be achieved without increasing the calculation time.
+The simplest way to demonstrate this might be implementing Verlet integration instead of using the simple Euler method.
+\end{itemize}
+
+\item \textbf{Extended Problems}
+
+\begin{itemize}
+\item \textbf{Non-stationary sun}
+
+It might be interesting to abandon the simplification of a stationary sun, as it nicely illustrates Newton's third law.
+Here it might also be advisable to reduce the mass of the sun to have a visible result.
+
+\item \textbf{Three-body problem}
+
+Ambitious students could even include another planet and see how the two planets interact.
+This is especially interesting, when discussing the perihelion motion of mercury, as it is mainly due to the influence of the other planets.
+Only a smaller part is due to general relativity.
+
+\end{itemize}
+
+\end{itemize}
 

### 2017-06-30 tex/MercuryNumerical.tex
Merged versions and included a diff pdf

@@
 \begin{equation}
-|\vec v(t)| \gg \frac12|\vec a(t)|\Delta t = \frac{1}{2m}\vec F(\vec x)\Delta t\ . 
+|\vec v(t)| \gg \frac12|\vec a(t)|\Delta t = \frac{1}{2m} F(\vec x(t))\Delta t\ . 
 \label{eq:check}
@@
 Both, for the numerical treatment and the intuitive understanding, it is useful to select parameters in a "natural range".
-A particular useful choice for this problem is given by expressing time intervals in days $T_0 = 1d$ {\bf These are earth days, I assume, right?} and distances in $R_0 = 10^{10}$ m.
+A particular useful choice for this problem is given by expressing time intervals in days $T_0 = 1d$, where the time refers to earth days, and distances in $R_0 = 10^{10}$ m.
 With this choice, the initial distance of mercury to the sun, the size of the initial velocity of mercury and the acceleration prefactor become

### 2017-07-01 tex/MercuryNumerical.tex
Worked through section 1-3

@@
 
+\usepackage[prependcaption, textsize=tiny]{todonotes}
+
 \newcommand{\python}[0]{\texttt{Python} }
@@
 \item effective tools to estimate the result a priori as an important cross check;
-\item the visualization of numerical results using V-Phython.
+\item the visualization of numerical results using \texttt{VPython}.
 \end{itemize}
@@
 concepts of differential equations their discretization is discussed based on Newtons law of gravitation and possible
-extensions thereof. Afterwards the visualization of the resulting trajectories using V-Phython is introduced and
+extensions thereof. Afterwards the visualization of the resulting trajectories using \texttt{VPython} is introduced and
 applied to the problem at hand. In particular tools are developed to extract the relevant quantity from the result
@@
 in particular its orientation in space) the location is parametrized
-by a single three vector $\vec x(t)$. As will become clear below in order to describe and control the dynamics of a physical
+by a single three vector $\vec x(t)$. As will become clear below, in order to describe and control the dynamics of a physical
 object in addition we need to be able to calculate its velocity, $\vec v(t)$, related to the location via
+%
+\todo{I see here a potential conflict with the implementation because we use $x(t-\Delta t)$ in order to get $v$ and so on}
+%
 \begin{equation}\label{eq:x_update}
@@
 If the mass does not change~\footnote{A well known example where $m$ does change with time is a 
-rocket, whose mass decreases as the rocket rises.} with time this reduces to the well known
+rocket, whose mass decreases as the rocket rises.} with time this, reduces to the well known
 \begin{equation}
@@
 given in Eqs.~(\ref{eq:vdef}) and (\ref{eq:adef}):
+%
+\todo{See previous comment.}
+%
 \begin{enumerate}
@@
 \begin{equation}
-|\vec v(t)| \gg \frac12|\vec a(t)|\Delta t = \frac{1}{2m} F(\vec x(t))\Delta t\ . 
+|\vec v(t)| \gg \frac12|\vec a(t)|\Delta t = \frac{1}{2m} |\vec F(\vec x(t)) |\Delta t\ . 
 \label{eq:check}
@@
 \begin{equation}
-\vec F_N(\vec x) = \frac{G_N m M_\odot}{r^2} \frac{\vec r}{r}\ ,
+\vec F_N(\vec x) = \frac{G_N m M_\odot}{r^2} \frac{\vec x}{r}\ ,
 \end{equation}
@@
 is infinitely more heavy than the planet and located at the center of the coordinate system. 
-Although this is not exact, since $m/M_\odot\sim 10^{-8}$ this is
-a pretty good approximation. For later convenience we introduce the Schwarzschildradius of the sun
+Although this is not exact, since $m/M_\odot\sim 10^{-8}$ this is a pretty good approximation.
+%
+\todo{This possibly can become a problem for our accuracy because it is at the 10\% level of our expected result, right? Same for $c$.}
+%
+For later convenience we introduce the Schwarzschildradius of the sun
 \begin{equation}
@@
 In general an attractive force that vanishes at large distances leads,
- depending on the initial conditions,
-  either to closed orbits or open orbits --- 
-  in the latter case the planet simply disappears from the sun. Depending on time available
-for the course  it might be helpful to explain to the students the origin of this, however, since the most
-transparent reasoning needs an integration as well as the concept of angular momentum
-not necessarily familiar to the whole class (details are given in the appendix {\bf to be done}), 
-it appears in general more appropriate to work out together with the students 
-that a moving body can not be kept by a force if its momentum is too large.
- Then they may try to find closed orbits in the simulation, e.g., by varying
-the start velocity while keeping the start location fixed.
+%
+\todo{I am confused. I thought only $1/r$ and the harmonic oscillator potential lead to closed orbits (Bertrand's theorem).}
+%
+ depending on the initial conditions, either to closed orbits or open orbits --- in the latter case the planet simply disappears from the sun.
+Depending on time available for the course  it might be helpful to explain to the students the origin of this, however, since the most transparent reasoning needs an integration as well as the concept of angular momentum not necessarily familiar to the whole class (details are given in the appendix
+%
+\todo{\bf Include this appendix part...}
+%
+), 
+it appears in general more appropriate to work out together with the students that a moving body can not be kept by a force if its momentum is too large.
+Then they may try to find closed orbits in the simulation, e.g., by varying the start velocity while keeping the start location fixed.
 
-The closed orbits that emerge from a force that scales as $1/r^2$
-are elliptic and in particular fixed in space. In particular the point of closest approach of the planet
-to the sun, the perihelion, does not move.
-  However, as soon
-as the potential is different, the perihelion moves. Because of this the behavior of the perihelion
+The closed orbits that emerge from a potential that scales as $1/r$ are elliptic and fixed in space.
+In particular the point of closest approach of the planet to the sun, the perihelion, does not move.
+However, as soon as the potential deviates from $1/r$, the perihelion moves. Because of this, the behavior of the perihelion
 is a very sensitive probe of the gravitational potential.
@@
 their gravitational force also acts on Mercury. However, a
-residual motion of $(43)''$ in 100 earth years remained unexplained, until Einstein quantified the
+residual motion of $\delta \phi_M = (43)''$ in 100 earth years remained unexplained, until Einstein quantified the
 predictions of General Relativity to this particular observable.
 
-{\bf Here should be a brief description of the relation between Newtons dynamics and General Relativity}
+%
+\todo{\bf Here should be a brief description of the relation between Newtons dynamics and General Relativity}
+%
 
-To allow for a movement of the perihelion also in the study presented here the
-potential of Eq.~(\ref{eq:newton}). We do this using the following ansatz
+We allow for a movement of the perihelion by modifying the potential of Eq.~(\ref{eq:newton}) with the following ansatz
 \begin{equation}
-\ddot{\vec x} = \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\left(1+\alpha\left(\frac{r_S}{r}\right)\right) \ ,
+\ddot{\vec x} = \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\left(1+\alpha\frac{r_S}{r}\right) \ ,
 \label{eq:newton_art}
 \end{equation}
-where $\alpha$ is some parameter. The idea is that the correction must be dimension less and
-should scale with some inverse power of $r$, for we still need to demand that the potential 
-vanishes at large distances. Since the only parameter of the system with dimension
-of a length is $r_S$, the correction should scale as $(r_S/r)$. When replacing $r$ by a typical
-distance of Mercury to the sun, $(r_S/r)\sim 10^{-7}$, it should be sufficient to include this kind
-of term to first order only. For comparison: the corresponding number for the venus, the neighbor
-of  Mercury, reads ...., which naturally explains why the effect is observable for Mercury only (IS IT - CHECK!)
+where $\alpha$ is some parameter related to the effects by general relativity.
+The idea is that the correction must be dimension-less and should scale with some inverse power of $r$, for we still need to demand that the potential vanishes at large distances.
+Since the only parameter of the system with dimension of a length is $r_S$, the correction should scale as $(r_S/r)$.
+When replacing $r$ by a typical distance of Mercury to the sun, $(r_S/r)\sim 10^{-7}$, it should be sufficient to include this kind of term to first order only.
+For comparison: the corresponding number for the Venus, the neighbor of Mercury, reads $r \sim 5 \cdot 10^{-8}$.
+Therefore the expected ratio of residual motion, favoring in the orbit period $T$, would result in
+\begin{equation}
+	\delta \phi_V \approx \frac{1}{T_V} \delta \phi_M T_M \frac{r_M}{r_V} \approx 9'' \, ,
+\end{equation}
+which precisely describes the experimental value of $9''$.
+
 %\end{itemize}
@@
   intuitive to understand and an open source language\footnote{See also appendix \ref{appendix:python} for an instruction 
-  how to set up python on different operating systems.}.  No prior knowledge of \python or any other programming language 
+  how to set up \python on different operating systems.}.  No prior knowledge of \python or any other programming language 
   is required because we explain all the steps necessary to create the simulation.  We also provide a working 
@@
 
-In python, this reads
+In \python, this reads
 \begin{lstlisting}
@@
  \end{equation}
-This is readily implemented in Vpython via
+This is readily implemented in \texttt{Vpython} via
 \begin{lstlisting}
@@
 Instead of providing the template to the students, they could build the program from scratch.
-Of course, this requires some basic knowledge in VPython, as they could acquire for example by working through a ball in the box example or the like.
+Of course, this requires some basic knowledge in \texttt{VPython}, as they could acquire for example by working through a ball in the box example or the like.
 They could even look up the needed parameters on their own.
@@
 \appendix
-\section{Instructions to install V-phython}
+\section{Instructions to install VPython}
 \label{appendix:python}

### 2017-07-05 tex/MercuryNumerical.tex
modified numerical implementation section

@@
 
+\usepackage{hyperref}
+\usepackage[space=true]{accsupp}
+% requires the latest version of package accsupp
+\newcommand{\copyablespace}{\BeginAccSupp{method=hex,unicode,ActualText=00A0}\ \EndAccSupp{}}
+
 \usepackage{ amssymb }
@@
 \definecolor{light-gray}{gray}{0.97}
+\definecolor{semilight-gray}{gray}{0.67}
 \definecolor{mygreen}{rgb}{0,0.6,0}
@@
   backgroundcolor=\color{light-gray},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}
-  basicstyle=\linespread{1}\footnotesize,        % the size of the fonts that are used for the code
+  basicstyle=\linespread{1}\footnotesize\ttfamily,        % the size of the fonts that are used for the code
   commentstyle=\color{mygreen},
@@
   tabsize=4,	                   % sets default tabsize to 2 spaces
-  rulecolor=\color{light-gray},
+  showtabs=true,
+  rulecolor=\color{black},
+  columns=fullflexible,
+  literate={\ }{{\copyablespace}}1
 }
@@
 
-\section{Introduction}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Introduction}\label{sec:intro}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\section{Trajectories, velocities, accelerations and Newtons second law}
-\label{sec:tva}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Trajectories, velocities, accelerations and Newtons second law}\label{sec:tva}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\section{Example: General Relativity and the perihelion motion of Mercury}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Example: General Relativity and the perihelion motion of Mercury}\label{sec:gr}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-%\end{itemize}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Numerical Implementation}\label{sec:Numerical Implementation}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\section{Numerical Implementation}
-\label{sec:Numerical Implementation}
+%------------------------------------------
+\subsection{Describing the Motion with Python}
+%------------------------------------------
+
 In this section we present the numerical implementation as well as the visualization of planetary trajectories and in
 particular
-  the perihelion motion of Mercury.  We select \python~\cite{} as  programming language, because \python is easy to learn, 
+the perihelion motion of Mercury.  We select \python~\cite{}\todo{Citation} as  programming language, because \python is easy to learn, 
 intuitive to understand and an open source language\footnote{See also appendix \ref{appendix:python} for an instruction 
@@
 is required because we explain all the steps necessary to create the simulation.  We also provide a working 
-  example \cite{} that can be used as template.
+example \cite{}
+%
+\todo{One can additional files to the arXiv upload} that can be used as template.
+%
 
@@
 As described above, here  one can simplify the problem by placing one planet (the "infinitely" heavy sun) in the center of the coordinate system and describe the motion of the other planet (mercury) in a two-dimensional plane.
-\textit{Why is this simplified picture sufficient to describe the motion}?
 
-Because the problem does not depend on the initial position of mercury on its orbit around the sun, we extract the parameters at the perihelion with $ \abs{\vec r_{MS}(0)} = 46 \cdot 10^6 km$ and $ \abs{\vec v_M(0)} = 39$ km/s \footnote{https://nssdc.gsfc.nasa.gov/planetary/factsheet/mercuryfact.html}.
+Because the problem does not depend on the initial position of mercury on its orbit around the sun, we extract the parameters at the perihelion with $ \abs{\vec r_{MS}(0)} = 46 \cdot 10^6$km and $ \abs{\vec v_M(0)} = 59$km/s \footnote{https://nssdc.gsfc.nasa.gov/planetary/factsheet/mercuryfact.html}.
 Since the computer does not understand physical units, one has to express each variable in an appropriate unit.
 Both, for the numerical treatment and the intuitive understanding, it is useful to select parameters in a "natural range".
-A particular useful choice for this problem is given by expressing time intervals in days $T_0 = 1d$, where the time refers to earth days, and distances in $R_0 = 10^{10}$ m.
+A particular useful choice for this problem is given by expressing time intervals in days $T_0 = 1$d, where $d$ refers to earth days, and distances in $R_0 = 10^{10}$ m.
 With this choice, the initial distance of mercury to the sun, the size of the initial velocity of mercury and the acceleration prefactor become
@@
 
+%
 \begin{figure}[htb]
@@
 \end{figure}
+%
 
@@
 rM0 = 4.6    # in units of R0
-	vM0 = 0.34   # in units of R0/T0
+vM0 = 0.51   # in units of R0/T0
 aM  = 0.99   # in units of R0/T0**2
+T   = 88.    # in units of T0
+rS  = 3.e-7  # in units of R0
 \end{lstlisting}
@@
 Next, we set up the initial directions, which will describe the motion in the two-dimensional plane.
-We will build on the existing \python module \texttt{Vpython}, which provides  an implementation for treating vectors as well as their visualization.
+We will build on the existing \python module \texttt{VPython}, which provides  an implementation for treating vectors as well as their visualization.
 The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.
@@
 \begin{lstlisting}
-	# Import the class vector from python
-	from vpython import vector
+# Import the class vector from vpythons module visual
+from visual import vector
 # Initialize distance and velocity vectors
 vec_rM0 = vector(0, rM0, 0)
-	vec_vM0 = vector(vm0, 0, 0)
+vec_vM0 = vector(vM0, 0, 0)
 \end{lstlisting}
@@
 However, we first need to fix the time step $\Delta t$ and use the estimate
-of Eq.~(\ref{eq:check}) as a guidance {\bf Careful: Check units of dt}
+of Eq.~(\ref{eq:check}) as a guidance \todo{\bf Careful: Check units of dt}
 \begin{lstlisting}
 # Definition of the time step
-        dt = 2 * vm0 / aM / 100
+dt = 2 * vM0 / aM / 1000
 \end{lstlisting}
-Here the factor $1/100$ makes sure that $\Delta t$ is indeed
-consistent with Eq.~(\ref{eq:check}).  {\bf We need to make sure that 'dt' is known
-to the subroutines}
+Here the factor $1/1000$ makes sure that $\Delta t$ is indeed consistent with Eq.~(\ref{eq:check}).
+%
+\todo{I need 1/1000 to get relatively closed orbits -- we have to crosscheck codes!}
+%
 Now we are in the position to calculate
@@
 def evolve_mercury(vec_rM_old, vec_vM_old, alpha):
-		<...Code...>
+	#<...Code...>
 	return vec_rM_new, vec_vM_new
@@
 \end{lstlisting}
-Because of \python{$\!'$}s syntax, it is necessary that the body of the function is indented relative to the definition statement {\bf I do not understand this sentence}.
+\python{$\!'$}s syntax enforces a clean programming style: it is necessary that the body of the function is indented relative to the definition statement of the function.
+This is visualized by the long underscore in the previous block.
+Furthermore, \python is an Interpreter language.
+Each line of the code is execute once the Interpreter passes them.
+For this reason, all the variables defined before the function are global and thus known to the function.
+The variables within the function (e.g. \texttt{vec\_rM\_old}, ...) are local and do not live beyond the scope of the function.
 
@@
 t     = 0
-	alpha=0
+alpha = 0
 # Execute the loop as long as t < T
@@
 The total runtime \texttt{T} is the amount of "virtual" days the simulations should run.
-To describe at least one full orbit, $T$ needs to be larger than $T>..$.
+To describe at least one full orbital period, $T$ needs to be larger than $T>88$d.
+With the previous choice for \texttt{dt}, this corresponds to roughly $N_T = 10^4$ evolution steps.
 
-{\bf Here we need to write something about visualizing the trajectories and how to run the code }
+%------------------------------------------
+\subsection{Visualizing the Motion with VPython}
+%------------------------------------------
+To start the visualization one has to include further objects of the \texttt{VPython} module \texttt{visual}.
+Additional to the vector class, one has to include
+\begin{lstlisting}
+from visual import vector, sphere, color, curve, rate, display
+\end{lstlisting}
+The class sphere will represent mercury and the sun in the simulation
+\begin{lstlisting}
+# define the initial coordinates; M = mercury, S = sun
+M = sphere( pos=vec_rM0,       radius=0.01, color=color.red   )
+S = sphere( pos=vector(0,0,0), radius=0.1,  color=color.yellow)
+# and the initial velocities
+M.velocity = vec_vM0
+S.velocity = vector(0,0,0)
+\end{lstlisting}
+We have placed the sun in the origin of our coordinate system and choose non-realistic radii sizes for visualization purposes.
 
-\section{Tests of stability}
+To further visualize the trajectory of mercury we add this property to our \texttt{mercury} object
+\begin{lstlisting}
+# add a visible trajectory to mercury
+M.trajectory = curve(color=color.black)
+\end{lstlisting}
+Next we tell python that the objects shall be drawn from this point on
+\begin{lstlisting}
+# display the objects from now on
+display(background=color.white)
+\end{lstlisting}
+and last but not least, instead of updating the object-unrelated vectors in the while loop, we update the object-related vectors which will eventually be drawn
+\begin{lstlisting}
+t     = 0
+alpha = 0
+# update the objects
+while t < T:
+	# set the frame rate: shows four earth days at once
+	rate(4*1000)
+	# update the drawn trajectory with the current position
+	M.trajectory.append(pos=M.pos)
+	# update the velocity and position
+	M.pos , M.velocity = evolve_mercury(M.pos , M.velocity , alpha)
+	# update tim
+	t = t + dt
+\end{lstlisting}
 
-\section{Tasks}
+If the starting values are chosen as advised in the previous section, then the students should end up with a trajectory as depicted in figure \ref{fig:MercuryOrbit}.
+To get the perihelion motion, the term from equation \ref{eq:newton_art} has to be be added.
+This is a good opportunity to let the students play with the size of $\alpha$ and get a feeling for the impact on the trajectories.
+However it might be advisable to not start out with the correct mercury values but e.g.
+a slightly higher initial distance between mercury and sun, since the perihelion motion is better visible then.
+This is illustrated in figure \ref{fig:MercuryOrbit}, where we used a value of $\alpha=10^5$ and $dt=0.1T_0$.
+%
+\todo{This is really to large for me to get these orbits. Also I need to run with $\alpha = 10^6$ to get such shapes...}
+%
+\begin{figure}[htb]
+	\centering
+	\includegraphics[width=.25\textwidth]{figs/MercuryOrbit.png} \quad
+	\includegraphics[width=.273\textwidth]{figs/PerihelionMotionLargeV.png} \quad
+	\includegraphics[width=.26\textwidth]{figs/PerihelionMotion.png}
+	\caption{\label{fig:MercuryOrbit} Output for $\alpha=0$ (left), $\alpha=10^5$ and bigger starting distance $r_{MS}(0)=6R_0$ (middle) and $\alpha=10^5$ with original starting values.}
+\end{figure}
+\todo{@CK: reproduce figure}
 
-Ideas for questions:
-\begin{itemize}
-	\item Mass and motion of sun
-	\item Estimate \texttt{dt} and also \texttt{T} in order to cover roughly two "mercury years".
-	\item What does happen when you change those units?
-	\item Compare values to the values in \cite{}.  Where does the difference come from?
-	\item Start with $\alpha=0$. What do you observe?
-\end{itemize}
+%------------------------------------------
+\subsection{Running the Program}
+%------------------------------------------
+The complete program needs to be saved as a python file, e.g. \texttt{mercury\_orbit.py}.
+Depending on the editor of choice, there might already be a execute functionality e.g. \textbf{bla...} for the recommended windows installation (see \ref{appendix:python}).
+%
+\todo{Look up the windows example...}
+%
+In case one uses a linux like system and wants to execute the code from a console/terminal, one has to to execute the command
+\begin{center}
+	\texttt{python 'address/to/python/file.py'}
+\end{center}
+If one already works in the file directory and has used the suggested naming scheme, this reads 
+\begin{center}
+	\texttt{python mercury\_orbit.py}
+\end{center}
+Here, the specified \texttt{python} interpreter needs to agree with the installed \texttt{VPython} version.
+For more information see \ref{appendix:python}.
 
-\section{Dimensional analysis}
+%------------------------------------------
+\subsection{Extracting the Perihelion Motion}
+%------------------------------------------
+
+\textit{What exactly do we want the students / advise the teachers to do?
+I think a nice approach would be to let the students implement the code to extract the perihelion motion and make a list of some values of alpha and the corresponding angles.
+Then discuss, that $\alpha=1$ is the natural size for $\alpha$ and that to extract the value of angle they need to exploit the linear dependence.
+Finally they can think on whether the value they get is reasonable and the teacher can explain the rest of section?
+What do you think?}\\
+
+To calculate the expected size of the perihelion motion for a modified gravitational force, the students need to be abel to extract the relative angle between the current mercury position and the starting position.
+There are several ways to do this, but the easiest is to extract the point where the distance to the sun gets minimal -- the definition of the perihelion\todo{why aphelion and not perihelion?
+I have changed the following code to the perihelion (because it is shorter).
+We can change it back if desired.} -- and compare the new position of the perihelion after $n$ turns to the original one.
+This can be done by looking for a minimal distance of mercury to the sun within a $while$-loop.
+
+Suppose we want to save the number of turns in the variable \texttt{turns}, the maximal number of turns in \texttt{max\_turns} (here e.g.
+10), the initial perihelion is already stored in the vector \texttt{vec\_rM0} and the goal is to find the final perihelion.
+In addition to find out when the perihelion is reached, the last two positions \texttt{vec\_r\_last} and \texttt{vec\_r\_before\_last} have to be known.
+Thus, if the last vector is smaller than the previous and the current vector, one has reached the perihelion again.
+The implementation can be realized as follows
+\begin{lstlisting}
+# set up vectors
+vec_r_last = vec_rM0
+turns      = 0
+max_turns  = 10
+# find perihelion after ten turns
+while turns < max_turns:
+	vec_r_before_last = vec_r_last
+	vec_r_last        = M.pos
+	#<...update mercury position...>
+	# check if at perihelion
+	if vec_r_before_last.mag > vec_r_last.mag < M.pos.mag:
+		turns = turns+1
+\end{lstlisting}
+When the loop stops, the last perihelion vector is stored in  \texttt{vec\_r\_last}.
+To compute the angle between two vectors one uses the formula
+ \begin{equation}
+ 	\sphericalangle(\vec{a}_0,\vec{a}) = \cos^{-1} \left( \frac{\vec{a}_0 \cdot \vec{a}}{|\vec{a}_0|\:|\vec{a}|} \right)
+ \end{equation}
+This is readily implemented in \texttt{VPython} via
+\begin{lstlisting}
+# import function to resolve angle
+from visual import acos, pi, dot
+angle = acos(
+	dot(vec_rM0, vec_r_last.mag) / (vec_rM0.mag * vec_r_last.mag)
+) / maxcounter * 380 / 2 / pi
+\end{lstlisting}
+
+\begin{figure}[htb]
+	\centering
+	\includegraphics[width=.5\textwidth]{figs/AlphaAngle.png}
+	\caption{\label{fig:AlphaAngle} Linear relation between $\alpha$ and the perihelion motion.}
+\end{figure}
+
+As explained above the natural value for $\alpha$ would be $1$.
+However if the student us this value for $\alpha$, they will find that there is close to no visual change in the trajectories and the distance between the first an the last perihelions is of the order of magnitude of one step in the calculation $v_{aphelion}\cdot dt$.
+Therefore, to estimate the size of the perihelion motion it is more advisable to use the fact, that there is an approximately linear dependence between $\alpha$ and the perihelion motion, if the angle is not to large.
+The students can convince themselves of this fact, by plotting some angles over $\alpha$.
+This can be done either by hand or using another program like e.g. Excel.
+One such plot can be found in figure \ref{fig:AlphaAngle}.
+
+To get to the angle at $\alpha=1$ one would ideally employ a linear regression (or ruler fit), but to get the order of magnitude right it is enough to consider only two points.
+Choosing $\alpha_1=1000$ and $\alpha_2=2000$ here, this would yield
+\begin{equation}
+	\Theta(\alpha=1) \approx \frac{0.201-0.100}{2000-1000} = 0.0001 = 0.36''
+\end{equation}
+so $3.63''\cdot 415\approx 150''$ per 100 earth years.
+To find out, whether this is a reasonable result, the students could first follow the dimensional analysis and then compare to the actual value for the mercury  as outlined above.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Tests of stability}\label{sec:stability}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Dimensional analysis}\label{sec:analysis}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\section{Discussion of Starting Values}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Tasks}\label{sec:tasks}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\textit{I am not quite sure what to add here to Christopher's discussion.}\\
+Ideas for questions:
+\begin{itemize}
+	\item Mass and motion of sun
+	\item Estimate \texttt{dt} and also \texttt{T} in order to cover roughly two "mercury years".
+	\item What does happen when you change those units?
+	\item Compare values to the values in \cite{}.  Where does the difference come from?
+	\item Start with $\alpha=0$. What do you observe?
+\end{itemize}
 
-If the starting values are chosen as advised in the previous section, then the students should end up with a trajectory as depicted in figure \ref{fig:MercuryOrbit}.
-To get the perihelion motion, the term from equation \ref{eq:newton_art} has to be be added.
-This is a good opportunity to let the students play with the size of $\alpha$ and get a feeling for the impact on the trajectories.
-However it might be advisable to not start out with the correct mercury values but e.g.
-a slightly higher initial distance between mercury and sun, since the perihelion motion is better visible then.
-This is illustrated in figure \ref{fig:MercuryOrbit}, where we used a value of $\alpha=10^5$ and $dt=0.1T_0$.
-
-\begin{figure}[htb]
-	\centering
-	\includegraphics[width=.25\textwidth]{figs/MercuryOrbit.png} \quad
-	\includegraphics[width=.273\textwidth]{figs/PerihelionMotionLargeV.png} \quad
-	\includegraphics[width=.26\textwidth]{figs/PerihelionMotion.png}
-	\caption{\label{fig:MercuryOrbit} Output for $\alpha=0$ (left), $\alpha=10^5$ and bigger starting distance $r_{MS}(0)=6R_0$ (middle) and $\alpha=10^5$ with original starting values.}
-\end{figure}
-
-\section{Extracting the Perihelion Motion}
-
-\textit{What exactly do we want the students / advise the teachers to do?
-I think a nice approach would be to let the students implement the code to extract the perihelion motion and make a list of some values of alpha and the corresponding angles.
-Then discuss, that $\alpha=1$ is the natural size for $\alpha$ and that to extract the value of angle they need to exploit the linear dependence.
-Finally they can think on whether the value they get is reasonable and the teacher can explain the rest of section?
-What do you think?}\\
-
-To calculate the expected size of the perihelion motion for a modified gravitational force, the students need to be abel to extract the angle.
-There are several ways to do this, but the easiest is to extract the point where the distance to the sun gets maximal, the aphelion, after $n$ turns and compare it to the original one.
-This can be done by using an $if$-statement in the $while$-loop.
-
-Suppose we want to save the number of turns in the variable $counter$, the maximal number of turns in $maxcounter$ (here e.g.
-10), the initial aphelion in the vector $a0$ and the final aphelion in the vector $a$.
-In addition to find out when the aphelion is reached, the last two positions $rlast$ and $rbeforelast$ have to be known.
-To set this up, before the while loop the following code needs to be included
-\begin{lstlisting}
-	r=r0
-	rlast=r0
-	counter=-1
-	maxcounter=10
-\end{lstlisting}
-Then the following code needs to be included in the $while$-loop
-\begin{lstlisting}
-    rbeforelast=rlast
-    rlast=r
-    r=r+v*dt
-    if (counter==-1): a0=rlast
-    if (rbeforelast.mag<rlast.mag and rlast.mag>r.mag): counter=counter+1
-    if counter==maxcounter:
-    	a=rlast
-    	break
-\end{lstlisting}
-and the condition for the $while$-loop should be set to $True$.
-The angle between two vectors is given by the formula
- \begin{equation}
- 	\sphericalangle(\vec{a}_0,\vec{a}) = \cos^{-1} \left( \frac{\vec{a}_0 \cdot \vec{a}}{|\vec{a}_0|\:|\vec{a}|} \right)
- \end{equation}
-This is readily implemented in \texttt{Vpython} via
-\begin{lstlisting}
-	angle=acos(dot(a0,a)/(a0.mag*a.mag))/maxcounter*180/pi
-	print("angle: ", (angle))
-\end{lstlisting}
-
-\begin{figure}[htb]
-	\centering
-	\includegraphics[width=.5\textwidth]{figs/AlphaAngle.png}
-	\caption{\label{fig:AlphaAngle} Linear relation between $\alpha$ and the perihelion motion.}
-\end{figure}
-
-As explained above the natural value for $\alpha$ would be $1$.
-However if the student try out this value, they will find, that there is no visual change in the trajectories and the distance between the first an the last aphelion is of the order of magnitude of one step in the calculation $v_{aphelion}\cdot dt$.
-Therefore to estimate the size of the perihelion motion it is more advisable to use the fact, that there is an approximately linear dependence between $\alpha$ and the perihelion motion, if the angle is not to large.
-The students can convince themselves of this fact, by plotting some angles over $\alpha$.
-This can be done either by hand or using another program like e.g.
-Excel.
-One such plot can be found in figure \ref{fig:AlphaAngle}.
-Then to get to the angle at $\alpha=1$ one would ideally employ a line of best fit, but to get the order of magnitude right it is enough to consider only two points.
-Choosing $\alpha_1=1000$ and $\alpha_2=2000$ here, this would yield
-\begin{equation}
-	\Theta(\alpha=1) \approx \frac{0.201-0.100}{2000-1000} = 0.0001 = 0.36''
-\end{equation}
-so $3.63''\cdot 415\approx 150''$ per 100 earth years.
-To find out, whether this is a reasonable result, the students could first follow the dimensional analysis and then compare to the actual value for the mercury  as outlined above.
-
-\section{Possible extensions}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Possible extensions}\label{sec:extensions}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-%\begin{table}
-%\caption{\label{jlab1}Journals to which this document applies, and macros for the abbreviated journal names in {\tt iopart.cls}. Macros for other journal titles are listed in appendix\,A.}
-%\footnotesize
-%\begin{tabular}{@{}llll}
-%\br
-%Short form of journal title&Macro name&Short form of journal title&Macro name\\
-%\mr
-%2D Mater.&\verb"\TDM"&Mater. Res. Express&\verb"\MRE"\\
-%\br
-%\end{tabular}\\
-%$^{a}$UK spelling is required; $^{b}$MSC classification may be used as well as PACS; $^{c}$titles of articles are required in journal references; $^{d}$Harvard-style references must be used (see section \ref{except}); $^{e}$final page numbers of articles are required in journal references.
-%
-%\end{table}
-%\normalsize
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\appendix
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\subsection{Acknowledgments}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Acknowledgments}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 Here they come
 
-\appendix
-\section{Instructions to install VPython}
-\label{appendix:python}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Instructions to install VPython}\label{appendix:python}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Mac}
+\href{http://vpython.org/contents/download_mac.html}{Download VPython for Mac}
 
-\section{The code}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{The code}\label{sec:code}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 Here we should put the code
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Bib
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \bibliographystyle{alpha din}
 \bibliography{MercuryNumerical}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2017-07-05 scripts/merkur_orbit.py

@@ -70,6 +70,8 @@ v_sonne_0  = vector(v_s,0,0)
 v_merkur_0 = vector(v_m ,0,0)
 print "v_Merkur v = {0:1.4e} AE / Tage ".format(v_m)
 
+print T, dt, r_merkur_0, v_merkur_0, G * m_sonne / r_merkur_0.mag**2
+
 # ----------------------------------
 ## (a) Berechnung des Orbits
 # ----------------------------------

### 2017-07-05 scripts/merkur_orbit_paper.py

@@ -1,0 +1,68 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+#  import the essential classes from vpythons module visual
+from visual import vector, sphere, color, curve, rate, display
+
+# Definition of parameters
+rM0 = 4.6   # in units of R0
+vM0 = 0.51  # in units of R0/T0
+aM  = 0.99  # in units of R0/T0**2
+T   = 88.   # in units of T0
+rS  = 3.e-7 # in units of R0
+
+# Definition of the time step
+dt = 2. * vM0 / aM / 1000
+
+# Define the update function
+def evolve_mercury(vec_rM_old , vec_vM_old , alpha ):
+	# get the magnitude of the acceleration
+	aMS        = aM * ( 1 + alpha * rS / vec_rM_old.mag ) / vec_rM_old.mag**2
+	# get the vector of the acceleration
+	vec_aMS    = - aMS * ( vec_rM_old / vec_rM_old.mag )
+	# get the new position vector of mercury
+	vec_rM_new = vec_rM_old + vec_vM_old * dt
+	# get the new velocity vector of mercury
+	vec_vM_new = vec_vM_old + vec_aMS    * dt
+	# return the new vectors
+	return vec_rM_new , vec_vM_new
+
+# Initialize distance and velocity vectors
+vec_rM0 = vector(0, rM0, 0) 
+vec_vM0 = vector(vM0, 0, 0)
+
+# display the objects from now on
+view = display(
+	background=color.white,
+	width=800,
+	height=1000,
+)
+view.center     = (0,0,0)
+view.userzoom = True
+view.userspin = True
+
+# define the initial coordinates; M= mercury, S = sun
+M = sphere(pos=vec_rM0,       radius=0.1, color=color.red   ) 
+S = sphere(pos=vector(0,0,0), radius=1.0,  color=color.yellow)
+# and the initial velocities
+M.velocity  = vec_vM0
+S.velocity  = vector(0 ,0 ,0)
+
+# add a visible trajectory to mercury
+M.trajectory = curve(color=color.black)
+
+t     = 0
+alpha = 0
+# update the objects
+while t < 30*T:
+	# set the frame rate: shows four earth days at once
+	rate(40*1000)
+	# update the drawn trajectory with the current position
+	M.trajectory.append(pos=M.pos)
+	# update the velocity and position
+	M.pos , M.velocity = evolve_mercury(M.pos , M.velocity , alpha)
+	# update tim
+	t = t + dt
+
+
+

### 2017-10-08 tex/MercuryNumerical.tex

@@
 	dot(vec_rM0, vec_r_last.mag) / (vec_rM0.mag * vec_r_last.mag)
-) / maxcounter * 380 / 2 / pi
+) / maxcounter * 360 / 2 / pi
 \end{lstlisting}
@@
 To find out, whether this is a reasonable result, the students could first follow the dimensional analysis and then compare to the actual value for the mercury  as outlined above.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\clearpage\subsection{Cross-Check Results}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\todo[inline]{Here I (CK) just make a few comments regarding the results.}
+First of all, I can confirm Inkas result up to a factor of 10.
+I believe this factor arises because we have analysed 10 years of motions (\verb|max_turns = 10|).
+See figure \ref{fcc1}.
+This result is quite nice, because we basically say the genreal relativity motion by mercury is caused by $\alpha \approx 3$.
+However there is a small issue: we do not have the numerical precision to claim this.
+\begin{figure}[htb]
+	\centering
+	\includegraphics[width=.4\textwidth]{figs/alpha_interpolation.pdf}
+	\caption{\label{fcc1}Interpolation for $\phi$ depending on $\alpha$ and $dt = 2 v_0 / a_0 / 5000$. }
+\end{figure}
+
+I have encountered, that we do not start perfectly at the perihelion or aphelion (depending on the boundary conditions): my simulations arrives at this point after roughly a tenth of a day and therefore I have an offset which needs to be subtracted off after each mercury year.
+More interestingly, this offset seems to depend on the tilmestep of the simulation (see figure \ref{fcc2}).
+\begin{figure}[htb]
+	\centering
+	\includegraphics[width=.4\textwidth]{figs/initial_time_shift.pdf}
+	\caption{\label{fcc2}Initial shift of perihelion for $\delta\theta = \delta\phi$ in degrees over $\alpha$ for different time steps (color).}
+\end{figure}
+
+One can also see that the size of the time spacing affects a potential uncertainty in the periodicity of the orbit when looking at the precession growth over different turns (see figure \ref{fcc3}).
+\begin{figure}[htb]
+	\centering
+	\includegraphics[width=.7\textwidth]{figs/precession_growth.pdf}
+	\caption{\label{fcc3}Precession growth of perihelion for $\delta\theta = \delta\phi$ in degrees over number of turns for different time steps (color).
+One should pay special attention to the oscillations within the growth for relatively large time steps.}
+\end{figure}
+
+Last but not least, for the precession interpolation to small $\alpha$, one should worry about the precision of the precession for small alpha.
+Even after subtracting of the results for zero turns, basically non of my results agree with zero for alpha equal to zero [to go from degrees per mercury year to arcsecs per 100 earth years, one has to multiply by $1.46 \times 10^{6}$.](see figure \ref{fcc4}).
+\begin{figure}[htb]
+	\centering
+	\includegraphics[width=.4\textwidth]{figs/alpha_dependence_0}
+	%
+	\includegraphics[width=.4\textwidth]{figs/alpha_dependence_1}
+	\caption{\label{fcc4}Perihelion precession $\delta\theta = \delta\phi$ in degrees over $\alpha$ without (left figure) or with (right figure) subtracting off the zero turn result.
+The uncertainty corresponds to the variation over different turns and the results are displayed for different time steps (color).}
+\end{figure}
+
+Because, after subtracting the initial angle as a reference, the slope of all the plots basically agree, I suggest the following prescription (as Inka actually did)
+\begin{enumerate}
+	\item Compute the precession for $\alpha = 0$ for several turns,
+	\item compute the precession for preferably large $\alpha > 1000$ for several turns,
+	\item verify that they are indeed on a line (linear relation),
+	\item subtract the zero value from the other values (since we know that it has to be zero and the uncertainty is coming from the time steps),
+	\item do the interpolation to $\delta \phi (\alpha_{GR}) = 43''$ for two values: $\delta \phi (0) = 0$ and $\delta \phi (\alpha_l) = \delta \phi_l$
+	\item repeat this for several $(\alpha_l, \delta \phi_l)$ and get the average result for $\alpha_{GR}$.	
+\end{enumerate}
+
+\clearpage
 

### 2017-10-25 scripts/old/merkur_orbit (1).py
worked out structure

@@ -1,145 +1,0 @@
-### Darstellung des Merkurorbits mit VPython
-
-# ----------------------------------
-## Laden der Hilfsmodule
-# ----------------------------------
-
-# Da es unglaublich viel Zeit kostet ein Programm für 3-dimensionale graphische Darstellung selber 
-# zu schreiben, lassen wir uns die Arbeit von "Vpython" abnehmen. Doch dazu müssen wir zunächst 
-# dieses Modul bereitstellen. Das erreicht man mit Hilfe des Befehls "import" oder "from .. import".
-
-from visual import *
-
-# Schreibt man 
-#   "from visual import *" 
-# bedeutet dies: "Lade alle Funktionen von Vpython so, dass man sie direkt verwenden kann: 
-#   "func = function(..)". 
-# Würde man 
-#   "import visual" 
-# schreiben, so müsste man die Funktion wie folgt aufrufen: 
-#   "func = visual.function(..)".
-
-# Eine Dokumentation von Modulen lässt sich durch den Befehl "help( modul_name )" aufrufen.
-#
-# Tipp: Will man eine Eingabe unterdrücken, so lässt sich dies auch mit "#" tun. Der darauf 
-# folgende Text wird dabei von Python überlesen.
-
-# help(visual)
-
-# ----------------------------------
-## Aufgabe 1: Merkurorbit)
-# ----------------------------------
-
-# Funktionen werden in Python wie folgt deklariert:
-# def func(x):
-#     .... Rechnung und Anderes ....
-# 
-# Diese wollen wir nun nutzen. Um die Umlaufbahn des Merkurs darzustellen, ist es sinvoll die 
-# Aufgabe in zwei Abschnitte zu Teilen:
-# (a) Berchnung des Orbits
-# (b) Zeichnung des Orbits
-# Für beide Abschnitte werden wir Funktionen nutzen.
-
-
-# ----------------------------------
-## Parameter
-# ----------------------------------
-
-# Physikalische Größen/Einheiten
-ms = 1.98892 * 1.e30 # Sonnenmassen
-AE = 1.49598 * 1.e11 # Astronomische Einheit
-T  = 60 * 60 * 24    # 1 Tag
-dt = 1./24*6         # 8 Stunden
-# Die Gravitationskonstante
-G  = 6.6738 * 1.e-11 * ms / AE**3 * T**2
[…]

### 2017-10-25 scripts/old/merkur_orbit 2.py/merkur_orbit.py

@@ -57,7 +57,7 @@ T  = 60 * 60 * 24    # 1 Tag
 dt = 1./24*6         # 6 Stunden
 # Die Gravitationskonstante
 G  = 6.6738 * 1.e-11 * ms / AE**3 * T**2
-print "Gravitationskonstante G = {0:1.4e} AE^3 * Tage^2 /M_Sonne".format(G)
+print("Gravitationskonstante G = {0:1.4e} AE^3 * Tage^2 /M_Sonne".format(G))
 # Die Massen der Planten (in Sonnenmassen)
 m_sonne  = 1.
 m_merkur = 1.66e-7
@@ -69,7 +69,7 @@ v_m = 58.98 * 1.e3 / AE * T
 v_s = - v_m * m_merkur / m_sonne # Impulserhaltung
 v_sonne_0  = vector(v_s,0,0)
 v_merkur_0 = vector(v_m ,0,0)
-print "v_Merkur v = {0:1.4e} AE / Tage ".format(v_m)
+print("v_Merkur v = {0:1.4e} AE / Tage ".format(v_m))
 
 # ----------------------------------
 ## (a) Berechnung des Orbits

### 2017-10-25 scripts/old/merkur_orbit 3.py/merkur_orbit.py

@@ -1,131 +1,0 @@
-# -*- coding: utf-8 -*-
-
-### Darstellung des Merkurorbits mit VPython
-
-# ----------------------------------
-## Laden der Hilfsmodule
-# ----------------------------------
-
-# Da es unglaublich viel Zeit kostet ein Programm für 3-dimensionale graphische Darstellung selber 
-# zu schreiben, lassen wir uns die Arbeit von "Vpython" abnehmen. Doch dazu müssen wir zunächst 
-# dieses Modul bereitstellen. Das erreicht man mit Hilfe des Befehls "import" oder "from .. import".
-
-from visual import *
-
-# Schreibt man 
-#   "from visual import *" 
-# bedeutet dies: "Lade alle Funktionen von Vpython so, dass man sie direkt verwenden kann: 
-#   "func = function(..)". 
-# Würde man 
-#   "import visual" 
-# schreiben, so müsste man die Funktion wie folgt aufrufen: 
-#   "func = visual.function(..)".
-
-# Eine Dokumentation von Modulen lässt sich durch den Befehl "help( modul_name )" aufrufen.
-#
-# Tipp: Will man eine Eingabe unterdrücken, so lässt sich dies auch mit "#" tun. Der darauf 
-# folgende Text wird dabei von Python überlesen.
-
-# help(visual)
-
-# ----------------------------------
-## Aufgabe 1: Merkurorbit)
-# ----------------------------------
-
-# Funktionen werden in Python wie folgt deklariert:
-# def func(x):
-#     .... Rechnung und Anderes ....
-# 
-# Diese wollen wir nun nutzen. Um die Umlaufbahn des Merkurs darzustellen, ist es sinvoll die 
-# Aufgabe in zwei Abschnitte zu Teilen:
-# (a) Berchnung des Orbits
-# (b) Zeichnung des Orbits
-# Für beide Abschnitte werden wir Funktionen nutzen.
-
-
-# ----------------------------------
-## Parameter
-# ----------------------------------
-
-# Für Merkur Parameter siehe http://nssdc.gsfc.nasa.gov/planetary/factsheet/mercuryfact.html
-# Für Sonnen Parameter siehe http://nssdc.gsfc.nasa.gov/planetary/factsheet/sunfact.html
-
-# Physikalische Größen/Einheiten
-ms = 1.9885 * 1.e30 # Sonnenmassen
[…]

### 2017-10-25 scripts/old/merkur_orbit 3.py/merkur_orbit_loesung.py

@@ -1,214 +1,0 @@
-# -*- coding: utf-8 -*-
-
-### Darstellung des Merkurorbits mit VPython
-
-# ----------------------------------
-## Laden der Hilfsmodule
-# ----------------------------------
-
-# Da es unglaublich viel Zeit kostet ein Programm fuer 3-dimensionale graphische Darstellung selber 
-# zu schreiben, lassen wir uns die Arbeit von "Vpython" abnehmen. Doch dazu muessen wir zunächst 
-# dieses Modul bereitstellen. Das erreicht man mit Hilfe des Befehls "import" oder "from .. import".
-
-from visual import *
-
-# Schreibt man 
-#   "from visual import *" 
-# bedeutet dies: "Lade alle Funktionen von Vpython so, dass man sie direkt verwenden kann: 
-#   "func = function(..)". 
-# Wuerde man 
-#   "import visual" 
-# schreiben, so muesste man die Funktion wie folgt aufrufen: 
-#   "func = visual.function(..)".
-
-# Eine Dokumentation von Modulen lässt sich durch den Befehl "help( modul_name )" aufrufen.
-#
-# Tipp: Will man eine Eingabe unterdruecken, so laesst sich dies auch mit "#" tun. Der darauf 
-# folgende Text wird dabei von Python ueberlesen.
-
-# help(visual)
-
-# ----------------------------------
-## Aufgabe 1: Merkurorbit)
-# ----------------------------------
-
-# Funktionen werden in Python wie folgt deklariert:
-# def func(x):
-#     .... Rechnung und Anderes ....
-# 
-# Diese wollen wir nun nutzen. Um die Umlaufbahn des Merkurs darzustellen, ist es sinvoll die 
-# Aufgabe in zwei Abschnitte zu Teilen:
-# (a) Berchnung des Orbits
-# (b) Zeichnung des Orbits
-# Fuer beide Abschnitte werden wir Funktionen nutzen.
-
-
-# ----------------------------------
-## Parameter
-# ----------------------------------
-
-# Fuer Merkur Parameter siehe http://nssdc.gsfc.nasa.gov/planetary/factsheet/mercuryfact.html
[…]

### 2017-10-25 scripts/merkur_orbit.py
JL scripts

@@ -1,0 +1,134 @@
+# -*- coding: utf-8 -*-
+
+### Darstellung des Merkurorbits mit VPython
+
+# ----------------------------------
+## Laden der Hilfsmodule
+# ----------------------------------
+
+# Da es unglaublich viel Zeit kostet ein Programm für 3-dimensionale graphische Darstellung selber
+# zu schreiben, lassen wir uns die Arbeit von "Vpython" abnehmen. Doch dazu müssen wir zunächst
+# dieses Modul bereitstellen. Das erreicht man mit Hilfe des Befehls "import" oder "from .. import".
+
+from visual import *
+
+# Schreibt man
+#   "from visual import *"
+# bedeutet dies: "Lade alle Funktionen von Vpython so, dass man sie direkt verwenden kann:
+#   "func = function(..)".
+# Würde man
+#   "import visual"
+# schreiben, so müsste man die Funktion wie folgt aufrufen:
+#   "func = visual.function(..)".
+
+# Eine Dokumentation von Modulen lässt sich durch den Befehl "help( modul_name )" aufrufen.
+#
+# Tipp: Will man eine Eingabe unterdrücken, so lässt sich dies auch mit "#" tun. Der darauf
+# folgende Text wird dabei von Python überlesen.
+
+# help(visual)
+
+# ----------------------------------
+## Aufgabe 1: Merkurorbit)
+# ----------------------------------
+
+# Funktionen werden in Python wie folgt deklariert:
+# def func(x):
+#     .... Rechnung und Anderes ....
+# 
+# Diese wollen wir nun nutzen. Um die Umlaufbahn des Merkurs darzustellen, ist es sinvoll die
+# Aufgabe in zwei Abschnitte zu Teilen:
+# (a) Berchnung des Orbits
+# (b) Zeichnung des Orbits
+# Für beide Abschnitte werden wir Funktionen nutzen.
+
+
+# ----------------------------------
+## Parameter
+# ----------------------------------
+
+# Für Merkur Parameter siehe http://nssdc.gsfc.nasa.gov/planetary/factsheet/mercuryfact.html
+# Für Sonnen Parameter siehe http://nssdc.gsfc.nasa.gov/planetary/factsheet/sunfact.html
+
+# Alle Größen sind in einem angepassten Einheitensystem gegeben:
+#  - Längen in Einheiten von R0 = 1e10 m
[…]

### 2017-10-25 scripts/merkur_orbit_loesung.py

@@ -1,0 +1,214 @@
+# -*- coding: utf-8 -*-
+
+### Darstellung des Merkurorbits mit VPython
+
+# ----------------------------------
+## Laden der Hilfsmodule
+# ----------------------------------
+
+# Da es unglaublich viel Zeit kostet ein Programm fuer 3-dimensionale graphische Darstellung selber 
+# zu schreiben, lassen wir uns die Arbeit von "Vpython" abnehmen. Doch dazu muessen wir zunächst 
+# dieses Modul bereitstellen. Das erreicht man mit Hilfe des Befehls "import" oder "from .. import".
+
+from visual import *
+
+# Schreibt man 
+#   "from visual import *" 
+# bedeutet dies: "Lade alle Funktionen von Vpython so, dass man sie direkt verwenden kann: 
+#   "func = function(..)". 
+# Wuerde man 
+#   "import visual" 
+# schreiben, so muesste man die Funktion wie folgt aufrufen: 
+#   "func = visual.function(..)".
+
+# Eine Dokumentation von Modulen lässt sich durch den Befehl "help( modul_name )" aufrufen.
+#
+# Tipp: Will man eine Eingabe unterdruecken, so laesst sich dies auch mit "#" tun. Der darauf 
+# folgende Text wird dabei von Python ueberlesen.
+
+# help(visual)
+
+# ----------------------------------
+## Aufgabe 1: Merkurorbit)
+# ----------------------------------
+
+# Funktionen werden in Python wie folgt deklariert:
+# def func(x):
+#     .... Rechnung und Anderes ....
+# 
+# Diese wollen wir nun nutzen. Um die Umlaufbahn des Merkurs darzustellen, ist es sinvoll die 
+# Aufgabe in zwei Abschnitte zu Teilen:
+# (a) Berchnung des Orbits
+# (b) Zeichnung des Orbits
+# Fuer beide Abschnitte werden wir Funktionen nutzen.
+
+
+# ----------------------------------
+## Parameter
+# ----------------------------------
+
+# Fuer Merkur Parameter siehe http://nssdc.gsfc.nasa.gov/planetary/factsheet/mercuryfact.html
[…]

### 2017-10-25 scripts/merkur_orbit_loesung.py
upadeted solution

@@ -6,104 +6,24 @@
 ## Laden der Hilfsmodule
 # ----------------------------------
 
-# Da es unglaublich viel Zeit kostet ein Programm fuer 3-dimensionale graphische Darstellung selber 
-# zu schreiben, lassen wir uns die Arbeit von "Vpython" abnehmen. Doch dazu muessen wir zunächst 
+# Da es unglaublich viel Zeit kostet ein Programm für 3-dimensionale graphische Darstellung selber
+# zu schreiben, lassen wir uns die Arbeit von "Vpython" abnehmen. Doch dazu müssen wir zunächst
 # dieses Modul bereitstellen. Das erreicht man mit Hilfe des Befehls "import" oder "from .. import".
 
 from visual import *
 
-# Schreibt man 
-#   "from visual import *" 
-# bedeutet dies: "Lade alle Funktionen von Vpython so, dass man sie direkt verwenden kann: 
-#   "func = function(..)". 
-# Wuerde man 
-#   "import visual" 
-# schreiben, so muesste man die Funktion wie folgt aufrufen: 
+# Schreibt man
+#   "from visual import *"
+# bedeutet dies: "Lade alle Funktionen von Vpython so, dass man sie direkt verwenden kann:
+#   "func = function(..)".
+# Würde man
+#   "import visual"
+# schreiben, so müsste man die Funktion wie folgt aufrufen:
 #   "func = visual.function(..)".
 
 # Eine Dokumentation von Modulen lässt sich durch den Befehl "help( modul_name )" aufrufen.
 #
-# Tipp: Will man eine Eingabe unterdruecken, so laesst sich dies auch mit "#" tun. Der darauf 
-# folgende Text wird dabei von Python ueberlesen.
-
-# help(visual)
-
-# ----------------------------------
-## Aufgabe 1: Merkurorbit)
-# ----------------------------------
-
-# Funktionen werden in Python wie folgt deklariert:
-# def func(x):
-#     .... Rechnung und Anderes ....
-# 
-# Diese wollen wir nun nutzen. Um die Umlaufbahn des Merkurs darzustellen, ist es sinvoll die 
-# Aufgabe in zwei Abschnitte zu Teilen:
[…]

### 2017-10-25 scripts/merkur_orbit.py
tested scripts plus tiny modifications

@@ -76,7 +76,7 @@ rS = 3.e-7
 # Dauer der Simulation in Erdentagen
 T = 88
 # Zeitschritt der Simulation in Erdentagen
-dt = 2*0.51*0.99/1000
+dt = 2*v_merkur_0.mag/G/100
 
 # ----------------------------------
 ## (a) Berechnung des Orbits

### 2017-10-25 scripts/merkur_orbit_loesung.py

@@ -76,7 +76,7 @@ rS = 3.e-7
 # Dauer der Simulation in Erdentagen
 T = 88
 # Zeitschritt der Simulation in Erdentagen
-dt = 2*0.51*0.99/1000
+dt = 2*v_merkur_0.mag/G/100
 
 # ----------------------------------
 ## (a) Berechnung des Orbits
@@ -86,11 +86,12 @@ dt = 2*0.51*0.99/1000
 # die aktuelle Position und Geschwindigkeit des Merkurs und der Sonne. Das Ergebnis der Rechnung
 # soll die jeweils neuen Positionen und Geschwindigkeiten nach einem Zeitintervall "dt" sein.
 
-def merkur_zeit_schritt(r_merkur, v_merkur, r_sonne, v_sonne):
+def merkur_zeit_schritt(r_merkur, v_merkur, r_sonne, v_sonne, alpha=0):
   # Berechne den Vektor der von Merkur nach Sonne zeigt
   r_ms = r_sonne - r_merkur
   # Berechne die Kraefte
   F_ms = - G * m_sonne * m_merkur / r_ms.mag**2 * norm(r_ms)
+  F_ms = F_ms - G * m_sonne * m_merkur * alpha / r_ms.mag**3 * norm(r_ms) * rS
   F_sm = - F_ms
   # Berechne die daraus resultierenden Geschwindigkeiten
   v_sonne_neu  = v_sonne  + F_ms / m_sonne  * dt
@@ -109,10 +110,10 @@ merkur_zeit_schritt(r_merkur_0, v_merkur_0, r_sonne_0, v_sonne_0)
 ## (b) Zeichnung des Orbits
 # ----------------------------------
 
-def zeichne_orbit():
+def zeichne_orbit(alpha=0):
   # "Erstelle Merkur und Sonne"
-  merkur = sphere( pos=r_merkur_0, radius=0.01, color=color.red   )
-  sonne  = sphere( pos=r_sonne_0 , radius=0.1, color=color.yellow)
+  merkur = sphere( pos=r_merkur_0, radius=0.2, color=color.red   )
+  sonne  = sphere( pos=r_sonne_0 , radius=0.8, color=color.yellow)
   # Erstelle die Bahnkurve fuer Merkur
   merkur.bahn = curve(color=color.white)
   # Weise den Planeten ihre Anfangsgeschwindigkeiten zu
@@ -122,12 +123,12 @@ def zeichne_orbit():
   t = 0
   while t < 88*6:
[…]

### 2017-11-30 tex/MercuryNumerical.tex
worked out structure
inka edits

--- 2017-07-05 tex/MercuryNumerical.tex
@@
 \usepackage[prependcaption, textsize=tiny]{todonotes}
+\usepackage{gensymb}
 
@@
 perturbative approach (as in theoretical descriptions of nuclear particles at the fundamental level). 
- The goal of this paper is it to
-introduce a project that could be used to make numerical simulations themselves the topic of the class.
+ This paper introduces a project that can be used to demonstrate the power of numerical simulations.
 On the example of the perihelion motion of the planet Mercury the students are supposed to get in touch with
@@
 The course as well as this paper is structured as follows: after an introduction to Newtonian dynamics and the 
-concepts of differential equations their discretization is discussed based on Newtons law of gravitation and possible
+concepts of differential equations, their discretization is discussed based on Newtons law of gravitation and possible
 extensions thereof. Afterwards the visualization of the resulting trajectories using \texttt{VPython} is introduced and
@@
 We are convinced that in order to excite the students for numerical simulations it is compulsory to
-demonstrate their power on an example that they catches their interest. This purpose is served 
+demonstrate their power on an example that catches their interest. This purpose is served 
 perfectly by the case chosen here, since Einsteins equations of General Relativity are fascinating to
@@
 object in addition we need to be able to calculate its velocity, $\vec v(t)$, related to the location via
-%
-\todo{I see here a potential conflict with the implementation because we use $x(t-\Delta t)$ in order to get $v$ and so on}
-%
 \begin{equation}\label{eq:x_update}
@@
 If the mass does not change~\footnote{A well known example where $m$ does change with time is a 
-rocket, whose mass decreases as the rocket rises.} with time this, reduces to the well known
+rocket, whose mass decreases as the rocket rises.} with time, this reduces to the well known
 \begin{equation}
@@
 Note that in general the force could depend also on the time or the velocity. We here restrict ourselves to the
-case relevant for our example where the force depends on the location only. 
+case relevant for our example, where the force depends on the location only. 
 Therefore, 
@@
 the highest derivative is $n$) needs $n$ initial conditions specified. For $n=2$ those are often chosen as location and velocity
-at some defined time, but also to pick two location at different times is possible. }
+at some defined time, but one may as well pick two location at different times. }
 
-Clearly for this procedure to work $\Delta t$ must be sufficiently small. What this means depends on the process
+For this procedure to work $\Delta t$ must be sufficiently small. What this means depends on the process
 studied. One way to estimate, if $\Delta t$ is small enough, is to verify, if the relation
@@
 \begin{equation}
-\vec F_N(\vec x) = \frac{G_N m M_\odot}{r^2} \frac{\vec x}{r}\ ,
+\vec F_N(\vec x) = - \frac{G_N m M_\odot}{r^2} \frac{\vec x}{r}\ ,
 \end{equation}
-where $G_N=6.67\times 10^{-11}$ m$^3$kg$^{-1}$s${-2}$ is the Newtonian constant of gravitation,
+where $G_N=6.67\times 10^{-11}$ m$^3$kg$^{-1}$s$^{-2}$ is the Newtonian constant of gravitation,
 $m$ is the mass of Mercury and $M_\odot=2\times 10^{30}$ kg is the mass of the sun.
@@
 where $c=3\times 10^8$ m/s denotes the speed of light.
-Note the $r_S$ is the characteristic length scale of the gravitational field of the sun
+Note that $r_S$ is the characteristic length scale of the gravitational field of the sun
 for --- up to the prefactor --- one can not form another quantity with dimensions of a length from $G_N$, $M_\odot$ and
@@
 With this Newtons second law reads 
+\todo{Or would it be better to write: $\ddot{x} = \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)$ and $\ddot{\vec x} = - \ddot{x} \frac{\vec{x}}{r}$?}
 \begin{equation}
-\ddot{\vec x} = \frac{c^2}{2}\left(\frac{r_S}{r^2}\right) \ .
+\ddot{\vec x} = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right) \frac{\vec{x}}{r} \ .
 \label{eq:newton}
@@
 In general an attractive force that vanishes at large distances leads,
-%
-\todo{I am confused. I thought only $1/r$ and the harmonic oscillator potential lead to closed orbits (Bertrand's theorem).}
-%
- depending on the initial conditions, either to closed orbits or open orbits --- in the latter case the planet simply disappears from the sun.
-Depending on time available for the course  it might be helpful to explain to the students the origin of this, however, since the most transparent reasoning needs an integration as well as the concept of angular momentum not necessarily familiar to the whole class (details are given in the appendix
-%
-\todo{\bf Include this appendix part...}
-%
-), 
-it appears in general more appropriate to work out together with the students that a moving body can not be kept by a force if its momentum is too large.
+ depending on the initial conditions, either to bounded orbits or to open orbits --- in the latter case the planet simply disappears from the sun.
+Depending on time available for the course  it might be helpful to explain to the students the origin of this, however, since the most transparent reasoning needs an integration as well as the concept of angular momentum not necessarily familiar to the whole class, it appears in general more appropriate to work out together with the students that a moving body can not be kept by a force if its momentum is too large.
 Then they may try to find closed orbits in the simulation, e.g., by varying the start velocity while keeping the start location fixed.
 
-The closed orbits that emerge from a potential that scales as $1/r$ are elliptic and fixed in space.
+The bounded orbits that emerge from a potential that scales as $1/r$ are elliptic and fixed in space.
 In particular the point of closest approach of the planet to the sun, the perihelion, does not move.
@@
 residual motion of $\delta \phi_M = (43)''$ in 100 earth years remained unexplained, until Einstein quantified the
-predictions of General Relativity to this particular observable.
-
-%
-\todo{\bf Here should be a brief description of the relation between Newtons dynamics and General Relativity}
-%
+predictions of General Relativity to this particular observable. \todo{citation here}
 
@@
 \end{equation}
-where $\alpha$ is some parameter related to the effects by general relativity.
+where $\alpha$ is some parameter related to the effects by General Relativity.
 The idea is that the correction must be dimension-less and should scale with some inverse power of $r$, for we still need to demand that the potential vanishes at large distances.
 Since the only parameter of the system with dimension of a length is $r_S$, the correction should scale as $(r_S/r)$.
-When replacing $r$ by a typical distance of Mercury to the sun, $(r_S/r)\sim 10^{-7}$, it should be sufficient to include this kind of term to first order only.
-For comparison: the corresponding number for the Venus, the neighbor of Mercury, reads $r \sim 5 \cdot 10^{-8}$.
+Replacing $r$ by a typical distance of Mercury to the sun, $(r_S/r_{MS})\sim 10^{-7}$, suggests that it should be sufficient to include this kind of term to first order only.\footnote{Considering another order would lead qualitatively to the same result, since every change in the potential leads to a perihelion motion.
+Only the value of the pre-factor would change.}
+For comparison: the typical distance of the Venus, the neighbor of Mercury, to the sun is $(r_S/r_{VS}) \sim 5 \cdot 10^{-8}$.
 Therefore the expected ratio of residual motion, favoring in the orbit period $T$, would result in
 \begin{equation}
-	\delta \phi_V \approx \frac{1}{T_V} \delta \phi_M T_M \frac{r_M}{r_V} \approx 9'' \, ,
+	\delta \phi_V \approx \frac{1}{T_V} \delta \phi_M T_M \frac{r_{MS}}{r_{VS}} \approx 9'' \, ,
 \end{equation}
-which precisely describes the experimental value of $9''$.
+where $T_V$ ($T_M$) denotes the length of a venus (mercury) year. This precisely describes the experimental value of $9''$.
 
@@
 
-In this section we present the numerical implementation as well as the visualization of planetary trajectories and in
+In this section we present the numerical implementation as well as the visualisation of planetary trajectories and in
 particular
@@
 how to set up \python on different operating systems.}.  No prior knowledge of \python or any other programming language 
-is required because we explain all the steps necessary to create the simulation.  We also provide a working 
+is required, because we explain all the steps necessary to create the simulation.  We also provide a working 
 example \cite{}
@@
 
-To start the simulation one needs the "initial" distances and velocities, which are needed as input to their equation of motions (Eq.~(\ref{eq:x_update}) and Eq.~(\ref{eq:v_update})).
+To start the simulation one needs the "initial" distances and velocities, which are needed as input to the equation of motion (Eq.~(\ref{eq:x_update}) and Eq.~(\ref{eq:v_update})).
 As described above, here  one can simplify the problem by placing one planet (the "infinitely" heavy sun) in the center of the coordinate system and describe the motion of the other planet (mercury) in a two-dimensional plane.
@@
 Both, for the numerical treatment and the intuitive understanding, it is useful to select parameters in a "natural range".
-A particular useful choice for this problem is given by expressing time intervals in days $T_0 = 1$d, where $d$ refers to earth days, and distances in $R_0 = 10^{10}$ m.
+A useful choice for this problem is given e.g.
+by expressing distances in $R_0 = 10^{10}$ m and time intervals in days, $T_0 = 1$d, where $d$ refers to earth days, such that one mercury year is given by $T_M=88\,T_0$.
 With this choice, the initial distance of mercury to the sun, the size of the initial velocity of mercury and the acceleration prefactor become
+\todo{Maybe we should rename $a_M$ to avoid confusion about the fact, that it is not an acceleration?}
 \begin{equation}
@@
 vM0 = 0.51   # in units of R0/T0
-aM  = 0.99   # in units of R0/T0**2
-T   = 88.    # in units of T0
+aM  = 0.99   # in units of R0/T0**2 * R0**2
+TM   = 88.   # in units of T0
 rS  = 3.e-7  # in units of R0
@@
 The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.
-Because of our choice of initial conditions (we picked the initial vectors in the perihelion), the velocity of mercury is perpendicular to the vector which connects mercury and the sun (see figure \ref{fig:sun_merc}):
+Because of our choice of initial conditions (we picked the initial vectors in the perihelion), the velocity of mercury is perpendicular to the vector which connects mercury and sun (see figure \ref{fig:sun_merc}):
 \begin{lstlisting}
@@
 However, we first need to fix the time step $\Delta t$ and use the estimate
-of Eq.~(\ref{eq:check}) as a guidance \todo{\bf Careful: Check units of dt}
+of Eq.~(\ref{eq:check}) as a guidance
 \begin{lstlisting}
 # Definition of the time step
-dt = 2 * vM0 / aM / 1000
+dt = 2 * vM0 / aM / 10
 \end{lstlisting}
-Here the factor $1/1000$ makes sure that $\Delta t$ is indeed consistent with Eq.~(\ref{eq:check}).
-%
-\todo{I need 1/1000 to get relatively closed orbits -- we have to crosscheck codes!}
-%
+Here the factor $1/10$ makes sure that $\Delta t$ is indeed consistent with Eq.~(\ref{eq:check}).
 Now we are in the position to calculate
-location and velocity of the planet at $t_0+\Delta t$ using the following comands
+location and velocity of the planet at $t_0+\Delta t$ using the following commands \footnote{Note, that the order of calculating $a_{MS}$, $v_M$ and $r_M$ is not clear a priori.
+All prescriptions lead to the same result for sufficiently small $dt$, but the following order turns out to be numerically most stable.}
 \begin{lstlisting}
@@
 # Call the function
-vec_rM_new, vec_vM_new = do_time_step(vec_rM_old, vec_vM_old)
+vec_rM_new, vec_vM_new = evolve_mercury(vec_rM_old, vec_vM_old)
 \end{lstlisting}
@@
 Furthermore, \python is an Interpreter language.
-Each line of the code is execute once the Interpreter passes them.
+Each line of the code is executed when the Interpreter passes it.
 For this reason, all the variables defined before the function are global and thus known to the function.
-The variables within the function (e.g. \texttt{vec\_rM\_old}, ...) are local and do not live beyond the scope of the function.
+The variables within the function (e.g. \texttt{vec\_rM\_old}, ...) are local and do not live beyond the finalization of the function.
 
-Finally, we can express the evolution by a \texttt{while}-loop
+Finally, we can express the evolution by a \texttt{while}-loop \todo{Should we also add  e.g. T=2*TM here?}
 \begin{lstlisting}
@@
 The total runtime \texttt{T} is the amount of "virtual" days the simulations should run.
-To describe at least one full orbital period, $T$ needs to be larger than $T>88$d.
+To describe at least one full orbital period, $T$ needs to be larger than $T_M$.
 With the previous choice for \texttt{dt}, this corresponds to roughly $N_T = 10^4$ evolution steps.
@@
 \end{lstlisting}
-The class sphere will represent mercury and the sun in the simulation
+The class sphere will represent mercury and the sun in the simulation \todo{The pictures on the next page are done with sun radius 1.5 and mercury radius 0.5.
+Would it be ok, if we change the following radii to these values?}
 \begin{lstlisting}
@@
 \end{lstlisting}
-Next we tell python that the objects shall be drawn from this point on
+Next we tell python that the objects shall be drawn from this point on \todo{For some reason display does not work for me.
+However I like black as background color anyway.
+Would it be ok, if we leave out the display and change the color of the trajectory to white?}
 \begin{lstlisting}
@@
 \end{lstlisting}
-and last but not least, instead of updating the object-unrelated vectors in the while loop, we update the object-related vectors which will eventually be drawn
+and last but not least, instead of updating the object-unrelated vectors in the while loop, we update the object-related vectors which will eventually be drawn \todo{changed rate ok?}
 \begin{lstlisting}
@@
 while t < T:
-	# set the frame rate: shows four earth days at once
-	rate(4*1000)
+	# set the frame rate (you can choose a higher rate to accelerate the program)
+	rate(100)
 	# update the drawn trajectory with the current position
@@
 If the starting values are chosen as advised in the previous section, then the students should end up with a trajectory as depicted in figure \ref{fig:MercuryOrbit}.
-To get the perihelion motion, the term from equation \ref{eq:newton_art} has to be be added.
-This is a good opportunity to let the students play with the size of $\alpha$ and get a feeling for the impact on the trajectories.
-However it might be advisable to not start out with the correct mercury values but e.g.
-a slightly higher initial distance between mercury and sun, since the perihelion motion is better visible then.
-This is illustrated in figure \ref{fig:MercuryOrbit}, where we used a value of $\alpha=10^5$ and $dt=0.1T_0$.
-%
-\todo{This is really to large for me to get these orbits. Also I need to run with $\alpha = 10^6$ to get such shapes...}
-%
+To get the perihelion motion, the term from equation (\ref{eq:newton_art}) has to be be added by choosing $\alpha=0$.
+This is a good opportunity to let the students play with the size of $\alpha$ and get a feeling for its impact on the trajectories.
+However, it might be advisable to not start out with the correct mercury values but a more excentric trajectory by choosing e.g.
+$r_{MS}(0)=6R_0$.
+Then the perihelion motion is better visible.
+This is illustrated in figure \ref{fig:MercuryOrbit}, where we used a value of $\alpha=10^6$.
+
 \begin{figure}[htb]
@@
 The complete program needs to be saved as a python file, e.g. \texttt{mercury\_orbit.py}.
-Depending on the editor of choice, there might already be a execute functionality e.g. \textbf{bla...} for the recommended windows installation (see \ref{appendix:python}).
+Depending on the editor of choice, there might already be an execute functionality e.g. \textbf{bla...} for the recommended windows installation (see \ref{appendix:python}).
 %
@@
 
-\textit{What exactly do we want the students / advise the teachers to do?
+%\textit{What exactly do we want the students / advise the teachers to do?
 I think a nice approach would be to let the students implement the code to extract the perihelion motion and make a list of some values of alpha and the corresponding angles.
@@
 
-To calculate the expected size of the perihelion motion for a modified gravitational force, the students need to be abel to extract the relative angle between the current mercury position and the starting position.
-There are several ways to do this, but the easiest is to extract the point where the distance to the sun gets minimal -- the definition of the perihelion\todo{why aphelion and not perihelion?
-I have changed the following code to the perihelion (because it is shorter).
-We can change it back if desired.} -- and compare the new position of the perihelion after $n$ turns to the original one.
-This can be done by looking for a minimal distance of mercury to the sun within a $while$-loop.
+To calculate the expected size of the perihelion motion for a modified gravitational force, the students need to
+\begin{itemize}
+\item extract the position of the perihelion
+\item calculate the angle between the perihelions for successive turns and compute the average angle $\delta \Theta (\alpha)$
+\item repeat the above points for different values of $\alpha$ and confirm the linear dependence
+\item extrapolate to $\alpha=1$
+\end{itemize}
+\todo{I changed the text here in an effort to make the purpose and structure of this section more clear. However we can change it back, if you prefer the old version.}\\
 
-Suppose we want to save the number of turns in the variable \texttt{turns}, the maximal number of turns in \texttt{max\_turns} (here e.g.
-10), the initial perihelion is already stored in the vector \texttt{vec\_rM0} and the goal is to find the final perihelion.
-In addition to find out when the perihelion is reached, the last two positions \texttt{vec\_r\_last} and \texttt{vec\_r\_before\_last} have to be known.
-Thus, if the last vector is smaller than the previous and the current vector, one has reached the perihelion again.
-The implementation can be realized as follows
+There are several ways to find the position of the perihelion.
+The easiest one is to look for the point of minimal distance to the sun -  the definition of the perihelion.
+This can be done within the $while$-loop.
+To find out when the minimum is reached, the last two positions \texttt{vec\_r\_last} and \texttt{vec\_r\_before\_last} have to be known.
+If the length of the last vector is smaller than that of the previous one as well as the current vector, one has passed the perihelion and its position is given by \texttt{vec\_r\_last}.
+Suppose we want to save the position of the perihelion for a certain number of turns (here e.g.
+\texttt{max\_turns}=10) in a list \texttt{list\_perih}.
+\todo{I am not sure this is the best way to do this, since I need to introduce yet another concept (lists).
+However I could not think of an elegant solution...}
+The implementation can be realised as follows \todo{I have a problem here, since I cannot use M.pos in this way.
+For me M.pos seems to be a pointer, because after one turn \texttt{vec\_r\_before\_last}, \texttt{vec\_r\_last} and M.pos all have the same value always.
+Therefore I need to use another variable and set M.pos to this value in the end.
+Could you check, whether it is the same for you?
+If yes, I will change the code accordingly, if not we have to compare codes again...}
 \begin{lstlisting}
@@
 max_turns  = 10
-# find perihelion after ten turns
+list_perih=[]
+# find perihelion for each turn and print it out
 while turns < max_turns:
@@
 	if vec_r_before_last.mag > vec_r_last.mag < M.pos.mag:
+		list_perih.append(vec_r_last)
 		turns = turns+1
 \end{lstlisting}
-When the loop stops, the last perihelion vector is stored in  \texttt{vec\_r\_last}.
-To compute the angle between two vectors one uses the formula
+To compute the angle between two vectors one uses the following
+formula
  \begin{equation}
- 	\sphericalangle(\vec{a}_0,\vec{a}) = \cos^{-1} \left( \frac{\vec{a}_0 \cdot \vec{a}}{|\vec{a}_0|\:|\vec{a}|} \right)
+ 	\sphericalangle(\vec{v}_{n-1},\vec{v}_n) = \cos^{-1} \left( \frac{\vec{v}_{n-1} \cdot \vec{v}_n}{|\vec{v}_{n-1}|\:|\vec{v}_n|} \right)
  \end{equation}
@@
 angle = acos(
-	dot(vec_rM0, vec_r_last.mag) / (vec_rM0.mag * vec_r_last.mag)
-) / maxcounter * 380 / 2 / pi
+	dot(list_perih[n-1], list_perih[n]) / (list_perih[n-1].mag * list_perih[n].mag) 
+) * 180  / pi
 \end{lstlisting}
+To minimise the statistical errors these can be averaged over a few turns.
+Depending on the programming proficiency of the students and time constraints, this can be either done by hand or e.g.
+by implementing the following code
+\begin{lstlisting}
+sum_angle=0.
+for n in range(1, max_turns):
+    # calculate angle
+    angle = ...
+    sum_angle = sum_angle+angle
+# display the average
+print(sum_angle/(max_turns-1.))
+\end{lstlisting}
+
+As explained above (and further discussed in chapter \ref{sec:analysis}) the natural value for $\alpha$ would be of the order of $1$.
+However, if the student use this value for $\alpha$, they will find that the change in the trajectories is close to invisible and uncertainty is much larger than the result.
+Therefore, to estimate the size of the perihelion motion, it is more advisable to use the fact that there is an approximately linear dependence between $\alpha$ and $\delta \Theta$.\footnote{Of course for very large values of $\alpha$ this linear relation will eventually break down.
+We therefore advise to choose $\alpha<7\cdot 10^6$ at which point the perihelion motion exceeds $\delta \Theta=90^{\degree}$.}
+The students can convince themselves of this fact by plotting some angles they extracted from their simulation over $\alpha$.
+This can be done either by hand or using another program like e.g. Excel.
+One such plot can be found in figure \ref{fig:AlphaAngle}.
 
@@
 
-As explained above the natural value for $\alpha$ would be $1$.
-However if the student us this value for $\alpha$, they will find that there is close to no visual change in the trajectories and the distance between the first an the last perihelions is of the order of magnitude of one step in the calculation $v_{aphelion}\cdot dt$.
-Therefore, to estimate the size of the perihelion motion it is more advisable to use the fact, that there is an approximately linear dependence between $\alpha$ and the perihelion motion, if the angle is not to large.
-The students can convince themselves of this fact, by plotting some angles over $\alpha$.
-This can be done either by hand or using another program like e.g. Excel.
-One such plot can be found in figure \ref{fig:AlphaAngle}.
+To extract the perihelion motion at $\alpha = 1$, we use the fact that it would be zero at
+ $\alpha =0$. 
+Considering in addition one result from the numeric simulation, e.g. $\delta\Theta(\alpha = 10^6) \approx 10.1^{\degree}$, the sought-after angle can be calculated from the linear interpolation as
+\begin{equation}
+	\delta\Theta (\alpha=1) = \frac{10.1^{\degree}}{10^6} = 0.036''
+\end{equation}
+Note when comparing the results of the simulation with the experimental result that one earth year corresponds to $T_E/T_M \approx 365d/88d\approx 4.15$ mercury years.
+Therefore, the perihelion motion per 100 earth years is $0.036''\cdot 415\approx 15''$.
+To test whether this is a reasonable result, the students could first follow the dimensional analysis and then compare to the actual value for mercury as outlined above.
 
-To get to the angle at $\alpha=1$ one would ideally employ a linear regression (or ruler fit), but to get the order of magnitude right it is enough to consider only two points.
-Choosing $\alpha_1=1000$ and $\alpha_2=2000$ here, this would yield
-\begin{equation}
-	\Theta(\alpha=1) \approx \frac{0.201-0.100}{2000-1000} = 0.0001 = 0.36''
-\end{equation}
-so $3.63''\cdot 415\approx 150''$ per 100 earth years.
-To find out, whether this is a reasonable result, the students could first follow the dimensional analysis and then compare to the actual value for the mercury  as outlined above.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\clearpage\subsection{Cross-Check Results}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\todo[inline]{Here I (CK) just make a few comments regarding the results.}
+%First of all, I can confirm Inkas result up to a factor of 10.
+%I believe this factor arises because we have analysed 10 years of motions (\verb|max_turns = 10|).
+%See figure \ref{fcc1}.
+%This result is quite nice, because we basically say the genreal relativity motion by mercury is caused by $\alpha \approx 3$.
+%However there is a small issue: we do not have the numerical precision to claim this.
+%\begin{figure}[htb]
+%	\centering
+%	\includegraphics[width=.4\textwidth]{figs/alpha_interpolation.pdf}
+%	\caption{\label{fcc1}Interpolation for $\phi$ depending on $\alpha$ and $dt = 2 v_0 / a_0 / 5000$. }
+%\end{figure}
+%
+%I have encountered, that we do not start perfectly at the perihelion or aphelion (depending on the boundary conditions): my simulations arrives at this point after roughly a tenth of a day and therefore I have an offset which needs to be subtracted off after each mercury year.
+%More interestingly, this offset seems to depend on the tilmestep of the simulation (see figure \ref{fcc2}).
+%\begin{figure}[htb]
+%	\centering
+%	\includegraphics[width=.4\textwidth]{figs/initial_time_shift.pdf}
+%	\caption{\label{fcc2}Initial shift of perihelion for $\delta\theta = \delta\phi$ in degrees over $\alpha$ for different time steps (color).}
+%\end{figure}
+%
+%One can also see that the size of the time spacing affects a potential uncertainty in the periodicity of the orbit when looking at the precession growth over different turns (see figure \ref{fcc3}).
+%\begin{figure}[htb]
+%	\centering
+%	\includegraphics[width=.7\textwidth]{figs/precession_growth.pdf}
+%	\caption{\label{fcc3}Precession growth of perihelion for $\delta\theta = \delta\phi$ in degrees over number of turns for different time steps (color).
+One should pay special attention to the oscillations within the growth for relatively large time steps.}
+%\end{figure}
+%
+%Last but not least, for the precession interpolation to small $\alpha$, one should worry about the precision of the precession for small alpha.
+%Even after subtracting of the results for zero turns, basically non of my results agree with zero for alpha equal to zero [to go from degrees per mercury year to arcsecs per 100 earth years, one has to multiply by $1.46 \times 10^{6}$.](see figure \ref{fcc4}).
+%\begin{figure}[htb]
+%	\centering
+%	\includegraphics[width=.4\textwidth]{figs/alpha_dependence_0}
+%	%
+%	\includegraphics[width=.4\textwidth]{figs/alpha_dependence_1}
+%	\caption{\label{fcc4}Perihelion precession $\delta\theta = \delta\phi$ in degrees over $\alpha$ without (left figure) or with (right figure) subtracting off the zero turn result.
+The uncertainty corresponds to the variation over different turns and the results are displayed for different time steps (color).}
+%\end{figure}
+%
+%Because, after subtracting the initial angle as a reference, the slope of all the plots basically agree, I suggest the following prescription (as Inka actually did)
+%\begin{enumerate}
+%	\item Compute the precession for $\alpha = 0$ for several turns,
+%	\item compute the precession for preferably large $\alpha > 1000$ for several turns,
+%	\item verify that they are indeed on a line (linear relation),
+%	\item subtract the zero value from the other values (since we know that it has to be zero and the uncertainty is coming from the time steps),
+%	\item do the interpolation to $\delta \phi (\alpha_{GR}) = 43''$ for two values: $\delta \phi (0) = 0$ and $\delta \phi (\alpha_l) = \delta \phi_l$
+%	\item repeat this for several $(\alpha_l, \delta \phi_l)$ and get the average result for $\alpha_{GR}$.	
+%\end{enumerate}
+%
+%
+%\clearpage
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Tests of stability}\label{sec:stability}
+\section{Tests of stability and error analysis}\label{sec:stability}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+Physical experiments in general, as do numerical simulations of physics, always involve uncertainties.
+It is important for the students to recognise this fact and to understand, that the cause of the uncertainties lies in the finite time steps.
+The students should therefore explore the source and size of the errors arising in this simulation at least qualitatively.
+
+Consider figure \ref{fcc3}, where the angle of the perihelion relative to the y-axis is shown for the first 10 turns for different choices of $dt$ and $\alpha$.
+As expected, the correlation between the number of turns and $\Theta$ is linear and the slope depends solely on $\alpha$.
+However, there is an offset between the lines for different $dt$.
+The reason is, due to the uncertainties in the simulation, the starting position does not coincide with the first perihelion as one might expect.
+This does not generate a problem for the procedure described in the previous section
+since the starting position is not considered. Still, it is helpful to be aware of this fact.\\
+One can see nicely in figure \ref{fcc3} that the oscillations of the perihelion motion increases for growing time steps.
+
+\begin{figure}[htb]
+	\centering
+	\includegraphics[width=.7\textwidth]{figs/precession_growth.pdf}
+	\caption{\label{fcc3}Precession growth of perihelion for $\delta\theta = \delta\phi$ in degrees over number of turns for different time steps (color).
+One should pay special attention to the oscillations within the growth for relatively large time steps.}
+\end{figure} \todo{Could you redo this figure with some bigger time steps?
+Could we call the precession growth $\Theta(\alpha)$ instead of $\delta\Theta$?
+I think this would fit better with our notation so far...}
+
+To point the students to this problem, they can be asked to increase their time steps significantly and observe the effect on the trajectories.
+For this study, it is advantageous to set $\alpha =0$, since in this case the trajectories should be closed and therefore deviations from the ideal case can be spotted easily.
+In figure \ref{fig: LargeDT} the mercury trajectory for $dt=2 * v_M(0) / a_M \cdot 5$ is shown.
+One can clearly see the failure of the simulation to reproduce the physical trajectory.
+From examples like this it should be apparent that the time steps influence the accuracy of the simulation.
+Only in the limit $dt \rightarrow 0$ would the simulation perfectly reproduce the actual trajectory.
+\begin{figure}[htb]
+	\centering
+	\includegraphics[width=.3\textwidth]{figs/TrajectoryLargeDT.png}
+	\caption{\label{fig: LargeDT} Deviation of the trajectory from the analytic result (indicated in blue) for large time steps (here $dt=2 * v_M(0) / a_M \cdot 5$)}
+\end{figure}
+
+\todo{It would be nice if they could also get a quantitive estimate of the error, what do you think would be best?
+Maybe they can just look at the perihelion motion for $\alpha=0$ and take that as an estimate of the systematic error?}
 
@@
 for $r$ the average distance Mercury-sun, namely $\bar r=6\times 10^7$ km.
-Form this one estimates for the expected angular shift per orbit
+From this one estimates for the expected angular shift per orbit
 \begin{equation}

### 2017-11-30 tex/MercuryNumerical.tex
Inka+ck edits

@@
 %
-\documentclass[12pt]{iopart}
+\documentclass[12pt, UK english]{iopart}
 \newcommand{\gguide}{{\it Preparing graphics for IOP Publishing journals}}
@@
 
-\usepackage{ amssymb }
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Custom
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\usepackage[T1]{fontenc}           % Embed fonts
+\usepackage{textcomp,gensymb}      % \degree
+
+\usepackage{polyglossia}                         % Define languages (Umlaute)
+\setdefaultlanguage[variant=american]{english}   % Accent of main language
+\setotherlanguage{german}                        % We also have german words
+\usepackage[autostyle=true]{csquotes}            % Define quotation style
+\MakeOuterQuote{"}                               % Text wrapped in "..." is recognized as a quotation
+
+\usepackage{amssymb}                  % Math stuff
 \usepackage{graphicx}                 % Import Pictures + PDFs
 \usepackage[svgnames]{xcolor}         % Enabling colors by their 'svgnames'
-\usepackage{ulem}
+\usepackage{listings}                 % Code listings  See: https://en.wikibooks.org/wiki/LaTeX/Source_Code_Listings
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Definitions
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \definecolor{fzjblue}{HTML}{005B81}
@@
 \definecolor{mygreen}{rgb}{0,0.6,0}
-\usepackage{listings} %  See: https://en.wikibooks.org/wiki/LaTeX/Source_Code_Listings
+
+% Custom code listings environment
 \lstset{
@@
   tabsize=4,	                   % sets default tabsize to 2 spaces
-  showtabs=true,
+  showtabs=false,
   rulecolor=\color{black},
@@
 \usepackage[prependcaption, textsize=tiny]{todonotes}
-\usepackage{gensymb}
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Commands
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \newcommand{\python}[0]{\texttt{Python} }
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Here it starts
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{document}
 
-\title[A primer to numerical simulations]{A primer to numerical simulations: The perihelion
-motion of Mercury}
+\title[A primer to numerical simulations]{A primer to numerical simulations: The perihelion motion of Mercury}
 
-\author{I. Hammer, C. Hanhart, C. K\"orber and C. M\"uller}
+\author{I. Hammer, C. Hanhart, C. Körber and C. Müller}
 
-\address{Forschungszentrum J\"ulich}
+\address{Forschungszentrum Jülich}
 \ead{c.hanhart@fz-juelich.de}
@@
 \begin{abstract}
-Numerical simulations are playing an increasingly important role in modern science. In this work it is
-suggested to use a numerical study of the famous perihelion motion of the planet Mercury (one of the prime observables
-supporting Einsteins General Relativity) as a test case to teach numerical simulations to high school students.
-The paper includes details about the development of the code as well as a discussion of the visualization 
-of the results. In addition a method is discussed how to estimate the size of the effect a priori. 
+Numerical simulations are playing an increasingly important role in modern science.
+In this work it is suggested to use a numerical study of the famous perihelion motion of the planet Mercury (one of the prime observables supporting Einsteins General Relativity) as a test case to teach numerical simulations to high school students.
+The paper includes details about the development of the code as well as a discussion of the visualization of the results.
+In addition a method is discussed how to estimate the size of the effect a priori. 
 \end{abstract}
@@
 
+\listoftodos
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-Numerical simulations play a key role in modern physics for they allow one to tackle theoretical problems
-not accessible otherwise, e.g., since there are too many particle participating in the system (as in
-simulations for weather predictions) or the interactions are too complicated to allow for a systematic,
-perturbative approach (as in theoretical descriptions of nuclear particles at the fundamental level). 
+Numerical simulations play a key role in modern physics for they allow one to tackle theoretical problems not accessible otherwise, e.g., since there are too many particle participating in the system (as in simulations for weather predictions) or the interactions are too complicated to allow for a systematic, perturbative approach (as in theoretical descriptions of nuclear particles at the fundamental level).
 This paper introduces a project that can be used to demonstrate the power of numerical simulations.
@@
 \end{itemize}
-The course as well as this paper is structured as follows: after an introduction to Newtonian dynamics and the 
-concepts of differential equations, their discretization is discussed based on Newtons law of gravitation and possible
-extensions thereof. Afterwards the visualization of the resulting trajectories using \texttt{VPython} is introduced and
-applied to the problem at hand. In particular tools are developed to extract the relevant quantity from the result
-of the simulation.
-Finally the principle of dimensional analysis is presented as a tool to cross check, if the results of the simulation
-are sensible.
+The course as well as this paper is structured as follows: after an introduction to Newtonian dynamics and the concepts of differential equations, their discretization is discussed based on Newtons law of gravitation and possible extensions thereof.
+Afterwards the visualization of the resulting trajectories using \texttt{VPython} is introduced and applied to the problem at hand.
+In particular tools are developed to extract the relevant quantity from the result of the simulation.
+Finally the principle of dimensional analysis is presented as a tool to cross check, if the results of the simulation are sensible.
 
-We are convinced that in order to excite the students for numerical simulations it is compulsory to
-demonstrate their power on an example that catches their interest. This purpose is served 
-perfectly by the case chosen here, since Einsteins equations of General Relativity are fascinating to
-a very broad public. While their detailed study needs deeper knowledge in Differential Geometry, for
-the study outlined here very little math is necessary, such that high school students from about 
-10$^{\rm th}$ grade up should benefit from it. This was already demonstrated in the
-'Sch\"ulderakademie Teilchenphysik', where an early version of this course was tested successfully 
-on a goup of 10 German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade.
+We are convinced that in order to excite the students for numerical simulations it is compulsory to demonstrate their power on an example that catches their interest.
+This purpose is served perfectly by the case chosen here, since Einsteins equations of General Relativity are fascinating to a very broad public.
+While their detailed study needs deeper knowledge in Differential Geometry, for the study outlined here very little math is necessary, such that high school students from about 10$^{\rm th}$ grade up should benefit from it.
+This was already demonstrated in the "Schülderakademie Teilchenphysik", where an early version of this course was tested successfully on a group of 10 German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade.
 
@@
 
-We can say that we understand a physical system, if we can demonstrate that the assumed force leads to the trajectories
-observed. In other words, we need to show that we can calculate the location in space of the object of interest at any point
-in time, once the initial conditions are fixed properly. If we can neglect the finite size of this object (and
-in particular its orientation in space) the location is parametrized
-by a single three vector $\vec x(t)$. As will become clear below, in order to describe and control the dynamics of a physical
-object in addition we need to be able to calculate its velocity, $\vec v(t)$, related to the location via
+We can say that we understand a physical system, if we can demonstrate that the assumed force leads to the trajectories observed.
+In other words, we need to show that we can calculate the location in space of the object of interest at any point in time, once the initial conditions are fixed properly.
+If we can neglect the finite size of this object (and in particular its orientation in space) the location is parametrized by a single three vector $\vec r(t)$.
+As will become clear below, in order to describe and control the dynamics of a physical object in addition we need to be able to calculate its velocity, $\vec v(t)$, related to the location via
 \begin{equation}\label{eq:x_update}
-\vec x(t+\Delta t) = \vec x(t) + \vec v(t) \Delta t + ... \ , \label{eq:vdef}
+\vec r(t+\Delta t) = \vec r(t) + \vec v(t) \Delta t + ... \ , \label{eq:vdef}
 \end{equation}
@@
 \end{eqnarray}
-The dots in the above expressions indicate that there are in general additional terms that may be expressed with
-higher powers in $\Delta t$, however, for sufficiently small $\Delta t$ those can be safely neglected.
+The dots in the above expressions indicate that there are in general additional terms that may be expressed with higher powers in $\Delta t$, however, for sufficiently small $\Delta t$ those can be safely neglected.
 Thus we may define the time derivative via
 \begin{equation}
-\vec v(t) = \lim_{\Delta t\to 0} \frac{\Delta \vec x(t)}{\Delta t} =: \frac{d\vec x(t)}{dt}  = \dot{\vec  x}(t) \ ,
+\vec v(t) = \lim_{\Delta t\to 0} \frac{\Delta \vec r(t)}{\Delta t} =: \frac{d\vec r(t)}{dt}  = \dot{\vec  r}(t) \ ,
 \end{equation}
-where $\Delta \vec x(t)=\vec x(t+\Delta t)-\vec x(t)$ and we introduced with the last expression a common short hand notation
-for time derivatives. Analogously we get
+where $\Delta \vec r(t)=\vec r(t+\Delta t)-\vec r(t)$ and we introduced with the last expression a common short hand notation for time derivatives.
+Analogously we get
 \begin{eqnarray}
 \vec a(t) &=& \lim_{\Delta t\to 0} \frac{\Delta \vec v(t)}{\Delta t} =: \frac{d\vec v(t)}{dt}  = \dot{\vec v}(t) \ , \\
- &=& \frac{d^2\vec x(t)}{dt^2}  = \ddot{\vec x}(t) \ ,
+ &=& \frac{d^2\vec r(t)}{dt^2}  = \ddot{\vec r}(t) \ ,
 \end{eqnarray}
@@
 
-It was Newton who observed that if a body is at rest it will remain at rest, and if it is in motion it will remain in motion in a 
-constant velocity in a straight line, unless it is acted upon by some force --- this is known as Newtons first law. 
-Said differently: a force $\vec F$ shows up by changing the motion of some object. This is quantified in Newton's second law
+It was Newton who observed that if a body is at rest it will remain at rest, and if it is in motion it will remain in motion in a constant velocity in a straight line, unless it is acted upon by some force --- this is known as Newtons first law.
+Said differently: a force $\vec F$ shows up by changing the motion of some object.
+This is quantified in Newton's second law
 \begin{equation}
-\vec F(\vec x, t) = \frac{d}{dt}(m \vec v) \ .
+\vec F(\vec r, t) = \frac{d}{dt}(m \vec v) \ .
 \end{equation} 
-If the mass does not change~\footnote{A well known example where $m$ does change with time is a 
-rocket, whose mass decreases as the rocket rises.} with time, this reduces to the well known
+If the mass does not change~\footnote{%
+	A well known example where $m$ does change with time is a rocket, whose mass decreases as the rocket rises.
+} with time, this reduces to the well known
 \begin{equation}
-\vec F(\vec x) = m \vec a(t) = m\dot{\vec v}(t) = m\ddot{\vec{x}}(t) \ . \label{eq:newton2}
+\vec F(\vec r) = m \vec a(t) = m\dot{\vec v}(t) = m\ddot{\vec{r}}(t) \ . \label{eq:newton2}
 \end{equation} 
-Note that in general the force could depend also on the time or the velocity. We here restrict ourselves to the
-case relevant for our example, where the force depends on the location only. 
-Therefore, 
-as soon as the force $\vec F(\vec x)$ is known for all $\vec x$ one can in principle calculate the trajectory, e.g. solving eq.~(\ref{eq:newton2})
-for $\vec x(t)$.
-Sometimes this requires some advanced knowledge in math, sometimes no closed form solution exists. However, alternatively 
-one can calculate the whole trajectory
-of some test body that experiences this force by a successive application of the rules 
-given in Eqs.~(\ref{eq:vdef}) and (\ref{eq:adef}):
-%
-\todo{See previous comment.}
-%
+Note that in general the force could depend also on the time or the velocity.
+We here restrict ourselves to the case relevant for our example, where the force depends on the location only. 
+Therefore, as soon as the force $\vec F(\vec r)$ is known for all $\vec r$ one can in principle calculate the trajectory, e.g. solving Eq.~(\ref{eq:newton2}) for $\vec r(t)$.
+Sometimes this requires some advanced knowledge in math, sometimes no closed form solution exists.
+However, alternatively one can calculate the whole trajectory of some test body that experiences this force by a successive application of the rules given in eqs.~(\ref{eq:vdef}) and (\ref{eq:adef}):
 \begin{enumerate}
-\item For a given time $t$, where $\vec x(t)$ and $\vec v(t)$ are known, use Eq.~(\ref{eq:newton2}) to calculate $\vec a(t)$.
-\item Use Eq.~(\ref{eq:vdef}) to calculate $\vec x(t+\Delta t)$ and
+\item For a given time $t$, where $\vec r(t)$ and $\vec v(t)$ are known, use Eq.~(\ref{eq:newton2}) to calculate $\vec a(t)$.
+\item Use Eq.~(\ref{eq:vdef}) to calculate $\vec r(t+\Delta t)$ and
 \item then use Eq.~(\ref{eq:adef}) to calculate $\vec v(t+\Delta t)$.
@@
 \end{enumerate}
-Clearly to initiate the procedure at some time $t_0$ both $\vec x(t_0)$ as well as $\vec v(t_0)$ must be known --- the 
-trajectories depend on these initial conditions.~\footnote{In general a differential equation of $n^{\rm th}$ degree (where
-the highest derivative is $n$) needs $n$ initial conditions specified. For $n=2$ those are often chosen as location and velocity
-at some defined time, but one may as well pick two location at different times. }
+Clearly to initiate the procedure at some time $t_0$ both $\vec r(t_0)$ as well as $\vec v(t_0)$ must be known --- the  trajectories depend on these initial conditions~\footnote{%
+	In general a differential equation of $n^{\rm th}$ degree (where the highest derivative is $n$) needs $n$ initial conditions specified.
+	For $n=2$ those are often chosen as location and velocity at some defined time, but one may as well pick two location at different times.%
+}
 
-For this procedure to work $\Delta t$ must be sufficiently small. What this means depends on the process
-studied. One way to estimate, if $\Delta t$ is small enough, is to verify, if the relation
+For this procedure to work $\Delta t$ must be sufficiently small.
+What this means depends on the process studied.
+One way to estimate, if $\Delta t$ is small enough, is to verify, if the relation
 \begin{equation}
-|\vec v(t)| \gg \frac12|\vec a(t)|\Delta t = \frac{1}{2m} |\vec F(\vec x(t)) |\Delta t\ . 
+|\vec v(t)| \gg \frac12|\vec a(t)|\Delta t = \frac{1}{2m} |\vec F(\vec r(t)) |\Delta t\ . 
 \label{eq:check}
 \end{equation}
-holds, since the next term neglected in Eq.~(\ref{eq:vdef}) reads $(1/2)a(t)(\Delta t)^2$.  Eq.~({\ref{eq:check}})
-also shows that small (large) time steps are necessary (sufficient), if the force is strong (weak), since
-the time steps need to be small enough that all changes induced by the force get resolved. Clearly,
-a relation as Eq.~({\ref{eq:check}}) can only provide guidance and can not replace a careful numerical check of
-the solutions: A valid result has to be insensitive the concrete value $\Delta t$ chosen --- in particular replacing
-$\Delta t$ by $\Delta t/2$ should not change the result significantly.
+holds, since the next term neglected in Eq.~(\ref{eq:vdef}) reads $(1/2)a(t)(\Delta t)^2$.
+Eq.~({\ref{eq:check}}) also shows that small (large) time steps are necessary (sufficient), if the force is strong (weak), since the time steps need to be small enough that all changes induced by the force get resolved.
+Clearly, a relation as Eq.~({\ref{eq:check}}) can only provide guidance and can not replace a careful numerical check of the solutions: A valid result has to be insensitive the concrete value $\Delta t$ chosen --- in particular replacing $\Delta t$ by $\Delta t/2$ should not change the result significantly.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-For this concrete example the starting point for the force is Newtons law
-of gravitation
+For this concrete example the starting point for the force is Newtons law of gravitation
 \begin{equation}
-\vec F_N(\vec x) = - \frac{G_N m M_\odot}{r^2} \frac{\vec x}{r}\ ,
+\vec F_N(\vec r) = - \frac{G_N m M_\odot}{r^2} \frac{\vec r}{r}\ ,
 \end{equation}
-where $G_N=6.67\times 10^{-11}$ m$^3$kg$^{-1}$s$^{-2}$ is the Newtonian constant of gravitation,
-$m$ is the mass of Mercury and $M_\odot=2\times 10^{30}$ kg is the mass of the sun.
-In addition $r=|\vec x(t)|$ denotes the distance between sun and Mercury, when we assume that the sun
-is infinitely more heavy than the planet and located at the center of the coordinate system. 
+where $G_N=6.67\times 10^{-11}$ m$^3$kg$^{-1}$s$^{-2}$ is the Newtonian constant of gravitation, $m$ is the mass of Mercury and $M_\odot=2\times 10^{30}$ kg is the mass of the sun.
+In addition $r=|\vec r(t)|$ denotes the distance between sun and Mercury, when we assume that the sun is infinitely more heavy than the planet and located at the center of the coordinate system. 
 Although this is not exact, since $m/M_\odot\sim 10^{-8}$ this is a pretty good approximation.
-%
-\todo{This possibly can become a problem for our accuracy because it is at the 10\% level of our expected result, right? Same for $c$.}
-%
 For later convenience we introduce the Schwarzschildradius of the sun
@@
 where $c=3\times 10^8$ m/s denotes the speed of light.
-Note that $r_S$ is the characteristic length scale of the gravitational field of the sun
-for --- up to the prefactor --- one can not form another quantity with dimensions of a length from $G_N$, $M_\odot$ and
-$c$. This is what we need $r_S$ for later in the discussion. That the Schwarzschild radius is
-also an important quantity to characterize black holes is not relevant for the discussion
-at hand.
+Note that $r_S$ is the characteristic length scale of the gravitational field of the sun for --- up to the prefactor --- one can not form another quantity with dimensions of a length from $G_N$, $M_\odot$ and $c$.
+This is what we need $r_S$ for later in the discussion.
+That the Schwarzschild radius is also an important quantity to characterize black holes is not relevant for the discussion at hand.
 With this Newtons second law reads 
-\todo{Or would it be better to write: $\ddot{x} = \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)$ and $\ddot{\vec x} = - \ddot{x} \frac{\vec{x}}{r}$?}
 \begin{equation}
-\ddot{\vec x} = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right) \frac{\vec{x}}{r} \ .
+	\ddot{r}      = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right) \, , \qquad
+	\ddot{\vec r} =   \ddot{r} \, \frac{\vec{r}}{r} \, .
 \label{eq:newton}
 \end{equation}
-%\begin{itemize}
-%\item{\bf First task:}
-%Now the students should implement the program outlined in Sec.~\ref{sec:tva} and visualize the resulting
-%trajectories of Mercury for different start parameters.
-%Why is it sufficient to work in two dimensions using simply 
-%$\vec x(t)=(x(t),y(t))^T$?
-% When the parameters are chosen appropriately close
-%orbits should emerge that are fixed in time --- in particular the perihelion (the point of closest approach
-%of planet and sun) stays fixed in space.
-%\end{itemize}
-In general an attractive force that vanishes at large distances leads,
- depending on the initial conditions, either to bounded orbits or to open orbits --- in the latter case the planet simply disappears from the sun.
+In general an attractive force that vanishes at large distances leads,  depending on the initial conditions, either to bounded orbits or to open orbits --- in the latter case the planet simply disappears from the sun.
 Depending on time available for the course it might be helpful to explain to the students the origin of this, however, since the most transparent reasoning needs an integration as well as the concept of angular momentum not necessarily familiar to the whole class, it appears in general more appropriate to work out together with the students that a moving body can not be kept by a force if its momentum is too large.
@@
 In particular the point of closest approach of the planet to the sun, the perihelion, does not move.
-However, as soon as the potential deviates from $1/r$, the perihelion moves. Because of this, the behavior of the perihelion
-is a very sensitive probe of the gravitational potential.
+However, as soon as the potential deviates from $1/r$, the perihelion moves.
+Because of this, the behavior of the perihelion is a very sensitive probe of the gravitational potential.
 
 The observed perihelion motion of Mercury is nowadays determined as $(574.10\pm 0.65)''$ per
-100 earth years~\footnote{The symbol $''$ denotes 'arc seconds': 1$''=(1/3600)^o$. }, but it should be noted that the bulk of this number
+100 earth years~\footnote{The symbol $''$ denotes "arc seconds": 1$''=(1/3600)^o$. }, but it should be noted that the bulk of this number
 can be understood by the presence of other planets within the Newtonian theory, since
 their gravitational force also acts on Mercury. However, a
-residual motion of $\delta \phi_M = (43)''$ in 100 earth years remained unexplained, until Einstein quantified the
-predictions of General Relativity to this particular observable. \todo{citation here}
+residual motion of $\delta \Theta_M = (43)''$ in 100 earth years remained unexplained, until Einstein quantified the
+predictions of General Relativity to this particular observable. \todo{@CH:citation here}
 
@@
 \begin{equation}
-\ddot{\vec x} = \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\left(1+\alpha\frac{r_S}{r}\right) \ ,
+\ddot{\vec r} = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\left(1+\alpha\frac{r_S}{r}\right) \, \frac{\vec{r}}{r} \ ,
 \label{eq:newton_art}
@@
 \begin{equation}
-	\delta \phi_V \approx \frac{1}{T_V} \delta \phi_M T_M \frac{r_{MS}}{r_{VS}} \approx 9'' \, ,
+	\delta \Theta_V \approx \frac{1}{T_V} \delta \Theta_M T_M \frac{r_{MS}}{r_{VS}} \approx 9'' \, ,
 \end{equation}
@@
 particular
-the perihelion motion of Mercury.  We select \python~\cite{}\todo{Citation} as  programming language, because \python is easy to learn, 
-intuitive to understand and an open source language\footnote{See also appendix \ref{appendix:python} for an instruction 
-how to set up \python on different operating systems.}.  No prior knowledge of \python or any other programming language 
-is required, because we explain all the steps necessary to create the simulation.  We also provide a working 
-example \cite{}
-%
-\todo{One can additional files to the arXiv upload} that can be used as template.
-%
+the perihelion motion of Mercury.
+We select \python{}\footnote{\url{https://www.python.org}} as  programming language, because \python is easy to learn, intuitive to understand and an open source language.
+No prior knowledge of \python or any other programming language is required, as we explain all the necessary steps to create the simulation.
+In appendix \ref{appendix:python} we give instructions for setting up \python on different operating systems.
+We also provide a working example program that can be used as a template \url{this-upload-url}\todo{url once uploaded (arXiv)}.
 
-To start the simulation one needs the "initial" distances and velocities, which are needed as input to the equation of motion (Eq.~(\ref{eq:x_update}) and Eq.~(\ref{eq:v_update})).
+To start the simulation one needs the "initial" distances and velocities, which are needed as input to the equation of motion [Eq.~(\ref{eq:x_update}) and Eq.~(\ref{eq:v_update})].
 As described above, here  one can simplify the problem by placing one planet (the "infinitely" heavy sun) in the center of the coordinate system and describe the motion of the other planet (mercury) in a two-dimensional plane.
 
-Because the problem does not depend on the initial position of mercury on its orbit around the sun, we extract the parameters at the perihelion with $ \abs{\vec r_{MS}(0)} = 46 \cdot 10^6$km and $ \abs{\vec v_M(0)} = 59$km/s \footnote{https://nssdc.gsfc.nasa.gov/planetary/factsheet/mercuryfact.html}.
+Because the problem does not depend on the initial position of mercury on its orbit around the sun, we extract the parameters at the perihelion with $ \abs{\vec r_{MS}(0)} = 46 \cdot 10^6$km and $ \abs{\vec v_M(0)} = 59$km/s\footnote{\url{https://nssdc.gsfc.nasa.gov/planetary/factsheet/mercuryfact.html}}.
 Since the computer does not understand physical units, one has to express each variable in an appropriate unit.
@@
 by expressing distances in $R_0 = 10^{10}$ m and time intervals in days, $T_0 = 1$d, where $d$ refers to earth days, such that one mercury year is given by $T_M=88\,T_0$.
-With this choice, the initial distance of mercury to the sun, the size of the initial velocity of mercury and the acceleration prefactor become
-\todo{Maybe we should rename $a_M$ to avoid confusion about the fact, that it is not an acceleration?}
-\begin{equation}
-r_{MS}(0) = 4.6 R_0 \, , \quad
-v_{M}(0)  = 0.51 \frac{R_0}{T_0} \, ,  \quad
-a_M       = 0.99 \frac{R_0}{T_0^2} \frac{1}{\left(r_{MS}/R_0\right)^2} 
+With this choice, the initial distance of mercury to the sun, the size of the initial velocity of mercury and the acceleration pre-factors become
+\begin{eqnarray}
+r_{MS}(0)   &=& 4.6 R_0 \, , \quad
+v_{M}(0)    = 0.51 \frac{R_0}{T_0} \, ,  \\
+a_M(r_{MS}) &=& 1.01 \frac{R_0}{T_0^2} \frac{1}{\left(r_{MS}/R_0\right)^2} 
 \, .
-\end{equation}
+\end{eqnarray}
 
@@
 	\includegraphics[width=.4\textwidth]{figs/sun_merc.pdf}
-	\caption{\label{fig:sun_merc}Sun mercury system with relevant vectors.  Because the mercury is at its perihelion, its velocity is perpendicular to its direct connection vector with the sun.}
+	\todo[inline]{@CK: figure: coordinate system}
+	\caption{\label{fig:sun_merc}%
+		Sun mercury system with relevant vectors.
+		Because the mercury is at its perihelion, its velocity is perpendicular to its direct connection vector with the sun.%
+	}
 \end{figure}
@@
 vM0 = 0.51   # in units of R0/T0
-aM  = 0.99   # in units of R0/T0**2 * R0**2
+c_a = 1.01   # in units of R0/T0**2
 TM  = 88.    # in units of T0
@@
 # Import the class vector from vpythons module visual
-from visual import vector
+from vpython import vector
 # Initialize distance and velocity vectors
@@
 According to Eq.~(\ref{eq:newton}), the force which acts on mercury changes the velocity of mercury which eventually changes the position.
-However, we first need to fix the time step $\Delta t$ and use the estimate
-of Eq.~(\ref{eq:check}) as a guidance
+However, we first need to fix the time step $\Delta t$ and use the estimate of Eq.~(\ref{eq:check}) as a guidance
 \begin{lstlisting}
 # Definition of the time step
-dt = 2 * vM0 / aM / 10
+dt = 2 * vM0 / c_a / 10
 \end{lstlisting}
 Here the factor $1/10$ makes sure that $\Delta t$ is indeed consistent with Eq.~(\ref{eq:check}).
-Now we are in the position to calculate
-location and velocity of the planet at $t_0+\Delta t$ using the following commands \footnote{Note, that the order of calculating $a_{MS}$, $v_M$ and $r_M$ is not clear a priori.
-All prescriptions lead to the same result for sufficiently small $dt$, but the following order turns out to be numerically most stable.}
+Now we are in the position to calculate location and velocity of the planet at $t_0+\Delta t$ using the following commands \footnote{%
+	Note, that there are different notions for numerically integrating differential equations with different accuracys.
+	Thus, the order of updating $a_{MS}$, $v_M$ and $r_M$ is not unique, but all correct prescriptions lead to the same result for sufficiently small $dt$.
+	For this particular problem, the following definition is in good balance between numerically stable and inexpensive to compute.
+}%
 \begin{lstlisting}
 # Compute the strength of the acceleration
-aMS = aM * ( 1 + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
+aMS = c_a * ( 1 + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
 # Multiply by the direction
@@
 \end{lstlisting}
-Note the beauty of working with the predefined \texttt{vector} class: the basic vector operations are already implemented.
+Note basic vector operations are already implemented in the predefined \texttt{vector} class.
 The difference and sum of two vectors, or the scalar vector multiplication return vectors themselves.
-Also the magnitude of a vector -- \texttt{vector.mag} -- is a property of the vector and can be easily extracted.
+Also the magnitude of a vector --- \texttt{vector.mag} --- is a property of the vector and can be easily extracted.
 
-It is handy to use \texttt{Python}s \texttt{function}s to embed repeating structures ("DRY" -- Don't Repeat Yourself)
+It is handy to use \texttt{Python}s \texttt{function}s to embed repeating structures ("DRY" --- Don't Repeat Yourself)
 \begin{lstlisting}
@@
 \python{$\!'$}s syntax enforces a clean programming style: it is necessary that the body of the function is indented relative to the definition statement of the function.
-This is visualized by the long underscore in the previous block.
 Furthermore, \python is an Interpreter language.
@@
 
-Finally, we can express the evolution by a \texttt{while}-loop \todo{Should we also add  e.g. T=2*TM here?}
-\begin{lstlisting}
+Finally, we can express the evolution by a \texttt{while}-loop
+\begin{lstlisting}%[float,floatplacement=H]
 t     = 0
 alpha = 0
-# Execute the loop as long as t < T
-while t < T:
+# Execute the loop as long as t < 2*TM
+while t < 2*TM:
 	vec_rM, vec_vM = evolve_mercury(vec_rM, vec_vM, alpha)
@@
 Note the required indent of the loop structure similar to the indent of a function.
-In each iteration of the \texttt{while}-loop, the previous distance and velocity are used to compute the new values -- which directly overwrite the previous values and thus enter the next iteration.
-The total runtime \texttt{T} is the amount of "virtual" days the simulations should run.
-To describe at least one full orbital period, $T$ needs to be larger than $T_M$.
-With the previous choice for \texttt{dt}, this corresponds to roughly $N_T = 10^4$ evolution steps.
+In each iteration of the \texttt{while}-loop, the previous distance and velocity are used to compute the new values --- which directly overwrite the previous values and thus enter the next iteration.
+The total runtime \texttt{2*TM} is the amount of "virtual" days the simulations should run.
+To describe at least one full orbital period, this time needs to be larger than $T_M$.
+With the previous choice for \texttt{dt}, this corresponds to roughly $N_T \approx 2 \cdot 10^3$ evolution steps.
 
@@
 \begin{lstlisting}
-from visual import vector, sphere, color, curve, rate, display
+from vpython import vector, sphere, color, curve, rate
 \end{lstlisting}
-The class sphere will represent mercury and the sun in the simulation \todo{The pictures on the next page are done with sun radius 1.5 and mercury radius 0.5.
-Would it be ok, if we change the following radii to these values?}
+The class sphere will represent mercury and the sun in the simulation
 \begin{lstlisting}
 # define the initial coordinates; M = mercury, S = sun
-M = sphere( pos=vec_rM0,       radius=0.01, color=color.red   )
-S = sphere( pos=vector(0,0,0), radius=0.1,  color=color.yellow)
+M = sphere( pos=vec_rM0,       radius=0.5,  color=color.red   )
+S = sphere( pos=vector(0,0,0), radius=1.5,  color=color.yellow)
 # and the initial velocities
@@
 S.velocity = vector(0,0,0)
+# add a visible trajectory to mercury
+M.trajectory = curve(color=color.white)
 \end{lstlisting}
@@
 
-To further visualize the trajectory of mercury we add this property to our \texttt{mercury} object
-\begin{lstlisting}
-# add a visible trajectory to mercury
-M.trajectory = curve(color=color.black)
-\end{lstlisting}
-Next we tell python that the objects shall be drawn from this point on \todo{For some reason display does not work for me.
-However I like black as background color anyway.
-Would it be ok, if we leave out the display and change the color of the trajectory to white?}
-\begin{lstlisting}
-# display the objects from now on
-display(background=color.white)
-\end{lstlisting}
-and last but not least, instead of updating the object-unrelated vectors in the while loop, we update the object-related vectors which will eventually be drawn \todo{changed rate ok?}
+Last but not least, instead of updating the object-unrelated vectors in the while loop, we update the object-related vectors which will eventually be drawn
 \begin{lstlisting}
@@
 	M.pos , M.velocity = evolve_mercury(M.pos , M.velocity , alpha)
-	# update tim
+	# update time
 	t = t + dt
@@
 To get the perihelion motion, the term from equation (\ref{eq:newton_art}) has to be be added by choosing $\alpha=0$.
-This is a good opportunity to let the students play with the size of $\alpha$ and get a feeling for its impact on the trajectories.
-However, it might be advisable to not start out with the correct mercury values but a more excentric trajectory by choosing e.g.
-$r_{MS}(0)=6R_0$.
-Then the perihelion motion is better visible.
-This is illustrated in figure \ref{fig:MercuryOrbit}, where we used a value of $\alpha=10^6$.
+This is a good opportunity to let the students play with the size of $\alpha$ and get a feeling for its impact on the trajectories or analyze what happens when changing the time-steps $\Delta t$.
+However, it might be advisable to not start out with the correct mercury values but a more excentric trajectory by choosing e.g. $r_{MS}(0)=6R_0$.
+Then the perihelion motion is better visible. This is illustrated in figure \ref{fig:MercuryOrbit}, where we used a value of $\alpha=10^6$.
 
@@
 	\includegraphics[width=.26\textwidth]{figs/PerihelionMotion.png}
+	\todo[inline]{@CK: reproduce figure}
 	\caption{\label{fig:MercuryOrbit} Output for $\alpha=0$ (left), $\alpha=10^5$ and bigger starting distance $r_{MS}(0)=6R_0$ (middle) and $\alpha=10^5$ with original starting values.}
 \end{figure}
-\todo{@CK: reproduce figure}
-
-%------------------------------------------
-\subsection{Running the Program}
-%------------------------------------------
-The complete program needs to be saved as a python file, e.g. \texttt{mercury\_orbit.py}.
-Depending on the editor of choice, there might already be an execute functionality e.g. \textbf{bla...} for the recommended windows installation (see \ref{appendix:python}).
-%
-\todo{Look up the windows example...}
-%
-In case one uses a linux like system and wants to execute the code from a console/terminal, one has to to execute the command
-\begin{center}
-	\texttt{python 'address/to/python/file.py'}
-\end{center}
-If one already works in the file directory and has used the suggested naming scheme, this reads 
-\begin{center}
-	\texttt{python mercury\_orbit.py}
-\end{center}
-Here, the specified \texttt{python} interpreter needs to agree with the installed \texttt{VPython} version.
-For more information see \ref{appendix:python}.
 
@@
 \begin{itemize}
-\item extract the position of the perihelion
-\item calculate the angle between the perihelions for successive turns and compute the average angle $\delta \Theta (\alpha)$
-\item repeat the above points for different values of $\alpha$ and confirm the linear dependence
-\item extrapolate to $\alpha=1$
+\item extract multiple positions of the perihelion $\vec r_{MS} ( t_\mathrm{ph}^{(n)} )$ for a fixed value of $\alpha$ and simulations times which covers several orbits,
+\item calculate the angle between the perihelions for two successive turns and compute the average angle $\delta \Theta (\alpha)$ over all individually computed angles
+\item repeat the above points for different values of $\alpha$ to confirm the linear dependence of $\delta \Theta (\alpha)$
+\item extrapolate $\delta \Theta (\alpha)$ to $\alpha=1$
 \end{itemize}
-\todo{I changed the text here in an effort to make the purpose and structure of this section more clear. However we can change it back, if you prefer the old version.}\\
 
 There are several ways to find the position of the perihelion.
-The easiest one is to look for the point of minimal distance to the sun -  the definition of the perihelion.
+The easiest one is to look for the point of minimal distance to the sun --- the definition of the perihelion.
 This can be done within the $while$-loop.
-To find out when the minimum is reached, the last two positions \texttt{vec\_r\_last} and \texttt{vec\_r\_before\_last} have to be known.
-If the length of the last vector is smaller than that of the previous one as well as the current vector, one has passed the perihelion and its position is given by \texttt{vec\_r\_last}.
-Suppose we want to save the position of the perihelion for a certain number of turns (here e.g.
-\texttt{max\_turns}=10) in a list \texttt{list\_perih}.
-\todo{I am not sure this is the best way to do this, since I need to introduce yet another concept (lists).
-However I could not think of an elegant solution...}
-The implementation can be realised as follows \todo{I have a problem here, since I cannot use M.pos in this way.
-For me M.pos seems to be a pointer, because after one turn \texttt{vec\_r\_before\_last}, \texttt{vec\_r\_last} and M.pos all have the same value always.
-Therefore I need to use another variable and set M.pos to this value in the end.
-Could you check, whether it is the same for you?
-If yes, I will change the code accordingly, if not we have to compare codes again...}
+To find identify a minimal distance to the sun, one has to know the last two positions \texttt{vec\_r\_last} and \texttt{vec\_r\_before\_last} in addition to the current position.
+If the length of the \texttt{last} vector is smaller than that of the \texttt{before\_last} vector and smaller than the current vector, one has passed the perihelion and its position is given by \texttt{vec\_r\_last}.
+It is useful to save the position of the perihelion in a list \texttt{list\_perih} for a certain number of turns (here e.g. \texttt{max\_turns}=10).
+The implementation can be realized as follows
 \begin{lstlisting}
@@
 max_turns  = 10
-list_perih=[]
+list_perih = list()
+\end{lstlisting}
+\begin{lstlisting}
 # find perihelion for each turn and print it out
@@
 	vec_r_before_last = vec_r_last
-	vec_r_last        = M.pos
+	# store position of mercury in a new vector (since we will change M.pos)
+	vec_r_last = vector(M.pos)
 	#<...update mercury position...>
 	# check if at perihelion
-	if vec_r_before_last.mag > vec_r_last.mag < M.pos.mag:
+	if (vec_r_last.mag < M.pos.mag) and (vec_r_last.mag < vec_r_before_last.mag):
 		list_perih.append(vec_r_last)
@@
  \begin{equation}
- 	\sphericalangle(\vec{v}_{n-1},\vec{v}_n) = \cos^{-1} \left( \frac{\vec{v}_{n-1} \cdot \vec{v}_n}{|\vec{v}_{n-1}|\:|\vec{v}_n|} \right)
+ 	\sphericalangle(\vec{v}_{1},\vec{v}_2) = \mathrm{arccos} \left( \frac{\vec{v}_{1} \cdot \vec{v}_2}{|\vec{v}_{1}|\:|\vec{v}_2|} \right)
+	\, .
  \end{equation}
@@
 # import function to resolve angle
-from visual import acos, pi, dot
-angle = acos(
-	dot(list_perih[n-1], list_perih[n]) / (list_perih[n-1].mag * list_perih[n].mag) 
-) * 180  / pi
+from vpython import acos, pi, dot
+# define function for angle extraction
+def angle_between(v1, v2):
+	return acos( dot(v1, v2) / (v1.mag * v2.mag) ) * 180. / pi
 \end{lstlisting}
-To minimise the statistical errors these can be averaged over a few turns.
-Depending on the programming proficiency of the students and time constraints, this can be either done by hand or e.g.
-by implementing the following code
+To account for the statistical errors (e.g., numerical rounding errors) one can be average over a few turns.
+Depending on the programming proficiency of the students and time constraints, this can be either done by hand or, e.g., by implementing the following code
 \begin{lstlisting}
@@
 	# calculate angle
-    angle = ...
-    sum_angle = sum_angle+angle
+	sum_angle = sum_angle + angle_between(list_perih[n-1], list_perih[n])
 # display the average
@@
 
-As explained above (and further discussed in chapter \ref{sec:analysis}) the natural value for $\alpha$ would be of the order of $1$.
-However, if the student use this value for $\alpha$, they will find that the change in the trajectories is close to invisible and uncertainty is much larger than the result.
-Therefore, to estimate the size of the perihelion motion, it is more advisable to use the fact that there is an approximately linear dependence between $\alpha$ and $\delta \Theta$.\footnote{Of course for very large values of $\alpha$ this linear relation will eventually break down.
-We therefore advise to choose $\alpha<7\cdot 10^6$ at which point the perihelion motion exceeds $\delta \Theta=90^{\degree}$.}
-The students can convince themselves of this fact by plotting some angles they extracted from their simulation over $\alpha$.
-This can be done either by hand or using another program like e.g. Excel.
-One such plot can be found in figure \ref{fig:AlphaAngle}.
+As explained above [and further discussed in chapter \ref{sec:analysis}] the natural value for $\alpha$ would be of the order of $1$.
+However, if the students use this value for $\alpha$, they will find that the change in the trajectories is close to invisible and the numerical uncertainty is much larger than the result.
+Therefore, to estimate the size of the perihelion motion, it is more advisable to use the fact that there is a linear dependence between $\alpha$ and $\delta \Theta$\footnote{%
+	Of course for very large values of $\alpha$ this linear relation will eventually break down.
+	We therefore advise to choose $\alpha<7\cdot 10^6$ at which point the perihelion motion exceeds $\delta \Theta=90^{\degree}$.
+}.
+The students can convince themselves of this fact by plotting the angles $\delta \Theta$ they have extracted from their simulation over $\alpha$.
+This can be done either by hand, \python (e.g., with \texttt{matplotlib}: \url{https://matplotlib.org}), or using another program like, e.g., Excel.
+For instance figure \ref{fig:AlphaAngle} demonstrates such a plot.
 
@@
 	\centering
-	\includegraphics[width=.5\textwidth]{figs/AlphaAngle.png}
+	\includegraphics[width=.8\textwidth]{figs/AlphaAngle.png}
+	\todo[inline]{@CK: pdf figure + confirm @IH: distribute data}
 	\caption{\label{fig:AlphaAngle} Linear relation between $\alpha$ and the perihelion motion.}
@@
 
-To extract the perihelion motion at $\alpha = 1$, we use the fact that it would be zero at
- $\alpha =0$. 
-Considering in addition one result from the numeric simulation, e.g. $\delta\Theta(\alpha = 10^6) \approx 10.1^{\degree}$, the sought-after angle can be calculated from the linear interpolation as
+To extract the perihelion motion at $\alpha = 1$, we use the fact that $\delta \Theta$ is zero for $\alpha =0$ --- if one could completely eliminate numerical errors.
+Considering one result from the numeric simulation, e.g. $\delta\Theta(\alpha = 10^6) \approx 10.1^{\degree}$, the sought-after angle can be calculated by a linear interpolation as
 \begin{equation}
 	\delta\Theta (\alpha=1) = \frac{10.1^{\degree}}{10^6} = 0.036''
+	\, .
 \end{equation}
-Note when comparing the results of the simulation with the experimental result that one earth year corresponds to $T_E/T_M \approx 365d/88d\approx 4.15$ mercury years.
+Depending on the knowledge of the students, multiple points and also estimated numerical uncertainties could be used to extract this value by a linear regression.
+Note that, when comparing the results of the simulation with the experimental result, one Earth year corresponds to $T_E/T_M \approx 365d/88d\approx 4.15$ mercury years.
 Therefore, the perihelion motion per 100 earth years is $0.036''\cdot 415\approx 15''$.
 To test whether this is a reasonable result, the students could first follow the dimensional analysis and then compare to the actual value for mercury as outlined above.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\clearpage\subsection{Cross-Check Results}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\todo[inline]{Here I (CK) just make a few comments regarding the results.}
-%First of all, I can confirm Inkas result up to a factor of 10.
-%I believe this factor arises because we have analysed 10 years of motions (\verb|max_turns = 10|).
-%See figure \ref{fcc1}.
-%This result is quite nice, because we basically say the genreal relativity motion by mercury is caused by $\alpha \approx 3$.
-%However there is a small issue: we do not have the numerical precision to claim this.
-%\begin{figure}[htb]
-%	\centering
-%	\includegraphics[width=.4\textwidth]{figs/alpha_interpolation.pdf}
-%	\caption{\label{fcc1}Interpolation for $\phi$ depending on $\alpha$ and $dt = 2 v_0 / a_0 / 5000$. }
-%\end{figure}
-%
-%I have encountered, that we do not start perfectly at the perihelion or aphelion (depending on the boundary conditions): my simulations arrives at this point after roughly a tenth of a day and therefore I have an offset which needs to be subtracted off after each mercury year.
-%More interestingly, this offset seems to depend on the tilmestep of the simulation (see figure \ref{fcc2}).
-%\begin{figure}[htb]
-%	\centering
-%	\includegraphics[width=.4\textwidth]{figs/initial_time_shift.pdf}
-%	\caption{\label{fcc2}Initial shift of perihelion for $\delta\theta = \delta\phi$ in degrees over $\alpha$ for different time steps (color).}
-%\end{figure}
-%
-%One can also see that the size of the time spacing affects a potential uncertainty in the periodicity of the orbit when looking at the precession growth over different turns (see figure \ref{fcc3}).
-%\begin{figure}[htb]
-%	\centering
-%	\includegraphics[width=.7\textwidth]{figs/precession_growth.pdf}
-%	\caption{\label{fcc3}Precession growth of perihelion for $\delta\theta = \delta\phi$ in degrees over number of turns for different time steps (color).
-One should pay special attention to the oscillations within the growth for relatively large time steps.}
-%\end{figure}
-%
-%Last but not least, for the precession interpolation to small $\alpha$, one should worry about the precision of the precession for small alpha.
-%Even after subtracting of the results for zero turns, basically non of my results agree with zero for alpha equal to zero [to go from degrees per mercury year to arcsecs per 100 earth years, one has to multiply by $1.46 \times 10^{6}$.](see figure \ref{fcc4}).
-%\begin{figure}[htb]
-%	\centering
-%	\includegraphics[width=.4\textwidth]{figs/alpha_dependence_0}
-%	%
-%	\includegraphics[width=.4\textwidth]{figs/alpha_dependence_1}
-%	\caption{\label{fcc4}Perihelion precession $\delta\theta = \delta\phi$ in degrees over $\alpha$ without (left figure) or with (right figure) subtracting off the zero turn result.
-The uncertainty corresponds to the variation over different turns and the results are displayed for different time steps (color).}
-%\end{figure}
-%
-%Because, after subtracting the initial angle as a reference, the slope of all the plots basically agree, I suggest the following prescription (as Inka actually did)
-%\begin{enumerate}
-%	\item Compute the precession for $\alpha = 0$ for several turns,
-%	\item compute the precession for preferably large $\alpha > 1000$ for several turns,
-%	\item verify that they are indeed on a line (linear relation),
-%	\item subtract the zero value from the other values (since we know that it has to be zero and the uncertainty is coming from the time steps),
-%	\item do the interpolation to $\delta \phi (\alpha_{GR}) = 43''$ for two values: $\delta \phi (0) = 0$ and $\delta \phi (\alpha_l) = \delta \phi_l$
-%	\item repeat this for several $(\alpha_l, \delta \phi_l)$ and get the average result for $\alpha_{GR}$.	
-%\end{enumerate}
-%
-%
-%\clearpage
 
@@
 
-Physical experiments in general, as do numerical simulations of physics, always involve uncertainties.
-It is important for the students to recognise this fact and to understand, that the cause of the uncertainties lies in the finite time steps.
+Physical experiments in general, as do numerical simulations of physics, must always account for possible uncertainties.
+It is important for the students to recognize this fact and to understand, that the cause of the uncertainties lies, e.g., in the finite time steps.
 The students should therefore explore the source and size of the errors arising in this simulation at least qualitatively.
 
-Consider figure \ref{fcc3}, where the angle of the perihelion relative to the y-axis is shown for the first 10 turns for different choices of $dt$ and $\alpha$.
-As expected, the correlation between the number of turns and $\Theta$ is linear and the slope depends solely on $\alpha$.
+Consider figure \ref{fcc3}, where the angle of the perihelion relative to the $y$-axis is shown for the first $10$ turns for different choices of $dt$ and $\alpha$.
+As expected, the correlation between the number of turns and $\delta\Theta$ is linear and the slope depends solely on $\alpha$.
 However, there is an offset between the lines for different $dt$.
 The reason is, due to the uncertainties in the simulation, the starting position does not coincide with the first perihelion as one might expect.
-This does not generate a problem for the procedure described in the previous section
-since the starting position is not considered. Still, it is helpful to be aware of this fact.\\
+This does not generate a problem for the procedure described in the previous section since the starting position is not considered. Still, it is helpful to be aware of this fact.
 One can see nicely in figure \ref{fcc3} that the oscillations of the perihelion motion increases for growing time steps.
@@
 	\includegraphics[width=.7\textwidth]{figs/precession_growth.pdf}
-	\caption{\label{fcc3}Precession growth of perihelion for $\delta\theta = \delta\phi$ in degrees over number of turns for different time steps (color).
-One should pay special attention to the oscillations within the growth for relatively large time steps.}
-\end{figure} \todo{Could you redo this figure with some bigger time steps?
+	\todo[inline]{@CK: Could you redo this figure with some bigger time steps?
 Could we call the precession growth $\Theta(\alpha)$ instead of $\delta\Theta$?
 I think this would fit better with our notation so far...}
+	\caption{\label{fcc3}Precession growth of perihelion for $\delta\Theta$ in degrees over number of turns for different time steps (color).
+One should pay special attention to the oscillations within the growth for relatively large time steps.}
+\end{figure}
 
@@
 For this study, it is advantageous to set $\alpha =0$, since in this case the trajectories should be closed and therefore deviations from the ideal case can be spotted easily.
-In figure \ref{fig: LargeDT} the mercury trajectory for $dt=2 * v_M(0) / a_M \cdot 5$ is shown.
+In figure \ref{fig: LargeDT} the mercury trajectory for $dt=2 * v_M(0) / c_a \cdot 5$ is shown.
 One can clearly see the failure of the simulation to reproduce the physical trajectory.
 From examples like this it should be apparent that the time steps influence the accuracy of the simulation.
+One could quantify the deviation of $\delta\Theta$ from zero and use this to estimate the numerical systematic uncertainty for a given time step $dt$.
 Only in the limit $dt \rightarrow 0$ would the simulation perfectly reproduce the actual trajectory.
@@
 
-\todo{It would be nice if they could also get a quantitive estimate of the error, what do you think would be best?
-Maybe they can just look at the perihelion motion for $\alpha=0$ and take that as an estimate of the systematic error?}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-Dimensional analysis is not only a tool that allows one to cross check if the results of some simulation is
-of the right order of magnitude, it is also very helpful to identify unusual dynamics in some system.
+Dimensional analysis is not only a tool that allows one to cross check if the results of some simulation is of the right order of magnitude, it is also very helpful to identify unusual dynamics in some system.
 Especially the latter aspect should become clear from the discussion in this section.
 
-The idea of dimensional analysis is that in a system that can be controlled by expanding the relevant quantities
-(like the force) in some small parameter(s), the coefficients in the expansion should turn out to be of order unity (that
-means anything between about 0.1 and 10 is fine - but 0.01 or 100 is irritating) --- parameters in line with this
-are called 'natural'. Applied to the problem at hand
-given by Eq.~(\ref{eq:newton}) this statement implies that from naturalness one would expect that 
-the parameter $\alpha$ is of order 1 when one uses
-for $r$ the average distance Mercury-sun, namely $\bar r=6\times 10^7$ km.
+The idea of dimensional analysis is that in a system that can be controlled by expanding the relevant quantities (like the force) in some small parameter(s), the coefficients in the expansion should turn out to be of order unity (that means anything between about 0.1 and 10 is fine - but 0.01 or 100 is irritating) --- parameters in line with this are called "natural".
+Applied to the problem at hand given by Eq.~(\ref{eq:newton}) this statement implies that from naturalness one would expect that the parameter $\alpha$ is of order 1 when one uses for $r$ the average distance Mercury-sun, namely $\bar r=6\times 10^7$ km.
 From this one estimates for the expected angular shift per orbit
 \begin{equation}
-\delta \phi \simeq 2\pi\left(\frac{r_S}{\bar r}\right) = (\pi \times 10^{-7}) \ \mbox{rad} = (2\times 10^{-5})^o = (7\times 10^{-2}) ^{''} \ ,
+\delta \Theta \simeq 2\pi\left(\frac{r_S}{\bar r}\right) = (\pi \times 10^{-7}) \ \mbox{rad} = (2\times 10^{-5})^o = (7\times 10^{-2}) ^{''} \ ,
 \end{equation}
 which leads to a shift of about $30^{''}$ in 100 earth years to be compared to the empirical value of $43^{''}$.
- Thus indeed the
-amount of perihelion motion of Mercury is in line with expectations, $if$ --- and this is an important 'if' ---
-the Newtonian dynamics is simply the leading term of some more general underlying theory. In particular,
-no new scales enter in the correction terms.
+Thus indeed the amount of perihelion motion of Mercury is in line with expectations, $if$ --- and this is an important "if" --- the Newtonian dynamics is simply the leading term of some more general underlying theory.
+In particular, no new scales enter in the correction terms.
 
-Thanks to Einstein we indeed know that the conclusion formulated above is correct ---- the more general underlying
-theory is General Relativity and indeed Einstein was able to quantitatively explain the perihelion motion of Mercury
-from his equations. 
-On the other hand had we found a dramatic deviation of $\alpha$ from unity one would have concluded that 
-there is probably some other dynamics going on that drives this difference.
+Thanks to Einstein we indeed know that the conclusion formulated above is correct ---- the more general underlying theory is General Relativity and indeed Einstein was able to quantitatively explain the perihelion motion of Mercury from his equations.
+On the other hand had we found a dramatic deviation of $\alpha$ from unity one would have concluded that there is probably some other dynamics going on that drives this difference.
 
-Indeed, we now several of those hierarchy problems in modern physics: e.g. the so called QCD $\theta$ term,
-expected to be of natural size, is at present known to be at most $10^{-10}$. This smallness, called the
-strong CP problem, is so irritating that physicists like S. Weinberg even proposed that there must exist an
-additional particle, the axion, whose interactions are in charge of pushing $\theta$ even to zero and there
-are now intense searches for this axion going on various labs.
+Indeed, we now several of those hierarchy problems in modern physics: e.g., the so called QCD $\Theta$ term, expected to be of natural size, is at present known to be at most $10^{-10}$.
+This smallness, called the strong CP problem, is so irritating that physicists like S.
+Weinberg even proposed that there must exist an additional particle, the axion, whose interactions are in charge of pushing $\Theta$ even to zero and there are now intense searches for this axion going on various labs.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
+\todo{@IH: this could be a little bit more verbose. Maybe include in section below.}
 Ideas for questions:
@@
 
-In chapter \ref{sec:Numerical Implementation} we suggested a certain way to present the material.
-By using a template as well as a step by step instruction, the students are guided rather strictly through the problem solution.
-We presented this approach with young students and limited time in mind.
-However if the students are more experienced and enough time is available, it is certainly advisable to give the students space for exploring the problem independently.
-This could be achieved by the following changes or additions to the concept presented in chapter \ref{sec:Numerical Implementation}
+In chapter \ref{sec:Numerical Implementation} we suggested to present the material by using a template as well as a step by step instruction to guide the students through the problem solution.
+These instructions can be cut down or left out depending on available time and numerical/computational versatility of the students.
+This could be achieved by the following changes or additions to the concept presented in chapter \ref{sec:Numerical Implementation}.
 
@@
 Instead of providing the template to the students, they could build the program from scratch.
-Of course, this requires some basic knowledge in \texttt{VPython}, as they could acquire for example by working through a ball in the box example or the like.
-They could even look up the needed parameters on their own.
+Of course, this requires some basic knowledge in \texttt{VPython}, as they could acquire for example by working through a ball in the box example\cite{}\todo{@IH: cite ball in the box} or the like.
+Also, they could independently research the problem relevant parameters.
 \item \textbf{Why can we work in one plane?}
 
-In the code the third coordinate is never used. On the first glance, this could seem like a simplification. Let the students work out on their own, why this does not imply any loss of generality.
+In the code the third coordinate ($z$) is never used.
+On the first glance, this might seem like a simplification.
+The students could work out on their own, why this does not imply any loss of generality.
 \item \textbf{What is the impact of the different parameters?}
@@
 Especially if the needed parameters are not specified beforehand, the students might have to experiment a bit, before getting the correct trajectories.
-But even if they are given, it might be beneficial to encourage the students to play with a few parameters, like mass or starting velocity, and track their impact on the trajectories.
-This way the students get a much better feeling for the physics involved.
+But even if they are given, it might be beneficial to encourage the students to play with a few parameters, like the masses or the starting velocities, and observe their impact on the trajectories.
+This way the students get a better feeling for the physics involved.
 \end{itemize}
@@
 
-Simulations always involve a balance between the time needed for the calculations and the accuracy of the results achieved.
-Even though this is a rather simple example, it contains some opportunities to make this concept accessible to the students.
+Simulations always involve a balance between the time needed for the calculations and the demanded accuracy for the results.
+Even though this is a rather simple example, it contains some opportunities to make these concepts accessible to the students.
 
@@
 
-To make the students realise, that the finite time steps lead to inaccuracies in the trajectories encourage them to increase the time steps.
-This is best done in a program with unmodified gravitational force, as in this case the trajectories are closed ellipses and a deviation is most prominent.
+To make the students realize that the finite time steps lead to inaccuracies in the trajectories, one can propose to test this by increasing the size of the time steps.
+This is best done in a program with unmodified gravitational force ($\alpha=0$), since in this case the trajectories are closed ellipses and numerical errors are most prominently visible in form of deviations from a bounded orbit.
 If the students choose the time steps big enough, they should witness big deviations.
@@
 
-However in practice there is a limit to decreasing the time steps, because the time needed for the calculation grows simultaneously. By including
-import time
+However in practice there is a limit to decreasing the time steps, because the time needed for the calculation grows simultaneously.
+By including \texttt{import time}
 \begin{lstlisting}
@@
 the students can measure the time needed by their program.
-By varying dt they can validate, that there is indeed approximately an anti-proportional dependence.
-(Note: This only works if the time in the loop is increased by dt, so t=t+dt.)
+By varying $dt$ they can validate, that there is indeed approximately an anti-proportional dependence.
+(Note: This only works if the time in the loop is increased by $dt$, so $t=t+dt$.)
 \item \textbf{Verlet integration}
 
-Of course by optimising the code, an improvement in accuracy can be achieved without increasing the calculation time.
-The simplest way to demonstrate this might be implementing Verlet integration instead of using the simple Euler method.
+Of course by optimizing the code, an improvement in accuracy can be achieved without changing $dt$.
+The simplest way to demonstrate this might be given by the implementation of the Verlet\cite{}\todo{@IH: cite Verlet} integration instead of using the simple Euler method.
 \end{itemize}
@@
 It might be interesting to abandon the simplification of a stationary sun, as it nicely illustrates Newton's third law.
-Here it might also be advisable to reduce the mass of the sun to have a visible result.
+Here it might also be advisable to reduce the mass of the sun to have a more visible result.
 
@@
 
-Ambitious students could even include another planet and see how the two planets interact.
+Ambitious students could even include further planets and see how the different planets interact.
 This is especially interesting, when discussing the perihelion motion of mercury, as it is mainly due to the influence of the other planets.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\todo[inline]{@CK: installation python+vpython win+linux}
 \subsection{Mac}
+
 \href{http://vpython.org/contents/download_mac.html}{Download VPython for Mac}
+
+%------------------------------------------
+\subsection{Running the Program}
+%------------------------------------------
+\todo[inline]{@CK: running python win+linux}
+The complete program needs to be saved as a python file, e.g. \texttt{mercury\_orbit.py}.
+Depending on the editor of choice, there might already be an execute functionality.
+In case one uses a linux like operating system and can execute the code from a console/terminal by running the command
+\begin{center}
+	\texttt{python 'address/to/python/file.py'}
+\end{center}
+If one already works in the file directory and has used the suggested naming scheme, this reads 
+\begin{center}
+	\texttt{python mercury\_orbit.py}
+\end{center}
+Here, the specified \texttt{python} interpreter needs to agree with the installed \texttt{VPython} version.
+For more information see \ref{appendix:python}.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\todo[inline]{@CK: inline code + script + test code + confirm data}
 Here we should put the code
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\bibliographystyle{alpha din}
-\bibliography{MercuryNumerical}
+%\bibliographystyle{alphadin}
+%\bibliography{MercuryNumerical}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2017-12-01 tex/MercuryNumerical.tex
Added appendix

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\newcommand{\python}[0]{\texttt{Python} }
+\newcommand{\python}[0]{\texttt{Python}}
+\newcommand{\vpython}[0]{\texttt{VPython}}
 \newcommand{\abs}[1]{\left\vert #1 \right\vert}
+\newcommand{\code}[1]{{\scriptsize\colorbox{light-gray}{\texttt{#1}}}}
 
@@
 the perihelion motion of Mercury.
-We select \python{}\footnote{\url{https://www.python.org}} as  programming language, because \python is easy to learn, intuitive to understand and an open source language.
-No prior knowledge of \python or any other programming language is required, as we explain all the necessary steps to create the simulation.
-In appendix \ref{appendix:python} we give instructions for setting up \python on different operating systems.
-We also provide a working example program that can be used as a template \url{this-upload-url}\todo{url once uploaded (arXiv)}.
+We select \python{}\footnote{\url{https://www.python.org}} as  programming language, because \python{} is easy to learn, intuitive to understand and an open source language.
+No prior knowledge of \python{} or any other programming language is required, as we explain all the necessary steps to create the simulation.
+In appendix \ref{appendix:python} we give instructions for setting up \python{} on different operating systems.
+We also provide a working example program that can be used as a template \url{this-upload-url}\todo{@arXiv: online link}.
 
@@
 
-In \python, this reads
+In \python{}, this reads
 \begin{lstlisting}
@@
 Next, we set up the initial directions, which will describe the motion in the two-dimensional plane.
-We will build on the existing \python module \texttt{VPython}, which provides an implementation for treating vectors as well as their visualization.
+We will build on the existing \python{} module \texttt{VPython}, which provides an implementation for treating vectors as well as their visualization.
 The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.
@@
 \begin{lstlisting}
-# Define the function
+# Define the coordinate and velocity update function
 def evolve_mercury(vec_rM_old, vec_vM_old, alpha):
@@
 \python{$\!'$}s syntax enforces a clean programming style: it is necessary that the body of the function is indented relative to the definition statement of the function.
-Furthermore, \python is an Interpreter language.
+Furthermore, \python{} is an Interpreter language.
 Each line of the code is executed when the Interpreter passes it.
@@
 # define the initial coordinates; M = mercury, S = sun
-M = sphere( pos=vec_rM0,       radius=0.5,  color=color.red   )
-S = sphere( pos=vector(0,0,0), radius=1.5,  color=color.yellow)
+M = sphere(pos=vec_rM0,       radius=0.5,  color=color.red   )
+S = sphere(pos=vector(0,0,0), radius=1.5,  color=color.yellow)
 # and the initial velocities
@@
 # update the objects
-while t < T:
+while t < 2*TM:
 	# set the frame rate (you can choose a higher rate to accelerate the program)
@@
 list_perih = list()
-\end{lstlisting}
-\begin{lstlisting}
 # find perihelion for each turn and print it out
@@
 The students can convince themselves of this fact by plotting the angles $\delta \Theta$ they have extracted from their simulation over $\alpha$.
-This can be done either by hand, \python (e.g., with \texttt{matplotlib}: \url{https://matplotlib.org}), or using another program like, e.g., Excel.
+This can be done either by hand, \python{} (e.g., with \texttt{matplotlib}: \url{https://matplotlib.org}), or using another program like, e.g., Excel.
 For instance figure \ref{fig:AlphaAngle} demonstrates such a plot.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo[inline]{@CK: installation python+vpython win+linux}
-\subsection{Mac}
+In short, to run the above commands and scripts, one needs to install \python{} and \vpython{}.
+We recommend the newest \vpython{} version 7 or higher which best runs with the \python{} versions of 3.5 or higher.
+It is also possible to use \python{} 2.7 or higher, but the 3.x versions have better \vpython{} support.
 
-\href{http://vpython.org/contents/download_mac.html}{Download VPython for Mac}
+For the advanced users, one simply has to update \python{} to these versions and install \vpython{} via the package manager of choice, e.g, \texttt{pip} [see also \url{http://vpython.org}].
 
-%------------------------------------------
-\subsection{Running the Program}
-%------------------------------------------
-\todo[inline]{@CK: running python win+linux}
-The complete program needs to be saved as a python file, e.g. \texttt{mercury\_orbit.py}.
-Depending on the editor of choice, there might already be an execute functionality.
-In case one uses a linux like operating system and can execute the code from a console/terminal by running the command
-\begin{center}
-	\texttt{python 'address/to/python/file.py'}
-\end{center}
-If one already works in the file directory and has used the suggested naming scheme, this reads 
-\begin{center}
-	\texttt{python mercury\_orbit.py}
-\end{center}
-Here, the specified \texttt{python} interpreter needs to agree with the installed \texttt{VPython} version.
-For more information see \ref{appendix:python}.
+Below, we provide detailed descriptions how to install these versions of \python{} and \vpython{} for different operating systems.
+Independent of the operating systems, one has to execute the following steps:
+\begin{enumerate}
+	\item install and update \python{} (at best 3.x, x$\geq5$),
+	\item install and update a \python{} package manager,
+	\item install \vpython{} and dependencies.
+\end{enumerate}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Linux based systems (including macOS)}\label{appendix:python-max}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+All Linux based system come with an installation of \python{}.
+Thus, if not already at the newest version, one just have to update existing installations.
+This is best done by using a distribution manager.
+Native to Linux based systems is for example \texttt{dpkg} and \texttt{apt-get}, or \texttt{homebrew} and \texttt{MacPorts} for macOS.
+While \texttt{dpkg} and \texttt{apt-get} usually come with the linux installations, macOS user must install such distribution managers at first.
+For this we refer to \url{https://www.macports.org/install.php} or \url{https://docs.brew.sh/Installation.html}.
+At first one should open a terminal/console (bash).
+Depending on the installation, one usually needs superuser rights in order to update \python{}.
+Therefore one has to prepend \code{sudo} to the following commands and confirm them by entering the superuser password.
+The installations using a distribution manager are save and tested by the vast community.
+In the following, the successive order of commands to update python is presented for \texttt{apt-get} (Linux) and \texttt{MacPorts} (macOS).
+\begin{table}[h!]
+\centering
+\footnotesize
+\begin{tabular}{l || l | l}
+	& \texttt{apt-get} & \texttt{MacPorts} \\\hline
+	1. Update the distribution manager & \code{sudo apt-get update} & \code{sudo port selfupdate} \\
+	2. Install \& update \python{} & \code{sudo apt-get install python3} & \code{sudo port install python36} \\
+	3. Installing \python{} package manager & \code{sudo apt-get install python3-pip} & \code{sudo port install py36-pip}
+\end{tabular}
+\normalsize
+\end{table}
+
+Depending on the previously installed packages, this command might install further dependencies automatically (after confirmation).
+One should test if the installation was successful by typing \code{python3 --version} or \code{python3.6 --version} (depending on the installation name) which prints out the version of the python installation (and should be larger than 3.5).
+Next one must update the package manager to get the newest versions of python packages.
+This is done by running
+\begin{table}[h!]
+\centering
+\footnotesize
+\begin{tabular}{l || l | l}
+	& \texttt{apt-get} & \texttt{MacPorts} \\\hline
+	4. Update \code{pip} & \code{pip3 install --upgrade --user pip wheel} & \code{pip-3.6 install --upgrade --user pip wheel} \\
+\end{tabular}
+\normalsize
+\end{table}
+
+Note that this just upgrades the packages for the current user.
+If one wants the systemwide installation, one again has to prepend \code{sudo} and can ignore the \code{--user} flag (\code{sudo pip3 install --upgrade pip wheel}).
+Furthermore, depending on the installed version of \code{pip}, it could have a different name.
+This is best figured out by typing \code{pip} and pressing \texttt{tab}.
+Finally, one can install \vpython{} by running
+\begin{table}[h!]
+\centering
+\footnotesize
+\begin{tabular}{l || l | l}
+	& \texttt{apt-get} & \texttt{MacPorts} \\\hline
+	5. Install \vpython{} & \code{pip3 install --user vpython} & \code{pip-3.6 install --user vpython} \\
+\end{tabular}
+\normalsize
+\end{table}
+
+At this point, everything needed is installed and one can start the simulation.
+To test the \vpython{} installation and should run \code{python3} or \code{python3.6}, which starts a \python{} console.
+Next, one shall run \code{from vpython import *}, which loads the \vpython{} package and opens the default browser.
+Running \code{box()} should generate a white box in the browser window, which can be rotated using the mouse.
+The program is quit by running \code{quit()} or pressing \texttt{Ctrl-c}.
+
+In general, there are multiple ways to run python scripts.
+For the sake of this paper, we suggest two scenarios.
+
+The first scenario to run python scripts is that one writes and saves a \python{} file, e.g., \texttt{test.py}, which can be edited using any text editor of choice.
+To run the python script one must navigate the terminal to the containing folder using the \code{cd} (and \code{ls}) command.
+While \code{ls} prints out the folders and files in the current directory, \code{cd} changes the directory.
+The command \code{cd} either takes relative paths (directories which are in the current directory), e.g., \code{cd folder}, or absolute paths beginning with a \code{/}, e.g., \code{cd /Users/name/Desktop/}.
+Once one has navigated to folder which contains the python file --- e.g., the file name is listed when typing \code{ls} --- than one can type \code{python3 test.py} or \code{python3.6 test.py} to execute the script.
+Again, this script can be ended (if it does not end automatically) by pressing \texttt{Ctrl-c}.
+
+The second scenario to run \python{} scripts is given by the interactive \texttt{ipython} or \texttt{jupyter} notebooks.
+This feature is automatically installed when installing the newest version of \vpython{}.
+To start an interactive session, one has to run \code{jupyter notebook} (or equivalent, depending on the installation name.
+Again, this can be figured out by typing \code{jupyter} and pressing \texttt{tab}).
+This will print a link (containing the name \texttt{localhost}) and open the browser at the link-address.
+Within this browser window, one can freely navigate to the directory of choice and create interactive \python{} and \vpython{} sessions by clicking on existing files or creating new ones (upper right button \texttt{new}$\rightarrow$\vpython{}).
+
+We provide example files for both scenarios.\todo{@arXiv: online link}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Windows}\label{appendix:python-windows}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+As the first step, one must install \python{}, again there are several possibilities but we suggest \texttt{Anaconda} for an intuitive installation.
+Following the link \url{https://conda.io/docs/user-guide/install/download.html}, one has two possible (free) options, \texttt{Anaconda} and \texttt{Miniconda}.
+Both are equally sufficient, because one can install further updates after the initial installation.
+In the case of \texttt{Miniconda} (\url{https://conda.io/miniconda.html}) one shall download the newest \python{} 3.x (x $\geq5$) installer for windows and the right system type (32-bit or 64-bit).
+The system type can be found by pressing \texttt{Windows+I}, navigating to \texttt{System}$\rightarrow$\texttt{About} and looking at the \texttt{System type} entry.
+
+Installing one of the two executables creates a new windows application called \texttt{Anaconda Prompt}.
+Starting this application opens a terminal which we will user to install the required \python{} packages.
+At first, one should test if the \texttt{Anaconda} installation was successful by running the command \code{conda list}.
+This command should print a list of already installed packages.
+Also one should run \code{conda update conda} to make sure one has the most recent installation.
+
+To install \vpython{} one simply has to run \code{conda install -c vpython vpython}.
+The additional flag \code{-c vpython} tells \texttt{Anaconda} the "channel" where to look for the \vpython{} package.
+Running this command prints out a list of dependencies which must be installed in order for \vpython{} to work --- this must be confirmed.
+
+In principle this is already sufficient to run \vpython{} scripts.
+One could furthermore install \texttt{Anaconda}s native navigator \texttt{anaconda-navigator} by running \code{conda install anaconda-navigator}.
+This will install another application, \texttt{Anaconda Navigator}, which can be run from the Desktop and makes it easier to program in \python{}.
+
+Again there are several possibilities to run \python{} files on \texttt{Windows}.
+\begin{itemize}
+	\item One could create a file \texttt{test.py}, edit it with the editor of choice and specify that this file should be opened with \texttt{Python3}.
+		To do so, \texttt{right click} on  the file, select \texttt{open with} and locate the \texttt{Anaconda Python3} installation.
+		Programs can be closed by closing the running window.
+	\item Start the \texttt{Anaconda Navigator} application, open an interactive \texttt{qtconsole} and program within the newly opened terminal.
+		Programs can be closed by closing the running window.
+	\item Start the \texttt{Anaconda Navigator} application and click on the \texttt{jupyter notebook} button.
+		This opens an interactive \python{} session in the default web-browser.
+		Here one open existing notebooks or create new files (upper right button \texttt{new}$\rightarrow$\vpython{}).
+\end{itemize}
+
+To start and test the \vpython{} installation one should pick any out of the three options and execute the code
+\begin{lstlisting}
+from vpython import *
+box()
+\end{lstlisting}
+Running this code should open the default browser and render a white box on black background.
+Using the mouse (right click and drag), one should be able to rotate the box.
+It was found that the \texttt{Internet Explorer} is not able to properly render the animation.
+Here we suggest to use the \texttt{Firefox} web-browser as the default choice.
+
+Once this test case works, one can start to test the mercury orbit simulation.
+We provide example files for both \texttt{juypter notebooks} and regular \python{} files.
+
+\todo{@arXiv: online link}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\clearpage
 \section{The code}\label{sec:code}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo[inline]{@CK: inline code + script + test code + confirm data}
-Here we should put the code
+Below we provide an example script on how to program the visualization of the mercury orbit in \vpython{}.
+\lstinputlisting[language=Python]{mercury_orbit_base.py}
+Further examples and template files can be found online\cite{}.\todo[inline]{@arXiv: online link}
 

### 2017-12-01 tex/mercury_orbit_base.py

@@ -1,0 +1,41 @@
+# Import all objects from the vpython module
+from vpython import *
+
+# Definition of parameters 
+rM0 = 4.6    # in units of R0
+vM0 = 0.51   # in units of R0/T0
+c_a = 1.01   # in units of R0/T0**2
+TM  = 88.    # in units of T0 
+rS  = 3.e-7  # in units of R0
+
+# Initialize distance and velocity vectors
+vec_rM0 = vector(0, rM0, 0)
+vec_vM0 = vector(vM0, 0, 0)
+
+# Definition of the time step
+dt = 2 * vM0 / c_a / 10
+
+# Define the coordinate and velocity update
+def evolve_mercury(vec_rM_old, vec_vM_old, alpha):
+    aMS = c_a * ( 1 + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
+    vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
+    vec_vM_new = vec_vM_old + vec_aMS * dt
+    vec_rM_new = vec_rM_old + vec_vM_new * dt
+    return vec_rM_new, vec_vM_new
+
+# Define the initial coordinates; M = mercury, S = sun
+M = sphere(pos=vec_rM0,       radius=0.5,  color=color.red   )
+S = sphere(pos=vector(0,0,0), radius=1.5,  color=color.yellow)
+M.velocity = vec_vM0
+S.velocity = vector(0,0,0)
+# add a visible trajectory to mercury
+M.trajectory = curve(color=color.white)
+
+# Start the simulation
+t=0
+alpha = 0
+while t < TM:
+    rate(100)
+    M.trajectory.append(pos=M.pos)
+    M.pos , M.velocity = evolve_mercury(M.pos , M.velocity , alpha)
+    t = t + dt

### 2017-12-01 tex/MercuryNumerical.tex
upadeted orbit figures

@@
 \usepackage{listings}                 % Code listings  See: https://en.wikibooks.org/wiki/LaTeX/Source_Code_Listings
+
+\usepackage{subcaption} % subfigures
 
@@
 	\centering
-	\includegraphics[width=.25\textwidth]{figs/MercuryOrbit.png} \quad
-	\includegraphics[width=.273\textwidth]{figs/PerihelionMotionLargeV.png} \quad
-	\includegraphics[width=.26\textwidth]{figs/PerihelionMotion.png}
-	\todo[inline]{@CK: reproduce figure}
-	\caption{\label{fig:MercuryOrbit} Output for $\alpha=0$ (left), $\alpha=10^5$ and bigger starting distance $r_{MS}(0)=6R_0$ (middle) and $\alpha=10^5$ with original starting values.}
+	\begin{subfigure}[b]{0.3\textwidth}
+		\includegraphics[width=\textwidth]{figs/a0T5dt20.png}
+		\caption{\label{fig:MercuryOrbit-a0-small}}
+	\end{subfigure}
+	~
+	\begin{subfigure}[b]{0.3\textwidth}
+		\includegraphics[width=\textwidth]{figs/a6T5dt20.png}
+		\caption{\label{fig:MercuryOrbit-a6-small}}
+	\end{subfigure}
+	~
+	\begin{subfigure}[b]{0.3\textwidth}
+		\includegraphics[width=\textwidth]{figs/a6T10dt20.png}
+		\caption{\label{fig:MercuryOrbit-a6-big}}
+	\end{subfigure}
+	\caption{\label{fig:MercuryOrbit}%
+		Different mercury orbits for $\alpha=0$ (a), $\alpha = 10^{6}$ (b) and $\alpha = 10^{6}$ but $r_{MS}(0) = 6 R_0$ (c).
+		All computations where executed with $dt = 2 * v_{M}(0) / c_a / 10$.
+		Note that for different starting values of $r_{MS}(0)$, also the numbers of days for a "mercury year" changes.
+	}
 \end{figure}
@@
 	\centering
-	\includegraphics[width=.3\textwidth]{figs/TrajectoryLargeDT.png}
-	\caption{\label{fig: LargeDT} Deviation of the trajectory from the analytic result (indicated in blue) for large time steps (here $dt=2 * v_M(0) / a_M \cdot 5$)}
+	\includegraphics[width=.5\textwidth]{figs/num-err.png}
+	\caption{\label{fig: LargeDT}%
+		Deviation of the trajectory from precise numeric result (black line; $dt=2 v_M(0) / a_M / 20$) for large time steps (red line; $dt=2 v_M(0) / a_M \times 2$).
+		Both orbits are plotted for $T = 10 T_M$.
+	}
 \end{figure}

### 2017-12-01 analysis/generate_figs.py

@@ -1,0 +1,65 @@
+# Import all objects from the vpython module
+from vpython import *
+
+scene.background = color.white
+scene.width   = 1024
+scene.height  = 1024
+scene.center  = vector(0, -3, 0)
+scene.userspin = False
+scene.userzoom = False
+scene.fullscreen    = True
+
+# Definition of parameters 
+rM0 = 4.6    # in units of R0
+vM0 = 0.51   # in units of R0/T0
+c_a = 1.01   # in units of R0/T0**2
+TM  = 88.    # in units of T0 
+rS  = 3.e-7  # in units of R0
+
+# Initialize distance and velocity vectors
+vec_rM0 = vector(0, rM0, 0)
+vec_vM0 = vector(vM0, 0, 0)
+
+# Definition of the time step
+dt = 2 * vM0 / c_a / 20
+
+# Define the coordinate and velocity update
+def evolve_mercury(vec_rM_old, vec_vM_old, alpha):
+    aMS = c_a * ( 1 + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
+    vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
+    vec_vM_new = vec_vM_old + vec_aMS * dt
+    vec_rM_new = vec_rM_old + vec_vM_new * dt
+    return vec_rM_new, vec_vM_new
+
+# Define the initial coordinates; M = mercury, S = sun
+M = sphere(pos=vec_rM0,       radius=0.0,  color=color.red   )
+S = sphere(pos=vector(0,0,0), radius=1.5,  color=color.yellow)
+M.velocity = vec_vM0
+S.velocity = vector(0,0,0)
+# add a visible trajectory to mercury
+M.trajectory = curve(color=color.black, radius=0.05)
+
+# Start the simulation
+t=0
+alpha = 0
+while t < 2*TM:
+    rate(10000)
+    M.trajectory.append(pos=M.pos)
+    M.pos , M.velocity = evolve_mercury(M.pos , M.velocity , alpha)
+    t = t + dt
+
+# Define the initial coordinates; M = mercury, S = sun
+M = sphere(pos=vec_rM0,       radius=0.5,  color=color.red   )
+M.velocity = vec_vM0
+S.velocity = vector(0,0,0)
+# add a visible trajectory to mercury
+M.trajectory = curve(color=color.red, radius=0.01)
+
+# Start the simulation
[…]

### 2017-12-01 tex/mercury_orbit_base.py

@@ -34,7 +34,7 @@ M.trajectory = curve(color=color.white)
 # Start the simulation
 t=0
 alpha = 0
-while t < TM:
+while t < 2*TM:
     rate(100)
     M.trajectory.append(pos=M.pos)
     M.pos , M.velocity = evolve_mercury(M.pos , M.velocity , alpha)

### 2017-12-01 analysis/perihelion-full.py
added full example code and added coordinate system to figure

@@ -1,0 +1,93 @@
+# Import the class vector from vpythons module visual
+from vpython import *
+
+# set up display
+scene.width      = 1680
+scene.height     = 1024
+scene.background = color.white
+scene.center     = vector(0,-2,0)
+
+# Definition of parameters
+rM0 = 4.6    # in units of R0
+vM0 = 0.51   # in units of R0/T0
+aM  = 0.99   # in units of R0/T0**2
+TM  = 88.    # in units of T0
+rS  = 3.e-7  # in units of R0
+
+# Initialize distance and velocity vectors
+vec_rM0 = vector(0, rM0, 0)
+vec_vM0 = vector(vM0, 0, 0)
+
+# define the initial coordinates; M = mercury, S = sun
+M = sphere(pos=vec_rM0,       radius=0.5,  color=color.red   )
+S = sphere(pos=vector(0,0,0), radius=1.5,  color=color.yellow)
+# and the initial velocities
+M.velocity = vec_vM0
+S.velocity = vector(0,0,0)
+
+# add a visible trajectory to mercury
+M.trajectory = curve(color=color.black, radius=0.005)
+
+# Definition of the time step
+dt = 2. * vM0 / aM / 10
+
+# Define the function
+def evolve_mercury(vec_rM_old, vec_vM_old, alpha):
+    # Compute the strength of the acceleration      
+    aMS = aM * ( 1. + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
+    # Multiply by the direction
+    vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
+    # Update velocity vector
+    vec_vM_new = vec_vM_old + vec_aMS * dt
+    # Update position vector
+    vec_rM_new = vec_rM_old + vec_vM_new * dt
+    return vec_rM_new, vec_vM_new
+
+# define function for angle extraction
+def angle_between(v1, v2):
+    return acos( dot(v1, v2) / (v1.mag * v2.mag) ) * 180. / pi
+
+# Run parameters
+alpha      = 1.e6
+vec_r_last = vec_rM0
+turns      = 0
+max_turns  = 10
+list_perih = list()
+sum_angle  = 0.
+
+# find perihelion for each turn an print it out
+while turns < max_turns:
+    vec_r_before_last = vec_r_last
+    vec_r_last        = vector(M.pos)
+    # set the frame rate: shows four earth days at once
+    rate(100)
[…]

### 2017-12-01 tex/MercuryNumerical.tex
added full example code and added coordinate system to figure
plotted turn variation

@@
 \begin{equation}
-\vec v(t) = \lim_{\Delta t\to 0} \frac{\Delta \vec r(t)}{\Delta t} =: \frac{d\vec r(t)}{dt}  = \dot{\vec  r}(t) \ ,
+\vec v(t) = \lim_{\Delta t\to 0} \frac{\Delta \vec r(t)}{\Delta t} =: \frac{d\vec r(t)}{d t}  = \dot{\vec  r}(t) \ ,
 \end{equation}
@@
 	\centering
-	\includegraphics[width=.4\textwidth]{figs/sun_merc.pdf}
-	\todo[inline]{@CK: figure: coordinate system}
+	\includegraphics[width=.5\textwidth]{figs/sun_merc.pdf}
 	\caption{\label{fig:sun_merc}%
@@
 	Note, that there are different notions for numerically integrating differential equations with different accuracys.
-	Thus, the order of updating $a_{MS}$, $v_M$ and $r_M$ is not unique, but all correct prescriptions lead to the same result for sufficiently small $dt$.
+	Thus, the order of updating $a_{MS}$, $v_M$ and $r_M$ is not unique, but all correct prescriptions lead to the same result for sufficiently small $\Delta t$.
 	For this particular problem, the following definition is in good balance between numerically stable and inexpensive to compute.
@@
 		Different mercury orbits for $\alpha=0$ (a), $\alpha = 10^{6}$ (b) and $\alpha = 10^{6}$ but $r_{MS}(0) = 6 R_0$ (c).
-		All computations where executed with $dt = 2 * v_{M}(0) / c_a / 10$.
+		All computations where executed with $\Delta t = 2 v_{M}(0) / c_a / 20$.
 		Note that for different starting values of $r_{MS}(0)$, also the numbers of days for a "mercury year" changes.
@@
 	\centering
-	\includegraphics[width=.8\textwidth]{figs/AlphaAngle.png}
-	\todo[inline]{@CK: pdf figure + confirm @IH: distribute data}
-	\caption{\label{fig:AlphaAngle} Linear relation between $\alpha$ and the perihelion motion.}
+	\includegraphics{figs/alpha-angle.pdf}
+	\caption{\label{fig:AlphaAngle} Linear relation between $\alpha$ and the perihelion motion $\delta \Theta$ for $\Delta t=2v_M(0)/c_a/100$.}
 \end{figure}
@@
 To extract the perihelion motion at $\alpha = 1$, we use the fact that $\delta \Theta$ is zero for $\alpha =0$ --- if one could completely eliminate numerical errors.
-Considering one result from the numeric simulation, e.g. $\delta\Theta(\alpha = 10^6) \approx 10.1^{\degree}$, the sought-after angle can be calculated by a linear interpolation as
+Considering one result from the numeric simulation, e.g. $\delta\Theta(\alpha = 10^6) \approx 10.0^{\degree}$, the sought-after angle can be calculated by a linear interpolation as
 \begin{equation}
-	\delta\Theta (\alpha=1) = \frac{10.1^{\degree}}{10^6} = 0.036''
+	\delta\Theta (\alpha=1) = \frac{10.0^{\degree}}{10^6} = 0.036''
 	\, .
@@
 
-Consider figure \ref{fcc3}, where the angle of the perihelion relative to the $y$-axis is shown for the first $10$ turns for different choices of $dt$ and $\alpha$.
+Consider figure \ref{fcc3}, where the angle of the perihelion relative to the $y$-axis is shown for the first $10$ turns for different choices of $\Delta t$ and $\alpha$.
 As expected, the correlation between the number of turns and $\delta\Theta$ is linear and the slope depends solely on $\alpha$.
-However, there is an offset between the lines for different $dt$.
+However, there is an offset between the lines for different $\Delta t$.
 The reason is, due to the uncertainties in the simulation, the starting position does not coincide with the first perihelion as one might expect.
@@
 	\centering
-	\includegraphics[width=.7\textwidth]{figs/precession_growth.pdf}
-	\todo[inline]{@CK: Could you redo this figure with some bigger time steps?
-Could we call the precession growth $\Theta(\alpha)$ instead of $\delta\Theta$?
-I think this would fit better with our notation so far...}
-	\caption{\label{fcc3}Precession growth of perihelion for $\delta\Theta$ in degrees over number of turns for different time steps (color).
-One should pay special attention to the oscillations within the growth for relatively large time steps.}
+	\includegraphics[width=.99\textwidth]{figs/angular-variaton.pdf}
+	\todo[inline]{@all: where is this periodic oscillation coming from?}
+	\caption{\label{fcc3}Precession growth of perihelion $\delta\Theta$ in degrees depending on the number of turns $N_t$ for different time steps $\Delta t$ (color) and different values for $\alpha$ (columns).
+	Here, $\Delta t_0 \equiv 2 v_M(0)/c_a$.
+	One should pay special attention to the oscillations within the growth for relatively large time steps.
+}
 \end{figure}
@@
 For this study, it is advantageous to set $\alpha =0$, since in this case the trajectories should be closed and therefore deviations from the ideal case can be spotted easily.
-In figure \ref{fig: LargeDT} the mercury trajectory for $dt=2 * v_M(0) / c_a \cdot 5$ is shown.
+In figure \ref{fig: LargeDT} the mercury trajectory for $\Delta t=2 * v_M(0) / c_a \cdot 5$ is shown.
 One can clearly see the failure of the simulation to reproduce the physical trajectory.
 From examples like this it should be apparent that the time steps influence the accuracy of the simulation.
-One could quantify the deviation of $\delta\Theta$ from zero and use this to estimate the numerical systematic uncertainty for a given time step $dt$.
-Only in the limit $dt \rightarrow 0$ would the simulation perfectly reproduce the actual trajectory.
+One could quantify the deviation of $\delta\Theta$ from zero and use this to estimate the numerical systematic uncertainty for a given time step $\Delta t$.
+Only in the limit $\Delta t \rightarrow 0$ would the simulation perfectly reproduce the actual trajectory.
 \begin{figure}[htb]
@@
 	\caption{\label{fig: LargeDT}%
-		Deviation of the trajectory from precise numeric result (black line; $dt=2 v_M(0) / a_M / 20$) for large time steps (red line; $dt=2 v_M(0) / a_M \times 2$).
+		Deviation of the trajectory from precise numeric result (black line; $\Delta t=2 v_M(0) / a_M / 20$) for large time steps (red line; $\Delta t=2 v_M(0) / a_M \times 2$).
 		Both orbits are plotted for $T = 10 T_M$.
@@
 the students can measure the time needed by their program.
-By varying $dt$ they can validate, that there is indeed approximately an anti-proportional dependence.
-(Note: This only works if the time in the loop is increased by $dt$, so $t=t+dt$.)
+By varying $\Delta t$ they can validate, that there is indeed approximately an anti-proportional dependence.
+(Note: This only works if the time in the loop is increased by $\Delta t$, so $t=t+\Delta t$.)
 \item \textbf{Verlet integration}
 
-Of course by optimizing the code, an improvement in accuracy can be achieved without changing $dt$.
+Of course by optimizing the code, an improvement in accuracy can be achieved without changing $\Delta t$.
 The simplest way to demonstrate this might be given by the implementation of the Verlet\cite{}\todo{@IH: cite Verlet} integration instead of using the simple Euler method.

### 2017-12-01 analysis/perihelion-full.py

@@ -10,7 +10,7 @@ scene.center     = vector(0,-2,0)
 # Definition of parameters
 rM0 = 4.6    # in units of R0
 vM0 = 0.51   # in units of R0/T0
-aM  = 0.99   # in units of R0/T0**2
+aM  = 1.01   # in units of R0/T0**2
 TM  = 88.    # in units of T0
 rS  = 3.e-7  # in units of R0
 

### 2017-12-02 tex/MercuryNumerical.tex
minor updates

@@
 \usepackage{subcaption} % subfigures
+\usepackage{caption}    % itemize in caption
 
@@
 While their detailed study needs deeper knowledge in Differential Geometry, for the study outlined here very little math is necessary, such that high school students from about 10$^{\rm th}$ grade up should benefit from it.
-This was already demonstrated in the "Schülderakademie Teilchenphysik", where an early version of this course was tested successfully on a group of 10 German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade.
+This was already demonstrated in the "Schülderakademie Teilchenphysik", where an early version of this course was tested successfully on a group of 17 German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade.
 
@@
 We here restrict ourselves to the case relevant for our example, where the force depends on the location only. 
-Therefore, as soon as the force $\vec F(\vec r)$ is known for all $\vec r$ one can in principle calculate the trajectory, e.g. solving Eq.~(\ref{eq:newton2}) for $\vec r(t)$.
+Therefore, as soon as the force $\vec F(\vec r)$ is known for all $\vec r$ one can in principle calculate the trajectory, e.g., solving Eq.~(\ref{eq:newton2}) for $\vec r(t)$.
 Sometimes this requires some advanced knowledge in math, sometimes no closed form solution exists.
@@
 	For $n=2$ those are often chosen as location and velocity at some defined time, but one may as well pick two location at different times.%
-}
+}.
 
@@
 particular
-the perihelion motion of Mercury.
-We select \python{}\footnote{\url{https://www.python.org}} as  programming language, because \python{} is easy to learn, intuitive to understand and an open source language.
+the perihelion motion of Mercury.  We select \python{} (\url{https://www.python.org}) as  programming language, because \python{} is easy to learn, intuitive to understand and an open source language.
 No prior knowledge of \python{} or any other programming language is required, as we explain all the necessary steps to create the simulation.
@@
 		Sun mercury system with relevant vectors.
-		Because the mercury is at its perihelion, its velocity is perpendicular to its direct connection vector with the sun.%
+		Because mercury is at its perihelion, its velocity is perpendicular to its direct connection vector with the sun.%
 	}
@@
 # Definition of the time step
-dt = 2 * vM0 / c_a / 10
+dt = 2 * vM0 / c_a / 20
 \end{lstlisting}
-Here the factor $1/10$ makes sure that $\Delta t$ is indeed consistent with Eq.~(\ref{eq:check}).
-Now we are in the position to calculate location and velocity of the planet at $t_0+\Delta t$ using the following commands \footnote{%
+Here the factor $1/20$ makes sure that $\Delta t$ is indeed consistent with Eq.~(\ref{eq:check}).
+Now we are in the position to calculate location and velocity of the planet at $t_0+\Delta t$ using the following equations\footnote{%
 	Note, that there are different notions for numerically integrating differential equations with different accuracys.
-	Thus, the order of updating $a_{MS}$, $v_M$ and $r_M$ is not unique, but all correct prescriptions lead to the same result for sufficiently small $\Delta t$.
+	Thus, the ordering and exact equations for updating $a_{MS}$, $v_M$ and $r_M$ are not unique, but all correct prescriptions lead to the same result for sufficiently small $\Delta t$.
 	For this particular problem, the following definition is in good balance between numerically stable and inexpensive to compute.
@@
 def evolve_mercury(vec_rM_old, vec_vM_old, alpha):
-	#<...Code...>
+	# Compute the strength of the acceleration
+	aMS = c_a * ( 1 + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
+	# Multiply by the direction
+	vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
+	# Update velocity vector
+	vec_vM_new = vec_vM_old + vec_aMS * dt
+	# Update position vector
 	return vec_rM_new, vec_vM_new
@@
 \end{lstlisting}
-\python{$\!'$}s syntax enforces a clean programming style: it is necessary that the body of the function is indented relative to the definition statement of the function.
+\python{}s syntax enforces a clean programming style: it is necessary that the body of the function is indented (by an arbitrary but consistent amount of \texttt{space}s or \texttt{tab}s) relative to the definition statement of the function.
 Furthermore, \python{} is an Interpreter language.
@@
 For this reason, all the variables defined before the function are global and thus known to the function.
-The variables within the function (e.g. \texttt{vec\_rM\_old}, ...) are local and do not live beyond the finalization of the function.
+The variables within the function (e.g. \texttt{aMS}, \texttt{vec\_aMS}, $\dots$) are local and do not exist beyond the scope of the function.
 
-Finally, we can express the evolution by a \texttt{while}-loop
+Finally, we can describe the evolution by a \texttt{while}-loop
 \begin{lstlisting}%[float,floatplacement=H]
@@
 \end{lstlisting}
-where for the start we set the parameter $\alpha$ to zero in order first study the properties of the pure 
+where for the start we set the parameter $\alpha=0$ in order first study the properties of the pure 
 $1/r^2$ force.
 Note the required indent of the loop structure similar to the indent of a function.
-In each iteration of the \texttt{while}-loop, the previous distance and velocity are used to compute the new values --- which directly overwrite the previous values and thus enter the next iteration.
+In each iteration of the \texttt{while}-loop, the previous distance and velocity are used to compute the new values --- which directly overwrites the previous values and thus enter the next iteration.
 The total runtime \texttt{2*TM} is the amount of "virtual" days the simulations should run.
@@
 With the previous choice for \texttt{dt}, this corresponds to roughly $N_T \approx 2 \cdot 10^3$ evolution steps.
+Note that the exact time it takes for mercury complete a full revolution depends on the initial coordinates and velocities as well as the accuracy of the computation.
+One might encourage students to analyze this in the beginning.
 
@@
 %------------------------------------------
-To start the visualization one has to include further objects of the \texttt{VPython} module \texttt{visual}.
+To start the visualization one has to \texttt{import} further objects from the \texttt{VPython} module.
 Additional to the vector class, one has to include
@@
 
-If the starting values are chosen as advised in the previous section, then the students should end up with a trajectory as depicted in figure \ref{fig:MercuryOrbit}.
-To get the perihelion motion, the term from equation (\ref{eq:newton_art}) has to be be added by choosing $\alpha=0$.
-This is a good opportunity to let the students play with the size of $\alpha$ and get a feeling for its impact on the trajectories or analyze what happens when changing the time-steps $\Delta t$.
-However, it might be advisable to not start out with the correct mercury values but a more excentric trajectory by choosing e.g. $r_{MS}(0)=6R_0$.
-Then the perihelion motion is better visible. This is illustrated in figure \ref{fig:MercuryOrbit}, where we used a value of $\alpha=10^6$.
-
+If the starting values are chosen as advised in the previous section, then the students should end up with a trajectory as depicted in figure \ref{fig:MercuryOrbit-a0-small}.
+At this point one might ask students to vary $\Delta t$ and observe its effects on the orbit [see also figure \ref{fig:MercuryOrbit-a0-small-dt-large}].
+To get the perihelion motion, the additional force term described in equation (\ref{eq:newton_art}) has to be "turned on" for $\alpha\neq0$ [figure \ref{fig:MercuryOrbit-a6-small}].
+This is a good opportunity to let the students play with the size of $\alpha$ and get a feeling for its impact on the trajectories.
+However, it might be advisable to not start out with the correct mercury values but a more excentric trajectory by choosing e.g. $r_{MS}(0)=6R_0$ [figure \ref{fig:MercuryOrbit-a6-big}].
+Then the perihelion motion is better visible.
 \begin{figure}[htb]
 	\centering
-	\begin{subfigure}[b]{0.3\textwidth}
+	\begin{subfigure}[c]{0.22\textwidth}
 		\includegraphics[width=\textwidth]{figs/a0T5dt20.png}
@@
 	~
-	\begin{subfigure}[b]{0.3\textwidth}
+	\begin{subfigure}[c]{0.22\textwidth}
+		\includegraphics[width=\textwidth]{figs/num-err.png}
+		\caption{\label{fig:MercuryOrbit-a0-small-dt-large}}
+	\end{subfigure}
+	~
+	\begin{subfigure}[c]{0.22\textwidth}
 		\includegraphics[width=\textwidth]{figs/a6T5dt20.png}
@@
 	~
-	\begin{subfigure}[b]{0.3\textwidth}
+	\begin{subfigure}[c]{0.22\textwidth}
 		\includegraphics[width=\textwidth]{figs/a6T10dt20.png}
@@
 	\end{subfigure}
-	\caption{\label{fig:MercuryOrbit}%
-		Different mercury orbits for $\alpha=0$ (a), $\alpha = 10^{6}$ (b) and $\alpha = 10^{6}$ but $r_{MS}(0) = 6 R_0$ (c).
-		All computations where executed with $\Delta t = 2 v_{M}(0) / c_a / 20$.
+	\captionsetup{singlelinecheck=off}
+	\caption[]{\label{fig:MercuryOrbit}
+		Different mercury orbits for:
+		\begin{itemize}
+		\item[(a)] $\alpha=0$ and $\Delta t = \Delta t_0  / 20$;
+		\item[(b)] $\alpha=0$ and $\Delta t = \Delta t_0  \times 2$, where the black line represents the orbit of (a) and the red line is the orbit for the larger time steps;
+		\item[(c)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$;
+		\item[(d)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$ but $r_{MS}(0) = 6 R_0$.
+		\end{itemize}
+		Here, the time steps are defined by $\Delta t_0 \equiv 2 v_{M}(0) / c_a$ and the images are screenshots of the simulation (for modified colors).
 		Note that for different starting values of $r_{MS}(0)$, also the numbers of days for a "mercury year" changes.
@@
 \begin{itemize}
-\item extract multiple positions of the perihelion $\vec r_{MS} ( t_\mathrm{ph}^{(n)} )$ for a fixed value of $\alpha$ and simulations times which covers several orbits,
-\item calculate the angle between the perihelions for two successive turns and compute the average angle $\delta \Theta (\alpha)$ over all individually computed angles
-\item repeat the above points for different values of $\alpha$ to confirm the linear dependence of $\delta \Theta (\alpha)$
-\item extrapolate $\delta \Theta (\alpha)$ to $\alpha=1$
+\item extract multiple positions of the perihelion $\vec r_{MS} ( t_\mathrm{ph}^{(n)} )$ for a fixed value of $\alpha$ and simulations times which covers several orbits;
+\item calculate the angle between the perihelions for two successive turns and compute the average angle $\delta \Theta (\alpha)$ over all individually computed angles;
+\item repeat the above points for different values of $\alpha$ to confirm the linear dependence of $\delta \Theta (\alpha)$;
+\item interpolate $\delta \Theta (\alpha)$ to $\alpha=1$ using that $\delta \Theta (0) = 0$ (modulo numerical errors).
 \end{itemize}
@@
 It is useful to save the position of the perihelion in a list \texttt{list\_perih} for a certain number of turns (here e.g. \texttt{max\_turns}=10).
-The implementation can be realized as follows
+The implementation can be realized as follows\footnote{%
+	Note that one stores \code{vec\_r\_last} by making a copy of \code{M.pos}: \code{vec\_r\_last = vector(M.pos)}.
+	This is essential because otherwise, if one changes \code{M.pos}, one would automatically change \code{vec\_r\_last} as well.
+}
 \begin{lstlisting}
@@
 \end{lstlisting}
-To account for the statistical errors (e.g., numerical rounding errors) one can be average over a few turns.
+To account for the statistical errors (e.g., numerical rounding errors) one can average over a few turns.
 Depending on the programming proficiency of the students and time constraints, this can be either done by hand or, e.g., by implementing the following code
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
+\todo[inline]{@all: where is this periodic oscillation coming from?}
 Physical experiments in general, as do numerical simulations of physics, must always account for possible uncertainties.
@@
 
-Consider figure \ref{fcc3}, where the angle of the perihelion relative to the $y$-axis is shown for the first $10$ turns for different choices of $\Delta t$ and $\alpha$.
-As expected, the correlation between the number of turns and $\delta\Theta$ is linear and the slope depends solely on $\alpha$.
+Consider figure \ref{fcc3}, where the angle of the perihelion growth $\delta \Theta$ is shown for the first $10$ turns for different choices of $\Delta t$ (colors) and $\alpha$ (columns).
+As expected, the correlation between the number of turns and $\delta\Theta$, for sufficiently small time-steps $\Delta t$, is constant on average and the values depend solely on $\alpha$.
 However, there is an offset between the lines for different $\Delta t$.
@@
 	\includegraphics[width=.99\textwidth]{figs/angular-variaton.pdf}
-	\todo[inline]{@all: where is this periodic oscillation coming from?}
 	\caption{\label{fcc3}Precession growth of perihelion $\delta\Theta$ in degrees depending on the number of turns $N_t$ for different time steps $\Delta t$ (color) and different values for $\alpha$ (columns).
@@
 For this study, it is advantageous to set $\alpha =0$, since in this case the trajectories should be closed and therefore deviations from the ideal case can be spotted easily.
-In figure \ref{fig: LargeDT} the mercury trajectory for $\Delta t=2 * v_M(0) / c_a \cdot 5$ is shown.
+In figure \ref{fig:MercuryOrbit-a0-small-dt-large} the mercury trajectory for $\Delta t=2 v_M(0) / c_a \times 2$ is shown.
 One can clearly see the failure of the simulation to reproduce the physical trajectory.
@@
 Only in the limit $\Delta t \rightarrow 0$ would the simulation perfectly reproduce the actual trajectory.
-\begin{figure}[htb]
-	\centering
-	\includegraphics[width=.5\textwidth]{figs/num-err.png}
-	\caption{\label{fig: LargeDT}%
-		Deviation of the trajectory from precise numeric result (black line; $\Delta t=2 v_M(0) / a_M / 20$) for large time steps (red line; $\Delta t=2 v_M(0) / a_M \times 2$).
-		Both orbits are plotted for $T = 10 T_M$.
-	}
-\end{figure}
 
@@
 
-Thanks to Einstein we indeed know that the conclusion formulated above is correct ---- the more general underlying theory is General Relativity and indeed Einstein was able to quantitatively explain the perihelion motion of Mercury from his equations.
+Thanks to Einstein we indeed know that the conclusion formulated above is correct --- the more general underlying theory is General Relativity and indeed Einstein was able to quantitatively explain the perihelion motion of Mercury from his equations.
 On the other hand had we found a dramatic deviation of $\alpha$ from unity one would have concluded that there is probably some other dynamics going on that drives this difference.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\todo{@all: Acknowledgments}
 Here they come
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-In short, to run the above commands and scripts, one needs to install \python{} and \vpython{}.
+In short, to run the above scripts, one needs to install \python{} and \vpython{}.
 We recommend the newest \vpython{} version 7 or higher which best runs with the \python{} versions of 3.5 or higher.
@@
 \begin{enumerate}
-	\item install and update \python{} (at best 3.x, x$\geq5$),
-	\item install and update a \python{} package manager,
+	\item install and update \python{} (at best 3.x, x$\geq5$);
+	\item install and update a \python{} package manager;
 	\item install \vpython{} and dependencies.
@@
 All Linux based system come with an installation of \python{}.
-Thus, if not already at the newest version, one just have to update existing installations.
+Thus, if not already at the newest version, one just has to update existing installations.
 This is best done by using a distribution manager.
@@
 For this we refer to \url{https://www.macports.org/install.php} or \url{https://docs.brew.sh/Installation.html}.
-At first one should open a terminal/console (bash).
+
+As the start one should open a terminal/console (bash).
 Depending on the installation, one usually needs superuser rights in order to update \python{}.
@@
 The installations using a distribution manager are save and tested by the vast community.
-In the following, the successive order of commands to update python is presented for \texttt{apt-get} (Linux) and \texttt{MacPorts} (macOS).
+In the following, the successive order of commands to update \python{} is presented for \texttt{apt-get} (Linux) and \texttt{MacPorts} (macOS), but are most similar for other distribution managers.
 \begin{table}[h!]
@@
 Depending on the previously installed packages, this command might install further dependencies automatically (after confirmation).
-One should test if the installation was successful by typing \code{python3 --version} or \code{python3.6 --version} (depending on the installation name) which prints out the version of the python installation (and should be larger than 3.5).
-Next one must update the package manager to get the newest versions of python packages.
+One should test if the installation was successful by typing \code{python3 --version} or \code{python3.6 --version} (depending on the installation name) which prints out the version of the \python{} installation (and should be larger than 3.5).
+Next one must update the package manager to get the newest versions of \python{} packages.
 This is done by running
@@
 \normalsize
-\end{table}
-
+\end{table}\\
 Note that this just upgrades the packages for the current user.
@@
 This is best figured out by typing \code{pip} and pressing \texttt{tab}.
+
 Finally, one can install \vpython{} by running
@@
 
-At this point, everything needed is installed and one can start the simulation.
-To test the \vpython{} installation and should run \code{python3} or \code{python3.6}, which starts a \python{} console.
+At this point, everything is installed and one can start the simulation.
+To test the \vpython{} installation one should run \code{python3} or \code{python3.6}, which starts a \python{} console.
 Next, one shall run \code{from vpython import *}, which loads the \vpython{} package and opens the default browser.
@@
 
-In general, there are multiple ways to run python scripts.
+In general, there are multiple ways to run \python{} scripts.
 For the sake of this paper, we suggest two scenarios.
-
-The first scenario to run python scripts is that one writes and saves a \python{} file, e.g., \texttt{test.py}, which can be edited using any text editor of choice.
-To run the python script one must navigate the terminal to the containing folder using the \code{cd} (and \code{ls}) command.
+\begin{enumerate}
+\item The first scenario to run \python{} scripts is that one writes and saves a \python{} file, e.g., \texttt{test.py}, which can be edited using any text editor of choice.
+	To run the \python{} script one must navigate the terminal to the containing folder using the \code{cd} (and \code{ls}) command.
 	While \code{ls} prints out the folders and files in the current directory, \code{cd} changes the directory.
 	The command \code{cd} either takes relative paths (directories which are in the current directory), e.g., \code{cd folder}, or absolute paths beginning with a \code{/}, e.g., \code{cd /Users/name/Desktop/}.
-Once one has navigated to folder which contains the python file --- e.g., the file name is listed when typing \code{ls} --- than one can type \code{python3 test.py} or \code{python3.6 test.py} to execute the script.
+	Once one has navigated to folder which contains the \python{} file --- e.g., the file name is listed when typing \code{ls} --- than one can type \code{python3 test.py} or \code{python3.6 test.py} to execute the script.
 	Again, this script can be ended (if it does not end automatically) by pressing \texttt{Ctrl-c}.
-
-The second scenario to run \python{} scripts is given by the interactive \texttt{ipython} or \texttt{jupyter} notebooks.
+\item The second scenario to run \python{} scripts is given by the interactive \texttt{ipython} or \texttt{jupyter} notebooks.
 	This feature is automatically installed when installing the newest version of \vpython{}.
@@
 	Within this browser window, one can freely navigate to the directory of choice and create interactive \python{} and \vpython{} sessions by clicking on existing files or creating new ones (upper right button \texttt{new}$\rightarrow$\vpython{}).
-
+\end{enumerate}
 We provide example files for both scenarios.\todo{@arXiv: online link}
@@
 Following the link \url{https://conda.io/docs/user-guide/install/download.html}, one has two possible (free) options, \texttt{Anaconda} and \texttt{Miniconda}.
-Both are equally sufficient, because one can install further updates after the initial installation.
+Both are equally sufficient.
+One can install further updates after the initial installation.
 In the case of \texttt{Miniconda} (\url{https://conda.io/miniconda.html}) one shall download the newest \python{} 3.x (x $\geq5$) installer for windows and the right system type (32-bit or 64-bit).
@@
 
-Installing one of the two executables creates a new windows application called \texttt{Anaconda Prompt}.
-Starting this application opens a terminal which we will user to install the required \python{} packages.
+Installing one of the two executables creates a new \texttt{Windows} application called \texttt{Anaconda Prompt}.
+Starting this application opens a terminal which will be used to install the required \python{} packages.
 At first, one should test if the \texttt{Anaconda} installation was successful by running the command \code{conda list}.
@@
 The additional flag \code{-c vpython} tells \texttt{Anaconda} the "channel" where to look for the \vpython{} package.
-Running this command prints out a list of dependencies which must be installed in order for \vpython{} to work --- this must be confirmed.
+Running this command prints out a list of dependencies which will be automatically installed in order for \vpython{} to work --- this must be confirmed.
 
@@
 		Programs can be closed by closing the running window.
+	\item One can run \code{python3} in the \texttt{Anaconda Prompt} application to start an interactive session.
+		Programs can be closed by closing the running window.
 	\item Start the \texttt{Anaconda Navigator} application, open an interactive \texttt{qtconsole} and program within the newly opened terminal.
@@
 		This opens an interactive \python{} session in the default web-browser.
-		Here one open existing notebooks or create new files (upper right button \texttt{new}$\rightarrow$\vpython{}).
+		Here, one can open existing notebooks or create new files (upper right button \texttt{new}$\rightarrow$\vpython{}).
 \end{itemize}
 
-To start and test the \vpython{} installation one should pick any out of the three options and execute the code
+To start and test the \vpython{} installation one should pick any out of these options and execute the code
 \begin{lstlisting}

### 2017-12-02 tex/mercury_orbit_base.py

@@ -13,7 +13,7 @@ vec_rM0 = vector(0, rM0, 0)
 vec_vM0 = vector(vM0, 0, 0)
 
 # Definition of the time step
-dt = 2 * vM0 / c_a / 10
+dt = 2 * vM0 / c_a / 20
 
 # Define the coordinate and velocity update
 def evolve_mercury(vec_rM_old, vec_vM_old, alpha):

### 2017-12-02 tex/MercuryNumerical.tex
imporved tex infrastructure --- now able to compile without xelatex

@@
 %
-\documentclass[12pt, UK english]{iopart}
+\documentclass[12pt,ngerman,american]{iopart}
 \newcommand{\gguide}{{\it Preparing graphics for IOP Publishing journals}}
@@
 
-\usepackage{hyperref}
+\usepackage[utf8]{inputenc}         % this is needed for Umlaute
+\usepackage[english,ngerman]{babel} % this is needed for Umlaute
+\usepackage[T1]{fontenc}            % this is needed for correct output of umlauts in pdf
+
+\usepackage[dvipsnames]{xcolor}         % Enabling colors by their 'svgnames'
+\usepackage[%
+  unicode=false,%
+  pdftitle={A primer to numerical simulations: The perihelion motion of Mercury},%
+  pdfauthor={I. Hammer, C. Hanhart, C. Körber and C. Müller},%
+  pdfkeywords={Numerical Simulations, Teaching, Perihelion Motion of Mercury, General Relativity},%
+  pdfsubject={A primer with the goal to teach general concepts of numerical simulations to high school students.
+The specific example is the perihelion motion of mercury due to General Relativity.},   % subject of the document
+  colorlinks=true,        % true: colored links
+  linkcolor=OliveGreen,          % color of internal links (change box color with linkbordercolor)
+  citecolor=magenta,        % color of links to bibliography
+  filecolor=YellowGreen,      % color of file links
+  urlcolor=blue           % color of external links
+]{hyperref}
 \usepackage[space=true]{accsupp}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\usepackage[T1]{fontenc}           % Embed fonts
+
 \usepackage{textcomp,gensymb}       % \degree
 
-\usepackage{polyglossia}                         % Define languages (Umlaute)
-\setdefaultlanguage[variant=american]{english}   % Accent of main language
-\setotherlanguage{german}                        % We also have german words
 \usepackage[autostyle=true]{csquotes}            % Define quotation style
@@
 \usepackage{graphicx}                 % Import Pictures + PDFs
-\usepackage[svgnames]{xcolor}         % Enabling colors by their 'svgnames'
 \usepackage{listings}                 % Code listings  See: https://en.wikibooks.org/wiki/LaTeX/Source_Code_Listings

### 2017-12-04 tex/MercuryNumerical.tex
explained oscilations

@@
 \end{lstlisting}
+Note that the "next" perihelion location is compared to the "previous" perihelion location in order to compute the perihelion motion.
+This is the case since, depending on the initial conditions and numerical uncertainty, one does not necessarily start at the perihelion and would therefore have a systematic error when comparing to initial coordinates.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo[inline]{@all: where is this periodic oscillation coming from?}
 Physical experiments in general, as do numerical simulations of physics, must always account for possible uncertainties.
@@
 
-Consider figure \ref{fcc3}, where the angle of the perihelion growth $\delta \Theta$ is shown for the first $10$ turns for different choices of $\Delta t$ (colors) and $\alpha$ (columns).
+Consider figure \ref{fcc3}, where the angle of the perihelion motion $\delta \Theta$ is shown for the first $10$ turns ($x$-axis) for different choices of $\Delta t$ (colors) and $\alpha$ (columns).
 As expected, the correlation between the number of turns and $\delta\Theta$, for sufficiently small time-steps $\Delta t$, is constant on average and the values depend solely on $\alpha$.
-However, there is an offset between the lines for different $\Delta t$.
-The reason is, due to the uncertainties in the simulation, the starting position does not coincide with the first perihelion as one might expect.
-This does not generate a problem for the procedure described in the previous section since the starting position is not considered. Still, it is helpful to be aware of this fact.
-One can see nicely in figure \ref{fcc3} that the oscillations of the perihelion motion increases for growing time steps.
+However, at $\alpha=0$, there is an offset between the data points for $\delta \Theta$ and zero for different values of $\Delta t$.
+Because one knows that for the regular gravitational potential, the orbit should be bounded, this is a numerical error.
+Indeed this can be viewed as a magnitude estimate for the numerical error for the perihelion motion $\delta \Theta$.
 
@@
 	\includegraphics[width=.99\textwidth]{figs/angular-variaton.pdf}
-	\caption{\label{fcc3}Precession growth of perihelion $\delta\Theta$ in degrees depending on the number of turns $N_t$ for different time steps $\Delta t$ (color) and different values for $\alpha$ (columns).
+	\caption{\label{fcc3}Motion of perihelion $\delta\Theta$ in degrees depending on the number of turns $N_t$ for different time steps $\Delta t$ (color) and different values for $\alpha$ (columns).
 	Here, $\Delta t_0 \equiv 2 v_M(0)/c_a$.
-	One should pay special attention to the oscillations within the growth for relatively large time steps.
+	Offsets of $\delta \Theta(\alpha=0)$ from zero can be viewed as an error magnitude estimated and oscillations in $\delta \Theta$ indicate to coarse time steps.
 }
@@
 
-To point the students to this problem, they can be asked to increase their time steps significantly and observe the effect on the trajectories.
+Furthermore, for large time-steps, one can observe that the values of the perihelion motion oscillates between discrete values at different turns.
+This indicates that the time-step is chosen to large for accurate computations and accordingly the grid for the perihelion vectors to coarse.
+The exact perihelion vector of Mercury and the Sun could potenitally lay between the discrete values computed for finite time-steps.
+If this is the case, then for some turns, the code would recognize the vector before the exact location as the perihelion and sometimes the vector behind the exact location as the perihelion.
+This results in the observed oscillations of $\delta \Theta$.
+One can show that this is the case if one also prints the angles between the previous perihelion and the vector before the new perihelion (\code{vec\_r\_before\_last}) and after the new perihelion (\code{M}).
+
+To point the students to this problem, they can be asked to adjust the size of the time-steps and and observe the effect on the trajectories.
+This should also be done for large values which supposed to be problematic because the effect is more observable.
 For this study, it is advantageous to set $\alpha =0$, since in this case the trajectories should be closed and therefore deviations from the ideal case can be spotted easily.
@@
 From examples like this it should be apparent that the time steps influence the accuracy of the simulation.
-One could quantify the deviation of $\delta\Theta$ from zero and use this to estimate the numerical systematic uncertainty for a given time step $\Delta t$.
 Only in the limit $\Delta t \rightarrow 0$ would the simulation perfectly reproduce the actual trajectory.

### 2017-12-07 analysis/perihelion-full.py
Clean up and spell check python script

@@ -1,53 +1,53 @@
-# Import the class vector from vpythons module visual
+# import everything from vpython (need graphics output, vectors, etc.)
 from vpython import *
 
 # set up display
 scene.width      = 1680
 scene.height     = 1024
 scene.background = color.white
-scene.center     = vector(0,-2,0)
+scene.center     = vector(0, -2, 0)
 
-# Definition of parameters
+# definition of parameters
 rM0 = 4.6    # in units of R0
 vM0 = 0.51   # in units of R0/T0
 aM  = 1.01   # in units of R0/T0**2
 TM  = 88.    # in units of T0
 rS  = 3.e-7  # in units of R0
 
-# Initialize distance and velocity vectors
+# initialize distance and velocity vectors of Mercury
 vec_rM0 = vector(0, rM0, 0)
 vec_vM0 = vector(vM0, 0, 0)
 
-# define the initial coordinates; M = mercury, S = sun
+# define graphical objects; M = mercury, S = sun
 M = sphere(pos=vec_rM0,       radius=0.5,  color=color.red   )
-S = sphere(pos=vector(0,0,0), radius=1.5,  color=color.yellow)
+S = sphere(pos=vector(0, 0, 0), radius=1.5,  color=color.yellow)
 # and the initial velocities
 M.velocity = vec_vM0
-S.velocity = vector(0,0,0)
+S.velocity = vector(0, 0, 0)
 
-# add a visible trajectory to mercury
+# add a visible trajectory to Mercury
 M.trajectory = curve(color=color.black, radius=0.005)
 
-# Definition of the time step
+# definition of the time step
 dt = 2. * vM0 / aM / 10
 
-# Define the function
+# function to advance Mercury in time
 def evolve_mercury(vec_rM_old, vec_vM_old, alpha):
-    # Compute the strength of the acceleration      
+    # compute the absolute value of the acceleration
     aMS = aM * ( 1. + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
-    # Multiply by the direction
+    # multiply by the direction to get the acceleration vector
     vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
-    # Update velocity vector
+    # update velocity vector
     vec_vM_new = vec_vM_old + vec_aMS * dt
-    # Update position vector
+    # update position vector
     vec_rM_new = vec_rM_old + vec_vM_new * dt
     return vec_rM_new, vec_vM_new
 
-# define function for angle extraction
[…]

### 2017-12-07 tex/MercuryNumerical.tex
Inka-> +Citations and removed tasks

@@
 \usepackage[T1]{fontenc}            % this is needed for correct output of umlauts in pdf
+\usepackage{cite}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Tasks}\label{sec:tasks}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo{@IH: this could be a little bit more verbose. Maybe include in section below.}
-Ideas for questions:
-\begin{itemize}
-	\item Mass and motion of sun
-	\item Estimate \texttt{dt} and also \texttt{T} in order to cover roughly two "mercury years".
-	\item What does happen when you change those units?
-	\item Compare values to the values in \cite{}.  Where does the difference come from?
-	\item Start with $\alpha=0$. What do you observe?
-\end{itemize}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Possible extensions}\label{sec:extensions}
@@
 Of course by optimizing the code, an improvement in accuracy can be achieved without changing $\Delta t$.
-The simplest way to demonstrate this might be given by the implementation of the Verlet\cite{}\todo{@IH: cite Verlet} integration instead of using the simple Euler method.
+The simplest way to demonstrate this might be given by the implementation of the Verlet (see e.g.~\cite{Hairer03geometricnumerical} and references therein) integration instead of using the simple Euler method.
 \end{itemize}
@@
 Here they come
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{References}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\bibliographystyle{alphadin}
+\bibliography{References}
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Bib
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\bibliographystyle{alphadin}
-%\bibliography{MercuryNumerical}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
 \end{document}

### 2017-12-07 tex/MercuryNumerical.tex
English ball in the box example

@@
 Instead of providing the template to the students, they could build the program from scratch.
-Of course, this requires some basic knowledge in \texttt{VPython}, as they could acquire for example by working through a ball in the box example\cite{}\todo{@IH: cite ball in the box} or the like.
+Of course, this requires some basic knowledge in \texttt{VPython}, as they could acquire for example by working through a ball in the box example (see~\url{http://www.vpython.org/contents/docs/VPython_Intro.pdf}) or the like.
 Also, they could independently research the problem relevant parameters.

### 2017-12-07 tex/MercuryNumerical.tex
author list: todo source files (license) and citations for urls

@@
 
-\author{I. Hammer, C. Hanhart, C. Körber and C. Müller}
-
-\address{Forschungszentrum Jülich}
-\ead{c.hanhart@fz-juelich.de}
+\author{
+	C.~Körber$^{1}$,
+	I.~Hammer$^{1}$,
+	??$^{3}$,
+	J.-L.~Wynen$^{1}$,
+	C.~Müller$^{2}$ and 
+	C.~Hanhart$^{1}$
+}
+\address{
+	$^1$ \textit{Institute für Kernphysik and Institute for Advanced Simulations, Forschungszentrum Jülich, D-52425 Jülich, Germany}\\
+	$^2$ \textit{??}\\
+	$^2$ \textit{??}
+}
+\ead{c.koerber@fz-juelich.de, c.hanhart@fz-juelich.de}
 \vspace{10pt}

### 2017-12-08 scripts/publish/perihelion.py
Add perihelion.py to publish

@@ -1,0 +1,91 @@
+"""
+
+"""
+
+# import everything from vpython (need graphics output, vectors, etc.)
+from vpython import *
+
+# set up display
+scene.width      = 1680
+scene.height     = 1024
+scene.background = color.white
+scene.center     = vector(0, -2, 0)
+
+# definition of parameters
+# values computed using https://nssdc.gsfc.nasa.gov/planetary/factsheet
+rM0 = 4.6    # initial radius of Mercury orbit, in units of R0
+vM0 = 0.51   # initial orbital speed of Mercury, in units of R0/T0
+aM  = 1.01   # base acceleration of Mercur, in units of R0/T0**2
+rS  = 3.e-7  # Schwarzschild radius of Sun, in units of R0
+
+# initialize distance and velocity vectors of Mercury (at perihelion)
+vec_rM0 = vector(0, rM0, 0)
+vec_vM0 = vector(vM0, 0, 0)
+
+# define graphical objects; M = mercury, S = sun
+M = sphere(pos=vec_rM0,       radius=0.5,  color=color.red   )
+S = sphere(pos=vector(0, 0, 0), radius=1.5,  color=color.yellow)
+# and the initial velocities
+M.velocity = vec_vM0
+S.velocity = vector(0, 0, 0)
+
+# add a visible trajectory to Mercury
+M.trajectory = curve(color=color.black, radius=0.005)
+
+# definition of the time step
+dt = 2. * vM0 / aM / 10
+
+# function to advance Mercury in time
+def evolve_mercury(vec_rM_old, vec_vM_old, alpha):
+    # compute the absolute value of the acceleration
+    aMS = aM * ( 1. + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
+    # multiply by the direction to get the acceleration vector
+    vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
+    # update velocity vector
+    vec_vM_new = vec_vM_old + vec_aMS * dt
+    # update position vector
+    vec_rM_new = vec_rM_old + vec_vM_new * dt
+    return vec_rM_new, vec_vM_new
+
+# define function for angle extraction (result is in degrees)
+def angle_between(v1, v2):
+    return acos( dot(v1, v2) / (v1.mag * v2.mag) ) * 180. / pi
+
+# run parameters
+alpha      = 1.e6
+vec_r_last = vec_rM0
+turns      = 0
+max_turns  = 10
[…]

### 2017-12-08 scripts/publish/base_solution.py
Add base solution

@@ -1,0 +1,75 @@
+"""
+Solution to the base problem.
+
+Simulates the movement of Mercury and displays its position and a trajectory at
+regular intervals.
+The additional term ~ alpha/r**3 is included in the force. In order to simulate
+without it, simply set alpha = 0 before the while loop.
+"""
+
+# import everything from vpython (need graphics output, vectors, etc.)
+from vpython import *
+
+# set up display
+scene.width      = 1680
+scene.height     = 1024
+scene.background = color.white
+scene.center     = vector(0, -2, 0)
+
+# definition of parameters
+# values computed using https://nssdc.gsfc.nasa.gov/planetary/factsheet
+rM0 = 4.6    # initial radius of Mercury orbit, in units of R0
+vM0 = 0.51   # initial orbital speed of Mercury, in units of R0/T0
+aM  = 1.01   # base acceleration of Mercur, in units of R0/T0**2
+rS  = 3.e-7  # Schwarzschild radius of Sun, in units of R0
+
+# initialize distance and velocity vectors of Mercury (at perihelion)
+vec_rM0 = vector(0, rM0, 0)
+vec_vM0 = vector(vM0, 0, 0)
+
+# define graphical objects; M = mercury, S = sun
+M = sphere(pos=vec_rM0,       radius=0.5,  color=color.red   )
+S = sphere(pos=vector(0, 0, 0), radius=1.5,  color=color.yellow)
+# and the initial velocities
+M.velocity = vec_vM0
+S.velocity = vector(0, 0, 0)
+
+# add a visible trajectory to Mercury
+M.trajectory = curve(color=color.black, radius=0.005)
+
+def evolve_mercury(vec_rM_old, vec_vM_old, alpha):
+    """
+    Advance Mercury in time by one step of length dt.
+    Arguments:
+         - vec_rM_old: old position vector of Mercury
+         - vec_vM_old: old velocity vector of Mercury
+         - alpha: strength of 1/r**3 term in force
+    Returns:
+         - vec_rM_new: new position vector of Mercury
+         - vec_vM_new: new velocity vector of Mercury
+    """
+
+    # compute the absolute value of the acceleration
+    aMS = aM * ( 1. + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
+    # multiply by the direction to get the acceleration vector
+    vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
+    # update velocity vector
[…]

### 2017-12-08 scripts/publish/template.py
Add template

@@ -1,0 +1,72 @@
+"""
+Template for the simulation of Mercuries orbit.
+
+Sets up the graphics display and defines basic parameters like the position and
+velocity of Mercury. Also provides a basic structure for the program to get started
+quickly.
+
+All parts that have to be filled in are marked by '### TODO' and brief descriptions
+of the tasks.
+
+Basis run parameters are given just before the while loop at the end. They can be
+changed freely but the values given have proven to yield good results.
+"""
+
+# import everything from vpython (need graphics output, vectors, etc.)
+from vpython import *
+
+# set up display
+scene.width      = 1680
+scene.height     = 1024
+scene.background = color.white
+scene.center     = vector(0, -2, 0)
+
+# definition of parameters
+# values computed using https://nssdc.gsfc.nasa.gov/planetary/factsheet
+rM0 = 4.6    # initial radius of Mercury orbit, in units of R0
+vM0 = 0.51   # initial orbital speed of Mercury, in units of R0/T0
+aM  = 1.01   # base acceleration of Mercur, in units of R0/T0**2
+rS  = 3.e-7  # Schwarzschild radius of Sun, in units of R0
+
+# initialize distance and velocity vectors of Mercury (at perihelion)
+vec_rM0 = vector(0, rM0, 0)
+vec_vM0 = vector(vM0, 0, 0)
+
+### TODO
+#  - Define graphical objects that you want to display.
+#    Give them initial positions and velocities.
+#  - Enable drawing of trajectory of Mercury.
+
+def evolve_mercury(vec_rM_old, vec_vM_old, alpha):
+    """
+    Advance Mercury in time by one step of length dt.
+    Arguments:
+         - vec_rM_old: old position vector of Mercury
+         - vec_vM_old: old velocity vector of Mercury
+         - alpha: strength of 1/r**3 term in force
+    Returns:
+         - vec_rM_new: new position vector of Mercury
+         - vec_vM_new: new velocity vector of Mercury
+    """
+
+    ### TODO
+    #  - Calculate new position and velocity of Mercury here.
+    # vec_rM_new = ??
+    # vec_vM_new = ??
+
+    return vec_rM_new, vec_vM_new
+
+# run parameters
+dt = 2. * vM0 / aM / 10  # time step
+alpha   = 1.e6           # strength of 1/r**3 term
+time = 0                 # current simulation time
+max_time = 1000*dt       # maximum simulation time
+
+# run the simulation for a given time and draw trajectory
[…]

### 2017-12-08 scripts/publish/perihelion.py
Add docstring to perihelion solution

@@ -1,5 +1,12 @@
 """
+Extension of the base solution to measure and display the preihelion.
 
+Extends base_solution.py by keeping track of the location of the perihelion
+of Mercury. Computes and outputs the angle by which it changes over
+the course of the simulation.
+
+The stopping criterion is different than in base_solution.py. Uses a fixed number
+of revolutions around the sun instead of a fixed run time.
 """
 
 # import everything from vpython (need graphics output, vectors, etc.)
@@ -32,11 +39,18 @@ S.velocity = vector(0, 0, 0)
 # add a visible trajectory to Mercury
 M.trajectory = curve(color=color.black, radius=0.005)
 
-# definition of the time step
-dt = 2. * vM0 / aM / 10
-
-# function to advance Mercury in time
 def evolve_mercury(vec_rM_old, vec_vM_old, alpha):
+    """
+    Advance Mercury in time by one step of length dt.
+    Arguments:
+         - vec_rM_old: old position vector of Mercury
+         - vec_vM_old: old velocity vector of Mercury
+         - alpha: strength of 1/r**3 term in force
+    Returns:
+         - vec_rM_new: new position vector of Mercury
+         - vec_vM_new: new velocity vector of Mercury
+    """
+
     # compute the absolute value of the acceleration
     aMS = aM * ( 1. + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
     # multiply by the direction to get the acceleration vector
@@ -47,17 +61,18 @@ def evolve_mercury(vec_rM_old, vec_vM_old, alpha):
     vec_rM_new = vec_rM_old + vec_vM_new * dt
     return vec_rM_new, vec_vM_new
 
-# define function for angle extraction (result is in degrees)
 def angle_between(v1, v2):
+    """Compute angle between two vectors. Result is in degrees."""
     return acos( dot(v1, v2) / (v1.mag * v2.mag) ) * 180. / pi
 
 # run parameters
-alpha      = 1.e6
-vec_r_last = vec_rM0
-turns      = 0
-max_turns  = 10
-list_perih = list()
-sum_angle  = 0.
+dt = 2. * vM0 / aM / 10   # time step
+alpha      = 1.e6         # strength of 1/r**3 term
+vec_r_last = vec_rM0      # previous position of Mercury
+turns      = 0            # number of completed turns
+max_turns  = 10           # maximum number of turns
+list_perih = list()       # list of perihelion locations
[…]

### 2017-12-11 scripts/publish/base_solution.py
Change time step to the one in the text

@@ -20,15 +20,15 @@ scene.center     = vector(0, -2, 0)
 # values computed using https://nssdc.gsfc.nasa.gov/planetary/factsheet
 rM0 = 4.6    # initial radius of Mercury orbit, in units of R0
 vM0 = 0.51   # initial orbital speed of Mercury, in units of R0/T0
-aM  = 1.01   # base acceleration of Mercur, in units of R0/T0**2
+c_a = 1.01   # base acceleration of Mercur, in units of R0/T0**2
 rS  = 3.e-7  # Schwarzschild radius of Sun, in units of R0
 
 # initialize distance and velocity vectors of Mercury (at perihelion)
 vec_rM0 = vector(0, rM0, 0)
 vec_vM0 = vector(vM0, 0, 0)
 
-# define graphical objects; M = mercury, S = sun
-M = sphere(pos=vec_rM0,       radius=0.5,  color=color.red   )
+# define graphical objects; M = Mercury, S = Sun
+M = sphere(pos=vec_rM0,         radius=0.5,  color=color.red   )
 S = sphere(pos=vector(0, 0, 0), radius=1.5,  color=color.yellow)
 # and the initial velocities
 M.velocity = vec_vM0
@@ -50,7 +50,7 @@ def evolve_mercury(vec_rM_old, vec_vM_old, alpha):
     """
 
     # compute the absolute value of the acceleration
-    aMS = aM * ( 1. + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
+    aMS = c_a * ( 1. + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
     # multiply by the direction to get the acceleration vector
     vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
     # update velocity vector
@@ -60,10 +60,10 @@ def evolve_mercury(vec_rM_old, vec_vM_old, alpha):
     return vec_rM_new, vec_vM_new
 
 # run parameters
-dt = 2. * vM0 / aM / 10  # time step
-alpha   = 1.e6           # strength of 1/r**3 term
-time = 0                 # current simulation time
-max_time = 1000*dt       # maximum simulation time
+dt = 2. * vM0 / c_a / 20  # time step
+alpha = 1.e6              # strength of 1/r**3 term
+time = 0                  # current simulation time
+max_time = 1000*dt        # maximum simulation time
 
 # run the simulation for a given time and draw trajectory
 while time < max_time:

### 2017-12-11 scripts/publish/template.py

@@ -25,7 +25,7 @@ scene.center     = vector(0, -2, 0)
 # values computed using https://nssdc.gsfc.nasa.gov/planetary/factsheet
 rM0 = 4.6    # initial radius of Mercury orbit, in units of R0
 vM0 = 0.51   # initial orbital speed of Mercury, in units of R0/T0
-aM  = 1.01   # base acceleration of Mercur, in units of R0/T0**2
+c_a = 1.01   # base acceleration of Mercur, in units of R0/T0**2
 rS  = 3.e-7  # Schwarzschild radius of Sun, in units of R0
 
 # initialize distance and velocity vectors of Mercury (at perihelion)
@@ -57,10 +57,10 @@ def evolve_mercury(vec_rM_old, vec_vM_old, alpha):
     return vec_rM_new, vec_vM_new
 
 # run parameters
-dt = 2. * vM0 / aM / 10  # time step
-alpha   = 1.e6           # strength of 1/r**3 term
-time = 0                 # current simulation time
-max_time = 1000*dt       # maximum simulation time
+dt = 2. * vM0 / c_a / 20  # time step
+alpha   = 1.e6            # strength of 1/r**3 term
+time = 0                  # current simulation time
+max_time = 1000*dt        # maximum simulation time
 
 # run the simulation for a given time and draw trajectory
 while time < max_time:

### 2017-12-11 tex/MercuryNumerical.tex

@@
   pdftitle={A primer to numerical simulations: The perihelion motion of Mercury},%
-  pdfauthor={I. Hammer, C. Hanhart, C. Körber and C. Müller},%
+  pdfauthor={C. Körber, I. Hammer, , J.-L. Wynen, C. Müller and C. Hanhart},%
   pdfkeywords={Numerical Simulations, Teaching, Perihelion Motion of Mercury, General Relativity},%
@@
 particular
-the perihelion motion of Mercury.  We select \python{} (\url{https://www.python.org}) as  programming language, because \python{} is easy to learn, intuitive to understand and an open source language.
+the perihelion motion of Mercury.  We select \python{} \cite{Python} as  programming language, because \python{} is easy to learn, intuitive to understand and an open source language.
 No prior knowledge of \python{} or any other programming language is required, as we explain all the necessary steps to create the simulation.
@@
 
-Because the problem does not depend on the initial position of mercury on its orbit around the sun, we extract the parameters at the perihelion with $ \abs{\vec r_{MS}(0)} = 46 \cdot 10^6$km and $ \abs{\vec v_M(0)} = 59$km/s\footnote{\url{https://nssdc.gsfc.nasa.gov/planetary/factsheet/mercuryfact.html}}.
+Because the problem does not depend on the initial position of mercury on its orbit around the sun, we extract the parameters at the perihelion with $ \abs{\vec r_{MS}(0)} = 46 \cdot 10^6$km and $ \abs{\vec v_M(0)} = 59$km/s\cite{MercuryFactSheet}.
 Since the computer does not understand physical units, one has to express each variable in an appropriate unit.
@@
 The students can convince themselves of this fact by plotting the angles $\delta \Theta$ they have extracted from their simulation over $\alpha$.
-This can be done either by hand, \python{} (e.g., with \texttt{matplotlib}: \url{https://matplotlib.org}), or using another program like, e.g., Excel.
+This can be done either by hand, \python{} (e.g., with \texttt{matplotlib} \cite{Matplotlib}), or using another program like, e.g., Excel.
 For instance figure \ref{fig:AlphaAngle} demonstrates such a plot.
@@
 Instead of providing the template to the students, they could build the program from scratch.
-Of course, this requires some basic knowledge in \texttt{VPython}, as they could acquire for example by working through a ball in the box example (see~\url{http://www.vpython.org/contents/docs/VPython_Intro.pdf}) or the like.
+Of course, this requires some basic knowledge in \texttt{VPython}, as they could acquire for example by working through a ball in the box example (see \cite{VPythonIntro}) or the like.
 Also, they could independently research the problem relevant parameters.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\bibliographystyle{alphadin}
+\bibliographystyle{unsrt}%{alphadin}
 \bibliography{References}
@@
 
-For the advanced users, one simply has to update \python{} to these versions and install \vpython{} via the package manager of choice, e.g, \texttt{pip} [see also \url{http://vpython.org}].
+For the advanced users, one simply has to update \python{} to these versions and install \vpython{} via the package manager of choice, e.g, \texttt{pip} (see also \cite{VPython}).
 
@@
 While \texttt{dpkg} and \texttt{apt-get} usually come with the linux installations, macOS user must install such distribution managers at first.
-For this we refer to \url{https://www.macports.org/install.php} or \url{https://docs.brew.sh/Installation.html}.
+For this we refer to \cite{MacPorts} or \cite{Homebrew}.
 
@@
 As the first step, one must install \python{}, again there are several possibilities but we suggest \texttt{Anaconda} for an intuitive installation.
-Following the link \url{https://conda.io/docs/user-guide/install/download.html}, one has two possible (free) options, \texttt{Anaconda} and \texttt{Miniconda}.
+Following the link \cite{conda}, one has two possible (free) options, \texttt{Anaconda} and \texttt{Miniconda}.
 Both are equally sufficient.
 One can install further updates after the initial installation.
-In the case of \texttt{Miniconda} (\url{https://conda.io/miniconda.html}) one shall download the newest \python{} 3.x (x $\geq5$) installer for windows and the right system type (32-bit or 64-bit).
+In the case of \texttt{Miniconda} \cite{Miniconda} one shall download the newest \python{} 3.x (x $\geq5$) installer for windows and the right system type (32-bit or 64-bit).
 The system type can be found by pressing \texttt{Windows+I}, navigating to \texttt{System}$\rightarrow$\texttt{About} and looking at the \texttt{System type} entry.

### 2017-12-20 tex/MercuryNumerical.tex
added new tex version

@@
 \usepackage[utf8]{inputenc}         % this is needed for Umlaute
-\usepackage[english,ngerman]{babel} % this is needed for Umlaute
+\usepackage[main=american]{babel} % this is needed for Umlaute
 \usepackage[T1]{fontenc}            % this is needed for correct output of umlauts in pdf
@@
   pdftitle={A primer to numerical simulations: The perihelion motion of Mercury},%
-  pdfauthor={C. Körber, I. Hammer, , J.-L. Wynen, C. Müller and C. Hanhart},%
+  pdfauthor={C. K\"orber, I. Hammer, J.-L. Wynen, J. Heuer, C. M\"uller and C. Hanhart},%
   pdfkeywords={Numerical Simulations, Teaching, Perihelion Motion of Mercury, General Relativity},%
@@
 \usepackage{subcaption} % subfigures
-\usepackage{caption}    % itemize in caption
 
@@
 \author{
-	C.~Körber$^{1}$,
+	C.~K\"orber$^{1}$,
 	I.~Hammer$^{1}$,
-	??$^{3}$,
 	J.-L.~Wynen$^{1}$,
-	C.~Müller$^{2}$ and 
+	J.~Heuer$^{2}$,
+	C.~M\"uller$^{3}$ and 
 	C.~Hanhart$^{1}$
@@
 \address{
-	$^1$ \textit{Institute für Kernphysik and Institute for Advanced Simulations, Forschungszentrum Jülich, D-52425 Jülich, Germany}\\
-	$^2$ \textit{??}\\
-	$^2$ \textit{??}
+	$^1$ \textit{Institute f\"ur Kernphysik (IKP-3) and Institute for Advanced Simulations (IAS-4), Forschungszentrum J\"ulich, D-52425 J\"ulich, Germany}\\
+	$^2$ \textit{Institut f\"ur Neurowissenschaften und Medizin (INM-4), Forschungszentrum J\"ulich, D-52425 J\"ulich, Germany}\\
+	$^3$ \textit{JuLab, Forschungszentrum J\"ulich, D-52425 J\"ulich, Germany}
 }
@@
 
-\listoftodos
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 Numerical simulations play a key role in modern physics for they allow one to tackle theoretical problems not accessible otherwise, e.g., since there are too many particle participating in the system (as in simulations for weather predictions) or the interactions are too complicated to allow for a systematic, perturbative approach (as in theoretical descriptions of nuclear particles at the fundamental level).
-This paper introduces a project that can be used to demonstrate the power of numerical simulations.
+This paper introduces a project that allows one to demonstrate the power of numerical simulations.
 On the example of the perihelion motion of the planet Mercury the students are supposed to get in touch with
@@
 \end{itemize}
-The course as well as this paper is structured as follows: after an introduction to Newtonian dynamics and the concepts of differential equations, their discretization is discussed based on Newtons law of gravitation and possible extensions thereof.
-Afterwards the visualization of the resulting trajectories using \texttt{VPython} is introduced and applied to the problem at hand.
-In particular tools are developed to extract the relevant quantity from the result of the simulation.
+The course as well as this paper is structured as follows: after an introduction to Newtonian dynamics and the concept of differential equations, their 
+discretization is discussed based on Newtons law of gravitation and possible extensions thereof.
+Afterwards the visualization of the resulting trajectories using \texttt{VPython} is introduced and applied to the problem at hand. In addition,
+ tools are developed to extract the relevant quantity from the result of the simulation.
 Finally the principle of dimensional analysis is presented as a tool to cross check, if the results of the simulation are sensible.
@@
 While their detailed study needs deeper knowledge in Differential Geometry, for the study outlined here very little math is necessary, such that high school students from about 10$^{\rm th}$ grade up should benefit from it.
-This was already demonstrated in the "Schülderakademie Teilchenphysik", where an early version of this course was tested successfully on a group of 17 German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade.
+This was already demonstrated in the "Schülderakademie Teilchenphysik", where this course was tested successfully on two groups consisting of in total 24 
+German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade.
 
@@
 
-We can say that we understand a physical system, if we can demonstrate that the assumed force leads to the trajectories observed.
+A physical system is understood, if the assumed force leads to the trajectories observed.
 In other words, we need to show that we can calculate the location in space of the object of interest at any point in time, once the initial conditions are fixed properly.
@@
 Sometimes this requires some advanced knowledge in math, sometimes no closed form solution exists.
-However, alternatively one can calculate the whole trajectory of some test body that experiences this force by a successive application of the rules given in eqs.~(\ref{eq:vdef}) and (\ref{eq:adef}):
+However, alternatively one can calculate the whole trajectory of some test body that experiences this force by a successive application of the rules given in Eqs.~(\ref{eq:vdef}) and (\ref{eq:adef}):
 \begin{enumerate}
@@
 Note that $r_S$ is the characteristic length scale of the gravitational field of the sun for --- up to the prefactor --- one can not form another quantity with dimensions of a length from $G_N$, $M_\odot$ and $c$.
-This is what we need $r_S$ for later in the discussion.
 That the Schwarzschild radius is also an important quantity to characterize black holes is not relevant for the discussion at hand.
@@
 \begin{equation}
-	\ddot{r}      = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right) \, , \qquad
-	\ddot{\vec r} =   \ddot{r} \, \frac{\vec{r}}{r} \, .
+	\ddot{\vec r}      = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\frac{\vec r}{r} \, .
 \label{eq:newton}
 \end{equation}
-In general an attractive force that vanishes at large distances leads,  depending on the initial conditions, either to bounded orbits or to open orbits --- in the latter case the planet simply disappears from the sun.
-Depending on time available for the course it might be helpful to explain to the students the origin of this, however, since the most transparent reasoning needs an integration as well as the concept of angular momentum not necessarily familiar to the whole class, it appears in general more appropriate to work out together with the students that a moving body can not be kept by a force if its momentum is too large.
-Then they may try to find closed orbits in the simulation, e.g., by varying the start velocity while keeping the start location fixed.
+In general an attractive force that vanishes at large distances leads,  depending on the initial conditions, either to bounded or to open orbits. 
+In the latter case the planet simply disappears from the sun, since a given force can capture only bodies with not too large momenta.
+The students may study those scenarios within their simulations  by varying the start velocity while keeping the start location fixed.
 
@@
 residual motion of $\delta \Theta_M = (43)''$ in 100 earth years remained unexplained, until Einstein quantified the
-predictions of General Relativity to this particular observable. \todo{@CH:citation here}
+predictions of General Relativity to this particular observable~\cite{Einstein}.
 
-We allow for a movement of the perihelion by modifying the potential of Eq.~(\ref{eq:newton}) with the following ansatz
+To allow for a movement of the perihelion we need to modify the force that appears in Eq.~(\ref{eq:newton}).
+This modification should be such that it depends on $r$ and it vanishes for $r\to \infty$.
+Based on the discussion presented above we may multiply the right hand side with a factor
+$(1+\alpha(r_S/r))$, where $\alpha$ denotes some dimensionless parameter. 
+However, besides $r_S$, which is the characteristic length scale for the gravitational potential
+of the sun, one can define one more parameter, which is characteristic for the dynamics of mercury, namely
 \begin{equation}
-\ddot{\vec r} = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\left(1+\alpha\frac{r_S}{r}\right) \, \frac{\vec{r}}{r} \ ,
+h=\frac{\vec L\,^2}{m^2c^2}= \frac{(\vec r\times \dot{\vec r} \, )^2}{c^2} \ ,
+\end{equation}
+where $\vec L$ denotes the angular momentum, which is a constant of motion for central potentials.
+It is easily verified that $h$ carries dimensions of a length squared.  Therefore we may use  the following ansatz
+for the modified potential
+\begin{equation}
+\ddot{\vec r} = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\left(1+\alpha\frac{r_S}{r}+\beta\frac{h}{r^2}\right) \, \frac{\vec{r}}{r} \ .
 \label{eq:newton_art}
 \end{equation}
-where $\alpha$ is some parameter related to the effects by General Relativity.
-The idea is that the correction must be dimension-less and should scale with some inverse power of $r$, for we still need to demand that the potential vanishes at large distances.
-Since the only parameter of the system with dimension of a length is $r_S$, the correction should scale as $(r_S/r)$.
-Replacing $r$ by a typical distance of Mercury to the sun, $(r_S/r_{MS})\sim 10^{-7}$, suggests that it should be sufficient to include this kind of term to first order only.\footnote{Considering another order would lead qualitatively to the same result, since every change in the potential leads to a perihelion motion.
-Only the value of the pre-factor would change.}
-For comparison: the typical distance of the Venus, the neighbor of Mercury, to the sun is $(r_S/r_{VS}) \sim 5 \cdot 10^{-8}$.
-Therefore the expected ratio of residual motion, favoring in the orbit period $T$, would result in
-\begin{equation}
-	\delta \Theta_V \approx \frac{1}{T_V} \delta \Theta_M T_M \frac{r_{MS}}{r_{VS}} \approx 9'' \, ,
-\end{equation}
-where $T_V$ ($T_M$) denotes the length of a venus (mercury) year. This precisely describes the experimental value of $9''$.
+For the system at hand $h$ may be estimated from the parameters of mercury at its perihelion ---  $\abs{\vec v_M(0)} = 59$km/s and 
+$ r_{MS}=\abs{\vec r_{MS}(0)} = 46 \cdot 10^6$km~\cite{MercuryFactSheet}, where both $\vec r$ and $\dot{\vec r}$ are perpendicular to each other. 
+We thus find that $(h/ r_{MS}^2) = 4\times10^{-6}$. Furthermore $(r_S/r_{MS})=6\times 10^{-6}$. 
+Therefore one may expect from both correction terms an effect similar in size. Furthermore, any term higher order in either $h$ or $r_S$
+should be suppressed by additional six orders of magnitude and thus can be safely neglected.
+The parameters $\alpha$ and $\beta$ could now either be extracted from a fit to data or be calculated from the underlying
+theory --- here General relativity. However, since we have only a single number to study, the angular motion of the perihelion,
+we can only fix either $\alpha$ or $\beta$.
+
+%For comparison: the typical distance of the Venus, the neighbor of Mercury, to the sun is $(r_S/r_{VS}) \sim 5 \cdot 10^{-8}$.
+%Therefore the expected ratio of residual motion, favoring in the orbit period $T$, would result in
+%\begin{equation}
+%	\delta \Theta_V \approx \frac{1}{T_V} \delta \Theta_M T_M \frac{r_{MS}}{r_{VS}} \approx 9'' \, ,
+%\end{equation}
+%where $T_V$ ($T_M$) denotes the length of a venus (mercury) year. This precisely describes the experimental value of $9''$.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{acknowledgments}
+
+This work is supported in part by NSFC and DFG through funds provided to the
+Sino--German CRC110 ``Symmetries and the Emergence of Structure in QCD''.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\section*{References}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \appendix
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Acknowledgments}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo{@all: Acknowledgments}
-Here they come
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section*{References}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \bibliographystyle{unsrt}%{alphadin}

### 2017-12-20 tex/MercuryNumerical.tex
added new figures and beta parameter to scripts

@@
 \begin{equation}
-h=\frac{\vec L\,^2}{m^2c^2}= \frac{(\vec r\times \dot{\vec r} \, )^2}{c^2} \ ,
+a^2=\frac{\vec L\,^2}{m^2c^2}= \frac{(\vec r\times \dot{\vec r} \, )^2}{c^2} \ ,
 \end{equation}
 where $\vec L$ denotes the angular momentum, which is a constant of motion for central potentials.
-It is easily verified that $h$ carries dimensions of a length squared.  Therefore we may use  the following ansatz
+It is easily verified that $a^2$ carries dimensions of a length squared.  Therefore we may use  the following ansatz
 for the modified potential
 \begin{equation}
-\ddot{\vec r} = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\left(1+\alpha\frac{r_S}{r}+\beta\frac{h}{r^2}\right) \, \frac{\vec{r}}{r} \ .
+\ddot{\vec r} = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\left(1+\alpha\frac{r_S}{r}+\beta\frac{a^2}{r^2}\right) \, \frac{\vec{r}}{r} \ .
 \label{eq:newton_art}
 \end{equation}
-For the system at hand $h$ may be estimated from the parameters of mercury at its perihelion ---  $\abs{\vec v_M(0)} = 59$km/s and 
+For the system at hand $a^2$ may be estimated from the parameters of mercury at its perihelion ---  $\abs{\vec v_M(0)} = 59$km/s and 
 $ r_{MS}=\abs{\vec r_{MS}(0)} = 46 \cdot 10^6$km~\cite{MercuryFactSheet}, where both $\vec r$ and $\dot{\vec r}$ are perpendicular to each other. 
-We thus find that $(h/ r_{MS}^2) = 4\times10^{-6}$. Furthermore $(r_S/r_{MS})=6\times 10^{-6}$. 
-Therefore one may expect from both correction terms an effect similar in size. Furthermore, any term higher order in either $h$ or $r_S$
+We thus find that $(a^2/ r_{MS}^2) = 4\cdot10^{-8}$. Furthermore $(r_S/r_{MS})=6\cdot 10^{-8}$. 
+Therefore one may expect from both correction terms an effect similar in size. Furthermore, any term higher order in either $a$ or $r_S$
 should be suppressed by additional six orders of magnitude and thus can be safely neglected.
 The parameters $\alpha$ and $\beta$ could now either be extracted from a fit to data or be calculated from the underlying
-theory --- here General relativity. However, since we have only a single number to study, the angular motion of the perihelion,
+theory --- here General relativity with the result
+\begin{equation}\label{eq:art-prediction}
+	\alpha = 0 \, , \quad \beta = 3 \, .
+\end{equation}
+
+However, since we have only a single number to study, the angular motion of the perihelion,
 we can only fix either $\alpha$ or $\beta$.
@@
 TM  = 88.    # in units of T0
-rS  = 3.e-7  # in units of R0
+rS  = 3.0e-7 # in units of R0
+a2  = 8.2e-7 # in units of R0**2
 \end{lstlisting}
@@
 # Compute the strength of the acceleration
-aMS = c_a * ( 1 + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
+temp = 1 + alpha * rS / vec_rM_old.mag + beta * a2 / vec_rM_old.mag**2
+aMS  = c_a * temp / vec_rM_old.mag**2
 # Multiply by the direction
@@
 # Define the coordinate and velocity update function
-def evolve_mercury(vec_rM_old, vec_vM_old, alpha):
+def evolve_mercury(vec_rM_old, vec_vM_old, alpha, beta):
 	# Compute the strength of the acceleration
-	aMS = c_a * ( 1 + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
+	temp = 1 + alpha * rS / vec_rM_old.mag + beta * a2 / vec_rM_old.mag**2
+	aMS  = c_a * temp / vec_rM_old.mag**2
 	# Multiply by the direction
@@
 # Call the function
-vec_rM_new, vec_vM_new = evolve_mercury(vec_rM_old, vec_vM_old)
+vec_rM_new, vec_vM_new = evolve_mercury(vec_rM_old, vec_vM_old, 0.0, 0.0)
 \end{lstlisting}
@@
 \begin{lstlisting}%[float,floatplacement=H]
-t     = 0
-alpha = 0
+t     = 0.0
+alpha = 0.0
+beta  = 0.0
 # Execute the loop as long as t < 2*TM
 while t < 2*TM:
-	vec_rM, vec_vM = evolve_mercury(vec_rM, vec_vM, alpha)
+	vec_rM, vec_vM = evolve_mercury(vec_rM, vec_vM, alpha, beta)
 	t = t + dt
 \end{lstlisting}
-where for the start we set the parameter $\alpha=0$ in order first study the properties of the pure 
+where for the start we set the parameter $\alpha=0 = \beta$ in order first study the properties of the pure 
 $1/r^2$ force.
@@
 \begin{lstlisting}
-t     = 0
-alpha = 0
+t     = 0.0
+alpha = 0.0
+beta  = 0.0
 # update the objects
@@
 	# update the velocity and position
-	M.pos , M.velocity = evolve_mercury(M.pos , M.velocity , alpha)
+	M.pos , M.velocity = evolve_mercury(M.pos , M.velocity , alpha, beta)
 	# update time
@@
 	\caption[]{\label{fig:MercuryOrbit}
-		Different mercury orbits for:
+		Different mercury orbits for $\beta=0$ and
 		\begin{itemize}
@@
 What do you think?}\\
-
+To extract the linear dependence of the perihelion motion depending on $\alpha$ and $\beta$, we recommend to set one parameter to zero and vary the other.
+Both parameter have similar effects on the perihelion motion.
+As pointed out in Eq. \ref{eq:art-prediction}, general relativity predicts $(\alpha=0, \beta=3)$.
 To calculate the expected size of the perihelion motion for a modified gravitational force, the students need to
 \begin{itemize}
-\item extract multiple positions of the perihelion $\vec r_{MS} ( t_\mathrm{ph}^{(n)} )$ for a fixed value of $\alpha$ and simulations times which covers several orbits;
+\item extract multiple positions of the perihelion $\vec r_{MS} ( t_\mathrm{ph}^{(n)} )$ for a fixed value of $\alpha$ and $\beta$ and simulations times which covers several orbits;
 \item calculate the angle between the perihelions for two successive turns and compute the average angle $\delta \Theta (\alpha)$ over all individually computed angles;
-\item repeat the above points for different values of $\alpha$ to confirm the linear dependence of $\delta \Theta (\alpha)$;
-\item interpolate $\delta \Theta (\alpha)$ to $\alpha=1$ using that $\delta \Theta (0) = 0$ (modulo numerical errors).
+\item repeat the above points for different values of $\alpha$ and $\beta$ to confirm the linear dependence of $\delta \Theta (\alpha, \beta)$;
+\item interpolate $\delta \Theta (\alpha, \beta)$ to $(\alpha=1,\beta=0)$ or $(\alpha=0,\beta=3)$ using that $\delta \Theta (0,0) = 0$ (modulo numerical errors).
 \end{itemize}
@@
 	\centering
-	\includegraphics{figs/alpha-angle.pdf}
-	\caption{\label{fig:AlphaAngle} Linear relation between $\alpha$ and the perihelion motion $\delta \Theta$ for $\Delta t=2v_M(0)/c_a/100$.}
+	\begin{subfigure}[c]{0.49\textwidth}
+		\includegraphics[width=\textwidth]{figs/alpha-angle.pdf}
+		\caption{\label{fig:angle-alpha}}
+	\end{subfigure}
+	\begin{subfigure}[c]{0.49\textwidth}
+		\includegraphics[width=\textwidth]{figs/beta-angle.pdf}
+		\caption{\label{fig:beta-alpha}}
+	\end{subfigure}
+	\caption{\label{fig:AlphaAngle} Linear relation between $\alpha$ [$\beta$] and and the perihelion motion $\delta \Theta$ for $\Delta t=2v_M(0)/c_a/100$ and $\beta=0$ [$\alpha=0$].}
 \end{figure}
 
-To extract the perihelion motion at $\alpha = 1$, we use the fact that $\delta \Theta$ is zero for $\alpha =0$ --- if one could completely eliminate numerical errors.
-Considering one result from the numeric simulation, e.g. $\delta\Theta(\alpha = 10^6) \approx 10.0^{\degree}$, the sought-after angle can be calculated by a linear interpolation as
+To extract the perihelion motion at $\alpha = 0$ and $\beta = 3$, we use the fact that $\delta \Theta$ is zero for $\alpha =0 = \beta$ --- if one could completely eliminate numerical errors.
+Considering one result from the numeric simulation, e.g., $\delta\Theta(\alpha = 0, \beta=10^6) \approx 10.7^{\degree}$, the sought-after angle can be calculated by a linear interpolation as
 \begin{equation}
-	\delta\Theta (\alpha=1) = \frac{10.0^{\degree}}{10^6} = 0.036''
+	\delta\Theta (\alpha=0, \beta=3) = \frac{10.6^{\degree}}{10^6} \times 3 = 0.116''
 	\, .
@@
 Note that, when comparing the results of the simulation with the experimental result, one Earth year corresponds to $T_E/T_M \approx 365d/88d\approx 4.15$ mercury years.
-Therefore, the perihelion motion per 100 earth years is $0.036''\cdot 415\approx 15''$.
-To test whether this is a reasonable result, the students could first follow the dimensional analysis and then compare to the actual value for mercury as outlined above.
+Therefore, the perihelion motion per 100 earth years is $0.116''\cdot 415\approx 48.2''$.
+This is close to experimental value of $\delta\Theta_{\mathrm{exp}} = 43.0''$.
 
@@
 	Here, $\Delta t_0 \equiv 2 v_M(0)/c_a$.
-	Offsets of $\delta \Theta(\alpha=0)$ from zero can be viewed as an error magnitude estimated and oscillations in $\delta \Theta$ indicate to coarse time steps.
+	Offsets of $\delta \Theta(\alpha=0, \beta=0)$ from zero can be viewed as an error magnitude estimated and oscillations in $\delta \Theta$ indicate to coarse time steps.
 }
@@
 \begin{equation}
-\delta \Theta \simeq 2\pi\left(\frac{r_S}{\bar r}\right) = (\pi \times 10^{-7}) \ \mbox{rad} = (2\times 10^{-5})^o = (7\times 10^{-2}) ^{''} \ ,
+\delta \Theta \simeq 2\pi\left(\frac{r_S}{\bar r}\right) = (\pi \times 10^{-7}) \ \mbox{rad} = (2\cdot 10^{-5})^o = (7\cdot 10^{-2}) ^{''} \ ,
 \end{equation}
-which leads to a shift of about $30^{''}$ in 100 earth years to be compared to the empirical value of $43^{''}$.
+or
+\begin{equation}
+\delta \Theta \simeq 2\pi\left(\frac{a^2}{\bar r^2}\right) = (\pi \times 5 \cdot 10^{-8}) \ \mbox{rad} = (\cdot 10^{-5})^o = (4\cdot 10^{-2}) ^{''} \ ,
+\end{equation}
+which leads to a shift of about $30^{''}$ [$15^{''}$] in 100 earth years to be compared to the empirical value of $43^{''}$.
 Thus indeed the amount of perihelion motion of Mercury is in line with expectations, $if$ --- and this is an important "if" --- the Newtonian dynamics is simply the leading term of some more general underlying theory.

### 2017-12-20 tex/mercury_orbit_base.py

@@ -6,7 +6,8 @@ rM0 = 4.6    # in units of R0
 vM0 = 0.51   # in units of R0/T0
 c_a = 1.01   # in units of R0/T0**2
 TM  = 88.    # in units of T0 
-rS  = 3.e-7  # in units of R0
+rS  = 3.0e-7 # in units of R0
+a2  = 8.2e-7 # in units of R0**2
 
 # Initialize distance and velocity vectors
 vec_rM0 = vector(0, rM0, 0)
@@ -16,8 +17,9 @@ vec_vM0 = vector(vM0, 0, 0)
 dt = 2 * vM0 / c_a / 20
 
 # Define the coordinate and velocity update
-def evolve_mercury(vec_rM_old, vec_vM_old, alpha):
-    aMS = c_a * ( 1 + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
+def evolve_mercury(vec_rM_old, vec_vM_old, alpha, beta):
+    temp = 1 + alpha * rS / vec_rM_old.mag + beta * a2 / vec_rM_old.mag**2
+    aMS  = c_a * temp / vec_rM_old.mag**2
     vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
     vec_vM_new = vec_vM_old + vec_aMS * dt
     vec_rM_new = vec_rM_old + vec_vM_new * dt
@@ -32,10 +34,11 @@ S.velocity = vector(0,0,0)
 M.trajectory = curve(color=color.white)
 
 # Start the simulation
-t=0
-alpha = 0
+t     = 0.0
+alpha = 0.0
+beta  = 0.0
 while t < 2*TM:
     rate(100)
     M.trajectory.append(pos=M.pos)
-    M.pos , M.velocity = evolve_mercury(M.pos , M.velocity , alpha)
+    M.pos , M.velocity = evolve_mercury(M.pos , M.velocity , alpha, beta)
     t = t + dt

### 2018-01-09 tex/MercuryNumerical.tex
C.H. modifications. Smoothening out beta part.

@@
   pdfsubject={A primer with the goal to teach general concepts of numerical simulations to high school students.
-The specific example is the perihelion motion of mercury due to General Relativity.},   % subject of the document
+The specific example is the perihelion motion of Mercury due to General Relativity.},   % subject of the document
   colorlinks=true,        % true: colored links
@@
 The paper includes details about the development of the code as well as a discussion of the visualization of the results.
-In addition a method is discussed how to estimate the size of the effect a priori. 
+In addition a method is discussed how to estimate the size of the effect a priori that at the same time also can be employed to
+double check the results found numerically. 
 \end{abstract}
@@
  tools are developed to extract the relevant quantity from the result of the simulation.
-Finally the principle of dimensional analysis is presented as a tool to cross check, if the results of the simulation are sensible.
+Finally the principle of dimensional analysis is presented as a tool not only to estimate some effect but also to cross check, if the results of the simulation are sensible.
 
-We are convinced that in order to excite the students for numerical simulations it is compulsory to demonstrate their power on an example that catches their interest.
+We are convinced that in order to excite students for numerical simulations it is compulsory to demonstrate their power on an example that catches their interest.
 This purpose is served perfectly by the case chosen here, since Einsteins equations of General Relativity are fascinating to a very broad public.
 While their detailed study needs deeper knowledge in Differential Geometry, for the study outlined here very little math is necessary, such that high school students from about 10$^{\rm th}$ grade up should benefit from it.
-This was already demonstrated in the "Schülderakademie Teilchenphysik", where this course was tested successfully on two groups consisting of in total 24 
-German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade.
+This was already demonstrated in the "Schülerakademie Teilchenphysik", where this course was tested successfully on two groups consisting of in total 24 
+German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade in 2015 and 2017.
 
@@
 Eq.~({\ref{eq:check}}) also shows that small (large) time steps are necessary (sufficient), if the force is strong (weak), since the time steps need to be small enough that all changes induced by the force get resolved.
-Clearly, a relation as Eq.~({\ref{eq:check}}) can only provide guidance and can not replace a careful numerical check of the solutions: A valid result has to be insensitive the concrete value $\Delta t$ chosen --- in particular replacing $\Delta t$ by $\Delta t/2$ should not change the result significantly.
+Clearly, a relation as Eq.~({\ref{eq:check}}) can only provide guidance and can not replace a careful numerical check of the solutions: A valid result has to be insensitive the concrete value chosen for $\Delta t$ --- in particular replacing $\Delta t$ by $\Delta t/2$ should not change the result significantly.
 
@@
 r_S=\frac{2G_N  M_\odot}{c^2} = 3 \ \mbox{km} \ ,
+\label{rsdef}
 \end{equation}
@@
 In particular the point of closest approach of the planet to the sun, the perihelion, does not move.
-However, as soon as the potential deviates from $1/r$, the perihelion moves.
+However, when a potential that vanishes for $r\to \infty$ deviates from $1/r$, the perihelion moves.
 Because of this, the behavior of the perihelion is a very sensitive probe of the gravitational potential.
 
-The observed perihelion motion of Mercury is nowadays determined as $(574.10\pm 0.65)''$ per
-100 earth years~\footnote{The symbol $''$ denotes "arc seconds": 1$''=(1/3600)^o$. }, but it should be noted that the bulk of this number
+The observed perihelion motion of Mercury is nowadays determined as 
+$$(574.10\pm 0.65)'' \ \mbox{per
+100 earth years}$$
+where the symbol $''$ denotes "arc seconds": 1$''=(1/3600)^o$. The bulk of this number
 can be understood by the presence of other planets within the Newtonian theory, since
 their gravitational force also acts on Mercury. However, a
-residual motion of $\delta \Theta_M = (43)''$ in 100 earth years remained unexplained, until Einstein quantified the
+residual motion of 
+\begin{equation}\delta \Theta_M = (43)'' \ \mbox{in 100 earth years}\label{delT}\end{equation}
+ remained unexplained, until Einstein quantified the
 predictions of General Relativity to this particular observable~\cite{Einstein}.
@@
 However, besides $r_S$, which is the characteristic length scale for the gravitational potential
-of the sun, one can define one more parameter, which is characteristic for the dynamics of mercury, namely
+of the sun, one can define one more parameter, which is characteristic for the dynamics of Mercury, namely
 \begin{equation}
 a^2=\frac{\vec L\,^2}{m^2c^2}= \frac{(\vec r\times \dot{\vec r} \, )^2}{c^2} \ ,
+\label{a2def}
 \end{equation}
@@
 \end{equation}
-For the system at hand $a^2$ may be estimated from the parameters of mercury at its perihelion ---  $\abs{\vec v_M(0)} = 59$km/s and 
-$ r_{MS}=\abs{\vec r_{MS}(0)} = 46 \cdot 10^6$km~\cite{MercuryFactSheet}, where both $\vec r$ and $\dot{\vec r}$ are perpendicular to each other. 
+For the system at hand $a^2$ may be estimated from the parameters of Mercury at its perihelion: The corresponding velocity is 
+ $\dot{\vec r}(t=0)=\abs{\vec v_M(0)} = 59$~km/s (here we
+already indicate that below we start the simulation at the perihelion) and the closest distance between sun and Mercury is
+$ r_{MS}=\abs{\vec r_{MS}(0)} = 46 \cdot 10^6$km~\cite{MercuryFactSheet}. Note that at the perihelion $\vec r$ and $\dot{\vec r}$ are perpendicular to each other. 
 We thus find that $(a^2/ r_{MS}^2) = 4\cdot10^{-8}$. Furthermore $(r_S/r_{MS})=6\cdot 10^{-8}$. 
-Therefore one may expect from both correction terms an effect similar in size. Furthermore, any term higher order in either $a$ or $r_S$
-should be suppressed by additional six orders of magnitude and thus can be safely neglected.
-The parameters $\alpha$ and $\beta$ could now either be extracted from a fit to data or be calculated from the underlying
-theory --- here General relativity with the result
+Therefore (for a more detailed discussion about the underlying logic we refer to Sec.~\ref{sec:analysis}) one may expect from both correction terms an effect similar in size.
+Furthermore, any term higher order in either $a$ or $r_S$
+should be suppressed by additional seven orders of magnitude and thus can be safely neglected.
+The parameters $\alpha$ and $\beta$ can be extracted from a fit to data --- this will be done below. 
+However, since we have with the perihelion shift (see Eq.~(\ref{delT})) only a single number to study,
+we can only fix either $\alpha$ or $\beta$ (or a linear combination thereof).
+At the same time the parameters can also be calculated from the assumed underlying theory and the actual values
+are in general specific for this theory. It turns out that
+General relativity gives~\cite{Einstein}
 \begin{equation}\label{eq:art-prediction}
@@
 \end{equation}
-
-However, since we have only a single number to study, the angular motion of the perihelion,
-we can only fix either $\alpha$ or $\beta$.
+Thus we may also use the simulation explained below to calculate the perihelion motion of Mercury from the input
+values given in Eq.~(\ref{eq:art-prediction}).
 
@@
 %\end{equation}
-%where $T_V$ ($T_M$) denotes the length of a venus (mercury) year. This precisely describes the experimental value of $9''$.
+%where $T_V$ ($T_M$) denotes the length of a venus (Mercury) year. This precisely describes the experimental value of $9''$.
 
@@
 No prior knowledge of \python{} or any other programming language is required, as we explain all the necessary steps to create the simulation.
-In appendix \ref{appendix:python} we give instructions for setting up \python{} on different operating systems.
+In  \ref{appendix:python} we give instructions for setting up \python{} on different operating systems.
 We also provide a working example program that can be used as a template \url{this-upload-url}\todo{@arXiv: online link}.
 
-To start the simulation one needs the "initial" distances and velocities, which are needed as input to the equation of motion [Eq.~(\ref{eq:x_update}) and Eq.~(\ref{eq:v_update})].
-As described above, here  one can simplify the problem by placing one planet (the "infinitely" heavy sun) in the center of the coordinate system and describe the motion of the other planet (mercury) in a two-dimensional plane.
+To start the simulation one needs the "initial" distance and velocity [Eq.~(\ref{eq:x_update}) and Eq.~(\ref{eq:v_update})].
+As described above, here  one can simplify the problem by placing one planet (the "infinitely" heavy sun) in the center of the coordinate system and describe the motion of the other planet (Mercury) in a two-dimensional plane.
 
-Because the problem does not depend on the initial position of mercury on its orbit around the sun, we extract the parameters at the perihelion with $ \abs{\vec r_{MS}(0)} = 46 \cdot 10^6$km and $ \abs{\vec v_M(0)} = 59$km/s\cite{MercuryFactSheet}.
+Since the trajectories do not depend on where on the orbit of Mercury the simulation is started, we 
+use as initial (for $t=0$) parameters the values at the perihelion with $ \abs{\vec r_{MS}(0)} = 46 \cdot 10^6$~km and $ \abs{\vec v_M(0)} = 59$~km/s\cite{MercuryFactSheet}.
 Since the computer does not understand physical units, one has to express each variable in an appropriate unit.
-Both, for the numerical treatment and the intuitive understanding, it is useful to select parameters in a "natural range".
-A useful choice for this problem is given e.g.
-by expressing distances in $R_0 = 10^{10}$ m and time intervals in days, $T_0 = 1$d, where $d$ refers to earth days, such that one mercury year is given by $T_M=88\,T_0$.
-With this choice, the initial distance of mercury to the sun, the size of the initial velocity of mercury and the acceleration pre-factors become
+Both, for the numerical treatment and the intuitive understanding, it is useful to select parameters in a "natural range", e.g., 
+by expressing distances in $R_0 = 10^{10}$ m and time intervals in days, $T_0 = 1$d, where $d$ refers to earth days, such that one Mercury year is given by $T_M=88\,T_0$.
+With this choice, the initial distance of Mercury to the sun, the size of the initial velocity of Mercury and the acceleration pre-factors become
 \begin{eqnarray}
@@
 	\caption{\label{fig:sun_merc}%
-		Sun mercury system with relevant vectors.
-		Because mercury is at its perihelion, its velocity is perpendicular to its direct connection vector with the sun.%
+		Sun Mercury system with relevant vectors.
+		Because Mercury is at its perihelion, its velocity is perpendicular to its direct connection vector with the sun.%
 	}
@@
 rS  = 3.0e-7 # in units of R0
-a2  = 8.2e-7 # in units of R0**2
+a2  = 8.2e-7 # in units of R0**2 \ ,
 \end{lstlisting}
+where the last two quantities refer to the Schwarzschildradius of the sun and the parameter $a^2$ defined in Eqs. (\ref{rsdef}) and (\ref{a2def}), respectively.
 So far we only fixed the length of the vectors.
@@
 The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.
-Because of our choice of initial conditions (we picked the initial vectors in the perihelion), the velocity of mercury is perpendicular to the vector which connects mercury and sun (see figure \ref{fig:sun_merc}):
+Because of our choice of initial conditions (we picked the initial vectors in the perihelion), the velocity of Mercury is perpendicular to the vector which connects Mercury and sun (see figure \ref{fig:sun_merc}):
 \begin{lstlisting}
@@
 vec_rM0 = vector(0, rM0, 0)
-vec_vM0 = vector(vM0, 0, 0)
+vec_vM0 = vector(vM0, 0, 0) \ .
 \end{lstlisting}
-According to Eq.~(\ref{eq:newton}), the force which acts on mercury changes the velocity of mercury which eventually changes the position.
-However, we first need to fix the time step $\Delta t$ and use the estimate of Eq.~(\ref{eq:check}) as a guidance
+To fix the time step $\Delta t$ we use the estimate of Eq.~(\ref{eq:check}) as a guidance to write
 \begin{lstlisting}
 # Definition of the time step
-dt = 2 * vM0 / c_a / 20
+dt = 2 * vM0 / c_a / 20 \ .
 \end{lstlisting}
@@
 Now we are in the position to calculate location and velocity of the planet at $t_0+\Delta t$ using the following equations\footnote{%
-	Note, that there are different notions for numerically integrating differential equations with different accuracys.
-	Thus, the ordering and exact equations for updating $a_{MS}$, $v_M$ and $r_M$ are not unique, but all correct prescriptions lead to the same result for sufficiently small $\Delta t$.
+	Note, that there are different notions for numerically integrating differential equations with different accuracies.
+	Thus, the ordering and exact equations for updating $a_{MS}$, $v_M$ and $r_{MS}$ are not unique, but all correct prescriptions lead to the same result for sufficiently small $\Delta t$.
 	For this particular problem, the following definition is in good balance between numerically stable and inexpensive to compute.
@@
 # Define the coordinate and velocity update function
-def evolve_mercury(vec_rM_old, vec_vM_old, alpha, beta):
+def evolve_Mercury(vec_rM_old, vec_vM_old, alpha, beta):
 	# Compute the strength of the acceleration
@@
 # Call the function
-vec_rM_new, vec_vM_new = evolve_mercury(vec_rM_old, vec_vM_old, 0.0, 0.0)
+vec_rM_new, vec_vM_new = evolve_Mercury(vec_rM_old, vec_vM_old, 0.0, 0.0)
 \end{lstlisting}
@@
 while t < 2*TM:
-	vec_rM, vec_vM = evolve_mercury(vec_rM, vec_vM, alpha, beta)
+	vec_rM, vec_vM = evolve_Mercury(vec_rM, vec_vM, alpha, beta)
 	t = t + dt
@@
 Note the required indent of the loop structure similar to the indent of a function.
-In each iteration of the \texttt{while}-loop, the previous distance and velocity are used to compute the new values --- which directly overwrites the previous values and thus enter the next iteration.
+In each iteration of the \texttt{while}-loop, the previous distance and velocity are used to compute the new values, 
+which directly overwrites the previous values to enter the next iteration.
 The total runtime \texttt{2*TM} is the amount of "virtual" days the simulations should run.
@@
 With the previous choice for \texttt{dt}, this corresponds to roughly $N_T \approx 2 \cdot 10^3$ evolution steps.
-Note that the exact time it takes for mercury complete a full revolution depends on the initial coordinates and velocities as well as the accuracy of the computation.
-One might encourage students to analyze this in the beginning.
+Note that the exact time it takes for Mercury to complete a full revolution depends on the initial coordinates and velocities, the accuracy of the computation
+(controlled by the value of $\Delta t$) as well as the computing power employed.
+One should encourage the students to analyze this in the beginning.
 
@@
 \end{lstlisting}
-The class sphere will represent mercury and the sun in the simulation
+The class sphere will represent Mercury and the sun in the simulation
 \begin{lstlisting}
-# define the initial coordinates; M = mercury, S = sun
+# define the initial coordinates; M = Mercury, S = sun
 M = sphere(pos=vec_rM0,       radius=0.5,  color=color.red   )
@@
 S.velocity = vector(0,0,0)
-# add a visible trajectory to mercury
+# add a visible trajectory to Mercury
 M.trajectory = curve(color=color.white)
@@
 We have placed the sun in the origin of our coordinate system and choose non-realistic radii sizes for visualization purposes.
-
-Last but not least, instead of updating the object-unrelated vectors in the while loop, we update the object-related vectors which will eventually be drawn
+Last but not least, one should use in the code presented above directly the vectors related to the visualization. Thus
+this piece of code now reads
 \begin{lstlisting}
@@
 	# update the velocity and position
-	M.pos , M.velocity = evolve_mercury(M.pos , M.velocity , alpha, beta)
+	M.pos , M.velocity = evolve_Mercury(M.pos , M.velocity , alpha, beta)
 	# update time
@@
 
-If the starting values are chosen as advised in the previous section, then the students should end up with a trajectory as depicted in figure \ref{fig:MercuryOrbit-a0-small}.
-At this point one might ask students to vary $\Delta t$ and observe its effects on the orbit [see also figure \ref{fig:MercuryOrbit-a0-small-dt-large}].
-To get the perihelion motion, the additional force term described in equation (\ref{eq:newton_art}) has to be "turned on" for $\alpha\neq0$ [figure \ref{fig:MercuryOrbit-a6-small}].
-This is a good opportunity to let the students play with the size of $\alpha$ and get a feeling for its impact on the trajectories.
-However, it might be advisable to not start out with the correct mercury values but a more excentric trajectory by choosing e.g. $r_{MS}(0)=6R_0$ [figure \ref{fig:MercuryOrbit-a6-big}].
-Then the perihelion motion is better visible.
+If the starting values are chosen as advised in the previous section, the students should end up with a trajectory as depicted in figure \ref{fig:MercuryOrbit-a0-small}.
+At this point one might ask students to vary $\Delta t$ and observe the effect of this on the orbit [see also figure \ref{fig:MercuryOrbit-a0-small-dt-large}].
+To get the perihelion motion, the additional force term described in Eq.~(\ref{eq:newton_art}) has to be "turned on", e.g., by choosing  $\alpha\neq0$ 
+[figure \ref{fig:MercuryOrbit-a6-small}].
+This is a good opportunity to let the students play with the size of $\alpha$ and get a feeling for its impact on the trajectories. In particular
+they should discover that values of $\alpha$ of the order of $10^6$ are necessary to get a visible effect.
+For a nicely visible  perihelion motion it is advisable to not  use the correct Mercury values but a more excentric trajectory by choosing, e.g., $r_{MS}(0)=6R_0$ [figure \ref{fig:MercuryOrbit-a6-big}].
 \begin{figure}[htb]
@@
 	\caption[]{\label{fig:MercuryOrbit}
-		Different mercury orbits for $\beta=0$ and
+		Different Mercury orbits for $\beta=0$ and
 		\begin{itemize}
@@
 		Here, the time steps are defined by $\Delta t_0 \equiv 2 v_{M}(0) / c_a$ and the images are screenshots of the simulation (for modified colors).
-		Note that for different starting values of $r_{MS}(0)$, also the numbers of days for a "mercury year" changes.
+		Note that for different starting values of $r_{MS}(0)$, also the numbers of days for a "Mercury year" changes.
 	}
@@
 What do you think?}\\
-To extract the linear dependence of the perihelion motion depending on $\alpha$ and $\beta$, we recommend to set one parameter to zero and vary the other.
+To extract the dependence of the amount of perihelion motion 
+ on the parameters $\alpha$ and $\beta$, we recommend to set one parameter to zero and to vary the other.
 Both parameter have similar effects on the perihelion motion.
-As pointed out in Eq. \ref{eq:art-prediction}, general relativity predicts $(\alpha=0, \beta=3)$.
-To calculate the expected size of the perihelion motion for a modified gravitational force, the students need to
+To calculate the size of the perihelion motion for a modified gravitational force, the students need to
 \begin{itemize}
-\item extract multiple positions of the perihelion $\vec r_{MS} ( t_\mathrm{ph}^{(n)} )$ for a fixed value of $\alpha$ and $\beta$ and simulations times which covers several orbits;
-\item calculate the angle between the perihelions for two successive turns and compute the average angle $\delta \Theta (\alpha)$ over all individually computed angles;
-\item repeat the above points for different values of $\alpha$ and $\beta$ to confirm the linear dependence of $\delta \Theta (\alpha, \beta)$;
-\item interpolate $\delta \Theta (\alpha, \beta)$ to $(\alpha=1,\beta=0)$ or $(\alpha=0,\beta=3)$ using that $\delta \Theta (0,0) = 0$ (modulo numerical errors).
+\item extract multiple positions of the perihelion $\vec r_{MS} ( t_\mathrm{ph}^{(n)} )$ for a fixed value of $\alpha$ and $\beta$ and simulations times which covers several orbits.
+\item calculate the angle between the perihelions for two successive turns and compute the average angle $\delta \Theta (\alpha)$ over all individually computed angles.
+\item repeat the steps given above for different values of $\alpha$ (or $\beta$). The students should find a linear dependence of $\delta \Theta$ on $\alpha$ or $\beta$.
+\item interpolate $\delta \Theta (\alpha, \beta)$ to small values of $\alpha$ (or $\beta$) --- including  $\delta \Theta(0,0)=0$.
 \end{itemize}
@@
 	vec_r_before_last = vec_r_last
-	# store position of mercury in a new vector (since we will change M.pos)
+	# store position of Mercury in a new vector (since we will change M.pos)
 	vec_r_last = vector(M.pos)
-	#<...update mercury position...>
+	#<...update Mercury position...>
 	# check if at perihelion
@@
 \end{lstlisting}
+where the factor $180/\pi$ in the last line converts the unit of the angle from radians into degrees.
 To account for the statistical errors (e.g., numerical rounding errors) one can average over a few turns.
@@
 \end{lstlisting}
-Note that the "next" perihelion location is compared to the "previous" perihelion location in order to compute the perihelion motion.
-This is the case since, depending on the initial conditions and numerical uncertainty, one does not necessarily start at the perihelion and would therefore have a systematic error when comparing to initial coordinates.
+Note that the "next" perihelion location is compared to the "previous" perihelion location in order to compute the perihelion motion,
+since, depending on the initial conditions and numerical uncertainty, one does not necessarily start exactly at the perihelion and 
+would therefore introduce a systematic error by comparing to initial coordinates.
 
@@
 
-To extract the perihelion motion at $\alpha = 0$ and $\beta = 3$, we use the fact that $\delta \Theta$ is zero for $\alpha =0 = \beta$ --- if one could completely eliminate numerical errors.
+As an example we  extract the perihelion motion using the parameters provided by General Relativity, namely $\alpha = 0$ and $\beta = 3$.
 Considering one result from the numeric simulation, e.g., $\delta\Theta(\alpha = 0, \beta=10^6) \approx 10.7^{\degree}$, the sought-after angle can be calculated by a linear interpolation as
@@
 Depending on the knowledge of the students, multiple points and also estimated numerical uncertainties could be used to extract this value by a linear regression.
-Note that, when comparing the results of the simulation with the experimental result, one Earth year corresponds to $T_E/T_M \approx 365d/88d\approx 4.15$ mercury years.
+Note that, when comparing the results of the simulation with the experimental result, one Earth year corresponds to $T_E/T_M \approx 365d/88d\approx 4.15$ Mercury years.
 Therefore, the perihelion motion per 100 earth years is $0.116''\cdot 415\approx 48.2''$.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-Physical experiments in general, as do numerical simulations of physics, must always account for possible uncertainties.
-It is important for the students to recognize this fact and to understand, that the cause of the uncertainties lies, e.g., in the finite time steps.
-The students should therefore explore the source and size of the errors arising in this simulation at least qualitatively.
+Measurements as well as numerical simulations in physics should always be accompanied by estimates of the corresponding uncertainties.
+It is important for the students to recognize this fact and to understand that one cause of the uncertainties lies, e.g., in the finite time steps.
+The students should therefore explore sources and sizes of inaccuracies arising in this simulation at least qualitatively.
 
-Consider figure \ref{fcc3}, where the angle of the perihelion motion $\delta \Theta$ is shown for the first $10$ turns ($x$-axis) for different choices of $\Delta t$ (colors) and $\alpha$ (columns).
-As expected, the correlation between the number of turns and $\delta\Theta$, for sufficiently small time-steps $\Delta t$, is constant on average and the values depend solely on $\alpha$.
-However, at $\alpha=0$, there is an offset between the data points for $\delta \Theta$ and zero for different values of $\Delta t$.
-Because one knows that for the regular gravitational potential, the orbit should be bounded, this is a numerical error.
-Indeed this can be viewed as a magnitude estimate for the numerical error for the perihelion motion $\delta \Theta$.
+Consider figure \ref{fcc3}, where the angle of the perihelion motion $\delta \Theta$ is shown for the first $10$ turns ($x$-axis) for different choices of $\Delta t$ (colors) and $\beta$ (columns).
+In all cases $\alpha=0$.
+As expected, the correlation between the number of turns and $\delta\Theta$, for sufficiently small time-steps $\Delta t$, is constant on average and the values 
+depend solely on $\beta$.
+However, even at $\beta=0$, there is an offset between the data points for $\delta \Theta$ and zero for different values of $\Delta t$.
+Because one knows that for the regular gravitational potential the perihelion is stable, this offset is a numerical error.
+Indeed this offset can be viewed as an estimate for the numerical error for the perihelion motion $\delta \Theta$.
 
@@
 	\includegraphics[width=.99\textwidth]{figs/angular-variaton.pdf}
-	\caption{\label{fcc3}Motion of perihelion $\delta\Theta$ in degrees depending on the number of turns $N_t$ for different time steps $\Delta t$ (color) and different values for $\alpha$ (columns).
+	\caption{\label{fcc3}Motion of perihelion $\delta\Theta$ in degrees depending on the number of turns $N_t$ for different time steps $\Delta t$ (color) and different values for $\beta$ (columns).
 	Here, $\Delta t_0 \equiv 2 v_M(0)/c_a$.
@@
 Furthermore, for large time-steps, one can observe that the values of the perihelion motion oscillates between discrete values at different turns.
-This indicates that the time-step is chosen to large for accurate computations and accordingly the grid for the perihelion vectors to coarse.
-The exact perihelion vector of Mercury and the Sun could potenitally lay between the discrete values computed for finite time-steps.
-If this is the case, then for some turns, the code would recognize the vector before the exact location as the perihelion and sometimes the vector behind the exact location as the perihelion.
+This indicates that the time-step is chosen too large for an accurate computation and accordingly the grid for the perihelion vectors too coarse.
+The exact perihelion vector of Mercury and  Sun should lay between the discrete values computed for finite time-steps, since for 
+some turns, the code recognizes the vector before the exact location as the perihelion and sometimes the vector behind the exact location.
 This results in the observed oscillations of $\delta \Theta$.
-One can show that this is the case if one also prints the angles between the previous perihelion and the vector before the new perihelion (\code{vec\_r\_before\_last}) and after the new perihelion (\code{M}).
 
-To point the students to this problem, they can be asked to adjust the size of the time-steps and and observe the effect on the trajectories.
-This should also be done for large values which supposed to be problematic because the effect is more observable.
-For this study, it is advantageous to set $\alpha =0$, since in this case the trajectories should be closed and therefore deviations from the ideal case can be spotted easily.
-In figure \ref{fig:MercuryOrbit-a0-small-dt-large} the mercury trajectory for $\Delta t=2 v_M(0) / c_a \times 2$ is shown.
+To point the students to this problem, they can be asked to vary the size of the time-steps and and observe the effect on the trajectories.
+This should also be done for large values which are expected to be problematic.
+For this study, it is advantageous to set both $\alpha$ and $\beta$ to zero, 
+since in this case the trajectories should be closed and therefore deviations from the correct case can be spotted easily.
+In figure \ref{fig:MercuryOrbit-a0-small-dt-large} the Mercury trajectory for $\Delta t=2 v_M(0) / c_a \times 2$ is shown.
 One can clearly see the failure of the simulation to reproduce the physical trajectory.
 From examples like this it should be apparent that the time steps influence the accuracy of the simulation.
-Only in the limit $\Delta t \rightarrow 0$ would the simulation perfectly reproduce the actual trajectory.
+Only in the limit $\Delta t \rightarrow 0$ the simulation reproduces the actual trajectory. Thus, in any numerical simulation
+one always has to identify a proper compromise between numerical accuracy and time spent for the simulation 
+(clearly for $\Delta t\to 0$ the computing time goes to inifinity).
 
@@
 The idea of dimensional analysis is that in a system that can be controlled by expanding the relevant quantities (like the force) in some small parameter(s), the coefficients in the expansion should turn out to be of order unity (that means anything between about 0.1 and 10 is fine - but 0.01 or 100 is irritating) --- parameters in line with this are called "natural".
-Applied to the problem at hand given by Eq.~(\ref{eq:newton}) this statement implies that from naturalness one would expect that the parameter $\alpha$ is of order 1 when one uses for $r$ the average distance Mercury-sun, namely $\bar r=6\times 10^7$ km.
-From this one estimates for the expected angular shift per orbit
+Applied to the problem at hand given by Eq.~(\ref{eq:newton}) this statement implies that from naturalness one would expect the parameter $\alpha$ and $\beta$  of order 1
+ when one uses for $r$ the average distance Mercury-sun, namely $\bar r=6\times 10^7$ km.
+From this one estimates for the expected angular shift per orbit for $\alpha=1$ and $\beta=0$
 \begin{equation}
@@
 \end{equation}
-or
+or for $\alpha=0$ and $\beta=1$
 \begin{equation}
-\delta \Theta \simeq 2\pi\left(\frac{a^2}{\bar r^2}\right) = (\pi \times 5 \cdot 10^{-8}) \ \mbox{rad} = (\cdot 10^{-5})^o = (4\cdot 10^{-2}) ^{''} \ ,
+\delta \Theta \simeq 2\pi\left(\frac{a^2}{\bar r^2}\right) = (\pi \times 5 \cdot 10^{-8}) \ \mbox{rad} = (1 \cdot 10^{-5})^o = (4\cdot 10^{-2}) ^{''} \ ,
 \end{equation}
@@
 Thanks to Einstein we indeed know that the conclusion formulated above is correct --- the more general underlying theory is General Relativity and indeed Einstein was able to quantitatively explain the perihelion motion of Mercury from his equations.
-On the other hand had we found a dramatic deviation of $\alpha$ from unity one would have concluded that there is probably some other dynamics going on that drives this difference.
+On the other hand had we found the need of 
+ a dramatic deviation of the parameters from unity one would have concluded that there is probably some other dynamics going on that drives this difference.
 
-Indeed, we now several of those hierarchy problems in modern physics: e.g., the so called QCD $\Theta$ term, expected to be of natural size, is at present known to be at most $10^{-10}$.
-This smallness, called the strong CP problem, is so irritating that physicists like S.
-Weinberg even proposed that there must exist an additional particle, the axion, whose interactions are in charge of pushing $\Theta$ even to zero and there are now intense searches for this axion going on various labs.
+Indeed, nowadays there are  several of those hierarchy problems in modern physics: 
+e.g., the so called QCD $\Theta$ term, expected to be of natural size, is at present known to be at most $10^{-10}$.
+This smallness, called the strong CP problem, is so irritating that physicists like S. Weinberg even proposed that there must 
+exist an additional particle, the axion, whose interactions are in charge of pushing $\Theta$ even to zero.
+At present various
+intense experimental searches for this axion are going on at various labs.
 
@@
 Of course, this requires some basic knowledge in \texttt{VPython}, as they could acquire for example by working through a ball in the box example (see \cite{VPythonIntro}) or the like.
-Also, they could independently research the problem relevant parameters.
+Also, they could independently research the parameters relevant for the system.
 \item \textbf{Why can we work in one plane?}
@@
 
-Especially if the needed parameters are not specified beforehand, the students might have to experiment a bit, before getting the correct trajectories.
+Especially if the parameters are not specified beforehand, the students might have to experiment a bit, before getting the correct trajectories.
 But even if they are given, it might be beneficial to encourage the students to play with a few parameters, like the masses or the starting velocities, and observe their impact on the trajectories.
@@
 \begin{itemize}
-\item \textbf{Consider error due to finite time steps}
-
-To make the students realize that the finite time steps lead to inaccuracies in the trajectories, one can propose to test this by increasing the size of the time steps.
-This is best done in a program with unmodified gravitational force ($\alpha=0$), since in this case the trajectories are closed ellipses and numerical errors are most prominently visible in form of deviations from a bounded orbit.
-If the students choose the time steps big enough, they should witness big deviations.
-This exercise stresses the point, that by choosing increasingly smaller time steps, deviations from the physical trajectories can be reduced.
 \item \textbf{Measure calculation time}
 
-However in practice there is a limit to decreasing the time steps, because the time needed for the calculation grows simultaneously.
+In practice there is a limit to decreasing the time steps, because the time needed for the calculation grows simultaneously.
 By including \texttt{import time}
@@
 Ambitious students could even include further planets and see how the different planets interact.
-This is especially interesting, when discussing the perihelion motion of mercury, as it is mainly due to the influence of the other planets.
-Only a smaller part is due to general relativity.
+This is especially interesting, when discussing the perihelion motion of Mercury, as it is mainly due to the influence of the other planets.
+Only a smaller part is due to General Relativity.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{acknowledgments}
+\section*{Acknowledgments}
 
@@
 
-Once this test case works, one can start to test the mercury orbit simulation.
+Once this test case works, one can start to test the Mercury orbit simulation.
 We provide example files for both \texttt{juypter notebooks} and regular \python{} files.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-Below we provide an example script on how to program the visualization of the mercury orbit in \vpython{}.
-\lstinputlisting[language=Python]{mercury_orbit_base.py}
+Below we provide an example script on how to program the visualization of the Mercury orbit in \vpython{}.
+\lstinputlisting[language=Python]{Mercury_orbit_base.py}
 Further examples and template files can be found online\cite{}.\todo[inline]{@arXiv: online link}

### 2018-01-09 tex/MercuryNumerical.tex
updated figures

@@
 rS  = 3.0e-7 # in units of R0
-a2  = 8.2e-7 # in units of R0**2 \ ,
+a2  = 8.2e-7 # in units of R0**2
 \end{lstlisting}
@@
 vec_rM0 = vector(0, rM0, 0)
-vec_vM0 = vector(vM0, 0, 0) \ .
+vec_vM0 = vector(vM0, 0, 0)
 \end{lstlisting}
@@
 # Definition of the time step
-dt = 2 * vM0 / c_a / 20 \ .
+dt = 2 * vM0 / c_a / 20
 \end{lstlisting}
@@
 \begin{equation}
-	\delta\Theta (\alpha=0, \beta=3) = \frac{10.6^{\degree}}{10^6} \times 3 = 0.116''
+	\delta\Theta (\alpha=0, \beta=3) = \frac{10.7^{\degree}}{10^6} \times 3 = 0.116''
 	\, .

### 2018-01-09 tex/MercuryNumerical.tex
Fix name of sample script file

@@
 Below we provide an example script on how to program the visualization of the Mercury orbit in \vpython{}.
-\lstinputlisting[language=Python]{Mercury_orbit_base.py}
+\lstinputlisting[language=Python]{mercury_orbit_base.py}
 Further examples and template files can be found online\cite{}.\todo[inline]{@arXiv: online link}

### 2018-01-12 tex/MercuryNumerical.tex
My revision of the paper

@@
 \address{
-	$^1$ \textit{Institute f\"ur Kernphysik (IKP-3) and Institute for Advanced Simulations (IAS-4), Forschungszentrum J\"ulich, D-52425 J\"ulich, Germany}\\
+	$^1$ \textit{Institut f\"ur Kernphysik (IKP-3) and Institute for Advanced Simulations (IAS-4), Forschungszentrum J\"ulich, D-52425 J\"ulich, Germany}\\
 	$^2$ \textit{Institut f\"ur Neurowissenschaften und Medizin (INM-4), Forschungszentrum J\"ulich, D-52425 J\"ulich, Germany}\\
@@
 The paper includes details about the development of the code as well as a discussion of the visualization of the results.
-In addition a method is discussed how to estimate the size of the effect a priori that at the same time also can be employed to
-double check the results found numerically.
+In addition a method is discussed how to estimate the size of the effect a priori. Which allows to double check the results found numerically.
 \end{abstract}
@@
 
-Numerical simulations play a key role in modern physics for they allow one to tackle theoretical problems not accessible otherwise, e.g., since there are too many particle participating in the system (as in simulations for weather predictions) or the interactions are too complicated to allow for a systematic, perturbative approach (as in theoretical descriptions of nuclear particles at the fundamental level).
+Numerical simulations play a key role in modern physics for they allow one to tackle theoretical problems not accessible otherwise.
+This might be the case because there are too many particles participating in the system (as in simulations for weather predictions) or the interactions are too complicated to allow for a systematic, perturbative approach (as in theoretical descriptions of nuclear particles at the fundamental level).
 This paper introduces a project that allows one to demonstrate the power of numerical simulations.
-On the example of the perihelion motion of the planet Mercury the students are supposed to get in touch with
+On the example of the perihelion motion of the planet Mercury the students are supposed to learn about
 \begin{itemize}
@@
 \item effective tools to estimate the result a priori as an important cross check;
-\item the visualization of numerical results using \texttt{VPython}.
+\item the visualization of numerical results using \vpython{}.
 \end{itemize}
-The course as well as this paper is structured as follows: after an introduction to Newtonian dynamics and the concept of differential equations, their
-discretization is discussed based on Newtons law of gravitation and possible extensions thereof.
-Afterwards the visualization of the resulting trajectories using \texttt{VPython} is introduced and applied to the problem at hand. In addition,
+The course \todo{what course?} as well as this paper is structured as follows: after an introduction to Newtonian dynamics and the concept of differential equations, their
+discretization is discussed based on Newton's law of gravitation and possible extensions thereof.
+Afterwards the visualization of the resulting trajectories using \vpython{} is introduced and applied to the problem at hand. In addition,
  tools are developed to extract the relevant quantity from the result of the simulation.
-Finally the principle of dimensional analysis is presented as a tool not only to estimate some effect but also to cross check, if the results of the simulation are sensible.
+Finally, the principle of dimensional analysis is presented as a tool not only to estimate some effect but also to cross check whether the results of the simulation are sensible.
 
@@
 This purpose is served perfectly by the case chosen here, since Einsteins equations of General Relativity are fascinating to a very broad public.
-While their detailed study needs deeper knowledge in Differential Geometry, for the study outlined here very little math is necessary, such that high school students from about 10$^{\rm th}$ grade up should benefit from it.
-This was already demonstrated in the "Schülerakademie Teilchenphysik", where this course was tested successfully on two groups consisting of in total 24
-German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade in 2015 and 2017.
+Their detailed study needs a deep knowledge of Differential Geometry.
+For the study outlined here, however, very little math is necessary, such that high school students from about 10$^{\rm th}$ grade up should benefit from it.
+This was already demonstrated in the "Sch\"ulerakademie Teilchenphysik", where this course was tested successfully on two groups consisting in total of 24 German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade in 2015 and 2017\todo{do foreigners know what that means?
+what are typical ages of students?
+did grade 13 even exist in 2015?}.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Trajectories, velocities, accelerations and Newtons second law}\label{sec:tva}
+\section{Trajectories, velocities, accelerations and Newton's second law}\label{sec:tva}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-A physical system is understood, if the assumed force leads to the trajectories observed.
+A physical system is said to be understood if the assumed force acting on it leads to the observed trajectories.
 In other words, we need to show that we can calculate the location in space of the object of interest at any point in time, once the initial conditions are fixed properly.
-If we can neglect the finite size of this object (and in particular its orientation in space) the location is parametrized by a single three vector $\vec r(t)$.
-As will become clear below, in order to describe and control the dynamics of a physical object in addition we need to be able to calculate its velocity, $\vec v(t)$, related to the location via
+If we can neglect the finite size of this object (and in particular its orientation in space), the location is parametrized by a single three vector $\vec r(t)$.
+Additionally, we need to be able to calculate the objects velocity $\vec v(t)$ and acceleration $a(t)$ in order to describe and control its dynamics; this will become clear below.
+The velocity describes a change of location over time which, for some infinitesimally small $\Delta t$, can be written as
 \begin{equation}\label{eq:x_update}
-\vec r(t+\Delta t) = \vec r(t) + \vec v(t) \Delta t + ... \ , \label{eq:vdef}
+\vec r(t+\Delta t) = \vec r(t) + \vec v(t) \Delta t + \ldots \ . \label{eq:vdef}
 \end{equation}
-for some infinitesimally small $\Delta t$ and the acceleration, $\vec a(t)$ that describes the change of the velocity
+Similarly, the acceleration describes a change in velocity, i.e.
 \begin{eqnarray}\label{eq:v_update}
-\vec v(t+\Delta t) = \vec v(t) + \vec a(t) \Delta t + ...\ . \label{eq:adef}
+\vec v(t+\Delta t) = \vec v(t) + \vec a(t) \Delta t + \ldots\ . \label{eq:adef}
 \end{eqnarray}
-The dots in the above expressions indicate that there are in general additional terms that may be expressed with higher powers in $\Delta t$, however, for sufficiently small $\Delta t$ those can be safely neglected.
+The dots in these expressions indicate additional terms that may be expressed in terms of higher powers in $\Delta t$.
+While they are needed in general, for sufficiently small $\Delta t$ they can be safely neglected.
 Thus we may define the time derivative via
@@
 \end{equation}
-where $\Delta \vec r(t)=\vec r(t+\Delta t)-\vec r(t)$ and we introduced with the last expression a common short hand notation for time derivatives.
+where $\Delta \vec r(t)=\vec r(t+\Delta t)-\vec r(t)$ and we introduced a common short hand notation for time derivatives in the last expression.
 Analogously we get
@@
 \end{eqnarray}
-where we introduced in the second line the second derivative.
+where we introduced the second derivative in the second line.
 
-It was Newton who observed that if a body is at rest it will remain at rest, and if it is in motion it will remain in motion in a constant velocity in a straight line, unless it is acted upon by some force --- this is known as Newtons first law.
-Said differently: a force $\vec F$ shows up by changing the motion of some object.
+It was Newton who observed that if a body is at rest, it will remain at rest, and if it is in motion it will remain in motion at a constant velocity in a straight line, unless it is acted upon by some force --- this is known as Newton's first law.
+Formulated differently: a force $\vec F$ expresses itself by changing the motion of some object.
 This is quantified in Newton's second law
@@
 If the mass does not change~\footnote{%
-	A well known example where $m$ does change with time is a rocket, whose mass decreases as the rocket rises.
+	A well known example where $m$ does change with time is a rocket, whose mass decreases as fuel is burned.
 } with time, this reduces to the well known
@@
 Note that in general the force could depend also on the time or the velocity.
-We here restrict ourselves to the case relevant for our example, where the force depends on the location only.
+Here, we restrict ourselves to the case relevant for our example, where the force depends on the location only.
 Therefore, as soon as the force $\vec F(\vec r)$ is known for all $\vec r$ one can in principle calculate the trajectory, e.g., solving Eq.~(\ref{eq:newton2}) for $\vec r(t)$.
-Sometimes this requires some advanced knowledge in math, sometimes no closed form solution exists.
+Sometimes this requires some advanced knowledge of mathematics and sometimes no closed form solution exists.
 However, alternatively one can calculate the whole trajectory of some test body that experiences this force by a successive application of the rules given in Eqs.~(\ref{eq:vdef}) and (\ref{eq:adef}):
 \begin{enumerate}
-\item For a given time $t$, where $\vec r(t)$ and $\vec v(t)$ are known, use Eq.~(\ref{eq:newton2}) to calculate $\vec a(t)$.
-\item Use Eq.~(\ref{eq:vdef}) to calculate $\vec r(t+\Delta t)$ and
-\item then use Eq.~(\ref{eq:adef}) to calculate $\vec v(t+\Delta t)$.
+\item For a given time $t$, where $\vec r(t)$ and $\vec v(t)$ are known, use the force to calculate $\vec a(t) = \vec F(\vec r(t)) / m$, see Eq.~(\ref{eq:newton2}).
+\item Use Eq.~(\ref{eq:vdef}) to calculate $\vec r(t+\Delta t)$.
+\item Use Eq.~(\ref{eq:adef}) to calculate $\vec v(t+\Delta t)$.
 \item Go back to (i) with $t\to t+\Delta t$.
@@
 Clearly to initiate the procedure at some time $t_0$ both $\vec r(t_0)$ as well as $\vec v(t_0)$ must be known --- the  trajectories depend on these initial conditions~\footnote{%
-	In general a differential equation of $n^{\rm th}$ degree (where the highest derivative is $n$) needs $n$ initial conditions specified.
-	For $n=2$ those are often chosen as location and velocity at some defined time, but one may as well pick two location at different times.%
+	In general, a differential equation of $n^{\rm th}$ degree (where the highest derivative is order $n$) needs $n$ initial conditions specified.
+	For $n=2$ those are often chosen as location and velocity at some starting time, but one may as well pick two locations at different times.%
 }.
 
-For this procedure to work $\Delta t$ must be sufficiently small.
-What this means depends on the process studied.
-One way to estimate, if $\Delta t$ is small enough, is to verify, if the relation
+This procedure can only work if $\Delta t$ is sufficiently small.
+What this means depends on the process one studies.
+One way to estimate whether $\Delta t$ is small enough, is to verify whether the relation
 \begin{equation}
-|\vec v(t)| \gg \frac12|\vec a(t)|\Delta t = \frac{1}{2m} |\vec F(\vec r(t)) |\Delta t\ .
+|\vec v(t)| \gg \frac12|\vec a(t)|\Delta t = \frac{1}{2m} |\vec F(\vec r(t)) |\Delta t\
 \label{eq:check}
 \end{equation}
-holds, since the next term neglected in Eq.~(\ref{eq:vdef}) reads $(1/2)a(t)(\Delta t)^2$.
+holds. This follows from Eq.~(\ref{eq:vdef}) where the first term that we neglected reads $(1/2)a(t){(\Delta t)}^2$.
 Eq.~({\ref{eq:check}}) also shows that small (large) time steps are necessary (sufficient), if the force is strong (weak), since the time steps need to be small enough that all changes induced by the force get resolved.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-For this concrete example the starting point for the force is Newtons law of gravitation
+For this concrete example the starting point for the force is Newton's law of gravitation
 \begin{equation}
@@
 \end{equation}
-where $G_N=6.67\times 10^{-11}$ m$^3$kg$^{-1}$s$^{-2}$ is the Newtonian constant of gravitation, $m$ is the mass of Mercury and $M_\odot=2\times 10^{30}$ kg is the mass of the sun.
-In addition $r=|\vec r(t)|$ denotes the distance between sun and Mercury, when we assume that the sun is infinitely more heavy than the planet and located at the center of the coordinate system.
-Although this is not exact, since $m/M_\odot\sim 10^{-8}$ this is a pretty good approximation.
-For later convenience we introduce the Schwarzschildradius of the sun
+where $G_N=6.67\times 10^{-11}$ m$^3$kg$^{-1}$s$^{-2}$ is the Newtonian constant of gravitation, $m$ is the mass of Mercury and $M_\odot=2\times 10^{30}$ kg is the mass of the Sun.
+In addition $r=|\vec r(t)|$ denotes the distance between Sun and Mercury when we assume that the Sun is infinitely heavier than Mercury and located at the center of the coordinate system.
+Although this is not exact, it is a pretty good approximation because $m/M_\odot\sim 10^{-8}$.
+For later convenience we introduce the Schwarzschild radius of the Sun
 \begin{equation}
-r_S=\frac{2G_N  M_\odot}{c^2} = 3 \ \mbox{km} \ ,
-\label{rsdef}
+r_S=\frac{2G_N  M_\odot}{c^2} = 3 \ \mbox{km} \ , \label{rsdef}
 \end{equation}
 where $c=3\times 10^8$ m/s denotes the speed of light.
-Note that $r_S$ is the characteristic length scale of the gravitational field of the sun for --- up to the prefactor --- one can not form another quantity with dimensions of a length from $G_N$, $M_\odot$ and $c$.
-That the Schwarzschild radius is also an important quantity to characterize black holes is not relevant for the discussion at hand.
-With this Newtons second law reads
+Note that $r_S$ is the characteristic length scale of the gravitational field of the Sun for --- up to the prefactor --- one can not form another quantity with dimensions of a length from $G_N$, $M_\odot$ and $c$.
+The Schwarzschild radius is also an important quantity to characterize black holes; this is however not relevant for the discussion at hand.
+With this, Newton's second law reads
 \begin{equation}
-	\ddot{\vec r}      = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\frac{\vec r}{r} \, .
-\label{eq:newton}
+\ddot{\vec r}      = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\frac{\vec r}{r} \, . \label{eq:newton}
 \end{equation}
-In general an attractive force that vanishes at large distances leads,  depending on the initial conditions, either to bounded or to open orbits.
-In the latter case the planet simply disappears from the sun, since a given force can capture only bodies with not too large momenta.
+An attractive force that vanishes at large distances leads in general, depending on the initial conditions, either to bounded or to open orbits.
+In the latter case the planet simply disappears from the Sun, since a given force can capture only bodies with small enough momenta.
 The students may study those scenarios within their simulations by varying the start velocity while keeping the start location fixed.
 
-The bounded orbits that emerge from a potential that scales as $1/r$ are elliptic and fixed in space.
-In particular the point of closest approach of the planet to the sun, the perihelion, does not move.
-However, when a potential that vanishes for $r\to \infty$ deviates from $1/r$, the perihelion moves.
-Because of this, the behavior of the perihelion is a very sensitive probe of the gravitational potential.
+The bounded orbits that emerge from a potential that scales as $1/r$ (which corresponds to a force that scales as $1/r^2$) are elliptic and fixed in space.
+In particular the point of closest approach of the planet to the Sun, the perihelion, does not move.
+However, when a potential that vanishes for $r\to \infty$ deviates from $1/r$, the perihelion does move.
+This makes the behavior of the perihelion a very sensitive probe of the gravitational potential.
 
-The observed perihelion motion of Mercury is nowadays determined as
-$$(574.10\pm 0.65)'' \ \mbox{per
-100 earth years}$$
-where the symbol $''$ denotes "arc seconds": 1$''=(1/3600)^o$. The bulk of this number
+The observed perihelion motion of Mercury is nowadays determined as \todo{reference?}
+\[(574.10\pm 0.65)'' \ \mbox{per 100 earth years}\]
+where the symbol $''$ denotes "arc seconds": 1$''={(1/3600)}^o$. The bulk of this number
 can be understood by the presence of other planets within the Newtonian theory, since
@@
 residual motion of
-\begin{equation}\delta \Theta_M = (43)'' \ \mbox{in 100 earth years}\label{delT}\end{equation}
- remained unexplained, until Einstein quantified the
-predictions of General Relativity to this particular observable~\cite{Einstein}.
+\begin{equation}
+\delta \Theta_M = (43)'' \ \mbox{in 100 earth years} \label{delT}
+\end{equation}
+remained unexplained, until Einstein quantified the predictions of General Relativity to this particular observable~\cite{Einstein}.
 
-To allow for a movement of the perihelion we need to modify the force that appears in Eq.~(\ref{eq:newton}).
+To allow for a movement of the perihelion we need to modify the force that led to Eq.~(\ref{eq:newton}).
 This modification should be such that it depends on $r$ and it vanishes for $r\to \infty$.
-Based on the discussion presented above we may multiply the right hand side with a factor
+Based on the discussion presented above we may multiply the right hand side of~(\ref{eq:newton}) with a factor
 $(1+\alpha(r_S/r))$, where $\alpha$ denotes some dimensionless parameter.
-However, besides $r_S$, which is the characteristic length scale for the gravitational potential
-of the sun, one can define one more parameter, which is characteristic for the dynamics of Mercury, namely
+This seems like a natural way to parametrize the additional potential because it adds the smallest possible deviation from a $1/r$ potential expressed relative to the characteristic length $r_S$.
+However, besides $r_S$, which characterizes the gravitational potential of the Sun, there is also a characteristic parameter for the dynamics of Mercury:
 \begin{equation}
-a^2=\frac{\vec L\,^2}{m^2c^2}= \frac{(\vec r\times \dot{\vec r} \, )^2}{c^2} \ ,
-\label{a2def}
+r_a^2 := \frac{\vec L\,^2}{m^2c^2}= \frac{{(\vec r\times \dot{\vec r} \, )}^2}{c^2} \ , \label{a2def}
 \end{equation}
 where $\vec L$ denotes the angular momentum, which is a constant of motion for central potentials.
-It is easily verified that $a^2$ carries dimensions of a length squared.  Therefore we may use  the following ansatz
-for the modified potential
+It is easily verified that $r_a^2$ carries dimensions of length squared.
+We therefore have to add one more term to the potential; in analogy to the one discussed above, we use $\beta (r_a^2/r^2)$.
+Combining everything, we use the following ansatz for the modified equation of motion:
 \begin{equation}
-\ddot{\vec r} = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\left(1+\alpha\frac{r_S}{r}+\beta\frac{a^2}{r^2}\right) \, \frac{\vec{r}}{r} \ .
+\ddot{\vec r} = - \frac{c^2}{2}\frac{r_S}{r^2}\left(1+\alpha\frac{r_S}{r}+\beta\frac{r_a^2}{r^2}\right) \, \frac{\vec{r}}{r} \ .
 \label{eq:newton_art}
 \end{equation}
-For the system at hand $a^2$ may be estimated from the parameters of Mercury at its perihelion: The corresponding velocity is
- $\dot{\vec r}(t=0)=\abs{\vec v_M(0)} = 59$~km/s (here we
-already indicate that below we start the simulation at the perihelion) and the closest distance between sun and Mercury is
+For the system at hand, $r_a^2$ may be estimated from the parameters of Mercury at its perihelion: The corresponding velocity is
+ $\dot{\vec r}(t=0)=\abs{\vec v_M(0)} = 59$~km/s (here we already indicate that we will start the simulation at the perihelion) and the closest distance between Sun and Mercury is
 $ r_{MS}=\abs{\vec r_{MS}(0)} = 46 \cdot 10^6$km~\cite{MercuryFactSheet}. Note that at the perihelion $\vec r$ and $\dot{\vec r}$ are perpendicular to each other.
-We thus find that $(a^2/ r_{MS}^2) = 4\cdot10^{-8}$. Furthermore $(r_S/r_{MS})=6\cdot 10^{-8}$.
-Therefore (for a more detailed discussion about the underlying logic we refer to Sec.~\ref{sec:analysis}) one may expect from both correction terms an effect similar in size.
-Furthermore, any term higher order in either $a$ or $r_S$
-should be suppressed by additional seven orders of magnitude and thus can be safely neglected.
+We thus find that $(r_a^2/ r_{MS}^2) = 4\cdot10^{-8}$. Furthermore $(r_S/r_{MS})=6\cdot 10^{-8}$.
+Therefore (for a more detailed discussion about the underlying logic we refer to Sec.~\ref{sec:analysis}) one may expect from both correction terms an effect of similar size.
+Furthermore, any term of higher order in either $a$ or $r_S$
+should be suppressed by additional seven orders of magnitude and thus can be neglected safely.
 The parameters $\alpha$ and $\beta$ can be extracted from a fit to data --- this will be done below.
-However, since we have with the perihelion shift (see Eq.~(\ref{delT})) only a single number to study,
+However, since we have only a single number to study, the perihelion shift (see Eq.~(\ref{delT})),
 we can only fix either $\alpha$ or $\beta$ (or a linear combination thereof).
-At the same time the parameters can also be calculated from the assumed underlying theory and the actual values
-are in general specific for this theory. It turns out that
+The parameters can also be calculated from the underlying theory one uses. The actual values depend on the specific theory;
 General relativity gives~\cite{Einstein}
-\begin{equation}\label{eq:art-prediction}
-	\alpha = 0 \, , \quad \beta = 3 \, .
+\begin{equation}
+	\alpha = 0 \, , \quad \beta = 3 \, . \label{eq:art-prediction}
 \end{equation}
@@
 
-%For comparison: the typical distance of the Venus, the neighbor of Mercury, to the sun is $(r_S/r_{VS}) \sim 5 \cdot 10^{-8}$.
-%Therefore the expected ratio of residual motion, favoring in the orbit period $T$, would result in
-%\begin{equation}
+% For comparison: the typical distance of the Venus, the neighbor of Mercury, to the sun is $(r_S/r_{VS}) \sim 5 \cdot 10^{-8}$.
+% Therefore the expected ratio of residual motion, favoring in the orbit period $T$, would result in
+% \begin{equation}
 % 	\delta \Theta_V \approx \frac{1}{T_V} \delta \Theta_M T_M \frac{r_{MS}}{r_{VS}} \approx 9'' \, ,
-%\end{equation}
-%where $T_V$ ($T_M$) denotes the length of a venus (Mercury) year. This precisely describes the experimental value of $9''$.
+% \end{equation}
+% where $T_V$ ($T_M$) denotes the length of a venus (Mercury) year. This precisely describes the experimental value of $9''$.
 
@@
 In this section we present the numerical implementation as well as the visualisation of planetary trajectories and in
-particular
-the perihelion motion of Mercury.  We select \python{} \cite{Python} as  programming language, because \python{} is easy to learn, intuitive to understand and an open source language.
-No prior knowledge of \python{} or any other programming language is required, as we explain all the necessary steps to create the simulation.
-In  \ref{appendix:python} we give instructions for setting up \python{} on different operating systems.
+particular the perihelion motion of Mercury.
+We choose \python{}~\cite{Python} as programming language because \python{} is easy to learn, intuitive to understand and an open source language.
+No prior knowledge of \python{} or any other programming language is required, as we explain all necessary steps to create the simulation.
+In~\ref{appendix:python} we give instructions for setting up \python{} on different operating systems.
 We also provide a working example program that can be used as a template \url{this-upload-url}\todo{@arXiv: online link}.
@@
 To start the simulation one needs the "initial" distance and velocity [Eq.~(\ref{eq:x_update}) and Eq.~(\ref{eq:v_update})].
-As described above, here  one can simplify the problem by placing one planet (the "infinitely" heavy sun) in the center of the coordinate system and describe the motion of the other planet (Mercury) in a two-dimensional plane.
+As described above, here one can simplify the problem by placing one object (the "infinitely" heavy Sun) in the center of the coordinate system and keep it fixed.
 
 Since the trajectories do not depend on where on the orbit of Mercury the simulation is started, we
-use as initial (for $t=0$) parameters the values at the perihelion with $ \abs{\vec r_{MS}(0)} = 46 \cdot 10^6$~km and $ \abs{\vec v_M(0)} = 59$~km/s\cite{MercuryFactSheet}.
-Since the computer does not understand physical units, one has to express each variable in an appropriate unit.
-Both, for the numerical treatment and the intuitive understanding, it is useful to select parameters in a "natural range", e.g.,
-by expressing distances in $R_0 = 10^{10}$ m and time intervals in days, $T_0 = 1$d, where $d$ refers to earth days, such that one Mercury year is given by $T_M=88\,T_0$.
-With this choice, the initial distance of Mercury to the sun, the size of the initial velocity of Mercury and the acceleration pre-factors become
+use the values at the perihelion with $\abs{\vec r_{MS}(0)} = 46 \cdot 10^6\,\mathrm{km}$ and $\abs{\vec v_M(0)} = 59\,\mathrm{km/s}$\cite{MercuryFactSheet} as initial (for $t=0$) parameters.
+The computer does not understand physical units, hence one has to express each variable in an appropriate unit.
+Both for the numerical treatment and the intuitive understanding, it is useful to select parameters in a "natural range", e.g.,
+by expressing distances in $R_0 = 10^{10}\,\mathrm{m}$ and time intervals in days, $T_0 = 1\,\mathrm{d}$, where d refers to earth days. One Mercury year is given by $T_M=88\,T_0$.
+With this choice, the initial distance of Mercury to the Sun, the size of the initial velocity of Mercury and the acceleration prefactor (see Eq.~(\ref{eq:newton_art})) become
 \begin{eqnarray}
@@
 v_{M}(0)    = 0.51 \frac{R_0}{T_0} \, ,  \\
-a_M(r_{MS}) &=& 1.01 \frac{R_0}{T_0^2} \frac{1}{\left(r_{MS}/R_0\right)^2}
-\, .
+a_{MS}(r_{MS}) &=& \frac{c^2}{2}\frac{r_S}{r^2} = 1.01 \frac{R_0}{T_0^2} \frac{1}{{\left(r_{MS}/R_0\right)}^2}\, .
 \end{eqnarray}
@@
 	\caption{\label{fig:sun_merc}%
-		Sun Mercury system with relevant vectors.
-		Because Mercury is at its perihelion, its velocity is perpendicular to its direct connection vector with the sun.%
+		Sun-Mercury system with relevant vectors.
+		Mercury is at its perihelion, therefore its velocity is perpendicular to its direct connection vector with the sun.%
 	}
@@
 vM0 = 0.51   # in units of R0/T0
-c_a = 1.01   # in units of R0/T0**2
+c_a = 1.01   # in units of R0**3/T0**2
 TM  = 88.    # in units of T0
 rS  = 3.0e-7 # in units of R0
-a2  = 8.2e-7 # in units of R0**2
+ra2 = 8.2e-7 # in units of R0**2
 \end{lstlisting}
-where the last two quantities refer to the Schwarzschildradius of the sun and the parameter $a^2$ defined in Eqs. (\ref{rsdef}) and (\ref{a2def}), respectively.
-So far we only fixed the length of the vectors.
-Next, we set up the initial directions, which will describe the motion in the two-dimensional plane.
-We will build on the existing \python{} module \texttt{VPython}, which provides an implementation for treating vectors as well as their visualization.
+where $\texttt{c\_a} =a_{MS} r_{MS}^2$, the acceleration without the radius factor.
+The last two quantities refer to the Schwarzschild radius of the Sun and the parameter $r_a^2$ defined in Eqs. (\ref{rsdef}) and (\ref{a2def}), respectively.
+
+So far we only fixed the length of the vectors. Next, we set up the initial directions, which will describe the motion in space.
+We will build on the existing \python{} module \vpython{}, which provides an implementation for treating vectors as well as their visualization.
 The first object of interest is a \texttt{vector}, which takes three-dimensional coordinates as its input.
-Because of our choice of initial conditions (we picked the initial vectors in the perihelion), the velocity of Mercury is perpendicular to the vector which connects Mercury and sun (see figure \ref{fig:sun_merc}):
+With our choice of initial conditions (we picked the initial vectors in the perihelion), the velocity of Mercury is perpendicular to the vector which connects Mercury and Sun (see figure~\ref{fig:sun_merc}):
 \begin{lstlisting}
-# Import the class vector from vpythons module visual
+# Import the class vector from vpython
 from vpython import vector
@@
 \end{lstlisting}
-To fix the time step $\Delta t$ we use the estimate of Eq.~(\ref{eq:check}) as a guidance to write
+We use the estimate of Eq.~(\ref{eq:check}) as a guidance to fix the time step $\Delta t$ as
 \begin{lstlisting}
@@
 Here the factor $1/20$ makes sure that $\Delta t$ is indeed consistent with Eq.~(\ref{eq:check}).
-Now we are in the position to calculate location and velocity of the planet at $t_0+\Delta t$ using the following equations\footnote{%
+Now we are in the position to calculate location and velocity of the planet at $t_0+\Delta t$ using the following expressions\footnote{%
 	Note, that there are different notions for numerically integrating differential equations with different accuracies.
-	Thus, the ordering and exact equations for updating $a_{MS}$, $v_M$ and $r_{MS}$ are not unique, but all correct prescriptions lead to the same result for sufficiently small $\Delta t$.
-	For this particular problem, the following definition is in good balance between numerically stable and inexpensive to compute.
+	Thus, the ordering and exact expressions for updating $a_{MS}$, $v_M$ and $r_{MS}$ are not unique. However, all correct prescriptions lead to the same result for sufficiently small $\Delta t$.
+	For this particular problem, the definitions we show are in good balance between being numerically stable and inexpensive to compute.
 }%
@@
 # Compute the strength of the acceleration
-temp = 1 + alpha * rS / vec_rM_old.mag + beta * a2 / vec_rM_old.mag**2
+temp = 1 + alpha * rS / vec_rM_old.mag + beta * ra2 / vec_rM_old.mag**2
 aMS  = c_a * temp / vec_rM_old.mag**2
@@
 The difference and sum of two vectors, or the scalar vector multiplication return vectors themselves.
-Also the magnitude of a vector --- \texttt{vector.mag} --- is a property of the vector and can be easily extracted.
+Also the magnitude of a vector --- \code{vector.mag} --- is an attribute of the vector and can be easily extracted.
 
-It is handy to use \texttt{Python}s \texttt{function}s to embed repeating structures ("DRY" --- Don't Repeat Yourself)
+It is handy to use \texttt{Python}s functions to structure the program and hold repeating code ("DRY" --- Don't Repeat Yourself):
 \begin{lstlisting}
 # Define the coordinate and velocity update function
-def evolve_Mercury(vec_rM_old, vec_vM_old, alpha, beta):
+def evolve_mercury(vec_rM_old, vec_vM_old, alpha, beta):
     # Compute the strength of the acceleration
-	temp = 1 + alpha * rS / vec_rM_old.mag + beta * a2 / vec_rM_old.mag**2
+    temp = 1 + alpha * rS / vec_rM_old.mag + beta * ra2 / vec_rM_old.mag**2
     aMS  = c_a * temp / vec_rM_old.mag**2
@@
     # Update position vector
+    vec_rM_new = vec_rM_old + vec_vM_new * dt
     return vec_rM_new, vec_vM_new
@@
 \end{lstlisting}
-\python{}s syntax enforces a clean programming style: it is necessary that the body of the function is indented (by an arbitrary but consistent amount of \texttt{space}s or \texttt{tab}s) relative to the definition statement of the function.
+\python{}s syntax enforces a clean programming style: it is necessary that the body of the function is indented (by an arbitrary but consistent amount of \texttt{space}s\footnote{While \python{} allows using \texttt{tab}s as well, this can cause programming errors, because their displayed width depends on the text editor.}) relative to the definition statement of the function.
 Furthermore, \python{} is an Interpreter language.
 Each line of the code is executed when the Interpreter passes it.
-For this reason, all the variables defined before the function are global and thus known to the function.
-The variables within the function (e.g. \texttt{aMS}, \texttt{vec\_aMS}, $\dots$) are local and do not exist beyond the scope of the function.
+For this reason, if we define the parameters before the function, they can be used inside of the function.
+Variables defined within the function (e.g. \code{aMS}, \code{vec\_aMS}, $\dots$) are local and do not exist beyond the scope of the function; they can not be used after the \texttt{return} statement.
 
 Finally, we can describe the evolution by a \texttt{while}-loop
-\begin{lstlisting}%[float,floatplacement=H]
+\begin{lstlisting}
 t     = 0.0
@@
 beta  = 0.0
+# Set position and velocity to their starting points
+vec_rM = vec_rM0
+vec_vM = vec_vM0
 # Execute the loop as long as t < 2*TM
 while t < 2*TM:
-	vec_rM, vec_vM = evolve_Mercury(vec_rM, vec_vM, alpha, beta)
+    # Update position and velocity
+    vec_rM, vec_vM = evolve_mercury(vec_rM, vec_vM, alpha, beta)
+    # Advance time by one step
     t = t + dt
 \end{lstlisting}
-where for the start we set the parameter $\alpha=0 = \beta$ in order first study the properties of the pure
+where for the start we set the parameters $\alpha=0 = \beta$ in order to first study the properties of the pure
 $1/r^2$ force.
@@
 In each iteration of the \texttt{while}-loop, the previous distance and velocity are used to compute the new values,
-which directly overwrites the previous values to enter the next iteration.
+directly overwriting the previous values.
 The total runtime \texttt{2*TM} is the amount of "virtual" days the simulations should run.
@@
 To start the visualization one has to \texttt{import} further objects from the \texttt{VPython} module.
-Additional to the vector class, one has to include
+We change the import statement from before to
 \begin{lstlisting}
@@
 \end{lstlisting}
-The class sphere will represent Mercury and the sun in the simulation
+The class sphere will represent Mercury and the Sun in the simulation
 \begin{lstlisting}
-# define the initial coordinates; M = Mercury, S = sun
+# Define the initial coordinates; M = Mercury, S = Sun
 M = sphere(pos=vec_rM0,       radius=0.5,  color=color.red   )
 S = sphere(pos=vector(0,0,0), radius=1.5,  color=color.yellow)
-# and the initial velocities
+# And the initial velocities
 M.velocity = vec_vM0
 S.velocity = vector(0,0,0)
-# add a visible trajectory to Mercury
+# Add a visible trajectory to Mercury
 M.trajectory = curve(color=color.white)
 \end{lstlisting}
-We have placed the sun in the origin of our coordinate system and choose non-realistic radii sizes for visualization purposes.
-Last but not least, one should use in the code presented above directly the vectors related to the visualization. Thus
-this piece of code now reads
+We place the Sun in the origin of our coordinate system and choose non-realistic radii for visualization purposes.
+Last but not least, one should the vectors related to the visualization directly instead of extra vectors as was done before. Thus
+the \texttt{while}-loop becomes
 \begin{lstlisting}
@@
 beta  = 0.0
-# update the objects
+# Execute the loop as long as t < 2*TM
 while t < 2*TM:
-	# set the frame rate (you can choose a higher rate to accelerate the program)
+    # Set the frame rate (you can choose a higher rate to accelerate the program)
     rate(100)
-	# update the drawn trajectory with the current position
+    # Update the drawn trajectory with the current position
     M.trajectory.append(pos=M.pos)
-	# update the velocity and position
-	M.pos , M.velocity = evolve_Mercury(M.pos , M.velocity , alpha, beta)
-	# update time
+    # Update velocity and position
+    M.pos, M.velocity = evolve_mercury(M.pos , M.velocity , alpha, beta)
+    # Advance time by one step
     t = t + dt
@@
 
-If the starting values are chosen as advised in the previous section, the students should end up with a trajectory as depicted in figure \ref{fig:MercuryOrbit-a0-small}.
-At this point one might ask students to vary $\Delta t$ and observe the effect of this on the orbit [see also figure \ref{fig:MercuryOrbit-a0-small-dt-large}].
+If the starting values are chosen as advised in the previous section, the students should end up with a trajectory as depicted in figure~\ref{fig:MercuryOrbit-a0-small}.
+At this point one might ask students to vary $\Delta t$ and observe the effect of this on the orbit [see also figure~\ref{fig:MercuryOrbit-a0-small-dt-large}].
 To get the perihelion motion, the additional force term described in Eq.~(\ref{eq:newton_art}) has to be "turned on", e.g., by choosing  $\alpha\neq0$
-[figure \ref{fig:MercuryOrbit-a6-small}].
+[figure~\ref{fig:MercuryOrbit-a6-small}].
 This is a good opportunity to let the students play with the size of $\alpha$ and get a feeling for its impact on the trajectories. In particular
 they should discover that values of $\alpha$ of the order of $10^6$ are necessary to get a visible effect.
-For a nicely visible  perihelion motion it is advisable to not  use the correct Mercury values but a more excentric trajectory by choosing, e.g., $r_{MS}(0)=6R_0$ [figure \ref{fig:MercuryOrbit-a6-big}].
+For a nicely visible perihelion motion it is advisable to not use the correct Mercury values but a more excentric trajectory by choosing, e.g., $r_{MS}(0)=6R_0$ [figure~\ref{fig:MercuryOrbit-a6-big}].
 \begin{figure}[htb]
@@
 		\end{itemize}
-		Here, the time steps are defined by $\Delta t_0 \equiv 2 v_{M}(0) / c_a$ and the images are screenshots of the simulation (for modified colors).
+		Here, the time steps are defined by $\Delta t_0 \equiv 2 v_{M}(0) / c_a$ and the images are screenshots of the simulation (with modified colors).
 		Note that for different starting values of $r_{MS}(0)$, also the numbers of days for a "Mercury year" changes.
@@
 What do you think?}\\
-To extract the dependence of the amount of perihelion motion
- on the parameters $\alpha$ and $\beta$, we recommend to set one parameter to zero and to vary the other.
-Both parameter have similar effects on the perihelion motion.
-To calculate the size of the perihelion motion for a modified gravitational force, the students need to
+Both $\alpha$ and $\beta$ have similar effects on the perihelion motion.
+We recommend setting one of them to zero and varying the other when analyzing the perihelion motion in dependence on the parameters.
+In order to calculate the size of the perihelion motion for a modified gravitational force, the students need to
 \begin{itemize}
-\item extract multiple positions of the perihelion $\vec r_{MS} ( t_\mathrm{ph}^{(n)} )$ for a fixed value of $\alpha$ and $\beta$ and simulations times which covers several orbits.
-\item calculate the angle between the perihelions for two successive turns and compute the average angle $\delta \Theta (\alpha)$ over all individually computed angles.
+\item extract multiple positions of the perihelion $\vec r_{MS} ( t_\mathrm{ph}^{(n)} )$ for a fixed value of $\alpha$ and $\beta$. The  simulation time needs to cover several revolutions.
+\item calculate the angle between the perihelions for each pair of successive turns and compute the average angle $\delta \Theta (\alpha)$ over all individually computed angles.
 \item repeat the steps given above for different values of $\alpha$ (or $\beta$). The students should find a linear dependence of $\delta \Theta$ on $\alpha$ or $\beta$.
@@
 This can be done within the $while$-loop.
-To find identify a minimal distance to the sun, one has to know the last two positions \texttt{vec\_r\_last} and \texttt{vec\_r\_before\_last} in addition to the current position.
-If the length of the \texttt{last} vector is smaller than that of the \texttt{before\_last} vector and smaller than the current vector, one has passed the perihelion and its position is given by \texttt{vec\_r\_last}.
-It is useful to save the position of the perihelion in a list \texttt{list\_perih} for a certain number of turns (here e.g. \texttt{max\_turns}=10).
+To identify a minimal distance to the sun, one has to know the last two positions \code{vec\_rM\_past} and \code{vec\_rM\_before\_last} in addition to the current position.
+If the length of the \texttt{last} vector is smaller than that of the \texttt{before\_last} vector and smaller than the current vector, one has passed the perihelion and its position is given by \code{vec\_r\_last}.
+It is useful to save the position of the perihelion in a list \code{list\_perih} for a certain number of turns (here e.g.\ \code{max\_turns}=10).
 The implementation can be realized as follows\footnote{%
@@
 \begin{lstlisting}
-# set up vectors
-vec_r_last = vec_rM0
+# Set up vectors
+vec_rM_last = vec_rM0
 turns      = 0
@@
 list_perih = list()
-# find perihelion for each turn and print it out
+# Find perihelion for each turn and print it out
 while turns < max_turns:
-	vec_r_before_last = vec_r_last
-	# store position of Mercury in a new vector (since we will change M.pos)
-	vec_r_last = vector(M.pos)
+    vec_rM_before_last = vec_rM_last
+    # Store position of Mercury in a new vector (since we will change M.pos)
+    vec_rM_last = vector(M.pos)
     #<...update Mercury position...>
-	# check if at perihelion
-	if (vec_r_last.mag < M.pos.mag) and (vec_r_last.mag < vec_r_before_last.mag):
-		list_perih.append(vec_r_last)
-		turns = turns+1
+    # Check if at perihelion
+    if (vec_rM_last.mag < M.pos.mag) and (vec_rM_last.mag < vec_rM_before_last.mag):
+        list_perih.append(vec_rM_last)
+        turns = turns + 1
 \end{lstlisting}
-To compute the angle between two vectors one uses the following
-formula
+The angle between two vectors can be computed using
  \begin{equation}
@@
 \begin{lstlisting}
-# import function to resolve angle
+# Import functions to compute angle
 from vpython import acos, pi, dot
-# define function for angle extraction
+# Define function for angle extraction
 def angle_between(v1, v2):
@@
 where the factor $180/\pi$ in the last line converts the unit of the angle from radians into degrees.
-To account for the statistical errors (e.g., numerical rounding errors) one can average over a few turns.
-Depending on the programming proficiency of the students and time constraints, this can be either done by hand or, e.g., by implementing the following code
+To account for the statistical errors (e.g.\ numerical rounding errors) one can average over a few turns.
+Depending on the programming proficiency of the students and time constraints, this can be done either by hand or, e.g., by implementing the following code
 \begin{lstlisting}
@@
 for n in range(1, max_turns):
-	# calculate angle
+    # Calculate angle
     sum_angle = sum_angle + angle_between(list_perih[n-1], list_perih[n])
-# display the average
-print(sum_angle/(max_turns-1.))
+# Display the average
+print(sum_angle/(max_turns-1))
 \end{lstlisting}
-Note that the "next" perihelion location is compared to the "previous" perihelion location in order to compute the perihelion motion,
-since, depending on the initial conditions and numerical uncertainty, one does not necessarily start exactly at the perihelion and
-would therefore introduce a systematic error by comparing to initial coordinates.
+Note that the perihelion motion is computed from the locations stored in \code{list\_perih} and not based on the initial position.
+This is important because depending on the initial conditions and numerical uncertainties, the simulation does not necessarily start in the perihelion.
 
-As explained above [and further discussed in chapter \ref{sec:analysis}] the natural value for $\alpha$ would be of the order of $1$.
+As explained above [and further discussed in chapter~\ref{sec:analysis}] the natural value for $\alpha$ would be of the order of $1$.
 However, if the students use this value for $\alpha$, they will find that the change in the trajectories is close to invisible and the numerical uncertainty is much larger than the result.
-Therefore, to estimate the size of the perihelion motion, it is more advisable to use the fact that there is a linear dependence between $\alpha$ and $\delta \Theta$\footnote{%
+It is therefore more advisable to use the fact that there is a linear dependence between $\alpha$ and $\delta \Theta$ to estimate the size of the perihelion motion\footnote{%
 	Of course for very large values of $\alpha$ this linear relation will eventually break down.
@@
 The students can convince themselves of this fact by plotting the angles $\delta \Theta$ they have extracted from their simulation over $\alpha$.
-This can be done either by hand, \python{} (e.g., with \texttt{matplotlib} \cite{Matplotlib}), or using another program like, e.g., Excel.
-For instance figure \ref{fig:AlphaAngle} demonstrates such a plot.
+This can be done either by hand, \python{} (e.g., with \texttt{matplotlib}~\cite{Matplotlib}), or using another program like, e.g., Excel.
+For instance figure~\ref{fig:AlphaAngle} demonstrates such a plot.
 
@@
 	\end{subfigure}
-	\caption{\label{fig:AlphaAngle} Linear relation between $\alpha$ [$\beta$] and and the perihelion motion $\delta \Theta$ for $\Delta t=2v_M(0)/c_a/100$ and $\beta=0$ [$\alpha=0$].}
+	\caption{\label{fig:AlphaAngle} Linear relation between $\alpha$ (a), $\beta$ (b) and the perihelion motion $\delta \Theta$ for $\Delta t=2v_M(0)/c_a/100$ and the other parameter set to zero.}
 \end{figure}
@@
 Therefore, the perihelion motion per 100 earth years is $0.116''\cdot 415\approx 48.2''$.
-This is close to experimental value of $\delta\Theta_{\mathrm{exp}} = 43.0''$.
+This is close to the experimental value of $\delta\Theta_{\mathrm{exp}} = 43.0''$.
 
@@
 
-Consider figure \ref{fcc3}, where the angle of the perihelion motion $\delta \Theta$ is shown for the first $10$ turns ($x$-axis) for different choices of $\Delta t$ (colors) and $\beta$ (columns).
+There are many potential sources of uncertainties in a simulation and discussing all of them is be beyond the scope of this work.
+Instead, we focus on the most accessible source, numerical errors due to finite time steps $\Delta t$.
+Additional sources of errors include the omission of terms in Eqs.~(\ref{eq:vdef}) and~(\ref{eq:adef}) and the infinite mass approximation of the Sun (i.e.\ keeping the Sun's position fixed).
+
+Consider figure~\ref{fcc3}, where the angle of the perihelion motion $\delta \Theta$ is shown for the first $10$ turns ($x$-axis) for different choices of $\Delta t$ (colors) and $\beta$ (columns).
 In all cases $\alpha=0$.
-As expected, the correlation between the number of turns and $\delta\Theta$, for sufficiently small time-steps $\Delta t$, is constant on average and the values
-depend solely on $\beta$.
-However, even at $\beta=0$, there is an offset between the data points for $\delta \Theta$ and zero for different values of $\Delta t$.
-Because one knows that for the regular gravitational potential the perihelion is stable, this offset is a numerical error.
-Indeed this offset can be viewed as an estimate for the numerical error for the perihelion motion $\delta \Theta$.
+As expected, $\delta\Theta$ is approximately constant for sufficiently small time-steps $\Delta t$ and its value depends solely on $\beta$.
+$\delta\Theta$ deviates from the expected value of zero for $\alpha = \beta = 0$.
+Using the data points in figure~\ref{fig:AlphaAngle} and extrapolating to $\alpha=\beta=0$ without enforcing $\delta\Theta(0,0) = 0$ leads to a similar offset.
+This offset is a numerical error and can be taken as an estimate for the error of the simulation.
 
@@
 	Here, $\Delta t_0 \equiv 2 v_M(0)/c_a$.
-	Offsets of $\delta \Theta(\alpha=0, \beta=0)$ from zero can be viewed as an error magnitude estimated and oscillations in $\delta \Theta$ indicate to coarse time steps.
+	Offsets of $\delta \Theta(\alpha=0, \beta=0)$ from zero can be used as an estimated of the magnitude of the error. Oscillations in $\delta \Theta$ indicate too coarse time steps.
 }
@@
 
-Furthermore, for large time-steps, one can observe that the values of the perihelion motion oscillates between discrete values at different turns.
-This indicates that the time-step is chosen too large for an accurate computation and accordingly the grid for the perihelion vectors too coarse.
-The exact perihelion vector of Mercury and  Sun should lay between the discrete values computed for finite time-steps, since for
-some turns, the code recognizes the vector before the exact location as the perihelion and sometimes the vector behind the exact location.
-This results in the observed oscillations of $\delta \Theta$.
+Furthermore, one can observe that for large time-steps the values of the perihelion motion oscillate between discrete values at different turns.
+This can be explained as follows: The sample code we present to find the perihelion vectors only finds the position closest to the Sun.
+This is not the true perihelion but only an approximation to it. Sometimes the program finds a point before and sometimes after the perihelion which causes the observed oscillations.
+The precision of this approximation rises as the time-step shrinks.
+Hence, strong visible oscillations in $\delta\Theta$ indicate a too large time-step.
 
-To point the students to this problem, they can be asked to vary the size of the time-steps and and observe the effect on the trajectories.
-This should also be done for large values which are expected to be problematic.
-For this study, it is advantageous to set both $\alpha$ and $\beta$ to zero,
-since in this case the trajectories should be closed and therefore deviations from the correct case can be spotted easily.
-In figure \ref{fig:MercuryOrbit-a0-small-dt-large} the Mercury trajectory for $\Delta t=2 v_M(0) / c_a \times 2$ is shown.
+This problem can be pointed out to the students by asking them to vary the size of the time-steps and observe the effect on the trajectories.
+This should also be done for a wide range of values to show both good and problematic regimes.
+It is advantageous to set both $\alpha$ and $\beta$ to zero for this study, since in this case the trajectories should be closed and deviations from the correct case can be spotted easily.
+Figure~\ref{fig:MercuryOrbit-a0-small-dt-large} shows the Mercury trajectory for $\Delta t=2 v_M(0) / c_a \times 2$.
 One can clearly see the failure of the simulation to reproduce the physical trajectory.
 From examples like this it should be apparent that the time steps influence the accuracy of the simulation.
-Only in the limit $\Delta t \rightarrow 0$ the simulation reproduces the actual trajectory. Thus, in any numerical simulation
+The simulation reproduces the actual trajectory only in the limit $\Delta t \rightarrow 0$. Thus, in any numerical simulation
 one always has to identify a proper compromise between numerical accuracy and time spent for the simulation
-(clearly for $\Delta t\to 0$ the computing time goes to inifinity).
+(clearly for $\Delta t\to 0$ the computing time goes to infinity).
 
@@
 
-The idea of dimensional analysis is that in a system that can be controlled by expanding the relevant quantities (like the force) in some small parameter(s), the coefficients in the expansion should turn out to be of order unity (that means anything between about 0.1 and 10 is fine - but 0.01 or 100 is irritating) --- parameters in line with this are called "natural".
-Applied to the problem at hand given by Eq.~(\ref{eq:newton}) this statement implies that from naturalness one would expect the parameter $\alpha$ and $\beta$  of order 1
- when one uses for $r$ the average distance Mercury-sun, namely $\bar r=6\times 10^7$ km.
+The idea of dimensional analysis is that in a system that can be controlled by expanding the relevant quantities (like the force) in some small parameter(s), the coefficients in the expansion should turn out to be of order one (that means anything between about 0.1 and 10 is fine --- but 0.01 or 100 is irritating); parameters in line with this are called "natural".
+Applied to the problem at hand given by Eq.~(\ref{eq:newton}) this statement implies that from naturalness one would expect the parameters $\alpha$ and $\beta$ of order one when $r$ is about the average distance Mercury-Sun, namely $\bar r=6\times 10^7\,\mathrm{km}$.
 From this one estimates for the expected angular shift per orbit for $\alpha=1$ and $\beta=0$
@@
 \end{equation}
-which leads to a shift of about $30^{''}$ [$15^{''}$] in 100 earth years to be compared to the empirical value of $43^{''}$.
-Thus indeed the amount of perihelion motion of Mercury is in line with expectations, $if$ --- and this is an important "if" --- the Newtonian dynamics is simply the leading term of some more general underlying theory.
-In particular, no new scales enter in the correction terms.
+which leads to a shift of about $30^{''}$ and $15^{''}$, respectively, in 100 earth years. These are to be compared to the empirical value of $43^{''}$.
+Thus the amount of perihelion motion of Mercury is indeed in line with expectations, \textit{if} --- and this is an important "if" --- the Newtonian dynamics is simply the leading term of some more general underlying theory.
+In particular, no new scales enter in the correction terms in addition to $r_S$ and $r_a^2$.
 
-Thanks to Einstein we indeed know that the conclusion formulated above is correct --- the more general underlying theory is General Relativity and indeed Einstein was able to quantitatively explain the perihelion motion of Mercury from his equations.
-On the other hand had we found the need of
- a dramatic deviation of the parameters from unity one would have concluded that there is probably some other dynamics going on that drives this difference.
+Thanks to Einstein we indeed know that the conclusion formulated above is correct; the more general underlying theory is General Relativity and Einstein was able to explain the perihelion motion of Mercury from his equations quantitatively.
+On the other hand, had we found the need for a dramatic deviation of the parameters from unity one would have concluded that there is probably some other dynamics that drives this difference.
 
-Indeed, nowadays there are  several of those hierarchy problems in modern physics:
-e.g., the so called QCD $\Theta$ term, expected to be of natural size, is at present known to be at most $10^{-10}$.
-This smallness, called the strong CP problem, is so irritating that physicists like S. Weinberg even proposed that there must
-exist an additional particle, the axion, whose interactions are in charge of pushing $\Theta$ even to zero.
-At present various
-intense experimental searches for this axion are going on at various labs.
+Indeed, nowadays there are  several of those hierarchy problems in modern physics: e.g., the so called QCD $\Theta$ term, expected to be of natural size, is at present known to be at most $10^{-10}$.
+This smallness, called the strong CP problem, is so irritating that physicists like S. Weinberg even proposed that there must exist an additional particle, the axion.
+Its interactions would even push $\Theta$ to zero.
+At present various intense experimental searches for this axion are going on at various labs.
 
@@
 
-In chapter \ref{sec:Numerical Implementation} we suggested to present the material by using a template as well as a step by step instruction to guide the students through the problem solution.
+In chapter~\ref{sec:Numerical Implementation} we suggested to present the material by using a template as well as a step by step instruction to guide the students through the problem solution.
 These instructions can be cut down or left out depending on available time and numerical/computational versatility of the students.
@@
 Instead of providing the template to the students, they could build the program from scratch.
-Of course, this requires some basic knowledge in \texttt{VPython}, as they could acquire for example by working through a ball in the box example (see \cite{VPythonIntro}) or the like.
+Of course, this requires some basic knowledge in \vpython{}, which they could acquire for example by working through "a ball in a box" (see~\cite{VPythonIntro}) or similar tutorials.
 Also, they could independently research the parameters relevant for the system.
@@
 
-In the code the third coordinate ($z$) is never used.
+In the code the third coordinate of all vectors is set to zero and never used.
 On the first glance, this might seem like a simplification.
-The students could work out on their own, why this does not imply any loss of generality.
+This choice is however possible without loss of generality because we are dealing with a central potential.
+The students could work out the reason behind this choice on their own.
 \item \textbf{What is the impact of the different parameters?}
@@
 
-Simulations always involve a balance between the time needed for the calculations and the demanded accuracy for the results.
+Simulations always involve a balance between time needed for the calculations and the demanded accuracy for the results.
 Even though this is a rather simple example, it contains some opportunities to make these concepts accessible to the students.
@@
 In practice there is a limit to decreasing the time steps, because the time needed for the calculation grows simultaneously.
-By including \texttt{import time}
+The following snippet shows how the run time of function \texttt{main} can be measured:
 \begin{lstlisting}
+import time
 start_time = time.time()
 main()
-print("--- %s seconds ---" % (time.time() - start_time))
+print("--- {} seconds ---".format(time.time() - start_time))
 \end{lstlisting}
-the students can measure the time needed by their program.
 By varying $\Delta t$ they can validate, that there is indeed approximately an anti-proportional dependence.
@@
 
-Of course by optimizing the code, an improvement in accuracy can be achieved without changing $\Delta t$.
+Obviously, an improvement in accuracy can be achieved by using a better algorithm without changing $\Delta t$.
 The simplest way to demonstrate this might be given by the implementation of the Verlet (see e.g.~\cite{Hairer03geometricnumerical} and references therein) integration instead of using the simple Euler method.
@@
 \begin{itemize}
-\item \textbf{Non-stationary sun}
+\item \textbf{Non-stationary Sun}
 
-It might be interesting to abandon the simplification of a stationary sun, as it nicely illustrates Newton's third law.
+It might be interesting to abandon the simplification of a stationary Sun, as it nicely illustrates Newton's third law.
 Here it might also be advisable to reduce the mass of the sun to have a more visible result.
@@
 \clearpage
-\section{The code}\label{sec:code}
+\section{The complete code}\label{sec:code}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-Below we provide an example script on how to program the visualization of the Mercury orbit in \vpython{}.
+Below we provide the complete code as developed above.
 \lstinputlisting[language=Python]{mercury_orbit_base.py}

### 2018-01-12 tex/mercury_orbit_base.py
Update complete Python script for paper

@@ -1,13 +1,13 @@
 # Import all objects from the vpython module
 from vpython import *
 
-# Definition of parameters 
+# Definition of parameters
 rM0 = 4.6    # in units of R0
 vM0 = 0.51   # in units of R0/T0
-c_a = 1.01   # in units of R0/T0**2
-TM  = 88.    # in units of T0 
+c_a = 1.01   # in units of R0**3/T0**2
+TM  = 88.    # in units of T0
 rS  = 3.0e-7 # in units of R0
-a2  = 8.2e-7 # in units of R0**2
+ra2 = 8.2e-7 # in units of R0**2
 
 # Initialize distance and velocity vectors
 vec_rM0 = vector(0, rM0, 0)
@@ -18,27 +18,36 @@ dt = 2 * vM0 / c_a / 20
 
 # Define the coordinate and velocity update
 def evolve_mercury(vec_rM_old, vec_vM_old, alpha, beta):
-    temp = 1 + alpha * rS / vec_rM_old.mag + beta * a2 / vec_rM_old.mag**2
+    # Compute the strength of the acceleration
+    temp = 1 + alpha * rS / vec_rM_old.mag + beta * ra2 / vec_rM_old.mag**2
     aMS  = c_a * temp / vec_rM_old.mag**2
+    # Multiply by the direction
     vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
+    # Update velocity vector
     vec_vM_new = vec_vM_old + vec_aMS * dt
+    # Update position vector
     vec_rM_new = vec_rM_old + vec_vM_new * dt
     return vec_rM_new, vec_vM_new
 
-# Define the initial coordinates; M = mercury, S = sun
+# Define the initial coordinates; M = mercury, S = Sun
 M = sphere(pos=vec_rM0,       radius=0.5,  color=color.red   )
 S = sphere(pos=vector(0,0,0), radius=1.5,  color=color.yellow)
+# And the initial velocities
 M.velocity = vec_vM0
 S.velocity = vector(0,0,0)
-# add a visible trajectory to mercury
+# Add a visible trajectory to mercury
 M.trajectory = curve(color=color.white)
 
-# Start the simulation
 t     = 0.0
 alpha = 0.0
 beta  = 0.0
+# Execute the loop as long as t < 2*TM
 while t < 2*TM:
+    # Set the frame rate (you can choose a higher rate to accelerate the program)
     rate(100)
[…]

### 2018-01-12 scripts/publish/base_solution.py
Remove "publish" scripts

@@ -1,75 +1,0 @@
-"""
-Solution to the base problem.
-
-Simulates the movement of Mercury and displays its position and a trajectory at
-regular intervals.
-The additional term ~ alpha/r**3 is included in the force. In order to simulate
-without it, simply set alpha = 0 before the while loop.
-"""
-
-# import everything from vpython (need graphics output, vectors, etc.)
-from vpython import *
-
-# set up display
-scene.width      = 1680
-scene.height     = 1024
-scene.background = color.white
-scene.center     = vector(0, -2, 0)
-
-# definition of parameters
-# values computed using https://nssdc.gsfc.nasa.gov/planetary/factsheet
-rM0 = 4.6    # initial radius of Mercury orbit, in units of R0
-vM0 = 0.51   # initial orbital speed of Mercury, in units of R0/T0
-c_a = 1.01   # base acceleration of Mercur, in units of R0/T0**2
-rS  = 3.e-7  # Schwarzschild radius of Sun, in units of R0
-
-# initialize distance and velocity vectors of Mercury (at perihelion)
-vec_rM0 = vector(0, rM0, 0)
-vec_vM0 = vector(vM0, 0, 0)
-
-# define graphical objects; M = Mercury, S = Sun
-M = sphere(pos=vec_rM0,         radius=0.5,  color=color.red   )
-S = sphere(pos=vector(0, 0, 0), radius=1.5,  color=color.yellow)
-# and the initial velocities
-M.velocity = vec_vM0
-S.velocity = vector(0, 0, 0)
-
-# add a visible trajectory to Mercury
-M.trajectory = curve(color=color.black, radius=0.005)
-
-def evolve_mercury(vec_rM_old, vec_vM_old, alpha):
-    """
-    Advance Mercury in time by one step of length dt.
-    Arguments:
-         - vec_rM_old: old position vector of Mercury
-         - vec_vM_old: old velocity vector of Mercury
-         - alpha: strength of 1/r**3 term in force
-    Returns:
-         - vec_rM_new: new position vector of Mercury
-         - vec_vM_new: new velocity vector of Mercury
-    """
-
-    # compute the absolute value of the acceleration
-    aMS = c_a * ( 1. + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
-    # multiply by the direction to get the acceleration vector
-    vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )
-    # update velocity vector
[…]

### 2018-01-12 scripts/publish/perihelion.py

@@ -1,106 +1,0 @@
-"""
-Extension of the base solution to measure and display the preihelion.
-
-Extends base_solution.py by keeping track of the location of the perihelion
-of Mercury. Computes and outputs the angle by which it changes over
-the course of the simulation.
-
-The stopping criterion is different than in base_solution.py. Uses a fixed number
-of revolutions around the sun instead of a fixed run time.
-"""
-
-# import everything from vpython (need graphics output, vectors, etc.)
-from vpython import *
-
-# set up display
-scene.width      = 1680
-scene.height     = 1024
-scene.background = color.white
-scene.center     = vector(0, -2, 0)
-
-# definition of parameters
-# values computed using https://nssdc.gsfc.nasa.gov/planetary/factsheet
-rM0 = 4.6    # initial radius of Mercury orbit, in units of R0
-vM0 = 0.51   # initial orbital speed of Mercury, in units of R0/T0
-c_a = 1.01   # base acceleration of Mercur, in units of R0/T0**2
-rS  = 3.e-7  # Schwarzschild radius of Sun, in units of R0
-
-# initialize distance and velocity vectors of Mercury (at perihelion)
-vec_rM0 = vector(0, rM0, 0)
-vec_vM0 = vector(vM0, 0, 0)
-
-# define graphical objects; M = Mercury, S = Sun
-M = sphere(pos=vec_rM0,         radius=0.5,  color=color.red   )
-S = sphere(pos=vector(0, 0, 0), radius=1.5,  color=color.yellow)
-# and the initial velocities
-M.velocity = vec_vM0
-S.velocity = vector(0, 0, 0)
-
-# add a visible trajectory to Mercury
-M.trajectory = curve(color=color.black, radius=0.005)
-
-def evolve_mercury(vec_rM_old, vec_vM_old, alpha):
-    """
-    Advance Mercury in time by one step of length dt.
-    Arguments:
-         - vec_rM_old: old position vector of Mercury
-         - vec_vM_old: old velocity vector of Mercury
-         - alpha: strength of 1/r**3 term in force
-    Returns:
-         - vec_rM_new: new position vector of Mercury
-         - vec_vM_new: new velocity vector of Mercury
-    """
-
-    # compute the absolute value of the acceleration
-    aMS = c_a * ( 1. + alpha * rS / vec_rM_old.mag  ) / vec_rM_old.mag**2
-    # multiply by the direction to get the acceleration vector
[…]

### 2018-01-12 scripts/publish/template.py

@@ -1,72 +1,0 @@
-"""
-Template for the simulation of Mercuries orbit.
-
-Sets up the graphics display and defines basic parameters like the position and
-velocity of Mercury. Also provides a basic structure for the program to get started
-quickly.
-
-All parts that have to be filled in are marked by '### TODO' and brief descriptions
-of the tasks.
-
-Basis run parameters are given just before the while loop at the end. They can be
-changed freely but the values given have proven to yield good results.
-"""
-
-# import everything from vpython (need graphics output, vectors, etc.)
-from vpython import *
-
-# set up display
-scene.width      = 1680
-scene.height     = 1024
-scene.background = color.white
-scene.center     = vector(0, -2, 0)
-
-# definition of parameters
-# values computed using https://nssdc.gsfc.nasa.gov/planetary/factsheet
-rM0 = 4.6    # initial radius of Mercury orbit, in units of R0
-vM0 = 0.51   # initial orbital speed of Mercury, in units of R0/T0
-c_a = 1.01   # base acceleration of Mercur, in units of R0/T0**2
-rS  = 3.e-7  # Schwarzschild radius of Sun, in units of R0
-
-# initialize distance and velocity vectors of Mercury (at perihelion)
-vec_rM0 = vector(0, rM0, 0)
-vec_vM0 = vector(vM0, 0, 0)
-
-### TODO
-#  - Define graphical objects that you want to display.
-#    Give them initial positions and velocities.
-#  - Enable drawing of trajectory of Mercury.
-
-def evolve_mercury(vec_rM_old, vec_vM_old, alpha):
-    """
-    Advance Mercury in time by one step of length dt.
-    Arguments:
-         - vec_rM_old: old position vector of Mercury
-         - vec_vM_old: old velocity vector of Mercury
-         - alpha: strength of 1/r**3 term in force
-    Returns:
-         - vec_rM_new: new position vector of Mercury
-         - vec_vM_new: new velocity vector of Mercury
-    """
-
-    ### TODO
-    #  - Calculate new position and velocity of Mercury here.
-    # vec_rM_new = ??
-    # vec_vM_new = ??
-
-    return vec_rM_new, vec_vM_new
-
-# run parameters
-dt = 2. * vM0 / c_a / 20  # time step
-alpha   = 1.e6            # strength of 1/r**3 term
-time = 0                  # current simulation time
-max_time = 1000*dt        # maximum simulation time
-
-# run the simulation for a given time and draw trajectory
[…]

### 2018-01-12 scripts/publish/three_body.py

@@ -1,82 +1,0 @@
-"""
-Extension of the base solution by simulating Venus as a second planet.
-
-Three body forces are neglected. Only the gravitational force of the Sun on
-Mercury and Venus is used.
-"""
-
-# import everything from vpython (need graphics output, vectors, etc.)
-from vpython import *
-
-# set up display
-scene.width      = 1680
-scene.height     = 1024
-scene.background = color.white
-scene.center     = vector(0, -2, 0)
-
-# definition of parameters
-# values computed using https://nssdc.gsfc.nasa.gov/planetary/factsheet
-rM0 = 4.6    # initial radius of Mercury orbit, in units of R0
-vM0 = 0.51   # initial orbital speed of Mercury, in units of R0/T0
-rV0 = 10.7   # initial radius of Venus orbit, in units of R0
-vV0 = 0.3    # initial orbital speed of Venus, in units of R0/T0
-c_a = 1.01   # base acceleration of Mercur, in units of R0/T0**2
-rS  = 3.e-7  # Schwarzschild radius of Sun, in units of R0
-
-# initialize distance and velocity vectors of Mercury and Venus (at perihelion)
-vec_rM0 = vector(0, rM0, 0)
-vec_vM0 = vector(vM0, 0, 0)
-vec_rV0 = vector(0, rV0, 0)
-vec_vV0 = vector(vV0, 0, 0)
-
-# define graphical objects; M = Mercury, V = Venus, S = Sun
-M = sphere(pos=vec_rM0,         radius=0.5,  color=color.red   )
-V = sphere(pos=vec_rV0,         radius=0.5,  color=color.orange)
-S = sphere(pos=vector(0, 0, 0), radius=1.5,  color=color.yellow)
-# and the initial velocities
-M.velocity = vec_vM0
-V.velocity = vec_vV0
-S.velocity = vector(0, 0, 0)
-
-# add a visible trajectory to Mercury and Venus
-M.trajectory = curve(color=color.black, radius=0.005)
-V.trajectory = curve(color=color.black, radius=0.005)
-
-def evolve_planet(vec_r_old, vec_v_old, alpha):
-    """
-    Advance one planet in time by one step of length dt.
-    Arguments:
-         - vec_r_old: old position vector of planet
-         - vec_v_old: old velocity vector of planet
-         - alpha: strength of 1/r**3 term in force
-    Returns:
-         - vec_r_new: new position vector of planet
-         - vec_v_new: new velocity vector of planet
-    """
-
[…]

### 2018-01-15 tex/MercuryNumerical.tex
added scripts url + tiny mods

@@
 \end{itemize}
-The course \todo{what course?} as well as this paper is structured as follows: after an introduction to Newtonian dynamics and the concept of differential equations, their
+We are convinced that in order to excite students for numerical simulations it is compulsory to demonstrate their power on an example that catches their interest.
+This purpose is served perfectly by the case chosen here, since Einsteins equations of General Relativity are fascinating to a very broad public.
+Their detailed study needs a deep knowledge of Differential Geometry.
+For the study outlined here, however, very little math is necessary, such that high school students familiar to derivatives benefit from it.
+This was already demonstrated in the "Sch\"ulerakademie Teilchenphysik", where this course was tested successfully on two groups consisting in total of 24 German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade in 2015 and 2017.
+
+The course as well as this paper is structured as follows: after an introduction to Newtonian dynamics and the concept of differential equations, their
 discretization is discussed based on Newton's law of gravitation and possible extensions thereof.
@@
 
-We are convinced that in order to excite students for numerical simulations it is compulsory to demonstrate their power on an example that catches their interest.
-This purpose is served perfectly by the case chosen here, since Einsteins equations of General Relativity are fascinating to a very broad public.
-Their detailed study needs a deep knowledge of Differential Geometry.
-For the study outlined here, however, very little math is necessary, such that high school students from about 10$^{\rm th}$ grade up should benefit from it.
-This was already demonstrated in the "Sch\"ulerakademie Teilchenphysik", where this course was tested successfully on two groups consisting in total of 24 German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade in 2015 and 2017\todo{do foreigners know what that means?
-what are typical ages of students?
-did grade 13 even exist in 2015?}.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-A physical system is said to be understood if the assumed force acting on it leads to the observed trajectories.
+A physical system is said to be understood if the assumed forces acting on it lead to the observed trajectories.
 In other words, we need to show that we can calculate the location in space of the object of interest at any point in time, once the initial conditions are fixed properly.
@@
 
-The observed perihelion motion of Mercury is nowadays determined as \todo{reference?}
-\[(574.10\pm 0.65)'' \ \mbox{per 100 earth years}\]
+The observed perihelion motion of Mercury is nowadays determined as
+\[(574.10\pm 0.65)'' \ \mbox{per 100 earth years \cite{RevModPhys.19.361}}\]
 where the symbol $''$ denotes "arc seconds": 1$''={(1/3600)}^o$. The bulk of this number
@@
 $(1+\alpha(r_S/r))$, where $\alpha$ denotes some dimensionless parameter.
-This seems like a natural way to parametrize the additional potential because it adds the smallest possible deviation from a $1/r$ potential expressed relative to the characteristic length $r_S$.
+This seems like a natural way to parametrize the additional potential because it adds the smallest possible deviation from an $1/r$ potential expressed relative to the characteristic length $r_S$.
 However, besides $r_S$, which characterizes the gravitational potential of the Sun, there is also a characteristic parameter for the dynamics of Mercury:
@@
 
-In this section we present the numerical implementation as well as the visualisation of planetary trajectories and in
+In this section we present the numerical implementation as well as the visualization of planetary trajectories and in
 particular the perihelion motion of Mercury.
@@
 No prior knowledge of \python{} or any other programming language is required, as we explain all necessary steps to create the simulation.
-In~\ref{appendix:python} we give instructions for setting up \python{} on different operating systems.
-We also provide a working example program that can be used as a template \url{this-upload-url}\todo{@arXiv: online link}.
+We provide online instructions for setting up \python{} and \vpython{} on different operating systems \cite{scripts}.
+In \ref{sec:code}, we also provide a working example as well as possible extensions and template files online.
 
@@
 If the length of the \texttt{last} vector is smaller than that of the \texttt{before\_last} vector and smaller than the current vector, one has passed the perihelion and its position is given by \code{vec\_r\_last}.
-It is useful to save the position of the perihelion in a list \code{list\_perih} for a certain number of turns (here e.g.\ \code{max\_turns}=10).
+It is useful to save the position of the perihelion in a list \code{list\_perih} for a certain number of turns (here e.g.\ \code{max\_turns = 10}).
 The implementation can be realized as follows\footnote{%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section*{Acknowledgments}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Instructions to install VPython}\label{appendix:python}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-In short, to run the above scripts, one needs to install \python{} and \vpython{}.
-We recommend the newest \vpython{} version 7 or higher which best runs with the \python{} versions of 3.5 or higher.
-It is also possible to use \python{} 2.7 or higher, but the 3.x versions have better \vpython{} support.
-
-For the advanced users, one simply has to update \python{} to these versions and install \vpython{} via the package manager of choice, e.g, \texttt{pip} (see also \cite{VPython}).
-
-Below, we provide detailed descriptions how to install these versions of \python{} and \vpython{} for different operating systems.
-Independent of the operating systems, one has to execute the following steps:
-\begin{enumerate}
-	\item install and update \python{} (at best 3.x, x$\geq5$);
-	\item install and update a \python{} package manager;
-	\item install \vpython{} and dependencies.
-\end{enumerate}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Linux based systems (including macOS)}\label{appendix:python-max}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-All Linux based system come with an installation of \python{}.
-Thus, if not already at the newest version, one just has to update existing installations.
-This is best done by using a distribution manager.
-Native to Linux based systems is for example \texttt{dpkg} and \texttt{apt-get}, or \texttt{homebrew} and \texttt{MacPorts} for macOS.
-While \texttt{dpkg} and \texttt{apt-get} usually come with the linux installations, macOS user must install such distribution managers at first.
-For this we refer to \cite{MacPorts} or \cite{Homebrew}.
-
-As the start one should open a terminal/console (bash).
-Depending on the installation, one usually needs superuser rights in order to update \python{}.
-Therefore one has to prepend \code{sudo} to the following commands and confirm them by entering the superuser password.
-The installations using a distribution manager are save and tested by the vast community.
-In the following, the successive order of commands to update \python{} is presented for \texttt{apt-get} (Linux) and \texttt{MacPorts} (macOS), but are most similar for other distribution managers.
-\begin{table}[h!]
-\centering
-\footnotesize
-\begin{tabular}{l || l | l}
-	& \texttt{apt-get} & \texttt{MacPorts} \\\hline
-	1. Update the distribution manager & \code{sudo apt-get update} & \code{sudo port selfupdate} \\
-	2. Install \& update \python{} & \code{sudo apt-get install python3} & \code{sudo port install python36} \\
-	3. Installing \python{} package manager & \code{sudo apt-get install python3-pip} & \code{sudo port install py36-pip}
-\end{tabular}
-\normalsize
-\end{table}
-
-Depending on the previously installed packages, this command might install further dependencies automatically (after confirmation).
-One should test if the installation was successful by typing \code{python3 --version} or \code{python3.6 --version} (depending on the installation name) which prints out the version of the \python{} installation (and should be larger than 3.5).
-Next one must update the package manager to get the newest versions of \python{} packages.
-This is done by running
-\begin{table}[h!]
-\centering
-\footnotesize
-\begin{tabular}{l || l | l}
-	& \texttt{apt-get} & \texttt{MacPorts} \\\hline
-	4. Update \code{pip} & \code{pip3 install --upgrade --user pip wheel} & \code{pip-3.6 install --upgrade --user pip wheel} \\
-\end{tabular}
-\normalsize
-\end{table}\\
-Note that this just upgrades the packages for the current user.
-If one wants the systemwide installation, one again has to prepend \code{sudo} and can ignore the \code{--user} flag (\code{sudo pip3 install --upgrade pip wheel}).
-Furthermore, depending on the installed version of \code{pip}, it could have a different name.
-This is best figured out by typing \code{pip} and pressing \texttt{tab}.
-
-Finally, one can install \vpython{} by running
-\begin{table}[h!]
-\centering
-\footnotesize
-\begin{tabular}{l || l | l}
-	& \texttt{apt-get} & \texttt{MacPorts} \\\hline
-	5. Install \vpython{} & \code{pip3 install --user vpython} & \code{pip-3.6 install --user vpython} \\
-\end{tabular}
-\normalsize
-\end{table}
-
-At this point, everything is installed and one can start the simulation.
-To test the \vpython{} installation one should run \code{python3} or \code{python3.6}, which starts a \python{} console.
-Next, one shall run \code{from vpython import *}, which loads the \vpython{} package and opens the default browser.
-Running \code{box()} should generate a white box in the browser window, which can be rotated using the mouse.
-The program is quit by running \code{quit()} or pressing \texttt{Ctrl-c}.
-
-In general, there are multiple ways to run \python{} scripts.
-For the sake of this paper, we suggest two scenarios.
-\begin{enumerate}
-\item The first scenario to run \python{} scripts is that one writes and saves a \python{} file, e.g., \texttt{test.py}, which can be edited using any text editor of choice.
-	To run the \python{} script one must navigate the terminal to the containing folder using the \code{cd} (and \code{ls}) command.
-	While \code{ls} prints out the folders and files in the current directory, \code{cd} changes the directory.
-	The command \code{cd} either takes relative paths (directories which are in the current directory), e.g., \code{cd folder}, or absolute paths beginning with a \code{/}, e.g., \code{cd /Users/name/Desktop/}.
-	Once one has navigated to folder which contains the \python{} file --- e.g., the file name is listed when typing \code{ls} --- than one can type \code{python3 test.py} or \code{python3.6 test.py} to execute the script.
-	Again, this script can be ended (if it does not end automatically) by pressing \texttt{Ctrl-c}.
-\item The second scenario to run \python{} scripts is given by the interactive \texttt{ipython} or \texttt{jupyter} notebooks.
-	This feature is automatically installed when installing the newest version of \vpython{}.
-	To start an interactive session, one has to run \code{jupyter notebook} (or equivalent, depending on the installation name.
-Again, this can be figured out by typing \code{jupyter} and pressing \texttt{tab}).
-	This will print a link (containing the name \texttt{localhost}) and open the browser at the link-address.
-	Within this browser window, one can freely navigate to the directory of choice and create interactive \python{} and \vpython{} sessions by clicking on existing files or creating new ones (upper right button \texttt{new}$\rightarrow$\vpython{}).
-\end{enumerate}
-We provide example files for both scenarios.\todo{@arXiv: online link}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Windows}\label{appendix:python-windows}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-As the first step, one must install \python{}, again there are several possibilities but we suggest \texttt{Anaconda} for an intuitive installation.
-Following the link \cite{conda}, one has two possible (free) options, \texttt{Anaconda} and \texttt{Miniconda}.
-Both are equally sufficient.
-One can install further updates after the initial installation.
-In the case of \texttt{Miniconda} \cite{Miniconda} one shall download the newest \python{} 3.x (x $\geq5$) installer for windows and the right system type (32-bit or 64-bit).
-The system type can be found by pressing \texttt{Windows+I}, navigating to \texttt{System}$\rightarrow$\texttt{About} and looking at the \texttt{System type} entry.
-
-Installing one of the two executables creates a new \texttt{Windows} application called \texttt{Anaconda Prompt}.
-Starting this application opens a terminal which will be used to install the required \python{} packages.
-At first, one should test if the \texttt{Anaconda} installation was successful by running the command \code{conda list}.
-This command should print a list of already installed packages.
-Also one should run \code{conda update conda} to make sure one has the most recent installation.
-
-To install \vpython{} one simply has to run \code{conda install -c vpython vpython}.
-The additional flag \code{-c vpython} tells \texttt{Anaconda} the "channel" where to look for the \vpython{} package.
-Running this command prints out a list of dependencies which will be automatically installed in order for \vpython{} to work --- this must be confirmed.
-
-In principle this is already sufficient to run \vpython{} scripts.
-One could furthermore install \texttt{Anaconda}s native navigator \texttt{anaconda-navigator} by running \code{conda install anaconda-navigator}.
-This will install another application, \texttt{Anaconda Navigator}, which can be run from the Desktop and makes it easier to program in \python{}.
-
-Again there are several possibilities to run \python{} files on \texttt{Windows}.
-\begin{itemize}
-	\item One could create a file \texttt{test.py}, edit it with the editor of choice and specify that this file should be opened with \texttt{Python3}.
-		To do so, \texttt{right click} on  the file, select \texttt{open with} and locate the \texttt{Anaconda Python3} installation.
-		Programs can be closed by closing the running window.
-	\item One can run \code{python3} in the \texttt{Anaconda Prompt} application to start an interactive session.
-		Programs can be closed by closing the running window.
-	\item Start the \texttt{Anaconda Navigator} application, open an interactive \texttt{qtconsole} and program within the newly opened terminal.
-		Programs can be closed by closing the running window.
-	\item Start the \texttt{Anaconda Navigator} application and click on the \texttt{jupyter notebook} button.
-		This opens an interactive \python{} session in the default web-browser.
-		Here, one can open existing notebooks or create new files (upper right button \texttt{new}$\rightarrow$\vpython{}).
-\end{itemize}
-
-To start and test the \vpython{} installation one should pick any out of these options and execute the code
-\begin{lstlisting}
-from vpython import *
-box()
-\end{lstlisting}
-Running this code should open the default browser and render a white box on black background.
-Using the mouse (right click and drag), one should be able to rotate the box.
-It was found that the \texttt{Internet Explorer} is not able to properly render the animation.
-Here we suggest to use the \texttt{Firefox} web-browser as the default choice.
-
-Once this test case works, one can start to test the Mercury orbit simulation.
-We provide example files for both \texttt{juypter notebooks} and regular \python{} files.
-
-\todo{@arXiv: online link}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \clearpage
@@
 Below we provide the complete code as developed above.
-\lstinputlisting[language=Python]{mercury_orbit_base.py}
-Further examples and template files can be found online\cite{}.\todo[inline]{@arXiv: online link}
+\lstinputlisting[language=Python,basicstyle=\scriptsize]{mercury_orbit_base.py}
+Further examples and template files can be found online \cite{scripts}.
 

### 2018-01-15 tex/MercuryNumerical.tex
code style bugfix

@@
 Below we provide the complete code as developed above.
-\lstinputlisting[language=Python,basicstyle=\scriptsize]{mercury_orbit_base.py}
+\lstinputlisting[language=Python,basicstyle=\linespread{1}\scriptsize\ttfamily]{mercury_orbit_base.py}
 Further examples and template files can be found online \cite{scripts}.

### 2018-01-15 tex/MercuryNumerical.tex
Updated simulation values
Update the simulation parameters and interpolation range in order to get closer to the numerical result by GR. I have al
so updated the example script, plots and added a short description of what I have been doing. This must still be complet
ed. Furthermore, the publication scripts must be updated and a summary must be written.

@@
 \end{equation}
-where $G_N=6.67\times 10^{-11}$ m$^3$kg$^{-1}$s$^{-2}$ is the Newtonian constant of gravitation, $m$ is the mass of Mercury and $M_\odot=2\times 10^{30}$ kg is the mass of the Sun.
+where $G_N=6.67\times 10^{-11}$ m$^3$kg$^{-1}$s$^{-2}$ is the Newtonian constant of gravitation, $m$ is the mass of Mercury and $M_\odot=1.99\times 10^{30}$ kg is the mass of the Sun.
 In addition $r=|\vec r(t)|$ denotes the distance between Sun and Mercury when we assume that the Sun is infinitely heavier than Mercury and located at the center of the coordinate system.
@@
 \begin{equation}
-r_S=\frac{2G_N  M_\odot}{c^2} = 3 \ \mbox{km} \ , \label{rsdef}
+r_S=\frac{2G_N  M_\odot}{c^2} = 2.95 \ \mbox{km} \ , \label{rsdef}
 \end{equation}
-where $c=3\times 10^8$ m/s denotes the speed of light.
+where $c=3.00\times 10^8$ m/s denotes the speed of light.
 Note that $r_S$ is the characteristic length scale of the gravitational field of the Sun for --- up to the prefactor --- one can not form another quantity with dimensions of a length from $G_N$, $M_\odot$ and $c$.
@@
 For the system at hand, $r_a^2$ may be estimated from the parameters of Mercury at its perihelion: The corresponding velocity is
- $\dot{\vec r}(t=0)=\abs{\vec v_M(0)} = 59$~km/s (here we already indicate that we will start the simulation at the perihelion) and the closest distance between Sun and Mercury is
-$ r_{MS}=\abs{\vec r_{MS}(0)} = 46 \cdot 10^6$km~\cite{MercuryFactSheet}. Note that at the perihelion $\vec r$ and $\dot{\vec r}$ are perpendicular to each other.
-We thus find that $(r_a^2/ r_{MS}^2) = 4\cdot10^{-8}$. Furthermore $(r_S/r_{MS})=6\cdot 10^{-8}$.
+ $\dot{\vec r}(t=0)=\abs{\vec v_M(0)} = 59.0$~km/s (here we already indicate that we will start the simulation at the perihelion) and the closest distance between Sun and Mercury is
+$ r_{MS}=\abs{\vec r_{MS}(0)} = 46.0 \cdot 10^6$km~\cite{MercuryFactSheet}. Note that at the perihelion $\vec r$ and $\dot{\vec r}$ are perpendicular to each other.
+We thus find that $(r_a^2/ r_{MS}^2) \sim 4\cdot10^{-8}$. Furthermore $(r_S/r_{MS})\sim6\cdot 10^{-8}$.
 Therefore (for a more detailed discussion about the underlying logic we refer to Sec.~\ref{sec:analysis}) one may expect from both correction terms an effect of similar size.
@@
 Both for the numerical treatment and the intuitive understanding, it is useful to select parameters in a "natural range", e.g.,
-by expressing distances in $R_0 = 10^{10}\,\mathrm{m}$ and time intervals in days, $T_0 = 1\,\mathrm{d}$, where d refers to earth days. One Mercury year is given by $T_M=88\,T_0$.
+by expressing distances in $R_0 = 10^{10}\,\mathrm{m}$ and time intervals in days, $T_0 = 1\,\mathrm{d}$, where d refers to earth days. One Mercury year is given by $T_M=88.0\,T_0$.
 With this choice, the initial distance of Mercury to the Sun, the size of the initial velocity of Mercury and the acceleration prefactor (see Eq.~(\ref{eq:newton_art})) become
 \begin{eqnarray}
-r_{MS}(0)   &=& 4.6 R_0 \, , \quad
-v_{M}(0)    = 0.51 \frac{R_0}{T_0} \, ,  \\
-a_{MS}(r_{MS}) &=& \frac{c^2}{2}\frac{r_S}{r^2} = 1.01 \frac{R_0}{T_0^2} \frac{1}{{\left(r_{MS}/R_0\right)}^2}\, .
+r_{MS}(0)   &=& 4.60 R_0 \, , \quad
+v_{M}(0)    = 0.510 \frac{R_0}{T_0} \, ,  \\
+a_{MS}(r_{MS}) &=& \frac{c^2}{2}\frac{r_S}{r^2} = 0.990 \frac{R_0}{T_0^2} \frac{1}{{\left(r_{MS}/R_0\right)}^2}\, .
 \end{eqnarray}
@@
 # Definition of parameters
-rM0 = 4.6    # in units of R0
-vM0 = 0.51   # in units of R0/T0
-c_a = 1.01   # in units of R0**3/T0**2
-TM  = 88.    # in units of T0
-rS  = 3.0e-7 # in units of R0
-ra2 = 8.2e-7 # in units of R0**2
+rM0 = 4.60    # initial radius of Mercury orbit, in units of R0
+vM0 = 5.10e-1 # initial orbital speed of Mercury, in units of R0/T0
+c_a = 9.90e-1 # base acceleration of Mercury, in units of R0**3/T0**2
+TM  = 8.80e+1 # Orbit period of Mercury
+rS  = 2.95e-7 # Schwarzschild radius of Sun,in units of R0
+ra2 = 8.19e-7 # specific angular momentum, in units of R0**2
 \end{lstlisting}
@@
 		\end{itemize}
-		Here, the time steps are defined by $\Delta t_0 \equiv 2 v_{M}(0) / c_a$ and the images are screenshots of the simulation (with modified colors).
+		Here, the time steps are defined by $\Delta t_0 \equiv 2 v_{M}(0) /a_M(0)$ and the images are screenshots of the simulation (with modified colors).
 		Note that for different starting values of $r_{MS}(0)$, also the numbers of days for a "Mercury year" changes.
@@
 	\end{subfigure}
-	\caption{\label{fig:AlphaAngle} Linear relation between $\alpha$ (a), $\beta$ (b) and the perihelion motion $\delta \Theta$ for $\Delta t=2v_M(0)/c_a/100$ and the other parameter set to zero.}
+	\caption{\label{fig:AlphaAngle} Linear relation between $\alpha$ (a), $\beta$ (b) and the perihelion motion $\delta \Theta$ for $\Delta t=2v_M(0)/a_M(0)/200$ and the other parameter set to zero.}
 \end{figure}
@@
 As an example we  extract the perihelion motion using the parameters provided by General Relativity, namely $\alpha = 0$ and $\beta = 3$.
-Considering one result from the numeric simulation, e.g., $\delta\Theta(\alpha = 0, \beta=10^6) \approx 10.7^{\degree}$, the sought-after angle can be calculated by a linear interpolation as
+Considering one result from the numeric simulation, e.g., $\delta\Theta(\alpha = 0, \beta=10^5) = 0.964^{\degree}$, the sought-after angle can be calculated by a linear interpolation as
 \begin{equation}
-	\delta\Theta (\alpha=0, \beta=3) = \frac{10.7^{\degree}}{10^6} \times 3 = 0.116''
+	\delta\Theta (\alpha=0, \beta=3) = \frac{0.964^{\degree}}{10^5} \times 3 = 0.104''
 	\, .
@@
 Depending on the knowledge of the students, multiple points and also estimated numerical uncertainties could be used to extract this value by a linear regression.
+Here, one has to find a proper range for the interpolation.
+Small values for $\alpha$ and $\beta$ usually come with relatively larger numerical errors while for larger values one cannot guarantee that the additional terms are sufficiently small.
+For this reason, we suggest to interpolate between zero and $\alpha_{\max}$, $\beta_{\max} \sim 10^5$.
+This will be further explained in the next section.
 Note that, when comparing the results of the simulation with the experimental result, one Earth year corresponds to $T_E/T_M \approx 365d/88d\approx 4.15$ Mercury years.
-Therefore, the perihelion motion per 100 earth years is $0.116''\cdot 415\approx 48.2''$.
+Therefore, the perihelion motion per 100 earth years is $0.104''\cdot 415\approx 43.2''$.
 This is close to the experimental value of $\delta\Theta_{\mathrm{exp}} = 43.0''$.
@@
 	\caption{\label{fcc3}Motion of perihelion $\delta\Theta$ in degrees depending on the number of turns $N_t$ for different time steps $\Delta t$ (color) and different values for $\beta$ (columns).
-	Here, $\Delta t_0 \equiv 2 v_M(0)/c_a$.
+	Here, $\Delta t_0 \equiv 2 v_M(0)/a_M(0)$.
 	Offsets of $\delta \Theta(\alpha=0, \beta=0)$ from zero can be used as an estimated of the magnitude of the error. Oscillations in $\delta \Theta$ indicate too coarse time steps.
@@
 It is advantageous to set both $\alpha$ and $\beta$ to zero for this study, since in this case the trajectories should be closed and deviations from the correct case can be spotted easily.
-Figure~\ref{fig:MercuryOrbit-a0-small-dt-large} shows the Mercury trajectory for $\Delta t=2 v_M(0) / c_a \times 2$.
+Figure~\ref{fig:MercuryOrbit-a0-small-dt-large} shows the Mercury trajectory for $\Delta t=2 v_M(0) /a_M(0) \times 2$.
 One can clearly see the failure of the simulation to reproduce the physical trajectory.
@@
 (clearly for $\Delta t\to 0$ the computing time goes to infinity).
+
+\todo[inline]{Explain fit range. From here until the next section is not complete yet!}
+\begin{figure}[htb]
+	\centering
+	\includegraphics[width=.99\textwidth]{figs/computation-precision.pdf}
+	\caption{\label{fcc4}Difference of numerically extracted perihelion angle relative to value computed from General Relativity for different time steps.
+Uncertainty bands correspond to equation \ref{eq:num-uncertainty}.
+}
+\end{figure}
+
+The total numerical uncertainty for the perihelion growth $\Delta \delta \theta (\alpha, \beta)$ at non-zero $\alpha$ and $\beta$ is estimated by the offset of the growth at $\alpha = 0 = \beta$, the oscillation of the growth at $\alpha = 0 = \beta$ as well as the oscillation of the growth at the computed values of $\alpha$ and $\beta$
+\begin{eqnarray}\label{eq:num-uncertainty}
+	\Delta \delta \Theta (\alpha, \beta) &=& \sqrt{
+		\delta \Theta_{std}^2 (0, 0) + \delta \Theta_{avg}^2 (0, 0)+ \delta \Theta_{std}^2 (\alpha, \beta)
+	} \, ,
+	\\
+	\delta \Theta_{avg} (\alpha, \beta)
+	&=& 
+	\frac{1}{N_{\mathrm{turns}}}
+	\sum\limits_{n=1}^{N_{\mathrm{turns}}} \delta \Theta_{n} (\alpha, \beta) \, ,
+	\\
+	\delta \Theta_{std}^2 (\alpha, \beta)
+	&=& 
+	\frac{1}{N_{\mathrm{turns}}}
+	\sum\limits_{n=1}^{N_{\mathrm{turns}}} \left[
+		\delta \Theta_{avg} (\alpha, \beta) - \delta \Theta_{n} (\alpha, \beta)
+	\right]^2 \, .
+\end{eqnarray}
+Thus, the extract numerical value is given by
+\begin{equation}
+	\delta \Theta (\alpha, \beta) = \delta \Theta_{avg} (\alpha, \beta)  \pm \Delta \delta \Theta (\alpha, \beta)
+	\, .
+\end{equation}
+
+The prediction coming from General Relativity for zero $\alpha$ is given by
+\begin{equation}
+	\delta \Theta_{GR} (\alpha=0, \beta) = 
+	\frac{1}{4} \frac{r_S^2}{r_a^2} \times \beta
+	+\frac{3}{32} \frac{ r_S^4}{ r_a^4} \times \beta ^2
+	+O\left(\frac{r_S}{r_a}\right)^6
+	\, .
+\end{equation}
 
@@
 
+\todo[inline]{Summary}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2018-01-15 tex/mercury_orbit_base.py

@@ -2,12 +2,12 @@
 from vpython import *
 
 # Definition of parameters
-rM0 = 4.6    # in units of R0
-vM0 = 0.51   # in units of R0/T0
-c_a = 1.01   # in units of R0**3/T0**2
-TM  = 88.    # in units of T0
-rS  = 3.0e-7 # in units of R0
-ra2 = 8.2e-7 # in units of R0**2
+rM0 = 4.60    # initial radius of Mercury orbit, in units of R0
+vM0 = 5.10e-1 # initial orbital speed of Mercury, in units of R0/T0
+c_a = 9.90e-1 # base acceleration of Mercury, in units of R0**3/T0**2
+TM  = 8.80e+1    # Orbit period of Mercury
+rS  = 2.95e-7 # Schwarzschild radius of Sun,in units of R0
+ra2 = 8.19e-7 # specific angular momentum, in units of R0**2
 
 # Initialize distance and velocity vectors
 vec_rM0 = vector(0, rM0, 0)

### 2018-01-16 tex/MercuryNumerical.tex
Updated uncertainty estimate section

@@
 
-\todo[inline]{Explain fit range. From here until the next section is not complete yet!}
+As a first estimate, one can approximate the numerical uncertainty of the the perihelion growth $\Delta \delta \Theta (\alpha, \beta)$ at non-zero $\alpha$ and $\beta$ by the offset of the growth at $\alpha = 0 = \beta$, the oscillation of the growth at $\alpha = 0 = \beta$ as well as the oscillation of the growth at the computed values for non-zero $\alpha$ and $\beta$
+\begin{eqnarray}\label{eq:num-uncertainty}
+	\Delta \delta \Theta (\alpha, \beta) &=& \sqrt{
+		\delta \Theta_{mean}^2 (0, 0) + 
+		\delta \Theta_{std}^2 (0, 0) + 
+		\delta \Theta_{std}^2 (\alpha, \beta)
+	} \, ,
+	\\
+	\delta \Theta_{mean} (\alpha, \beta)
+	&=& 
+	\frac{1}{N_{t}}
+	\sum\limits_{n=1}^{N_{t}} \delta \Theta_{n} (\alpha, \beta) \, ,
+	\\
+	\delta \Theta_{std}^2 (\alpha, \beta)
+	&=& 
+	\frac{1}{N_{t}}
+	\sum\limits_{n=1}^{N_{t}} \left[
+		\delta \Theta_{mean} (\alpha, \beta) - \delta \Theta_{n} (\alpha, \beta)
+	\right]^2 \, ,
+\end{eqnarray}
+where the standard deviations and mean values are obtained from the data sample of $N_t$ turns (orbits of mercury).
+Thus, the extract numerical value is given by
+\begin{equation}\label{eq:num-vals}
+	\delta \Theta (\alpha, \beta) = \delta \Theta_{mean} (\alpha, \beta)  \pm \Delta \delta \Theta (\alpha, \beta)
+	\, .
+\end{equation}
+Note that this has to be repeated for each different time step $\Delta t$.
+
+The absolute value of the numerical error mostly depends on the tilmestep $\Delta t$ and is roughly independent on the values of $\alpha$ and $\beta$.
+Therefore, it is more desirable to pick large values for $\alpha$ and $\beta$, because this increases the absolute value of the precision growth and thus decreases the relative numerical error.
+However, there is a competing effect which places upper bounds on the values of $\alpha$ and $\beta$.
+For instance, the prediction for the precision growth coming from General Relativity for zero $\alpha$ is given by
+\begin{equation}\label{eq:gr-prediction}
+	\delta \Theta_{GR} (\alpha=0, \beta) = 
+	2 \pi \left[
+	\frac{\beta}{4} \,\frac{r_S^2}{r_a^2}
+	+\mathcal{O}\left(\beta^2\,\frac{r_S^4}{r_a^4}\right)
+	\right]
+	\, .
+\end{equation}
+Thus, once $\beta$ is sufficiently large, quadratic corrections in $\beta$ become relevant and it is not possible to extract the perihelion growth of Mercury by performing a linear interpolation.
+We display these competing effects in Figure \ref{fcc4}.
+The relative difference between numerically extracted values and the General Relativity prediction for $\delta \Theta$ are plotted against $\beta$ for zero $\alpha$.
+While for smaller $\beta$, the relative numerical uncertainty grows, for larger $\beta \gtrsim 10^5$, the numerical values deviate from the linear prediction.
+This is expected, because the non-linear corrections start to enter the percent regime
+\begin{equation}
+	\frac{\beta^2 \, r_S^4 / r_a^4}{\beta \, r_S^2 / r_a^2} \Bigg|_{\beta = 10^5}
+	\sim
+	1 \%
+	\, .
+\end{equation}
+Identifying a parameter space for reliable and precise computations is a general challenge for numerical simulations.
 \begin{figure}[htb]
@@
 	\includegraphics[width=.99\textwidth]{figs/computation-precision.pdf}
-	\caption{\label{fcc4}Difference of numerically extracted perihelion angle relative to value computed from General Relativity for different time steps.
-Uncertainty bands correspond to equation \ref{eq:num-uncertainty}.
+	\caption{\label{fcc4}Difference of numerically extracted perihelion angle growth relative to value computed from General Relativity [Equation (\ref{eq:gr-prediction})] for different time steps.
+The error bands are the envelope for computed data points where the size of the error is determined according to Equation (\ref{eq:num-vals}).
 }
 \end{figure}
-
-The total numerical uncertainty for the perihelion growth $\Delta \delta \theta (\alpha, \beta)$ at non-zero $\alpha$ and $\beta$ is estimated by the offset of the growth at $\alpha = 0 = \beta$, the oscillation of the growth at $\alpha = 0 = \beta$ as well as the oscillation of the growth at the computed values of $\alpha$ and $\beta$
-\begin{eqnarray}\label{eq:num-uncertainty}
-	\Delta \delta \Theta (\alpha, \beta) &=& \sqrt{
-		\delta \Theta_{std}^2 (0, 0) + \delta \Theta_{avg}^2 (0, 0)+ \delta \Theta_{std}^2 (\alpha, \beta)
-	} \, ,
-	\\
-	\delta \Theta_{avg} (\alpha, \beta)
-	&=& 
-	\frac{1}{N_{\mathrm{turns}}}
-	\sum\limits_{n=1}^{N_{\mathrm{turns}}} \delta \Theta_{n} (\alpha, \beta) \, ,
-	\\
-	\delta \Theta_{std}^2 (\alpha, \beta)
-	&=& 
-	\frac{1}{N_{\mathrm{turns}}}
-	\sum\limits_{n=1}^{N_{\mathrm{turns}}} \left[
-		\delta \Theta_{avg} (\alpha, \beta) - \delta \Theta_{n} (\alpha, \beta)
-	\right]^2 \, .
-\end{eqnarray}
-Thus, the extract numerical value is given by
-\begin{equation}
-	\delta \Theta (\alpha, \beta) = \delta \Theta_{avg} (\alpha, \beta)  \pm \Delta \delta \Theta (\alpha, \beta)
-	\, .
-\end{equation}
-
-The prediction coming from General Relativity for zero $\alpha$ is given by
-\begin{equation}
-	\delta \Theta_{GR} (\alpha=0, \beta) = 
-	\frac{1}{4} \frac{r_S^2}{r_a^2} \times \beta
-	+\frac{3}{32} \frac{ r_S^4}{ r_a^4} \times \beta ^2
-	+O\left(\frac{r_S}{r_a}\right)^6
-	\, .
-\end{equation}
 

### 2018-01-16 tex/mercury_orbit_base.py
typos in example script

@@ -2,12 +2,12 @@
 from vpython import *
 
 # Definition of parameters
-rM0 = 4.60    # initial radius of Mercury orbit, in units of R0
-vM0 = 5.10e-1 # initial orbital speed of Mercury, in units of R0/T0
-c_a = 9.90e-1 # base acceleration of Mercury, in units of R0**3/T0**2
-TM  = 8.80e+1    # Orbit period of Mercury
+rM0 = 4.60    # Initial radius of Mercury orbit, in units of R0
+vM0 = 5.10e-1 # Initial orbital speed of Mercury, in units of R0/T0
+c_a = 9.90e-1 # Base acceleration of Mercury, in units of R0**3/T0**2
+TM  = 8.80e+1 # Orbit period of Mercury
 rS  = 2.95e-7 # Schwarzschild radius of Sun,in units of R0
-ra2 = 8.19e-7 # specific angular momentum, in units of R0**2
+ra2 = 8.19e-7 # Specific angular momentum, in units of R0**2
 
 # Initialize distance and velocity vectors
 vec_rM0 = vector(0, rM0, 0)

### 2018-01-19 tex/MercuryNumerical.tex
added summary (@CH)

@@
 The paper includes details about the development of the code as well as a discussion of the visualization of the results.
-In addition a method is discussed how to estimate the size of the effect a priori. Which allows to double check the results found numerically.
+In addition a method is discussed that allows one to estimate the size of the effect as well as the uncertainty of 
+the approach a priori. At the same time this enables the students to double check the results found numerically.
 \end{abstract}
@@
 This purpose is served perfectly by the case chosen here, since Einsteins equations of General Relativity are fascinating to a very broad public.
-Their detailed study needs a deep knowledge of Differential Geometry.
-For the study outlined here, however, very little math is necessary, such that high school students familiar to derivatives benefit from it.
+Their detailed study needs a deep understanding of Differential Geometry.
+For the course presented here, however, very little math is necessary, such that high school students familiar to derivatives will benefit from it.
 This was already demonstrated in the "Sch\"ulerakademie Teilchenphysik", where this course was tested successfully on two groups consisting in total of 24 German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade in 2015 and 2017.
@@
  tools are developed to extract the relevant quantity from the result of the simulation.
-Finally, the principle of dimensional analysis is presented as a tool not only to estimate some effect but also to cross check whether the results of the simulation are sensible.
+Finally, the principle of dimensional analysis is presented as a tool not only to estimate some effect as well as the 
+expected accuracy of a calculation, but also to cross check whether the results of the simulation are sensible.
+As is pointed out in the chapters below the course can be finished at some canonical points --- in particular we think
+that for most students a detailed study of the uncertainties of the numerical simulations might be too technical.
+However, we put also this discussion into the paper not only for completeness but also to provide an additional 
+challenge to the more advanced students.
 
@@
 
-A physical system is said to be understood if the assumed forces acting on it lead to the observed trajectories.
-In other words, we need to show that we can calculate the location in space of the object of interest at any point in time, once the initial conditions are fixed properly.
+A physical system is said to be understood, if the assumed forces acting on it lead to the observed trajectories.
+In other words, we need to show that we can calculate the location in space of the object of interest at any future point in time, once the initial conditions are fixed properly.
 If we can neglect the finite size of this object (and in particular its orientation in space), the location is parametrized by a single three vector $\vec r(t)$.
-Additionally, we need to be able to calculate the objects velocity $\vec v(t)$ and acceleration $a(t)$ in order to describe and control its dynamics; this will become clear below.
+Additionally, we need to be able to calculate the objects velocity $\vec v(t)$ and acceleration $\vec a(t)$ in order to describe and control its dynamics; this will become clear below.
 The velocity describes a change of location over time which, for some infinitesimally small $\Delta t$, can be written as
@@
 
-It was Newton who observed that if a body is at rest, it will remain at rest, and if it is in motion it will remain in motion at a constant velocity in a straight line, unless it is acted upon by some force --- this is known as Newton's first law.
+It was Newton who observed that, if a body is at rest, it will remain at rest, and if it is in motion it will remain in motion at a constant velocity in a straight line, 
+unless it is acted upon by some force --- this is known as Newton's first law.
 Formulated differently: a force $\vec F$ expresses itself by changing the motion of some object.
@@
 Here, we restrict ourselves to the case relevant for our example, where the force depends on the location only.
-Therefore, as soon as the force $\vec F(\vec r)$ is known for all $\vec r$ one can in principle calculate the trajectory, e.g., solving Eq.~(\ref{eq:newton2}) for $\vec r(t)$.
+Therefore, as soon as the force $\vec F(\vec r)$ is known for all $\vec r$ one can in principle calculate the trajectory by solving Eq.~(\ref{eq:newton2}) for $\vec r(t)$.
 Sometimes this requires some advanced knowledge of mathematics and sometimes no closed form solution exists.
@@
 
-This procedure can only work if $\Delta t$ is sufficiently small.
+This procedure can only work, if $\Delta t$ is sufficiently small.
 What this means depends on the process one studies.
@@
 Clearly, a relation as Eq.~({\ref{eq:check}}) can only provide guidance and can not replace a careful numerical check of the solutions: A valid result has to be insensitive the concrete value chosen for $\Delta t$ --- in particular replacing $\Delta t$ by $\Delta t/2$ should not change the result significantly.
+This issue will
+be discussed below in more detail.
 
@@
 where $G_N=6.67\times 10^{-11}$ m$^3$kg$^{-1}$s$^{-2}$ is the Newtonian constant of gravitation, $m$ is the mass of Mercury and $M_\odot=1.99\times 10^{30}$ kg is the mass of the Sun.
-In addition $r=|\vec r(t)|$ denotes the distance between Sun and Mercury when we assume that the Sun is infinitely heavier than Mercury and located at the center of the coordinate system.
+In addition $r=|\vec r(t)|$ denotes the distance between Sun and Mercury, when we assume that the Sun is infinitely heavier than Mercury and located at the center of the coordinate system.
 Although this is not exact, it is a pretty good approximation because $m/M_\odot\sim 10^{-8}$.
@@
 \begin{equation}
-\delta \Theta_M = (43)'' \ \mbox{in 100 earth years} \label{delT}
+\delta \Theta_M = (42.56\pm 0.94)'' \ \mbox{per 100 earth years \cite{RevModPhys.19.361}} \label{delT}
 \end{equation}
@@
  $\dot{\vec r}(t=0)=\abs{\vec v_M(0)} = 59.0$~km/s (here we already indicate that we will start the simulation at the perihelion) and the closest distance between Sun and Mercury is
-$ r_{MS}=\abs{\vec r_{MS}(0)} = 46.0 \cdot 10^6$km~\cite{MercuryFactSheet}. Note that at the perihelion $\vec r$ and $\dot{\vec r}$ are perpendicular to each other.
+$ r_{MS}=\abs{\vec r_{MS}(0)} = 46.0 \cdot 10^6$~km~\cite{MercuryFactSheet}. Note that at the perihelion $\vec r$ and $\dot{\vec r}$ are perpendicular to each other.
 We thus find that $(r_a^2/ r_{MS}^2) \sim 4\cdot10^{-8}$. Furthermore $(r_S/r_{MS})\sim6\cdot 10^{-8}$.
@@
 should be suppressed by additional seven orders of magnitude and thus can be neglected safely.
-The parameters $\alpha$ and $\beta$ can be extracted from a fit to data --- this will be done below.
+The parameters $\alpha$ and $\beta$ can be extracted from a fit to data.
 However, since we have only a single number to study, the perihelion shift (see Eq.~(\ref{delT})),
@@
 The parameters can also be calculated from the underlying theory one uses. The actual values depend on the specific theory;
-General relativity gives~\cite{Einstein}
+General Relativity gives~\cite{Einstein}
 \begin{equation}
@@
 This is a good opportunity to let the students play with the size of $\alpha$ and get a feeling for its impact on the trajectories. In particular
-they should discover that values of $\alpha$ of the order of $10^6$ are necessary to get a visible effect.
-For a nicely visible perihelion motion it is advisable to not use the correct Mercury values but a more excentric trajectory by choosing, e.g., $r_{MS}(0)=6R_0$ [figure~\ref{fig:MercuryOrbit-a6-big}].
+they should discover that values of $\alpha$ of the order of $10^6$ are necessary to get a nicely visible effect.
+For an even more enhanced perihelion motion it is advisable to not use the correct Mercury values but a more excentric trajectory by choosing, e.g., $r_{MS}(0)=6R_0$ [figure~\ref{fig:MercuryOrbit-a6-big}].
 \begin{figure}[htb]
@@
 
+This finalizes that basic course ---
+for the not so ambitious students this might be a good point to stop. Already up to here the students should have
+learned a lot about numerical physics, but the load of details was hopefully not too high. Moreover they have already the nice
+achievement that they were able to simulate numerically a perihelion motion. The chapters to come are increasingly technical
+and thus address more advanced students.
+
 %------------------------------------------
@@
 Both $\alpha$ and $\beta$ have similar effects on the perihelion motion.
-We recommend setting one of them to zero and varying the other when analyzing the perihelion motion in dependence on the parameters.
+We recommend to set one of them to zero and to vary the other, when analyzing the perihelion motion. 
 In order to calculate the size of the perihelion motion for a modified gravitational force, the students need to
@@
 
-As explained above [and further discussed in chapter~\ref{sec:analysis}] the natural value for $\alpha$ would be of the order of $1$.
+As explained above [and further discussed in chapter~\ref{sec:analysis}] the natural value for $\alpha$ (and $\beta$)  would be of the order of $1$.
 However, if the students use this value for $\alpha$, they will find that the change in the trajectories is close to invisible and the numerical uncertainty is much larger than the result.
-It is therefore more advisable to use the fact that there is a linear dependence between $\alpha$ and $\delta \Theta$ to estimate the size of the perihelion motion\footnote{%
-	Of course for very large values of $\alpha$ this linear relation will eventually break down.
-	We therefore advise to choose $\alpha<7\cdot 10^6$ at which point the perihelion motion exceeds $\delta \Theta=90^{\degree}$.
-}.
-The students can convince themselves of this fact by plotting the angles $\delta \Theta$ they have extracted from their simulation over $\alpha$.
+It is therefore more advisable to use the fact that there is a linear dependence between $\alpha$ and $\delta \Theta$ to estimate the size of the perihelion motion.
+On the other hand as soon as the value of $\alpha$ or $\beta$ get too large the effective ansatz to use only the leading terms 
+in the $(r_s/r)$ and $(r_a/r)^2$ expansion is no longer justified and the method of extraction might get unreliable.
+We therefore recommend for the simulations used to extract the final result for choose the parameters at most of the order of $10^5$ --- this
+value will better justified in chapter~\ref{sec:analysis}].
+
+The students should convince themselves of the linear relation between the angles $\delta \Theta$ and the paramters $\alpha$ or $\beta$.
 This can be done either by hand, \python{} (e.g., with \texttt{matplotlib}~\cite{Matplotlib}), or using another program like, e.g., Excel.
-For instance figure~\ref{fig:AlphaAngle} demonstrates such a plot.
+For instance figure~\ref{fig:AlphaAngle} demonstrates such plots.
 
@@
 This will be further explained in the next section.
-Note that, when comparing the results of the simulation with the experimental result, one Earth year corresponds to $T_E/T_M \approx 365d/88d\approx 4.15$ Mercury years.
+Note that, when comparing the results of the simulation with the experimental result, one Earth year corresponds to $T_E/T_M \approx 365~d/88~d\approx 4.15$ Mercury years.
 Therefore, the perihelion motion per 100 earth years is $0.104''\cdot 415\approx 43.2''$.
-This is close to the experimental value of $\delta\Theta_{\mathrm{exp}} = 43.0''$.
+This is fully consistent with the experimental value of $\delta\Theta_{\mathrm{exp}} = (42.56\pm 0.94)''$ (see Eq.~(\ref{delT})).
+
+This agreement is the second important achievement after which finishing the course appears natural: The students have understood quantitatively the
+perihelion motion of Mercury. The next section that finalizes the numerical investigation is even more technical and should
+be worked on by the most advanced students only.
 
@@
 Measurements as well as numerical simulations in physics should always be accompanied by estimates of the corresponding uncertainties.
-It is important for the students to recognize this fact and to understand that one cause of the uncertainties lies, e.g., in the finite time steps.
+It is important for the students to recognize this fact and to understand what the sources of uncertainties are.
 The students should therefore explore sources and sizes of inaccuracies arising in this simulation at least qualitatively.
@@
 There are many potential sources of uncertainties in a simulation and discussing all of them is be beyond the scope of this work.
-Instead, we focus on the most accessible source, numerical errors due to finite time steps $\Delta t$.
+Instead, we focus on the most accessible source: Numerical errors due to finite time steps $\Delta t$.
 Additional sources of errors include the omission of terms in Eqs.~(\ref{eq:vdef}) and~(\ref{eq:adef}) and the infinite mass approximation of the Sun (i.e.\ keeping the Sun's position fixed).
@@
 As expected, $\delta\Theta$ is approximately constant for sufficiently small time-steps $\Delta t$ and its value depends solely on $\beta$.
-$\delta\Theta$ deviates from the expected value of zero for $\alpha = \beta = 0$.
+However, contrary to the correct result $\delta\Theta$ deviates from zero for $\alpha = \beta = 0$ for all $\Delta t$.
 Using the data points in figure~\ref{fig:AlphaAngle} and extrapolating to $\alpha=\beta=0$ without enforcing $\delta\Theta(0,0) = 0$ leads to a similar offset.
@@
 	Offsets of $\delta \Theta(\alpha=0, \beta=0)$ from zero can be used as an estimated of the magnitude of the error. Oscillations in $\delta \Theta$ indicate too coarse time steps. 
+	Note the change in scale for the different plots.
 }
@@
 Furthermore, one can observe that for large time-steps the values of the perihelion motion oscillate between discrete values at different turns.
-This can be explained as follows: The sample code we present to find the perihelion vectors only finds the position closest to the Sun.
-This is not the true perihelion but only an approximation to it. Sometimes the program finds a point before and sometimes after the perihelion which causes the observed oscillations.
-The precision of this approximation rises as the time-step shrinks.
-Hence, strong visible oscillations in $\delta\Theta$ indicate a too large time-step.
+This can be explained as follows: The sample code we present to find the perihelion vectors only finds the position closest to the Sun amongst
+the discrete set of vectors evaluated at the discretized time steps and labels it as perihelion vector. However,
+this is only an approximation: Sometimes the program finds a point before and sometimes after the perihelion causing the oscillations observed.
+The precision of this approximation improves as the time-step shrinks.
+Hence, strongly visible oscillations in $\delta\Theta$ for a different number of turns indicate that the time-step used is too large.
 
@@
 Figure~\ref{fig:MercuryOrbit-a0-small-dt-large} shows the Mercury trajectory for $\Delta t=2 v_M(0) /a_M(0) \times 2$.
-One can clearly see the failure of the simulation to reproduce the physical trajectory.
+One can clearly see the failure of the simulation to reproduce the physical trajectory shown as the black, solid line.
 From examples like this it should be apparent that the time steps influence the accuracy of the simulation.
@@
 
-As a first estimate, one can approximate the numerical uncertainty of the the perihelion growth $\Delta \delta \Theta (\alpha, \beta)$ at non-zero $\alpha$ and $\beta$ by the offset of the growth at $\alpha = 0 = \beta$, the oscillation of the growth at $\alpha = 0 = \beta$ as well as the oscillation of the growth at the computed values for non-zero $\alpha$ and $\beta$
+As a first estimate, one can approximate the numerical uncertainty of the perihelion motion, $\Delta \delta \Theta (\alpha, \beta)$ at non-zero $\alpha$ and $\beta$, by its offset
+ at $\alpha = 0 = \beta$,  the amplitude of its oscillation at $\alpha = 0 = \beta$ as well as the amplitude of its oscillation at the non-zero values of $\alpha$ and $\beta$
+ used for the actual calculation.
 \begin{eqnarray}\label{eq:num-uncertainty}
@@
 
-The absolute value of the numerical error mostly depends on the tilmestep $\Delta t$ and is roughly independent on the values of $\alpha$ and $\beta$.
-Therefore, it is more desirable to pick large values for $\alpha$ and $\beta$, because this increases the absolute value of the precision growth and thus decreases the relative numerical error.
+The absolute value of the numerical error mostly depends on the time step $\Delta t$ and is roughly independent on the values of $\alpha$ and $\beta$.
+Therefore, it is more desirable to pick large values for $\alpha$ and $\beta$, because this increases the absolute value of the perihelion
+motion and thus decreases the relative numerical error.
 However, there is a competing effect which places upper bounds on the values of $\alpha$ and $\beta$.
@@
 \end{equation}
-Thus, once $\beta$ is sufficiently large, quadratic corrections in $\beta$ become relevant and it is not possible to extract the perihelion growth of Mercury by performing a linear interpolation.
+Thus, for large values of $\beta$, quadratic corrections in $\beta$ become relevant and it is not possible to extract the 
+perihelion growth of Mercury by performing a linear interpolation.
 We display these competing effects in Figure \ref{fcc4}.
 The relative difference between numerically extracted values and the General Relativity prediction for $\delta \Theta$ are plotted against $\beta$ for zero $\alpha$.
-While for smaller $\beta$, the relative numerical uncertainty grows, for larger $\beta \gtrsim 10^5$, the numerical values deviate from the linear prediction.
-This is expected, because the non-linear corrections start to enter the percent regime
-\begin{equation}
-	\frac{\beta^2 \, r_S^4 / r_a^4}{\beta \, r_S^2 / r_a^2} \Bigg|_{\beta = 10^5}
-	\sim
-	1 \%
-	\, .
-\end{equation}
+The value of $\beta$ we recommend for the extraction is of the order of $10^5$ as will be further motivated in the next section.
+As can be nicely seen in the figure: For smaller values of $\beta$, the relative numerical uncertainty grows, while for larger 
+ values $\beta \gtrsim 10^5$, the numerical values deviate from the assumed linear dependence on $\beta$.
 Identifying a parameter space for reliable and precise computations is a general challenge for numerical simulations.
@@
 
-Dimensional analysis is not only a tool that allows one to cross check if the results of some simulation is of the right order of magnitude, it is also very helpful to identify unusual dynamics in some system.
+Dimensional analysis is not only a tool that allows one to cross check, if the results of some simulation is of the right order of magnitude, it is also very helpful to identify unusual dynamics in some system.
 Especially the latter aspect should become clear from the discussion in this section.
@@
 The idea of dimensional analysis is that in a system that can be controlled by expanding the relevant quantities (like the force) in some small parameter(s), the coefficients in the expansion should turn out to be of order one (that means anything between about 0.1 and 10 is fine --- but 0.01 or 100 is irritating); parameters in line with this are called "natural".
-Applied to the problem at hand given by Eq.~(\ref{eq:newton}) this statement implies that from naturalness one would expect the parameters $\alpha$ and $\beta$ of order one when $r$ is about the average distance Mercury-Sun, namely $\bar r=6\times 10^7\,\mathrm{km}$.
-From this one estimates for the expected angular shift per orbit for $\alpha=1$ and $\beta=0$
+Applied to the problem at hand given by Eq.~(\ref{eq:newton}) this statement implies that from naturalness one would predict the parameters $\alpha$ and $\beta$ to be
+of order one.
+One expects that there must be a physics reason for unnatural parameters to appear.
+Employing for the problem at hand the average distance Mercury-Sun $\bar r=6\times 10^7\,\mathrm{km}$, the concept on 
+naturalness allows us to estimate the expected angular shift per orbit for, e.g., $\alpha=1$ and $\beta=0$
 \begin{equation}
@@
 which leads to a shift of about $30^{''}$ and $15^{''}$, respectively, in 100 earth years. These are to be compared to the empirical value of $43^{''}$.
-Thus the amount of perihelion motion of Mercury is indeed in line with expectations, \textit{if} --- and this is an important "if" --- the Newtonian dynamics is simply the leading term of some more general underlying theory.
-In particular, no new scales enter in the correction terms in addition to $r_S$ and $r_a^2$.
+Thus the amount of perihelion motion of Mercury is indeed in line with expectations, \textit{if} --- and this is an important "if" --- the 
+Newtonian dynamics is simply the leading term of some more general underlying theory. 
+ In particular, no new scales enter in the correction terms in addition to $r_S$ and $r_a^2$. This is by itself already an
+ interesting observation.
+ In case of General Relativity $\beta=3$ is a natural value 
+while $\alpha$ vanishes. This pattern is a therefore a non-trivial prediction of General Relativity and one should expect
+that alternative theories of gravitation generate non-vanishing values of $\alpha$.
 
-Thanks to Einstein we indeed know that the conclusion formulated above is correct; the more general underlying theory is General Relativity and Einstein was able to explain the perihelion motion of Mercury from his equations quantitatively.
-On the other hand, had we found the need for a dramatic deviation of the parameters from unity one would have concluded that there is probably some other dynamics that drives this difference.
+Since the dimensional analysis allows one to estimate with little effort a certain effect to be studied in numerical simulation
+one may also use it as a check of the numerical results: If the simulation had produced a result orders of magnitude different
+from the expectations of the dimensional analysis, there must be either a dynamical reason in the physical system that
+is worth to identify --- or the code underlying the simulation has an error.
 
-Indeed, nowadays there are  several of those hierarchy problems in modern physics: e.g., the so called QCD $\Theta$ term, expected to be of natural size, is at present known to be at most $10^{-10}$.
+It is even possible to push the idea of naturalness further to estimate the intrinsic uncertainty of a given study. 
+For the problem at hand we identified the expansion parameters $(r_s/r)$ and $(r_a/r)^2$ both being of order $10^{-7}$.
+Therefore, as long as natural parameters are employed in the simulation one expects corrections to be suppressed by
+seven orders of magnitude since those need to scale as $(r_s/r)^2$, $(r_a/r)^4$, or $(r_s r_a^2/r^3)$.
+In the simulations discussed above we observed that for $\beta=10^6$ the deviation of the result from the expectation
+of the underlying theory is 8\% ($cf$. Fig.~\ref{fcc4}). Even this is in line the estimate just discussed, since 
+for such large values of $\beta$ the effective expansion parameter is $\beta(r_a/r)^2\sim 10$\%. This is the 
+justification for limiting $\beta$ for a reliable extraction of the perihelion motion of Mercury to values below $10^5$.
+
+Please note, however, dimensional analysis only provides an order of magnitude estimate of a given effect. 
+While it can be used to cross check some explicit detailed evaluation it can by no means replace it, if one
+aims at precise results.
+
+We do not want to leave unmentioned that in modern physics the concept of naturalness plays a very important role. 
+It is regarded as a serious problem, when parameters deviate significantly from their natural values. 
+Nowadays there are  several of those hierarchy problems in modern physics: e.g., the so called QCD $\Theta$ term, 
+expected to be of natural size, is at present known to be at most $10^{-10}$.
 This smallness, called the strong CP problem, is so irritating that physicists like S. Weinberg even proposed that there must exist an additional particle, the axion.
@@
 Also, they could independently research the parameters relevant for the system.
-\item \textbf{Why can we work in one plane?}
+\item \textbf{Why can we work in a plane?}
 
@@
 \end{lstlisting}
-By varying $\Delta t$ they can validate, that there is indeed approximately an anti-proportional dependence.
+By varying $\Delta t$ the students can validate that there is indeed approximately an anti-proportional dependence.
 (Note: This only works if the time in the loop is increased by $\Delta t$, so $t=t+\Delta t$.)
@@
 
-\todo[inline]{Summary}
+\section{Summary}
+
+In this work a course is presented that should enable students to understand quantitatively the perihelion motion of Mercury
+by using a numerical simulation. At the same time the active participation in the course teaches the central role of 
+differential equations in theoretical physics, the basic concepts of how to use numerical simulations to find their
+solutions as well as the need to estimate the uncertainties of a given study.
+In addition the concepts of dimensional analysis and naturalness were introduced which not only allow for an
+estimate of a given effect a priori (to cross check the numerical results) but also to estimate the intrinsic uncertainty
+of the result.
+
+The course is structured such that students at different levels can stop after different achievements. The basic
+course contains the set-up of the numerical simulation and its visualization: The students will have observed
+the impacts of different potentials on the trajectories and some basic features of numerical studies.
+The more advanced students may proceed to extract the perihelion motion quantitatively from the 
+parameters provided by General Relativity. And finally the most advanced students may even follow the 
+discussion of the uncertainties of the simulation.
+
+We are convinced that the course presented in this paper is very well suited to teach high school students not
+only the power of numerical simulations but also the beauty of theoretical physics.
 
@@
 \clearpage
-\section{The complete code}\label{sec:code}
+\section{The code covered by the basic course (visualization of the trajectories)}\label{sec:code}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-Below we provide the complete code as developed above.
+Below we provide the complete code for part 1 as developed above.
 \lstinputlisting[language=Python,basicstyle=\linespread{1}\scriptsize\ttfamily]{mercury_orbit_base.py}

### 2018-01-23 tex/MercuryNumerical.tex
Group edditing session applied

@@
 the approach a priori. At the same time this enables the students to double check the results found numerically.
+The course is structured into a basic block and two further refinements which aim at more advanced students.
 \end{abstract}
@@
 
-Numerical simulations play a key role in modern physics for they allow one to tackle theoretical problems not accessible otherwise.
+Numerical simulations play a key role in modern physics because they allow one to tackle some theoretical problems not accessible otherwise.
 This might be the case because there are too many particles participating in the system (as in simulations for weather predictions) or the interactions are too complicated to allow for a systematic, perturbative approach (as in theoretical descriptions of nuclear particles at the fundamental level).
@@
 Their detailed study needs a deep understanding of Differential Geometry.
-For the course presented here, however, very little math is necessary, such that high school students familiar to derivatives will benefit from it.
+For the course presented here, however, very little math is necessary, such that high school students vaguely familiar to vector calculus and derivatives will benefit from it.
 This was already demonstrated in the "Sch\"ulerakademie Teilchenphysik", where this course was tested successfully on two groups consisting in total of 24 German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade in 2015 and 2017.
@@
 discretization is discussed based on Newton's law of gravitation and possible extensions thereof.
-Afterwards the visualization of the resulting trajectories using \vpython{} is introduced and applied to the problem at hand. In addition,
- tools are developed to extract the relevant quantity from the result of the simulation.
-Finally, the principle of dimensional analysis is presented as a tool not only to estimate some effect as well as the 
-expected accuracy of a calculation, but also to cross check whether the results of the simulation are sensible.
+Afterwards the visualization of the resulting trajectories using \vpython{} is introduced and applied to the problem at hand.
+In addition, tools are developed to extract the relevant quantity from the result of the simulation.
+Finally, the principle of dimensional analysis is presented as a tool to estimate the size of the effect studied as well as its expected accuracy.
+Furthermore the method allows one to cross check the results of the simulation.
 As is pointed out in the chapters below the course can be finished at some canonical points --- in particular we think
 that for most students a detailed study of the uncertainties of the numerical simulations might be too technical.
-However, we put also this discussion into the paper not only for completeness but also to provide an additional 
-challenge to the more advanced students.
+However, we include this discussion as well both for completeness and as an additional challenge to the more advanced students.
 
@@
 
-This procedure can only work, if $\Delta t$ is sufficiently small.
-What this means depends on the process one studies.
-One way to estimate whether $\Delta t$ is small enough, is to verify whether the relation
+This procedure can only work if $\Delta t$ is sufficiently small.
+One way to estimate whether $\Delta t$ is small enough is to verify whether the relation
 \begin{equation}
@@
 Eq.~({\ref{eq:check}}) also shows that small (large) time steps are necessary (sufficient), if the force is strong (weak), since the time steps need to be small enough that all changes induced by the force get resolved.
-Clearly, a relation as Eq.~({\ref{eq:check}}) can only provide guidance and can not replace a careful numerical check of the solutions: A valid result has to be insensitive the concrete value chosen for $\Delta t$ --- in particular replacing $\Delta t$ by $\Delta t/2$ should not change the result significantly.
-This issue will
-be discussed below in more detail.
+Clearly, a relation as Eq.~({\ref{eq:check}}) can only provide guidance and can not replace a careful numerical check of the solutions: A valid result has to be insensitive to the concrete value chosen for $\Delta t$ --- in particular replacing $\Delta t$ by $\Delta t/2$ should not change the result significantly.
+This issue will be discussed in more detail below.
 
@@
 In addition $r=|\vec r(t)|$ denotes the distance between Sun and Mercury, when we assume that the Sun is infinitely heavier than Mercury and located at the center of the coordinate system.
-Although this is not exact, it is a pretty good approximation because $m/M_\odot\sim 10^{-8}$.
+Although this is not exact, it is a good approximation because $m/M_\odot\sim 10^{-8}$.
 For later convenience we introduce the Schwarzschild radius of the Sun
@@
 \end{equation}
-remained unexplained, until Einstein quantified the predictions of General Relativity to this particular observable~\cite{Einstein}.
+remained unexplained, until Einstein quantified the predictions of General Relativity to this particular observable \cite{Einstein}
+\begin{equation}
+\delta \Theta_{GR} = (43.03\pm 0.03)'' \ \mbox{per 100 earth years \cite{RevModPhys.19.361}} \label{delT} \, .
+\end{equation}
 
@@
 Based on the discussion presented above we may multiply the right hand side of~(\ref{eq:newton}) with a factor
-$(1+\alpha(r_S/r))$, where $\alpha$ denotes some dimensionless parameter.
-This seems like a natural way to parametrize the additional potential because it adds the smallest possible deviation from an $1/r$ potential expressed relative to the characteristic length $r_S$.
+$(1+\alpha \, \frac{r_S}{r} )$, where $\alpha$ denotes some dimensionless parameter.
+This seems like a natural way to parametrize the additional potential because it adds the smallest possible deviation from a $1/r$ potential expressed relative to the characteristic length $r_S$.
 However, besides $r_S$, which characterizes the gravitational potential of the Sun, there is also a characteristic parameter for the dynamics of Mercury:
 \begin{equation}
-r_a^2 := \frac{\vec L\,^2}{m^2c^2}= \frac{{(\vec r\times \dot{\vec r} \, )}^2}{c^2} \ , \label{a2def}
+r_L^2 := \frac{\vec L\,^2}{m^2c^2}= \frac{{(\vec r\times \dot{\vec r} \, )}^2}{c^2} \ , \label{a2def}
 \end{equation}
 where $\vec L$ denotes the angular momentum, which is a constant of motion for central potentials.
-It is easily verified that $r_a^2$ carries dimensions of length squared.
-We therefore have to add one more term to the potential; in analogy to the one discussed above, we use $\beta (r_a^2/r^2)$.
+It is easily verified that $r_L^2$ carries dimensions of length squared.
+We therefore have to add one more term to the potential; in analogy to the one discussed above, we use $\beta \, \frac{r_L^2}{r^2}$.
 Combining everything, we use the following ansatz for the modified equation of motion:
 \begin{equation}
-\ddot{\vec r} = - \frac{c^2}{2}\frac{r_S}{r^2}\left(1+\alpha\frac{r_S}{r}+\beta\frac{r_a^2}{r^2}\right) \, \frac{\vec{r}}{r} \ .
+\ddot{\vec r} = - \frac{c^2}{2}\frac{r_S}{r^2}\left(1+\alpha\frac{r_S}{r}+\beta\frac{r_L^2}{r^2}\right) \, \frac{\vec{r}}{r} \ .
 \label{eq:newton_art}
 \end{equation}
-For the system at hand, $r_a^2$ may be estimated from the parameters of Mercury at its perihelion: The corresponding velocity is
- $\dot{\vec r}(t=0)=\abs{\vec v_M(0)} = 59.0$~km/s (here we already indicate that we will start the simulation at the perihelion) and the closest distance between Sun and Mercury is
-$ r_{MS}=\abs{\vec r_{MS}(0)} = 46.0 \cdot 10^6$~km~\cite{MercuryFactSheet}. Note that at the perihelion $\vec r$ and $\dot{\vec r}$ are perpendicular to each other.
-We thus find that $(r_a^2/ r_{MS}^2) \sim 4\cdot10^{-8}$. Furthermore $(r_S/r_{MS})\sim6\cdot 10^{-8}$.
-Therefore (for a more detailed discussion about the underlying logic we refer to Sec.~\ref{sec:analysis}) one may expect from both correction terms an effect of similar size.
-Furthermore, any term of higher order in either $a$ or $r_S$
+For the system at hand, $r_L^2$ may be estimated from the parameters of Mercury at its perihelion: The corresponding velocity is
+ $\dot{ r}(t=0)=\abs{\vec v_M(0)} = 59.0$~km/s (here we already indicate that we will start the simulation at the perihelion) and the closest distance between Sun and Mercury is
+$ r_{MS}=\abs{\vec r_{MS}(0)} = 46.0 \cdot 10^6$~km~\cite{MercuryFactSheet}. Note that at the perihelion $\vec r$ and $\dot{\vec r}$ are perpendicular to each other (see Figure \ref{fig:sun_merc}).
+We thus find that $(r_L^2/ r_{MS}^2) \sim 4\cdot10^{-8}$. Furthermore $(r_S/r_{MS})\sim6\cdot 10^{-8}$.
+Therefore one may expect from both correction terms an effect of similar size --- for a more detailed discussion about the underlying logic we refer to Sec.~\ref{sec:analysis}.
+Furthermore, any term of higher order in either $r_L^2$ or $r_S$
 should be suppressed by additional seven orders of magnitude and thus can be neglected safely.
@@
 Since the trajectories do not depend on where on the orbit of Mercury the simulation is started, we
-use the values at the perihelion with $\abs{\vec r_{MS}(0)} = 46 \cdot 10^6\,\mathrm{km}$ and $\abs{\vec v_M(0)} = 59\,\mathrm{km/s}$\cite{MercuryFactSheet} as initial (for $t=0$) parameters.
+use the values at the perihelion with $\abs{\vec r_{MS}(0)} = 46.0 \cdot 10^6\,\mathrm{km}$ and $\abs{\vec v_M(0)} = 59.0\,\mathrm{km/s}$\cite{MercuryFactSheet} as initial (for $t=0$) parameters.
 The computer does not understand physical units, hence one has to express each variable in an appropriate unit.
@@
 v_{M}(0)    = 0.510 \frac{R_0}{T_0} \, ,  \\
-a_{MS}(r_{MS}) &=& \frac{c^2}{2}\frac{r_S}{r^2} = 0.990 \frac{R_0}{T_0^2} \frac{1}{{\left(r_{MS}/R_0\right)}^2}\, .
+a_{MS}(r_{MS}) &=& \frac{c^2}{2}\frac{r_S}{r_{MS}^2} = 0.990 \frac{R_0}{T_0^2} \frac{1}{{\left(r_{MS}/R_0\right)}^2}\, .
 \end{eqnarray}
@@
 # Definition of parameters
-rM0 = 4.60    # initial radius of Mercury orbit, in units of R0
-vM0 = 5.10e-1 # initial orbital speed of Mercury, in units of R0/T0
-c_a = 9.90e-1 # base acceleration of Mercury, in units of R0**3/T0**2
+rM0 = 4.60    # Initial radius of Mercury orbit, in units of R0
+vM0 = 5.10e-1 # Initial orbital speed of Mercury, in units of R0/T0
+c_a = 9.90e-1 # Base acceleration of Mercury, in units of R0**3/T0**2
 TM  = 8.80e+1 # Orbit period of Mercury
 rS  = 2.95e-7 # Schwarzschild radius of Sun,in units of R0
-ra2 = 8.19e-7 # specific angular momentum, in units of R0**2
+rL2 = 8.19e-7 # Specific angular momentum, in units of R0**2
 \end{lstlisting}
 where $\texttt{c\_a} =a_{MS} r_{MS}^2$, the acceleration without the radius factor.
-The last two quantities refer to the Schwarzschild radius of the Sun and the parameter $r_a^2$ defined in Eqs. (\ref{rsdef}) and (\ref{a2def}), respectively.
+The last two quantities refer to the Schwarzschild radius of the Sun and the parameter $r_L^2$ defined in Eqs. (\ref{rsdef}) and (\ref{a2def}), respectively.
 
@@
 # Compute the strength of the acceleration
-temp = 1 + alpha * rS / vec_rM_old.mag + beta * ra2 / vec_rM_old.mag**2
+temp = 1 + alpha * rS / vec_rM_old.mag + beta * rL2 / vec_rM_old.mag**2
 aMS  = c_a * temp / vec_rM_old.mag**2
@@
     # Compute the strength of the acceleration
-    temp = 1 + alpha * rS / vec_rM_old.mag + beta * ra2 / vec_rM_old.mag**2
+    temp = 1 + alpha * rS / vec_rM_old.mag + beta * rL2 / vec_rM_old.mag**2
     aMS  = c_a * temp / vec_rM_old.mag**2
@@
 We place the Sun in the origin of our coordinate system and choose non-realistic radii for visualization purposes.
-Last but not least, one should the vectors related to the visualization directly instead of extra vectors as was done before. Thus
+Last but not least, one should use in the code the vectors directly related to the visualization. Thus
 the \texttt{while}-loop becomes
@@
 
-If the starting values are chosen as advised in the previous section, the students should end up with a trajectory as depicted in figure~\ref{fig:MercuryOrbit-a0-small}.
-At this point one might ask students to vary $\Delta t$ and observe the effect of this on the orbit [see also figure~\ref{fig:MercuryOrbit-a0-small-dt-large}].
-To get the perihelion motion, the additional force term described in Eq.~(\ref{eq:newton_art}) has to be "turned on", e.g., by choosing  $\alpha\neq0$
-[figure~\ref{fig:MercuryOrbit-a6-small}].
-This is a good opportunity to let the students play with the size of $\alpha$ and get a feeling for its impact on the trajectories. In particular
-they should discover that values of $\alpha$ of the order of $10^6$ are necessary to get a nicely visible effect.
-For an even more enhanced perihelion motion it is advisable to not use the correct Mercury values but a more excentric trajectory by choosing, e.g., $r_{MS}(0)=6R_0$ [figure~\ref{fig:MercuryOrbit-a6-big}].
+If the starting values are chosen as advised in the previous section, the students should end up with a trajectory as depicted in Figure~\ref{fig:MercuryOrbit-a0-small}.
+At this point one might ask students to vary $\Delta t$ and observe the effect of this on the orbit [see also Figure~\ref{fig:MercuryOrbit-a0-small-dt-large}].
+To get the perihelion motion, the additional force term described in Eq.~(\ref{eq:newton_art}) has to be "turned on", e.g., by choosing either $\alpha\neq0$ or $\beta\neq0$
+[Figure~\ref{fig:MercuryOrbit-a6-small}].
+This is a good opportunity to let the students play with the size of $\alpha$ or $\beta$ and get a feeling for its impact on the trajectories. In particular
+they should discover that values of $\alpha$ or $\beta$ larger than $10^5$ are necessary to get a visible effect.
+For an even more enhanced perihelion motion it is advisable to not use the correct Mercury values but a more excentric trajectory by choosing, e.g., $r_{MS}(0)=6R_0$ [Figure~\ref{fig:MercuryOrbit-a6-big}].
 \begin{figure}[htb]
@@
 
-This finalizes that basic course ---
+This finalizes the basic course ---
 for the not so ambitious students this might be a good point to stop. Already up to here the students should have
-learned a lot about numerical physics, but the load of details was hopefully not too high. Moreover they have already the nice
-achievement that they were able to simulate numerically a perihelion motion. The chapters to come are increasingly technical
+learned a lot about numerical physics by successfully simulating the perihelion motion. The chapters to come are increasingly technical
 and thus address more advanced students.
@@
 Note that the perihelion motion is computed from the locations stored in \code{list\_perih} and not based on the initial position.
-This is important because depending on the initial conditions and numerical uncertainties, the simulation does not necessarily start in the perihelion.
+This is important because depending on the initial conditions and numerical uncertainties, the simulation does not necessarily start in the exact perihelion.
 
-As explained above [and further discussed in chapter~\ref{sec:analysis}] the natural value for $\alpha$ (and $\beta$)  would be of the order of $1$.
+As explained above [and further discussed in Section~\ref{sec:analysis}] the natural value for $\alpha$ (and $\beta$)  would be of the order of $1$.
 However, if the students use this value for $\alpha$, they will find that the change in the trajectories is close to invisible and the numerical uncertainty is much larger than the result.
 It is therefore more advisable to use the fact that there is a linear dependence between $\alpha$ and $\delta \Theta$ to estimate the size of the perihelion motion.
-On the other hand as soon as the value of $\alpha$ or $\beta$ get too large the effective ansatz to use only the leading terms 
-in the $(r_s/r)$ and $(r_a/r)^2$ expansion is no longer justified and the method of extraction might get unreliable.
-We therefore recommend for the simulations used to extract the final result for choose the parameters at most of the order of $10^5$ --- this
-value will better justified in chapter~\ref{sec:analysis}].
+On the other hand as soon as the values of $\alpha$ or $\beta$ get too large the effective ansatz to use only the leading terms 
+in the $(r_s/r)$ and $(r_L/r)^2$ expansion is no longer justified and the method of extraction might get unreliable.
+At this point we therefore recommend to choose the parameters  $\alpha$ and $\beta$ at most of the order of $10^5$ --- this
+value will be better justified in Section~\ref{sec:analysis}.
 
-The students should convince themselves of the linear relation between the angles $\delta \Theta$ and the paramters $\alpha$ or $\beta$.
+The students should convince themselves that in this parameter range there is a linear relation between the angles $\delta \Theta$ and the parameters $\alpha$ or $\beta$:
+\begin{equation}\label{eq:theta-linear}
+	\delta\Theta (\alpha, \beta) = m_\alpha \cdot \alpha + m_\beta \cdot \beta
+	\, .
+\end{equation}
 This can be done either by hand, \python{} (e.g., with \texttt{matplotlib}~\cite{Matplotlib}), or using another program like, e.g., Excel.
-For instance figure~\ref{fig:AlphaAngle} demonstrates such plots.
+For instance Figure~\ref{fig:AlphaAngle} demonstrates such plots.
 
@@
 
-As an example we  extract the perihelion motion using the parameters provided by General Relativity, namely $\alpha = 0$ and $\beta = 3$.
-Considering one result from the numeric simulation, e.g., $\delta\Theta(\alpha = 0, \beta=10^5) = 0.964^{\degree}$, the sought-after angle can be calculated by a linear interpolation as
+As an example we  extract the perihelion motion using the parameters provided by General Relativity, namely $\alpha = 0$ and $\beta = 3$ evaluated with $\Delta t = \Delta t_0 / 200$.
+Considering one result from the numeric simulation, e.g., $\delta\Theta(\alpha = 0, \beta=10^5) = 0.964^{\degree}$, the sought-after angle can be calculated using Eq.~(\ref{eq:theta-linear})
 \begin{equation}
-	\delta\Theta (\alpha=0, \beta=3) = \frac{0.964^{\degree}}{10^5} \times 3 = 0.104''
+	\delta\Theta (\alpha=0, \beta=3) = \frac{0.964^{\degree}}{10^5} \cdot 3 = 0.104''
 	\, .
@@
 This will be further explained in the next section.
-Note that, when comparing the results of the simulation with the experimental result, one Earth year corresponds to $T_E/T_M \approx 365~d/88~d\approx 4.15$ Mercury years.
+Note that, when comparing the results of the simulation with the experimental result, one Earth year corresponds to $T_E/T_M \approx 365~\mathrm{d}/88~\mathrm{d}\approx 4.15$ Mercury years.
 Therefore, the perihelion motion per 100 earth years is $0.104''\cdot 415\approx 43.2''$.
-This is fully consistent with the experimental value of $\delta\Theta_{\mathrm{exp}} = (42.56\pm 0.94)''$ (see Eq.~(\ref{delT})).
+This is consistent with the observation $\delta\Theta_{\mathrm{M}} = (42.56\pm 0.94)''$ [see Eq.~(\ref{delT})].
 
@@
 
-Consider figure~\ref{fcc3}, where the angle of the perihelion motion $\delta \Theta$ is shown for the first $10$ turns ($x$-axis) for different choices of $\Delta t$ (colors) and $\beta$ (columns).
+Consider Figure~\ref{fcc3}, where the angle of the perihelion motion $\delta \Theta$ is shown for the first $10$ turns ($x$-axis) for different choices of $\Delta t$ (colors) and $\beta$ (columns).
 In all cases $\alpha=0$.
 As expected, $\delta\Theta$ is approximately constant for sufficiently small time-steps $\Delta t$ and its value depends solely on $\beta$.
-However, contrary to the correct result $\delta\Theta$ deviates from zero for $\alpha = \beta = 0$ for all $\Delta t$.
-Using the data points in figure~\ref{fig:AlphaAngle} and extrapolating to $\alpha=\beta=0$ without enforcing $\delta\Theta(0,0) = 0$ leads to a similar offset.
+However, contrary to the correct result $\delta\Theta$ deviates from zero for $\alpha = 0 = \beta$ for all $\Delta t$.
+Using the data points in Figure~\ref{fig:AlphaAngle} and extrapolating to $\alpha = 0 = \beta$ without enforcing $\delta\Theta(0,0) = 0$ leads to a similar offset.
 This offset is a numerical error and can be taken as an estimate for the error of the simulation.
@@
 	\includegraphics[width=.99\textwidth]{figs/angular-variaton.pdf}
-	\caption{\label{fcc3}Motion of perihelion $\delta\Theta$ in degrees depending on the number of turns $N_t$ for different time steps $\Delta t$ (color) and different values for $\beta$ (columns).
+	\caption{\label{fcc3}The motion of the perihelion $\delta\Theta$ in degrees depending on the number of turns $N_t$ for different time steps $\Delta t$ (color) and different values for $\beta$ (columns).
 	Here, $\Delta t_0 \equiv 2 v_M(0)/a_M(0)$.
-	Offsets of $\delta \Theta(\alpha=0, \beta=0)$ from zero can be used as an estimated of the magnitude of the error. Oscillations in $\delta \Theta$ indicate too coarse time steps. 
+	Offsets of $\delta \Theta(\alpha=0, \beta=0)$ from zero can be used as an estimate of the magnitude of the error. Oscillations in $\delta \Theta$ indicate too coarse time steps. 
 	Note the change in scale for the different plots.
@@
 	&=& 
-	\frac{1}{N_{t}}
+	\frac{1}{N_{t}-1}
 	\sum\limits_{n=1}^{N_{t}} \left[
@@
 where the standard deviations and mean values are obtained from the data sample of $N_t$ turns (orbits of mercury).
-Thus, the extract numerical value is given by
+Thus, the numerical value extracted is given by
 \begin{equation}\label{eq:num-vals}
@@
 
-The absolute value of the numerical error mostly depends on the time step $\Delta t$ and is roughly independent on the values of $\alpha$ and $\beta$.
-Therefore, it is more desirable to pick large values for $\alpha$ and $\beta$, because this increases the absolute value of the perihelion
+The absolute value of the numerical error mostly depends on the time step $\Delta t$ and is roughly independent of the values of $\alpha$ and $\beta$.
+Therefore, it is more desirable to pick large values for $\alpha$ and $\beta$, because this increases the absolute size of the perihelion
 motion and thus decreases the relative numerical error.
 However, there is a competing effect which places upper bounds on the values of $\alpha$ and $\beta$.
-For instance, the prediction for the precision growth coming from General Relativity for zero $\alpha$ is given by
+For instance, the prediction for the perihelion motion coming from General Relativity allowing for varying values of $\beta$ is given by
 \begin{equation}\label{eq:gr-prediction}
-	\delta \Theta_{GR} (\alpha=0, \beta) = 
+	\delta \Theta_{GR} (\beta) = 
 	2 \pi \left[
-	\frac{\beta}{4} \,\frac{r_S^2}{r_a^2}
-	+\mathcal{O}\left(\beta^2\,\frac{r_S^4}{r_a^4}\right)
+	\frac{\beta}{4} \,\frac{r_S^2}{r_L^2}
+	+\mathcal{O}\left(\beta^2\,\frac{r_S^4}{r_L^4}\right)
 	\right]
@@
 The value of $\beta$ we recommend for the extraction is of the order of $10^5$ as will be further motivated in the next section.
-As can be nicely seen in the figure: For smaller values of $\beta$, the relative numerical uncertainty grows, while for larger 
- values $\beta \gtrsim 10^5$, the numerical values deviate from the assumed linear dependence on $\beta$.
+As can be seen in the figure: For smaller values of $\beta$, the relative numerical uncertainty grows, while for larger 
+ values, $\beta \gtrsim 10^5$, the numerical values deviate from the assumed linear dependence on $\beta$.
 Identifying a parameter space for reliable and precise computations is a general challenge for numerical simulations.
@@
 	\includegraphics[width=.99\textwidth]{figs/computation-precision.pdf}
-	\caption{\label{fcc4}Difference of numerically extracted perihelion angle growth relative to value computed from General Relativity [Equation (\ref{eq:gr-prediction})] for different time steps.
-The error bands are the envelope for computed data points where the size of the error is determined according to Equation (\ref{eq:num-vals}).
+	\caption{\label{fcc4}Difference of numerically extracted perihelion motion relative to the value computed from General Relativity [Eq.~(\ref{eq:gr-prediction})] for different time steps.
+The error bands are computed according to Eq.~(\ref{eq:num-uncertainty}).
 }
@@
 
-Dimensional analysis is not only a tool that allows one to cross check, if the results of some simulation is of the right order of magnitude, it is also very helpful to identify unusual dynamics in some system.
+Dimensional analysis is not only a tool that allows one to cross check, if the results of simulations are of the right order of magnitude, it is also very helpful to identify unusual dynamics in some systems.
 Especially the latter aspect should become clear from the discussion in this section.
@@
 of order one.
-One expects that there must be a physics reason for unnatural parameters to appear.
-Employing for the problem at hand the average distance Mercury-Sun $\bar r=6\times 10^7\,\mathrm{km}$, the concept on 
+One expects that unnatural parameters point at dynamics not accounted for explicitly.
+Employing for the problem at hand the average distance Mercury-Sun $\bar r_{MS}=6\times 10^7\,\mathrm{km}$, the concept on 
 naturalness allows us to estimate the expected angular shift per orbit for, e.g., $\alpha=1$ and $\beta=0$
 \begin{equation}
-\delta \Theta \simeq 2\pi\left(\frac{r_S}{\bar r}\right) = (\pi \times 10^{-7}) \ \mbox{rad} = (2\cdot 10^{-5})^o = (7\cdot 10^{-2}) ^{''} \ ,
+\delta \Theta \simeq 2\pi\left(\frac{r_S}{\bar r_{MS}}\right) = (\pi \times 10^{-7}) \ \mbox{rad} = (2\cdot 10^{-5})^o = (7\cdot 10^{-2}) ^{''} \ ,
 \end{equation}
@@
 \begin{equation}
-\delta \Theta \simeq 2\pi\left(\frac{a^2}{\bar r^2}\right) = (\pi \times 5 \cdot 10^{-8}) \ \mbox{rad} = (1 \cdot 10^{-5})^o = (4\cdot 10^{-2}) ^{''} \ ,
+\delta \Theta \simeq 2\pi\left(\frac{a^2}{\bar r^2_{MS}}\right) = (\pi \times 5 \cdot 10^{-8}) \ \mbox{rad} = (1 \cdot 10^{-5})^o = (4\cdot 10^{-2}) ^{''} \ ,
 \end{equation}
@@
 Newtonian dynamics is simply the leading term of some more general underlying theory. 
- In particular, no new scales enter in the correction terms in addition to $r_S$ and $r_a^2$. This is by itself already an
+ In particular, no new scales enter in the correction terms in addition to $r_S$ and $r_L^2$. This is by itself already an
  interesting observation.
  In case of General Relativity $\beta=3$ is a natural value 
-while $\alpha$ vanishes. This pattern is a therefore a non-trivial prediction of General Relativity and one should expect
+while $\alpha$ vanishes. This pattern is therefore a non-trivial prediction of General Relativity and one should expect
 that alternative theories of gravitation generate non-vanishing values of $\alpha$.
@@
 one may also use it as a check of the numerical results: If the simulation had produced a result orders of magnitude different
-from the expectations of the dimensional analysis, there must be either a dynamical reason in the physical system that
-is worth to identify --- or the code underlying the simulation has an error.
+from the expectations of the dimensional analysis, there must be a dynamical reason in the physical system that deserves to be identified --- or the code underlying the simulation has an error.
 
 It is even possible to push the idea of naturalness further to estimate the intrinsic uncertainty of a given study. 
-For the problem at hand we identified the expansion parameters $(r_s/r)$ and $(r_a/r)^2$ both being of order $10^{-7}$.
+For the problem at hand we identified the expansion parameters $(r_s/r)$ and $(r_L/r)^2$ both being of order $10^{-7}$.
 Therefore, as long as natural parameters are employed in the simulation one expects corrections to be suppressed by
-seven orders of magnitude since those need to scale as $(r_s/r)^2$, $(r_a/r)^4$, or $(r_s r_a^2/r^3)$.
+seven orders of magnitude since those need to scale as $(r_s/r)^2$, $(r_L/r)^4$, or $(r_s r_L^2/r^3)$.
 In the simulations discussed above we observed that for $\beta=10^6$ the deviation of the result from the expectation
 of the underlying theory is 8\% ($cf$. Fig.~\ref{fcc4}). Even this is in line the estimate just discussed, since 
-for such large values of $\beta$ the effective expansion parameter is $\beta(r_a/r)^2\sim 10$\%. This is the 
+for such large values of $\beta$ the effective expansion parameter is $\beta(r_L/r)^2\sim 10$\%. This is the 
 justification for limiting $\beta$ for a reliable extraction of the perihelion motion of Mercury to values below $10^5$.
@@
 
-\item \textbf{Explore Problem autonomously}
+\item \textbf{Explore problem autonomously}
 
-In chapter~\ref{sec:Numerical Implementation} we suggested to present the material by using a template as well as a step by step instruction to guide the students through the problem solution.
+In Section~\ref{sec:Numerical Implementation} we suggested to present the material by using a template as well as a step by step instruction to guide the students through the problem solution \cite{scripts}.
 These instructions can be cut down or left out depending on available time and numerical/computational versatility of the students.
-This could be achieved by the following changes or additions to the concept presented in chapter \ref{sec:Numerical Implementation}.
+This could be achieved by the following changes or additions to the concept presented in Section~\ref{sec:Numerical Implementation}.
 
@@
 Instead of providing the template to the students, they could build the program from scratch.
-Of course, this requires some basic knowledge in \vpython{}, which they could acquire for example by working through "a ball in a box" (see~\cite{VPythonIntro}) or similar tutorials.
+Of course, this requires some basic knowledge in \vpython{}, which they could acquire for example by working through introductory materials (see~\cite{VPython}).
 Also, they could independently research the parameters relevant for the system.
@@
 (Note: This only works if the time in the loop is increased by $\Delta t$, so $t=t+\Delta t$.)
-\item \textbf{Verlet integration}
+\item \textbf{Verlet algorithm}
 
 Obviously, an improvement in accuracy can be achieved by using a better algorithm without changing $\Delta t$.
-The simplest way to demonstrate this might be given by the implementation of the Verlet (see e.g.~\cite{Hairer03geometricnumerical} and references therein) integration instead of using the simple Euler method.
+The simplest way to demonstrate this might be given by the implementation of the Verlet algorithm (see e.g.~\cite{Hairer03geometricnumerical} and references therein) instead of using the simple Euler method employed here for solving the differential equation.
 \end{itemize}
 
-\item \textbf{Extended Problems}
+\item \textbf{Extended problems}
 
@@
 It might be interesting to abandon the simplification of a stationary Sun, as it nicely illustrates Newton's third law.
-Here it might also be advisable to reduce the mass of the sun to have a more visible result.
+Here it might also be advisable to reduce the mass of the Sun to have a more visible result.
 
@@
 
-In this work a course is presented that should enable students to understand quantitatively the perihelion motion of Mercury
+In this work a course is presented that should enable advanced high school students to understand quantitatively the perihelion motion of Mercury
 by using a numerical simulation. At the same time the active participation in the course teaches the central role of 
@@
 course contains the set-up of the numerical simulation and its visualization: The students will have observed
-the impacts of different potentials on the trajectories and some basic features of numerical studies.
+the impact of different forces on the trajectories and some basic features of numerical studies.
 The more advanced students may proceed to extract the perihelion motion quantitatively from the 

### 2018-01-23 tex/mercury_orbit_base.py

@@ -7,7 +7,7 @@ vM0 = 5.10e-1 # Initial orbital speed of Mercury, in units of R0/T0
 c_a = 9.90e-1 # Base acceleration of Mercury, in units of R0**3/T0**2
 TM  = 8.80e+1 # Orbit period of Mercury
 rS  = 2.95e-7 # Schwarzschild radius of Sun,in units of R0
-ra2 = 8.19e-7 # Specific angular momentum, in units of R0**2
+rL2 = 8.19e-7 # Specific angular momentum, in units of R0**2
 
 # Initialize distance and velocity vectors
 vec_rM0 = vector(0, rM0, 0)
@@ -19,7 +19,7 @@ dt = 2 * vM0 / c_a / 20
 # Define the coordinate and velocity update
 def evolve_mercury(vec_rM_old, vec_vM_old, alpha, beta):
     # Compute the strength of the acceleration
-    temp = 1 + alpha * rS / vec_rM_old.mag + beta * ra2 / vec_rM_old.mag**2
+    temp = 1 + alpha * rS / vec_rM_old.mag + beta * rL2 / vec_rM_old.mag**2
     aMS  = c_a * temp / vec_rM_old.mag**2
     # Multiply by the direction
     vec_aMS = - aMS * ( vec_rM_old / vec_rM_old.mag )

### 2018-01-23 tex/figs/flowchart.tikz.tex
included example flowchart in tikz

\documentclass{article}
\usepackage{tikz}
\usetikzlibrary{shapes.geometric, arrows}
%
\tikzstyle{startstop} = [rectangle, rounded corners, minimum width=3cm, minimum height=1cm,text centered, draw=black, fill=red!30]
\tikzstyle{io} = [trapezium, trapezium left angle=70, trapezium right angle=110, minimum width=3cm, minimum height=1cm, text centered, draw=black, fill=blue!30]
\tikzstyle{process} = [rectangle, minimum width=3cm, minimum height=1cm, text centered, draw=black, fill=orange!30]
\tikzstyle{decision} = [diamond, minimum width=3cm, minimum height=1cm, text centered, draw=black, fill=green!30]
\tikzstyle{arrow} = [thick,->,>=stealth]
%
\begin{document}
%
\begin{tikzpicture}[node distance=2cm]
  \node (start) [startstop] {Start};
  \node (in1) [io, below of=start] {Input};
  \node (pro1) [process, below of=in1] {Process 1};
  \node (dec1) [decision, below of=pro1] {Decision 1};
  \draw [arrow] (start) -- (in1);
  \draw [arrow] (in1) -- (pro1);
  \draw [arrow] (pro1) -- (dec1);
\end{tikzpicture}
\end{document}

### 2018-01-30 tex/MercuryNumerical.tex
Added python and vpython version number to section 4
Added footnote for prior vpython 7.0 import statement

@@
 No prior knowledge of \python{} or any other programming language is required, as we explain all necessary steps to create the simulation.
+In the following, we present code examples which tested for \python{} version \texttt{3.6.4} and \vpython{} version \texttt{7.3.2}.
 We provide online instructions for setting up \python{} and \vpython{} on different operating systems \cite{scripts}.
@@
 vec_vM0 = vector(vM0, 0, 0)
-\end{lstlisting}
+\end{lstlisting}\footnote{Note that the statement \code{from vpython import vector} must be replaced by \code{from visual import vector} for \vpython{} versions prior \texttt{7.0}.}
 We use the estimate of Eq.~(\ref{eq:check}) as a guidance to fix the time step $\Delta t$ as

### 2018-02-01 tex/MercuryNumerical.tex
Added flowchart, adjusted code, changed footnote, removed \todo package

--- 2018-01-23 tex/MercuryNumerical.tex
@@
 
-\usepackage[prependcaption, textsize=tiny]{todonotes}
+\usepackage{tikz}
+\usetikzlibrary{shapes.geometric, arrows}
+%
+\tikzstyle{startstop} = [
+	rectangle,
+	rounded corners,
+	minimum width=3cm,
+	minimum height=1cm,
+	text width=4.5cm,
+	text centered,
+	draw=black,
+	fill=green!30
+]
+
+\tikzstyle{process} = [
+	rectangle,
+	minimum width=3cm,
+	minimum height=1cm,
+	text width=4.5cm,
+	text centered,
+	draw=black,
+	fill=blue!25
+]
+\tikzstyle{decision} = [
+	diamond,
+	minimum width=4.5cm,
+	minimum height=1cm,
+	text width=1.7cm,
+	text centered,
+	draw=black,
+	fill=red!30
+]
+\tikzstyle{arrow} = [thick,->,>=stealth]
 
@@
 \begin{equation}
-\delta \Theta_{GR} = (43.03\pm 0.03)'' \ \mbox{per 100 earth years \cite{RevModPhys.19.361}} \label{delT} \, .
+\delta \Theta_{GR} = (43.03\pm 0.03)'' \ \mbox{per 100 earth years \cite{RevModPhys.19.361}} \label{delTGR} \, .
 \end{equation}
@@
 vec_vM0 = vector(vM0, 0, 0)
-\end{lstlisting}
+\end{lstlisting}\footnote{Note that the code is designed for \vpython{} versions \texttt{7} and later.
+For \vpython{} versions prior \texttt{7}, e.g., the statement from \code{vpython import vector}  must be replaced by \code{from visual import vector}.}
 We use the estimate of Eq.~(\ref{eq:check}) as a guidance to fix the time step $\Delta t$ as
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-Below we provide the complete code for part 1 as developed above.
-\lstinputlisting[language=Python,basicstyle=\linespread{1}\scriptsize\ttfamily]{mercury_orbit_base.py}
+Below we provide the complete code as well as a flowchart for part 1 as developed above.
+\begin{figure}[htb]
+\centering
+\begin{tikzpicture}[node distance=1.8cm, every node/.style={scale=0.8}]
+  \node (start)  [startstop] {Start};
+  \node (proc1)  [process, below of=start] {Definition of\\ physical parameters};
+  \draw [arrow]  (start) -- (proc1);
+  \node (proc2)  [process, below of=proc1] {Initialize distance and velocity vectors};
+  \draw [arrow]  (proc1) -- (proc2);
+  \node (proc4)  [process, below of=proc2] {Activate drawing of trajectory};
+  \draw [arrow]  (proc2) -- (proc4);
+  \node (proc5)  [process, below of=proc4] {Definition of time step \texttt{dt}};
+  \draw [arrow]  (proc4) -- (proc5);
+  \node (proc6)  [process, below of=proc5] {Set \texttt{t=0}};
+  \draw [arrow]  (proc5) -- (proc6);
+  \node (proc6b) [process, below of=proc6] {Set \texttt{alpha} and \texttt{beta}};
+  \draw [arrow]  (proc6) -- (proc6b);
+  \node (dec1)   [decision, below of=proc6b, yshift=-0.7cm] {\texttt{t < Tmax}};
+  \draw [arrow]  (proc6b) -- (dec1);
+  \node (proc7)  [process, below of=dec1, yshift=-1.2cm] {Update Mercury trajectory};
+  \draw [arrow]  (dec1) -- node[anchor=west] {YES} (proc7) ;
+  \node (proc8)  [process, below of=proc7] {Calculate \texttt{vec\_aMS(t)}};
+  \draw [arrow]  (proc7) -- (proc8);
+  \node (proc9)  [process, below of=proc8] {Calculate \texttt{vec\_vM(t+dt)}};
+  \draw [arrow]  (proc8) -- (proc9);
+  \node (proc10) [process, below of=proc9] {Calculate \texttt{vec\_rM(t+dt)}};
+  \draw [arrow]  (proc9) -- (proc10);
+  \node (stop)   [startstop, right of=proc7, xshift=3.5cm] {Stop};
+  \node (proc11) [process, left of=proc7, xshift=-3.5cm] {\texttt{t=t+dt}};
+  \draw [arrow]  (dec1) -| node[anchor=north, xshift=-1.2cm] {NO} (stop);
+  \draw [arrow]  (proc10) -| (proc11);
+  \draw [arrow]  (proc11) |- (dec1);
+\end{tikzpicture}
+\caption{\label{fig:flow}Flowchart demonstrating the logical ordering of the example code.}
+\end{figure}
+
+\begin{figure}[htb]
+\lstinputlisting[language=Python,basicstyle=\linespread{1}\footnotesize\ttfamily]{mercury_orbit_base.py}
+\end{figure}
 Further examples and template files can be found online \cite{scripts}.

### 2018-02-01 tex/mercury_orbit_base.py

@@ -1,7 +1,7 @@
 # Import all objects from the vpython module
 from vpython import *
 
-# Definition of parameters
+# Definition of physical parameters
 rM0 = 4.60    # Initial radius of Mercury orbit, in units of R0
 vM0 = 5.10e-1 # Initial orbital speed of Mercury, in units of R0/T0
 c_a = 9.90e-1 # Base acceleration of Mercury, in units of R0**3/T0**2
@@ -9,9 +9,15 @@ TM  = 8.80e+1 # Orbit period of Mercury
 rS  = 2.95e-7 # Schwarzschild radius of Sun,in units of R0
 rL2 = 8.19e-7 # Specific angular momentum, in units of R0**2
 
-# Initialize distance and velocity vectors
-vec_rM0 = vector(0, rM0, 0)
-vec_vM0 = vector(vM0, 0, 0)
+# Define the initial coordinates; M = mercury, S = Sun
+M = sphere(pos=vector(0, rM0, 0), radius=0.5,  color=color.red   )
+S = sphere(pos=vector(0, 0, 0),   radius=1.5,  color=color.yellow)
+# And the initial velocities
+M.velocity = vector(vM0, 0, 0)
+S.velocity = vector(0, 0, 0)
+
+# Add a visible trajectory to mercury
+M.trajectory = curve(color=color.white)
 
 # Definition of the time step
 dt = 2 * vM0 / c_a / 20
@@ -28,15 +34,6 @@ def evolve_mercury(vec_rM_old, vec_vM_old, alpha, beta):
     # Update position vector
     vec_rM_new = vec_rM_old + vec_vM_new * dt
     return vec_rM_new, vec_vM_new
-
-# Define the initial coordinates; M = mercury, S = Sun
-M = sphere(pos=vec_rM0,       radius=0.5,  color=color.red   )
-S = sphere(pos=vector(0,0,0), radius=1.5,  color=color.yellow)
-# And the initial velocities
-M.velocity = vec_vM0
-S.velocity = vector(0,0,0)
-# Add a visible trajectory to mercury
-M.trajectory = curve(color=color.white)
 
 t     = 0.0
 alpha = 0.0

### 2018-02-05 tex/MercuryNumerical.tex
Added addition why no terms linear in rL plus additional ref

@@
 	J.-L.~Wynen$^{1}$,
-	J.~Heuer$^{2}$,
+	J.~Heuer$^{2}$\footnote{Present Address: Institut f\"ur Neurowissenschaften und Medizin (INM-4), Forschungszentrum J\"ulich, D-52425 J\"ulich, Germany},
 	C.~M\"uller$^{3}$ and
@@
 	$^1$ \textit{Institut f\"ur Kernphysik (IKP-3) and Institute for Advanced Simulations (IAS-4), Forschungszentrum J\"ulich, D-52425 J\"ulich, Germany}\\
-	$^2$ \textit{Institut f\"ur Neurowissenschaften und Medizin (INM-4), Forschungszentrum J\"ulich, D-52425 J\"ulich, Germany}\\
-	$^3$ \textit{JuLab, Forschungszentrum J\"ulich, D-52425 J\"ulich, Germany}
+	$^2$ \textit{Hochschule Hamm-Lippstadt, Marker Allee 76-78, 59063 Hamm, Germany}\\
+	$^3$ \textit{Sch\"ulerlabor JuLab, Forschungszentrum J\"ulich, D-52425 J\"ulich, Germany}
 }
@@
 We therefore have to add one more term to the potential; in analogy to the one discussed above, we use $\beta \, \frac{r_L^2}{r^2}$.
+Here one might wonder why this additional term is chosen proportional to $(r_L/r)^2$ and not to $(r_L/r)$. The reason for this choice
+is indeed not obvious: In general the correction terms added must be scalar quantities --- accordingly vectors can enter only
+as scalar products. Moreover, one is not allowed to use square-roots of those scalar products for these impose
+wrong mathematical properties to the equations of motion (for the case at hand, e.g., the derivative with
+respect to the velocity would not be defined at $\dot{\vec r}=0$) --- the only vector that is allowed to appear as its length in linear order
+is the radius, since this length is related to geometrical properties of the system.  
 Combining everything, we use the following ansatz for the modified equation of motion:
@@
 values given in Eq.~(\ref{eq:art-prediction}).
+Modifications to the Newtonian equation of gravity introduced to account for the perihelion motion of Mercury
+are discussed in great detail in a very pedagogical way also in Ref.~\cite{Wells:2011st}.
 
@@
 Dimensional analysis is not only a tool that allows one to cross check, if the results of simulations are of the right order of magnitude, it is also very helpful to identify unusual dynamics in some systems.
-Especially the latter aspect should become clear from the discussion in this section.
+Especially the latter aspect should become clear from the discussion in this section. The modifications to the Newtonian equation of gravity introduced to account for the perihelion motion of Mercury
+are discussed also in Ref.~\cite{Wells:2011st}. To get a deeper understanding of the concepts introduced in this section reading this article is highly recommended.
 
@@
 \begin{equation}
-\delta \Theta \simeq 2\pi\left(\frac{a^2}{\bar r^2_{MS}}\right) = (\pi \times 5 \cdot 10^{-8}) \ \mbox{rad} = (1 \cdot 10^{-5})^o = (4\cdot 10^{-2}) ^{''} \ ,
+\delta \Theta \simeq 2\pi\left(\frac{r_L^2}{\bar r^2_{MS}}\right) = (\pi \times 5 \cdot 10^{-8}) \ \mbox{rad} = (1 \cdot 10^{-5})^o = (4\cdot 10^{-2}) ^{''} \ ,
 \end{equation}

### 2018-04-17 tex/MercuryNumerical.tex
Added CHs reply and modifications to the PED ref

@@
 This was already demonstrated in the "Sch\"ulerakademie Teilchenphysik", where this course was tested successfully on two groups consisting in total of 24 German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade in 2015 and 2017.
+As an example on how the course can be implemented in practice we describe our own
+experiences with it in the second to last section.
 
@@
 
-A physical system is said to be understood, if the assumed forces acting on it lead to the observed trajectories.
+A  classical physics system is said to be understood, if the assumed forces acting on it lead to the observed trajectories.
 In other words, we need to show that we can calculate the location in space of the object of interest at any future point in time, once the initial conditions are fixed properly.
@@
 \begin{eqnarray}
-\vec a(t) &=& \lim_{\Delta t\to 0} \frac{\Delta \vec v(t)}{\Delta t} =: \frac{d\vec v(t)}{dt}  = \dot{\vec v}(t) \ , \\
- &=& \frac{d^2\vec r(t)}{dt^2}  = \ddot{\vec r}(t) \ ,
+\vec a(t) &=& \lim_{\Delta t\to 0} \frac{\Delta \vec v(t)}{\Delta t} =: \frac{d\vec v(t)}{dt}  = \dot{\vec v}(t)  = \ddot{\vec r}(t) \ ,
 \end{eqnarray}
-where we introduced the second derivative in the second line.
-
-It was Newton who observed that, if a body is at rest, it will remain at rest, and if it is in motion it will remain in motion at a constant velocity in a straight line, 
-unless it is acted upon by some force --- this is known as Newton's first law.
-Formulated differently: a force $\vec F$ expresses itself by changing the motion of some object.
-This is quantified in Newton's second law
+where we introduced the second derivative in the last step.
+
+The dynamics is controlled by Newton's second law
 \begin{equation}
@@
 \end{enumerate}
-Clearly to initiate the procedure at some time $t_0$ both $\vec r(t_0)$ as well as $\vec v(t_0)$ must be known --- the  trajectories depend on these initial conditions~\footnote{%
+Clearly to initiate the procedure at some time $t_0$ both $\vec r(t_0)$ as well as $\vec v(t_0)$ must be fixed --- the  trajectories depend on these initial conditions~\footnote{%
 	In general, a differential equation of $n^{\rm th}$ degree (where the highest derivative is order $n$) needs $n$ initial conditions specified.
@@
 }.
+\begin{figure}[htb]
+	\centering
+	\begin{subfigure}[c]{0.22\textwidth}
+		\includegraphics[width=\textwidth]{figs/a0T5dt20.png}
+		\caption{\label{fig:MercuryOrbit-a0-small}}
+	\end{subfigure}
+	~
+	\begin{subfigure}[c]{0.22\textwidth}
+		\includegraphics[width=\textwidth]{figs/num-err.png}
+		\caption{\label{fig:MercuryOrbit-a0-small-dt-large}}
+	\end{subfigure}
+	\captionsetup{singlelinecheck=off}
+	\caption[]{\label{fig:MercuryOrbit1}
+		Different Mercury orbits for the purely Newtonian gravity and
+		 $\Delta t = \Delta t_0  / 20$ for (a) and $\Delta t = \Delta t_0  \times 2$ for (b), 
+		 where the black line represents the orbit of (a) and the red line is the orbit for the larger time steps;
+		Here, the reference time step is given by $\Delta t_0$ defined via Eq.~(\ref{eq:check}) as explained
+		in the text.
+		The images are screenshots of the simulation that is described below (with modified colors).
+	}
+\end{figure}
 
@@
 \end{equation}
-holds. This follows from Eq.~(\ref{eq:vdef}) where the first term that we neglected reads $(1/2)a(t){(\Delta t)}^2$.
+holds. This follows from Eq.~(\ref{eq:vdef}) where the first term that was neglected reads $(1/2)a(t){(\Delta t)}^2$.
+The impact of the size of $\Delta t$ on the simulation is illustrated in Fig.~\ref{fig:MercuryOrbit1} where
+the force employed in Newtonian gravity as detailed in the next section. While for the
+calculations for the left panel $\Delta t$
+is chosen significantly below $\Delta t_0=2m |\vec v(0)| /   |\vec F(\vec r(0)) |$ ($cf.$ Eq.~(\ref{eq:check})), for the
+right panel it was chosen twice as large. Accordingly only the trajectory in the left channel shows the characteristic
+feature of a $1/r^2$--force of an ellipse fixed in space --- the trajectory in the right channel does not reproduce itself in successive revolutions.
+
 Eq.~({\ref{eq:check}}) also shows that small (large) time steps are necessary (sufficient), if the force is strong (weak), since the time steps need to be small enough that all changes induced by the force get resolved.
@@
 	\begin{subfigure}[c]{0.22\textwidth}
-		\includegraphics[width=\textwidth]{figs/a0T5dt20.png}
-		\caption{\label{fig:MercuryOrbit-a0-small}}
-	\end{subfigure}
-	~
-	\begin{subfigure}[c]{0.22\textwidth}
-		\includegraphics[width=\textwidth]{figs/num-err.png}
-		\caption{\label{fig:MercuryOrbit-a0-small-dt-large}}
-	\end{subfigure}
-	~
-	\begin{subfigure}[c]{0.22\textwidth}
 		\includegraphics[width=\textwidth]{figs/a6T5dt20.png}
@@
 		\begin{itemize}
-		\item[(a)] $\alpha=0$ and $\Delta t = \Delta t_0  / 20$;
-		\item[(b)] $\alpha=0$ and $\Delta t = \Delta t_0  \times 2$, where the black line represents the orbit of (a) and the red line is the orbit for the larger time steps;
-		\item[(c)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$;
-		\item[(d)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$ but $r_{MS}(0) = 6 R_0$.
+		\item[(a)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$;
+		\item[(b)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$ but $r_{MS}(0) = 6 R_0$.
 		\end{itemize}
@@
 \end{itemize}
+
+\section{Own experiences with the implementation of the course}
+
+The idea that a numerical study of the perihelion motion of Mercury is an interesting subject for high
+school students was taken from a project work of one of the authors (JH) presented at the 
+M\"adchen-Gymnasium J\"ulich (a German high school) in 11$^{\rm th}$ grade in 2014. Based on this study
+the course was developed as outlined in Sec.~\ref{sec:Numerical Implementation} to be used in 
+the Sch\"ulerakademie Teilchenphysik" in 2015, which aims at  high school students from 10$^{\rm th}$ to 
+13$^{\rm th}$ grade. This academy is financed by the German Research Society (DFG) as the outreach
+branch
+of a large research grant focussing on basic research in particle and nuclear physics.
+One of the authors (CH) is acting as a Principal Investigator for the outreach
+activities within this scheme. It runs biannually
+at the Science Center Overbach in J\"ulich
+for four days and has about 25 participants coming from various parts of Germany. 
+The academy comprises lectures, a tour to the particle accelerator
+COSY as well as the high performance computing facility at the Research Center J\"ulich as well
+as one full day of practical work. The course presented here
+ is offered as one option out of three on this day and was prepared
+by a lecture to the whole group explaining the basics of numerical simulations. 
+Given the setting the participating students
+are highly motivated and eager to understand. Typically about one half of the participants chose to
+work on the numerical project --- out of those about half had programmed a computer before while
+the other half did not yet have any programming experience. Most of the participants had already
+learned about derivatives in school. Given the very positive to enthusiastic 
+feed backs we received from the participants we decided to make the course in the form presented
+to a permanent part of the academy. It was used in the 2017 with the same success.
+
+The students worked in pairs. Each group had an own laptop provided by us, where the necessary
+software was preinstalled. Four of the authors (IH, CK, CM, JLW) acted as supervisors for
+the students. 
+The material covered in this one day is in most cases what is presented in Sec.~\ref{sec:Numerical Implementation}.
+Only very few participants managed to start thinking about how to actually extract quantitatively the
+perihelion motion --- not to speak of uncertainty estimates.
+Once the basic problem was tackled some students deviated from the suggested path and for
+instance studied the problem with one additional planet which was fine for us as well. This is
+a very interesting problem by itself for there are parameter ranges where the classical three-body
+system develops chaotic features.
+ The reason why we included the
+additional material in the paper is that we wanted to provide teachers with sufficient material to
+probably even fill a working group outside the general curriculum running for a longer period.
 

### 2018-04-17 ped-response/response.tex
Created latex version of response and adjusted main tex file
Reincluded figure of alpha=0 in sec 4. And made modifications to sec 8.

%Bira 06/04/11
%Rob 06/04/11
\documentclass[prc,preprint,showpacs,showkeys,nofootinbib,tightenlines]{revtex4}
\newcommand{\slashT}{\slash\hspace{-0.4em}T}

\usepackage{amsmath}

\begin{document}
\pagenumbering{gobble}

\section*{Letter to the Editor}

\noindent
Dear Editor, \\

We thank the referee for the positive and constructive report.
As detailed below we followed the suggestions made by the referee and hope that the improved version is now regarded appropriate for publication in Physics Education.

\begin{enumerate}
\item \textbf{Referee:} 
\textit{
Section 2 is largely unnecessary as it covers the basics of Newton's Laws and the Euler step method, which will be familiar to everyone.
This section could be shortened to the necessary equations and an introduction to the notation, together with a comment on time steps as this is important later.
}
\\
\\
\textbf{Response:}
We shortened the section somewhat, however, kept most of the equations. In particular, although straightforward, we regard it as useful for the logic of the presentation (and probably also for teaching the course) to introduce the derivatives via Eqs (1) and (2) and not directly via Eqs (3) and (4).

\item \textbf{Referee:} 
\textit{
figure 2 shows that the trajectory varying even when alpha and beta are zero.
It is worth explaining at this point (rather than later) that this is due to the finite time step.
Also it is worth noting if the trajectory change from one orbit to the next leads to a perihelion shift (also picked up later in more detail)
}
\\
\\
\textbf{Response:}
We moved half of Fig. 2 forward into Sec. 2. We thank the referee for this suggestion for it allows us to illustrate the statements immediately and quantitatively, which improves the presentation significantly.

\item \textbf{Referee:} 
\textit{
Finally, the only reason this paper is suitable for Phys Ed rather than, say a journal aimed at undergraduates, is that the authors claim it has been used successfully on students of high school age.
However, there is no supporting evidence of this, no indication of success criteria and no explanation of what was done.
In order to fit better with the idea of an education research journal at this level, perhaps the authors could include some feedback or anecdotal evidence even if they have no formal study of the efficacy of the trial?
}
\\
\\
\textbf{Response:}
We added the new Sec. 8 (Own experiences with the implementation of the course), where we now not only explain in which context we used (and will be using) the course, but also provide some information about the participants.

\end{enumerate}

\noindent
Sincerely,\\

	C.~K\"orber,
	I.~Hammer,
	J.-L.~Wynen,
	J.~Heuer,
	C.~M\"uller and
	C.~Hanhart
\newpage

\end{document}

### 2018-04-17 tex/MercuryNumerical.tex
Created latex version of response and adjusted main tex file
Reincluded figure of alpha=0 in sec 4. And made modifications to sec 8.

@@
 This was already demonstrated in the "Sch\"ulerakademie Teilchenphysik", where this course was tested successfully on two groups consisting in total of 24 German high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade in 2015 and 2017.
-As an example on how the course can be implemented in practice we describe our own
+As an example on how the course can be implemented in practice, we describe our own
 experiences with it in the second to last section.
@@
 	\begin{subfigure}[c]{0.22\textwidth}
+		\includegraphics[width=\textwidth]{figs/a0T5dt20.png}
+		\caption{\label{fig:MercuryOrbit-a6-small}}
+	\end{subfigure}
+	\begin{subfigure}[c]{0.22\textwidth}
 		\includegraphics[width=\textwidth]{figs/a6T5dt20.png}
@@
 		\begin{itemize}
-		\item[(a)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$;
-		\item[(b)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$ but $r_{MS}(0) = 6 R_0$.
+		\item[(a)] $\alpha=0 $ and $\Delta t = \Delta t_0 / 20$;
+		\item[(b)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$;
+		\item[(c)] $\alpha=10^{6}$ and $\Delta t = \Delta t_0 / 20$ but $r_{MS}(0) = 6 R_0$.
 		\end{itemize}
@@
 
-The idea that a numerical study of the perihelion motion of Mercury is an interesting subject for high
-school students was taken from a project work of one of the authors (JH) presented at the 
-M\"adchen-Gymnasium J\"ulich (a German high school) in 11$^{\rm th}$ grade in 2014. Based on this study
-the course was developed as outlined in Sec.~\ref{sec:Numerical Implementation} to be used in 
-the Sch\"ulerakademie Teilchenphysik" in 2015, which aims at  high school students from 10$^{\rm th}$ to 
-13$^{\rm th}$ grade. This academy is financed by the German Research Society (DFG) as the outreach
-branch
-of a large research grant focussing on basic research in particle and nuclear physics.
-One of the authors (CH) is acting as a Principal Investigator for the outreach
-activities within this scheme. It runs biannually
-at the Science Center Overbach in J\"ulich
-for four days and has about 25 participants coming from various parts of Germany. 
-The academy comprises lectures, a tour to the particle accelerator
-COSY as well as the high performance computing facility at the Research Center J\"ulich as well
-as one full day of practical work. The course presented here
- is offered as one option out of three on this day and was prepared
-by a lecture to the whole group explaining the basics of numerical simulations. 
-Given the setting the participating students
-are highly motivated and eager to understand. Typically about one half of the participants chose to
-work on the numerical project --- out of those about half had programmed a computer before while
-the other half did not yet have any programming experience. Most of the participants had already
-learned about derivatives in school. Given the very positive to enthusiastic 
-feed backs we received from the participants we decided to make the course in the form presented
-to a permanent part of the academy. It was used in the 2017 with the same success.
-
-The students worked in pairs. Each group had an own laptop provided by us, where the necessary
-software was preinstalled. Four of the authors (IH, CK, CM, JLW) acted as supervisors for
-the students. 
-The material covered in this one day is in most cases what is presented in Sec.~\ref{sec:Numerical Implementation}.
-Only very few participants managed to start thinking about how to actually extract quantitatively the
-perihelion motion --- not to speak of uncertainty estimates.
-Once the basic problem was tackled some students deviated from the suggested path and for
-instance studied the problem with one additional planet which was fine for us as well. This is
-a very interesting problem by itself for there are parameter ranges where the classical three-body
-system develops chaotic features.
- The reason why we included the
-additional material in the paper is that we wanted to provide teachers with sufficient material to
-probably even fill a working group outside the general curriculum running for a longer period.
+The idea that the numerical study of the perihelion motion of Mercury is an interesting subject for high school students was taken from a project work of one of the authors (JH).
+The project was eventually presented at the M\"adchen-Gymnasium J\"ulich (a German high school) in 11$^{\rm th}$ grade in 2014.
+Based on this study, the course was developed as outlined in Sec.~\ref{sec:Numerical Implementation} and used in the Sch\"ulerakademie Teilchenphysik in 2015, which aims at high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade.
+This academy is financed by the German Research Society (DFG) as the outreach branch of a research grant focussing on basic research in particle and nuclear physics.
+One of the authors (CH) is acting as a Principal Investigator for the outreach activities within this scheme.
+The Sch\"ulerakademie Teilchenphysik runs biannually at the Science Center Overbach in J\"ulich for four days and has about 25 participants coming from various parts of Germany.
+The academy comprises lectures, a tour to the particle accelerator COSY as well as the high performance computing facility at the Research Center J\"ulich as well as one full day of practical work.
+Given the setting the participating students are highly motivated and eager to understand.
+
+The course presented in this paper is offered as one out of three hands-on project options and was prepared by a lecture to the whole group explaining the basics of numerical simulations.
+Typically about one half of the participants chose to work on the numerical project --- out of those about one half had previous programming experience.
+Most of the participants had already learned about derivatives and vector calculus in school.
+Given the positive to enthusiastic feedback we have received from the participants on the numerical course, we decided to make the course a permanent part of the academy.
+It was used in the 2017 with the same success.
+
+The numerical course was allocated for one work day (seven hours) and the students worked in pairs.
+Each group worked on a laptop provided by us, where the necessary software was preinstalled.
+Four of the authors (IH, CK, CM, JLW) acted as supervisors for the students. 
+The course was initiated with a brief summary of previously presented methods, motivations and objectives of the simulation.
+In the first two hours the students were introduced to \python{} and \vpython{}.
+In the following part the students worked with an initial code template and were motivated to come up with own solutions to fulfill the objectives.
+A large part of the material covered during this day is presented in Sec.~\ref{sec:Numerical Implementation}.
+While a majority of the participants where able to visually simulate the motion of mercury including forces from GR, only a few participants started to implement the quantitative extraction of the perihelion motion --- not to speak of uncertainty estimates.
+Once the basic problem was tackled some students deviated from the suggested path and for instance studied the problem with one additional planet or different starting parameters.
+This is a very interesting problem by itself for there are parameter ranges where the classical three-body system develops chaotic features.
+We encouraged such explorations as well.
+The reason why we included the additional material in the paper is that we wanted to provide teachers with sufficient material to probably even fill a working group outside the general curriculum running for a longer period.
 

### 2018-04-17 tex/MercuryNumerical.tex
CH edits 2

@@
 The idea that the numerical study of the perihelion motion of Mercury is an interesting subject for high school students was taken from a project work of one of the authors (JH).
-The project was eventually presented at the M\"adchen-Gymnasium J\"ulich (a German high school) in 11$^{\rm th}$ grade in 2014.
+The project was presented at the M\"adchen-Gymnasium J\"ulich (a German high school) in 11$^{\rm th}$ grade in 2014.
 Based on this study, the course was developed as outlined in Sec.~\ref{sec:Numerical Implementation} and used in the Sch\"ulerakademie Teilchenphysik in 2015, which aims at high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade.
@@
 The Sch\"ulerakademie Teilchenphysik runs biannually at the Science Center Overbach in J\"ulich for four days and has about 25 participants coming from various parts of Germany.
-The academy comprises lectures, a tour to the particle accelerator COSY as well as the high performance computing facility at the Research Center J\"ulich as well as one full day of practical work.
+The academy comprises lectures, a tour to the particle accelerator COSY and the high performance computing facility at the Research Center J\"ulich as well as one full day of practical work.
 Given the setting the participating students are highly motivated and eager to understand.

### 2018-04-18 tex/MercuryNumerical.tex
Add my mustard to the text

@@
 	\caption[]{\label{fig:MercuryOrbit1}
-		Different Mercury orbits for the purely Newtonian gravity and
+		Different Mercury orbits for the purely Newtonian gravity~(\ref{eq:newton_grav}) and
 		 $\Delta t = \Delta t_0  / 20$ for (a) and $\Delta t = \Delta t_0  \times 2$ for (b),
-		 where the black line represents the orbit of (a) and the red line is the orbit for the larger time steps;
-		Here, the reference time step is given by $\Delta t_0$ defined via Eq.~(\ref{eq:check}) as explained
+		 where the black line represents the orbit of (a) and the red line is the orbit for the larger time steps.
+		The reference time step $\Delta t_0$ is given by Eq.~(\ref{eq:check}) as explained
 		in the text.
@@
 The impact of the size of $\Delta t$ on the simulation is illustrated in Fig.~\ref{fig:MercuryOrbit1} where
-the force employed in Newtonian gravity as detailed in the next section. While for the
+only Newtonian gravity was used. While for the
 calculations for the left panel $\Delta t$
 is chosen significantly below $\Delta t_0=2m |\vec v(0)| /   |\vec F(\vec r(0)) |$ ($cf.$ Eq.~(\ref{eq:check})), for the
-right panel it was chosen twice as large. Accordingly only the trajectory in the left channel shows the characteristic
-feature of a $1/r^2$--force of an ellipse fixed in space --- the trajectory in the right channel does not reproduce itself in successive revolutions.
+right panel it was chosen twice as large. Accordingly, only the trajectory in the left panel shows the characteristic
+feature of a $1/r^2$--force of an ellipse fixed in space. The trajectory in the right panel does not reproduce itself in successive revolutions.
 
@@
 \begin{equation}
-\vec F_N(\vec r) = - \frac{G_N m M_\odot}{r^2} \frac{\vec r}{r}\ ,
+\vec F_N(\vec r) = - \frac{G_N m M_\odot}{r^2} \frac{\vec r}{r}\ ,\label{eq:newton_grav}
 \end{equation}
@@
 In the following, we present code examples which tested for \python{} version \texttt{3.6.4} and \vpython{} version \texttt{7.3.2}.
-We provide online instructions for setting up \python{} and \vpython{} on different operating systems \cite{scripts}.
+We provide online instructions for setting up \python{} and \vpython{} on different operating systems~\cite{scripts}.
 In \ref{sec:code}, we also provide a working example as well as possible extensions and template files online.
@@
 
-In Section~\ref{sec:Numerical Implementation} we suggested to present the material by using a template as well as a step by step instruction to guide the students through the problem solution \cite{scripts}.
+In Section~\ref{sec:Numerical Implementation} we suggested to present the material by using a template as well as a step by step instruction to guide the students through the problem solution~\cite{scripts}.
 These instructions can be cut down or left out depending on available time and numerical/computational versatility of the students.
@@
 
-The idea that the numerical study of the perihelion motion of Mercury is an interesting subject for high school students was taken from a project work of one of the authors (JH).
+Work on the subject began with a project work of one of the authors (JH).
 The project was presented at the M\"adchen-Gymnasium J\"ulich (a German high school) in 11$^{\rm th}$ grade in 2014.
@@
 One of the authors (CH) is acting as a Principal Investigator for the outreach activities within this scheme.
-The Sch\"ulerakademie Teilchenphysik runs biannually at the Science Center Overbach in J\"ulich for four days and has about 25 participants coming from various parts of Germany.
+The Sch\"ulerakademie Teilchenphysik runs biennially at the Science Center Overbach in J\"ulich for four days and has about 25 participants coming from various parts of Germany.
 The academy comprises lectures, a tour to the particle accelerator COSY and the high performance computing facility at the Research Center J\"ulich as well as one full day of practical work.
-Given the setting the participating students are highly motivated and eager to understand.
-
-The course presented in this paper is offered as one out of three hands-on project options and was prepared by a lecture to the whole group explaining the basics of numerical simulations.
-Typically about one half of the participants chose to work on the numerical project --- out of those about one half had previous programming experience.
+Given the setting, the participating students are highly motivated and eager to understand.
+
+The course presented in this paper is offered as one out of three hands-on project options.
+A lecture to the whole group explaining the basics of numerical simulations laid down some foundations in addition to what is presented in this paper.
+Given the positive to enthusiastic feedback we have received from the participants on the course after its first installation in 2015, we decided to make the course a permanent part of the academy.
+In 2017 we received similarly positive feedback.
+About one half of the participants of each year chose to work on the numerical project --- out of those about one half had previous programming experience.
 Most of the participants had already learned about derivatives and vector calculus in school.
-Given the positive to enthusiastic feedback we have received from the participants on the numerical course, we decided to make the course a permanent part of the academy.
-It was used in the 2017 with the same success.
 
 The numerical course was allocated for one work day (seven hours) and the students worked in pairs.
-Each group worked on a laptop provided by us, where the necessary software was preinstalled.
-Four of the authors (IH, CK, CM, JLW) acted as supervisors for the students. 
+Each group worked on a laptop provided by us, where the necessary software was pre-installed.
+Four of the authors (IH, CK, CM, JLW) acted as supervisors for the students although not all at the same time. Each instance of the course was supervised by one to three people.
 The course was initiated with a brief summary of previously presented methods, motivations and objectives of the simulation.
 In the first two hours the students were introduced to \python{} and \vpython{}.
-In the following part the students worked with an initial code template and were motivated to come up with own solutions to fulfill the objectives.
+In the following part the students worked with an initial code template (see~\cite{scripts}) and were motivated to come up with own solutions to fulfill the objectives.
 A large part of the material covered during this day is presented in Sec.~\ref{sec:Numerical Implementation}.
-While a majority of the participants where able to visually simulate the motion of mercury including forces from GR, only a few participants started to implement the quantitative extraction of the perihelion motion --- not to speak of uncertainty estimates.
+With the exception of some younger students, the participants where able to simulate and visualize the motion of mercury including forces from GR.\@
+Only a few participants started to implement the quantitative extraction of the perihelion motion, uncertainty estimates were not a subject as the time frame was too short.
 Once the basic problem was tackled some students deviated from the suggested path and for instance studied the problem with one additional planet or different starting parameters.
@@
 We encouraged such explorations as well.
-The reason why we included the additional material in the paper is that we wanted to provide teachers with sufficient material to probably even fill a working group outside the general curriculum running for a longer period.
+We include the additional material in the paper because we want to provide teachers with sufficient material to fill a working group outside the general curriculum running for a longer period than one day.
 
@@
 \end{figure}
-Further examples and template files can be found online \cite{scripts}.
+Further examples and template files can be found online~\cite{scripts}.
 
